---
title: "Logistic mixed model: MASS bacteria data"
author: "Patrick Brown"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
---

## Introduction

The `bacteria` data set in **MASS** records repeated throat swabs for
*Haemophilus influenzae* on 50 children (Venables and Ripley 2002). We fit the
logistic mixed model from the `glmmPQL` help page:

```r
glmmPQL(y ~ trt + I(week > 2), random = ~ 1 | ID,
        family = binomial, data = bacteria)
```

using the built-in `binomial()` observation term and `iid()` random intercept.

## Data

```{r setup}
library(adlaplace)
library(MASS)
have_tmb <- requireNamespace("glmmTMB", quietly = TRUE)
have_inla <- requireNamespace("INLA", quietly = TRUE)
data(bacteria)
bacteria$yn <- as.integer(bacteria$y == "y")
bacteria$wk <- as.integer(bacteria$week > 2)
head(bacteria[, c("ID", "y", "trt", "week", "yn", "wk")])
```

## Fit with adlaplace

```{r fit}
fit <- adlaplace(
  binomial(yn) ~ trt + wk + iid(ID, init = 1),
  data = bacteria,
  num_shards = 20L
)
summary(fit)
```

```{r coef}
co <- coef(fit)
data.frame(
  parameter = names(co),
  estimate = unname(co)
)
```

The random-intercept standard deviation `ID_iid` is on the log scale during
optimization; exponentiate for the natural scale.

## glmmPQL reference

```{r pql}
pql <- glmmPQL(
  y ~ trt + I(week > 2), random = ~ 1 | ID,
  family = binomial, data = bacteria, verbose = FALSE
)
nlme::fixef(pql)
as.numeric(nlme::VarCorr(pql)["(Intercept)", "StdDev"])
```

## glmmTMB reference

```{r tmb, eval=have_tmb}
tmb <- glmmTMB::glmmTMB(
  yn ~ trt + wk + (1 | ID),
  family = binomial(), data = bacteria
)
glmmTMB::fixef(tmb)$cond
sqrt(as.numeric(glmmTMB::VarCorr(tmb)$cond$ID))
```

## INLA reference

```{r inla, eval=have_inla}
INLA::inla.setOption(num.threads = "1:1")
inl <- INLA::inla(
  yn ~ trt + wk + f(ID, model = "iid",
    hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01)))),
  family = "binomial", Ntrials = 1, data = bacteria
)
inl$summary.fixed[, "mean"]
INLA::inla.emarginal(
  function(x) 1 / sqrt(x),
  inl$marginals.hyperpar[["Precision for ID"]]
)
```

## Comparison table

```{r compare}
lab <- names(co)
adl <- c(
  `(Intercept)` = unname(co[lab == "intercept"]),
  trtdrug = unname(co[lab == "trt_linear"][1]),
  `trtdrug+` = unname(co[which(lab == "trt_linear")[2]]),
  `I(week>2)` = unname(co[lab == "wk_linear"]),
  `SD(ID)` = unname(co[lab == "ID_iid"])
)

fe <- nlme::fixef(pql)
pql_vec <- c(
  fe[["(Intercept)"]], fe[["trtdrug"]], fe[["trtdrug+"]],
  fe[["I(week > 2)TRUE"]],
  as.numeric(nlme::VarCorr(pql)["(Intercept)", "StdDev"])
)

comparison_table <- data.frame(
  Parameter = names(adl),
  adlaplace = as.numeric(adl),
  glmmPQL = pql_vec,
  check.names = FALSE
)

if (have_tmb) {
  tfe <- glmmTMB::fixef(tmb)$cond
  tmb_sd <- sqrt(as.numeric(glmmTMB::VarCorr(tmb)$cond$ID))
  comparison_table$glmmTMB <- c(
    tfe[["(Intercept)"]], tfe[["trtdrug"]], tfe[["trtdrug+"]],
    tfe[["wk"]], tmb_sd
  )
}

if (have_inla) {
  ife <- inl$summary.fixed[, "mean"]
  names(ife) <- rownames(inl$summary.fixed)
  inla_sd <- INLA::inla.emarginal(
    function(x) 1 / sqrt(x),
    inl$marginals.hyperpar[["Precision for ID"]]
  )
  comparison_table$INLA <- c(
    ife[["(Intercept)"]], ife[["trtdrug"]], ife[["trtdrug+"]],
    ife[["wk"]], inla_sd
  )
}

knitr::kable(comparison_table, digits = 3)
```

`adlaplace` and `glmmTMB` agree closely (both use an exact Laplace
approximation). Differences from `glmmPQL` are expected for a binary
response; `INLA` posterior means differ slightly because of its prior on the
random-intercept precision.
