---
title: "FEM Matérn methods: SPDE, B-splines, and covariance comparison"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{FEM Matérn methods: SPDE, B-splines, and covariance comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(adlaplaceFem)
library(Matrix)
```

# Continuous model: Matérn via Whittle SPDE

A zero-mean Gaussian field $u(\mathbf{s})$ on $\mathbb{R}^d$ with Matérn covariance

$$
\operatorname{Cov}\bigl(u(\mathbf{s}),u(\mathbf{s}')\bigr)
=
\frac{\sigma^2}{2^{\nu-1}\Gamma(\nu)}
(\kappa\|\mathbf{h}\|)^\nu K_\nu(\kappa\|\mathbf{h}\|),
\quad \mathbf{h}=\mathbf{s}-\mathbf{s}',
$$

is a stationary solution of

$$
\tau\,(\kappa^2-\Delta)^{\alpha/2}\,u(\mathbf{s})=\mathcal{W}(\mathbf{s}),
\qquad \alpha=\nu+d/2,
$$

with Gaussian white noise $\mathcal{W}$ (Whittle; Lindgren, Rue & Lindström 2011).
The continuous precision operator is $\tau^2(\kappa^2-\Delta)^\alpha$.

In $d=2$:

| α | ν | SPDE | Precision operator |
|---|----|------|--------------------|
| 2 | 1 | $\tau(\kappa^2-\Delta)u=\mathcal{W}$ | $\tau^2(\kappa^2-\Delta)^2$ |
| 3 | 2 | $\tau(\kappa^2-\Delta)^{3/2}u=\mathcal{W}$ | $\tau^2(\kappa^2-\Delta)^3$ |

On a bounded domain, homogeneous Neumann boundary conditions are used; pad the
domain by about half a correlation range in applications.

# Parameterization: range and sd

The FEM term exposes hyperparameters $(\mathrm{range},\mathrm{sd})$ (optimized as
$\log\mathrm{range}$, $\log\mathrm{sd}$), not the raw SPDE pair $(\kappa,\tau)$.
The public `range` is the same **practical range** as geostatsp/INLA,
$\rho=\sqrt{8\nu}/\kappa$. Internally the AD kernels convert

$$
\kappa = \sqrt{8\nu}/\mathrm{range},
\qquad
\tau =
\begin{cases}
1\big/\bigl(\kappa\,\sigma\sqrt{4\pi}\bigr) & \alpha=2\ (\nu=1),\\
1\big/\bigl(\kappa^2\,\sigma\sqrt{8\pi}\bigr) & \alpha=3\ (\nu=2),
\end{cases}
$$

with $\sigma=\mathrm{sd}$, so that $\mathrm{sd}$ is the **marginal standard
deviation** of the Matérn field $u(\mathbf{s})$ on the linear-predictor scale.
In 2D, $\nu=\alpha-1$.

| Quantity | Meaning | Units (loaloa) |
|----------|---------|----------------|
| Coordinates / knots | Spatial location | m |
| `range` $=\rho=\sqrt{8\nu}/\kappa$ | Practical range (≈0.1 correlation) | m |
| `sd` $=\sigma$ | Marginal SD of $u(\mathbf{s})$ | same as $u$ (e.g. logit) |
| $\tau$ | Internal SPDE precision scale | derived from $\kappa,\sigma$ |

So with coordinates in metres and $\nu=2$, `range = 1e5` means a practical range
of $100\,\mathrm{km}$ and $\kappa=4\times10^{-5}\,\mathrm{m}^{-1}$.

**geostatsp / INLA `glgm`** uses the same `(range, sd)` meaning, so coefficient
tables can be compared directly:

```r
cf <- coef(fit)
c(
  range = unname(cf[["geometry_matern_log_range"]]),
  sd = unname(cf[["geometry_matern_log_sd"]])
)
```

(`glgm` may label the range row `range/1000` when the summary is in kilometres;
convert by $\times 1000$ first.)

The Gram matrices $\mathbf{C}$, $\mathbf{G}$, … use the same coordinate units as
the knots, so `range` is already on that scale.

# Finite-element weights and Grams

Write $\widetilde{u}(\mathbf{s})=\boldsymbol{\psi}(\mathbf{s})^\top\mathbf{w}$.
The weight precision is assembled from

| Matrix | Entries |
|--------|---------|
| Mass $\mathbf{C}$ | $C_{ij}=\langle\psi_i,\psi_j\rangle$ |
| Stiffness $\mathbf{G}$ | $G_{ij}=\langle\nabla\psi_i,\nabla\psi_j\rangle$ |
| Biharmonic $\mathbf{G}_2$ | $(G_2)_{ij}=\langle\Delta\psi_i,\Delta\psi_j\rangle$ |
| Third-order $\mathbf{G}_3$ | $(G_3)_{ij}=\langle\nabla\Delta\psi_i,\nabla\Delta\psi_j\rangle$ |

Working formulas (higher-order B-splines; no mass lumping):

$$
\mathbf{Q}_2=\tau^2\bigl(\kappa^4\mathbf{C}+2\kappa^2\mathbf{G}+\mathbf{G}_2\bigr),
$$

$$
\mathbf{Q}_3=\tau^2\bigl(\kappa^6\mathbf{C}+3\kappa^4\mathbf{G}+3\kappa^2\mathbf{G}_2+\mathbf{G}_3\bigr).
$$

These replace Galerkin expressions involving $\mathbf{C}^{-1}$, which destroy sparsity.
Recommended B-spline degrees: $p=2$ for α = 2; $p=3$ for α = 3.

# Derivatives

The Gaussian prior on the weights $\mathbf{w}$ is

$$
\log\pi(\mathbf{w}\mid\theta)
=
\frac12\log\det\mathbf{Q}(\theta)
-\frac12\mathbf{w}^\top\mathbf{Q}(\theta)\mathbf{w}
-\frac{n}{2}\log(2\pi),
$$

with $\theta=(\log\mathrm{range},\log\mathrm{sd})$ and $n$ the number of weights.
The two terms are separate AD groups. Both see the same sparse
$\mathbf{Q}(\theta)=\sum_j c_j(\theta)\,\mathbf{M}_j$, where $\mathbf{M}_j$ runs over
$(\mathbf{C},\mathbf{G},\mathbf{G}_2)$ for $\alpha=2$ and also $\mathbf{G}_3$ for
$\alpha=3$. The scalars $c_j(\theta)$ are the coefficients in the formulas for
$\mathbf{Q}_2$ and $\mathbf{Q}_3$ above, after substituting $\kappa$ and $\tau$
from $(\mathrm{range},\mathrm{sd})$.

The Laplace approximation differentiates this density in three places.
The inner optimization needs the value, the gradient, and the Hessian in
$\mathbf{w}$. The outer optimization needs first and second derivatives in
$\theta$. The profile trace needs the contraction of $H^{-1}$ with third
derivatives of the joint density; for this prior those third derivatives are
nonzero only in $\theta$.

## Quadratic form

`random_fem_ssq_2` and `random_fem_ssq_3` contribute
$f=-\frac12\mathbf{w}^\top\mathbf{Q}\mathbf{w}$.
$\mathbf{Q}$ is linear in the $c_j$ and $f$ is exactly quadratic in
$\mathbf{w}$, so the derivatives are

$$
\begin{aligned}
\frac{\partial f}{\partial\mathbf{w}}
&= -\mathbf{Q}\mathbf{w},\\
\frac{\partial f}{\partial\theta_k}
&= -\frac12\sum_j\frac{\partial c_j}{\partial\theta_k}\,
  (\mathbf{w}^\top\mathbf{M}_j\mathbf{w}),\\
\frac{\partial^2 f}{\partial\mathbf{w}\,\partial\mathbf{w}^\top}
&= -\mathbf{Q},\\
\frac{\partial^2 f}{\partial\mathbf{w}\,\partial\theta_k}
&= -\frac{\partial\mathbf{Q}}{\partial\theta_k}\mathbf{w},\\
\frac{\partial^2 f}{\partial\theta_a\,\partial\theta_b}
&= -\frac12\sum_j
  \frac{\partial^2 c_j}{\partial\theta_a\,\partial\theta_b}\,
  (\mathbf{w}^\top\mathbf{M}_j\mathbf{w}).
\end{aligned}
$$

Every one of these is a pass over the stored upper triangle of the Grams:
combine the $c_j$ (or their $\theta$-derivatives) into $\mathbf{Q}$ or
$\partial\mathbf{Q}/\partial\theta_k$, then one quadratic form or one sparse
matrix-vector product. A two-input tape maps $\theta$ to the $c_j$ and supplies
$\partial c_j/\partial\theta$ and $\partial^2 c_j/\partial\theta^2$. The
products along the stored pattern of $\mathbf{Q}$ are not operations on that
tape.

Third derivatives in $\mathbf{w}$ are identically zero. In a direction
$\mathbf{v}$ in the weight space the only surviving third-order term is

$$
\frac{\partial}{\partial\theta_k}
\Bigl(\mathbf{v}^\top\frac{\partial^2 f}{\partial\mathbf{w}\,\partial\mathbf{w}^\top}\mathbf{v}\Bigr)
=
-\frac12\mathbf{v}^\top\frac{\partial\mathbf{Q}}{\partial\theta_k}\mathbf{v}.
$$

## Log-determinant atomic

`random_fem_det_2` and `random_fem_det_3` contribute
$\frac12\log\det\mathbf{Q}-\frac{n}{2}\log(2\pi)$.
This term does not depend on $\mathbf{w}$, so its inner gradient and Hessian
are zero. All of its derivatives are in $\theta$, obtained by differentiating
in the coefficients $x_j=c_j(\theta)$ and chaining back through the same
two-input map.

The atomic `fem_logdet` accepts the coefficient vector $\mathbf{x}$ (length 3
or 4) and returns $\log\det\mathbf{Q}(\mathbf{x})$ for
$\mathbf{Q}=\sum_j x_j\mathbf{M}_j$. The sparse LDL and the selected inverse
stay inside the atomic; the tape records only the call and the chain rule
from $\mathbf{x}$ to $\theta$.

One evaluation assembles the numeric values of $\mathbf{Q}$, factors it with
the symbolic LDL pattern stored on the payload, and, when a derivative is
required, forms the Takahashi selected inverse $\boldsymbol{\Sigma}$
($\mathbf{Q}^{-1}$ on the sparsity pattern of the factor). The identity
$d\log\det\mathbf{Q}=\operatorname{Tr}(\mathbf{Q}^{-1}\,d\mathbf{Q})$ then gives

$$
\frac{\partial\log\det\mathbf{Q}}{\partial x_j}
=
\operatorname{Tr}(\mathbf{Q}^{-1}\mathbf{M}_j)
=
\sum_k \omega_k (M_j)_k \Sigma_k,
$$

where the sum is over stored upper-triangle entries and $\omega_k$ is $1$ on
the diagonal and $2$ off it.

That selected inverse is `takahashi_davis`. For each column of $\mathbf{L}$,
from last to first, the column of $\boldsymbol{\Sigma}$ is scattered into a
work vector of length $n$, and the Takahashi update reads the vector by row
index. The stored result is $\mathbf{Q}^{-1}$ on the sparsity pattern of
$\mathbf{L}$. `fem_logdet` calls this function. A dual scalar in the work
vector is what produces the directional derivative below.

The same subset can be computed without a FEM model. From R,
`takahashi_davis(Q)` returns a `dgCMatrix` of those entries in the original
variable order. Pass `chol = fem_chol_pattern(Q)` when that symbolic factor
is already known. From C++, a package with `LinkingTo: adlaplaceFem`
includes `adlaplaceFem/takahashi_davis.hpp`. That function takes $\mathbf{L}$
and $\mathbf{D}$. A caller that has only $\mathbf{Q}$ factors it with
`chol_update_csc` from adlaplace first.

Because $\mathbf{Q}$ is linear in $\mathbf{x}$, the second derivative has no
$\operatorname{Tr}(\mathbf{Q}^{-1}\partial^2\mathbf{Q})$ term:

$$
\frac{\partial^2\log\det\mathbf{Q}}{\partial x_i\,\partial x_j}
=
-\operatorname{Tr}(\mathbf{Q}^{-1}\mathbf{M}_i\mathbf{Q}^{-1}\mathbf{M}_j).
$$

CppAD asks the atomic for this as a Hessian-vector product. The same LDL and
Takahashi code runs with a dual scalar $(a,\dot a)$, so one factorization
returns both $\partial\log\det\mathbf{Q}/\partial x_j$ and its directional
derivative. The prior density uses half of these derivatives.

# Tensor-product B-splines

On a rectangle, with 1-D B-splines $B_i^x$ and $B_j^y$,
$\psi_{ij}(x,y)=B_i^x(x)\,B_j^y(y)$ and the Grams factor as Kronecker sums

$$
\mathbf{C}=\mathbf{C}_x\otimes\mathbf{C}_y,
$$

$$
\mathbf{G}=\mathbf{G}_x\otimes\mathbf{C}_y+\mathbf{C}_x\otimes\mathbf{G}_y,
$$

$$
\mathbf{G}_2=\mathbf{G}_{2,x}\otimes\mathbf{C}_y+\mathbf{C}_x\otimes\mathbf{G}_{2,y}+2\,\mathbf{G}_x\otimes\mathbf{G}_y.
$$

# Hierarchical B-splines

For spatial domains where data are concentrated in subregions, build
hierarchical knot lines with [hb_knots()] and pass the result to `matern()`:

```r
kn <- hb_knots(
  outer = list(                          # level 0: padded domain
    xmin = ..., xmax = ..., ymin = ..., ymax = ...,
    resolution = c(dx, dy)
  ),
  inner = list(
    c(xmin, xmax, ymin, ymax),           # level 1: study region
    list(west_ext, east_ext)             # level 2: sibling rectangles
  ),
  fact = c(2, 4)
)
matern(geometry, knots = kn, shape = 2L)
```

`fact[i]` is the cell-size ratio from the previous level. Each inner extent is
snapped to that level's nested cell grid. This gives nested spaces
$V^0 \subset V^1 \subset \cdots \subset V^L$. A truncated hierarchical (THB)
basis keeps level-$\ell$ tensor B-splines active when their support lies in the
level-$\ell$ region but not in finer regions. Every active function embeds in
the finest space via a sparse map $\mathbf{S}$, so the Gram matrices become

$$
\mathbf{C}_H = \mathbf{S}^\top \mathbf{C}_L \mathbf{S},\quad
\mathbf{G}_H = \mathbf{S}^\top \mathbf{G}_L \mathbf{S},\ \ldots
$$

with $\mathbf{C}_L,\mathbf{G}_L,\ldots$ the usual Kronecker Grams on the finest
level. The SPDE precision formulas and C++ `random_fem_*` kernels are unchanged.
Inspect the layout with `hb_summary(hb_basis(kn, degree))`.

Cost scales with the finest global knot grid; local refinement adds resolution
where needed without refining the entire padded domain.

# Numerical comparison: Matérn vs FEM

Fix $\nu=1$ (SPDE order $\alpha=2$), $\kappa=3$, and marginal $\mathrm{sd}=1$.
The practical range is $\rho=\sqrt{8\nu}/\kappa$. The analytic covariance below uses
the Matérn kernel at that $(\kappa,\mathrm{sd})$. The FEM covariances use the
same $(\rho,\mathrm{sd})$ through [matern_fem_covariance()], which returns
$\mathbf{A}\mathbf{Q}^{-1}\mathbf{A}^\top$ at the geometry in `data`.

```{r params}
nu <- 1
kappa <- 3
sd <- 1
rho <- sqrt(8 * nu) / kappa
c(kappa = kappa, range = rho, sd = sd)
```

Evaluation sites (same for both grids):

```{r sites}
knots_coarse <- list(
  x = seq(-0.2, 1.2, by = 0.2),
  y = seq(-0.2, 1.2, by = 0.2)
)
knots_fine <- list(
  x = seq(-1, 2, by = 0.1),
  y = seq(-1, 2, by = 0.1)
)

set.seed(0)
sites <- data.frame(
  x = runif(100),
  y = runif(100)
)
sites$geometry <- as.matrix(sites[c("x", "y")])

matern_cov <- function(h, nu, kappa, sigma2 = 1) {
  h <- pmax(h, 1e-12)
  r <- kappa * h
  sigma2 * (2^(1 - nu) / gamma(nu)) * (r^nu) * besselK(r, nu)
}

D <- as.matrix(dist(sites[c("x", "y")]))
Sigma <- matern_cov(D, nu, kappa, sigma2 = sd^2)
diag(Sigma) <- sd^2
```

FEM covariance at the same sites, for two knot resolutions:

```{r compare}
term_coarse <- matern("geometry", knots = knots_coarse, shape = 1L)
term_fine <- matern("geometry", knots = knots_fine, shape = 1L)
Sig_coarse <- matern_fem_covariance(term_coarse, sites, range = rho, sd = sd)
Sig_fine <- matern_fem_covariance(term_fine, sites, range = rho, sd = sd)

rel_err <- function(S_hat, S) {
  sqrt(sum((S_hat - S)^2)) / sqrt(sum(S^2))
}

data.frame(
  grid = c("coarse (by = 0.2)", "fine (by = 0.1)"),
  frobenius_rel = c(rel_err(Sig_coarse, Sigma), rel_err(Sig_fine, Sigma)),
  max_abs = c(max(abs(Sig_coarse - Sigma)), max(abs(Sig_fine - Sigma))),
  mean_diag = c(mean(diag(Sig_coarse)), mean(diag(Sig_fine)))
)
```

Correlation versus distance (exact Matérn vs FEM):

```{r plot-corr, fig.width=7, fig.height=4}
upper <- upper.tri(D)
df <- data.frame(
  dist = D[upper],
  matern = Sigma[upper] / sqrt(diag(Sigma)[row(Sigma)[upper]] * diag(Sigma)[col(Sigma)[upper]]),
  coarse = Sig_coarse[upper] / sqrt(pmax(diag(Sig_coarse)[row(Sig_coarse)[upper]], 1e-8) *
    pmax(diag(Sig_coarse)[col(Sig_coarse)[upper]], 1e-8)),
  fine = Sig_fine[upper] / sqrt(pmax(diag(Sig_fine)[row(Sig_fine)[upper]], 1e-8) *
    pmax(diag(Sig_fine)[col(Sig_fine)[upper]], 1e-8))
)
df <- df[order(df$dist), ]
plot(df$dist, df$matern,
  type = "l", lwd = 2,
  ylim = c(0, 1),
  xlim = c(0, 1),
  xlab = "distance", ylab = "correlation",
  main = "Matern vs FEM correlations"
)
lines(df$dist, df$coarse, col = "tomato", lty = 2, lwd = 4)
lines(df$dist, df$fine, col = "steelblue", lty = 3, lwd = 2)
legend("topright", c("Matern", "FEM coarse", "FEM fine"),
  col = c("black", "tomato", "steelblue"), lty = c(1, 2, 3), lwd = c(2, 1, 1), bty = "n"
)
```

The finer knot grid tracks the Matérn correlation more closely (especially away from the boundary).

# References

Whittle (1954, 1963); Lindgren, Rue & Lindström (2011); Bolin & Lindgren (CSDA 2013);
Liu, Guillas & Lai (2016); Lindgren, Bolin & Rue (2022); Bakka (arXiv:1803.03765).
