```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')


dataMedSimple =
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01'
    & deaths$tempClean >=0
#    & deaths$ecoRegion %in% 2:6 #setdiff(1:20, c(1,7,19)) # regions 1, 7, 19 have some very low temperatures.
    , ]
save(dataMedSimple, file='.RData')

```




```{r tmbpackage, eval=FALSE}
library(hpoltest)
formula <- count ~ f(date, model = 'iid')  + 
  f(tempClean, model = 'hiwp', ref_value = 20, p=3, 
      knots = seq(0, 45, by=2.5), 
    group_var = ecoRegion)
  fit <- hnlm(formula, dataMedSimple, 
    cc_design =  ccDesign(time_var = 'date',
    strat_vars=c("cell","ecoRegion", "yearWeek3Dow"))
  )
  sdr <- sdreport(fit$obj, getJointPrecision=TRUE)
  mu <- fit$obj$env$last.par

```


```{r fit}
# R CMD INSTALL --clean --preclean hpoltest_0.0.1.tar.gz   --configure-args="--enable-openmp" --configure-vars="PKG_CPPFLAGS='-fopenmp' PKG_LIBS='-fopenmp'"
# devtools::install_github("samperochkin/hpoltest")
# ulimit -s unlimited 


library('hpolcc')

fitD <- hpolcc::hnlm(
  formula= medical ~ #f(date, model='iid') +
#    f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
#      knots = seq(0,32, by=2),
#      group_var = ecoRegion) + 
    f(tempClean, model='hiwp',ref_value = 20, p=3, 
      knots = seq(0, 45, by=5), 
    group_var = ecoRegion), 
  data=dataMedSimple[dataMedSimple$ecoRegion < 7,],
  dirichelet= TRUE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  for_dev=TRUE, verbose=TRUE
  )





fitD$config$transform_theta = TRUE
if(ps::ps_system_memory()$total > 5e10) {
  fitD$config$num_threads = parallel::detectCores()
} else {
  fitD$config$num_threads = 4
}


  Nbeta = nrow(fitD$tmb_data$XTp)
  Ngamma = nrow(fitD$tmb_data$ATp)

  Ntheta = length(fitD$theta_info$name)

  parameters = c(
    0.0, 0.01, # beta (slope, quadratic) 
    1, # sd global
    0.1, # sd local
    0.001, 0.001, # sd local slope, quad
    0.1[fitD$config$dirichelet] # dirichelet OD
    )

 gamma_start = runif(Ngamma, -0.01, 0.01) #mapOptim$solution[seq(Nbeta+1, len=Ngamma)]

    config = fitD$config
    config$transform_theta = FALSE
    config$verbose=FALSE

if(FALSE) {
    
 
             data=fitD$tmb_data
    control = list(maxit=10, start.trust.radius = 0.1,
      report.level=4, report.freq=1)

 x=c(parameters[1:Nbeta], gamma_start, parameters[-(1:Nbeta)])

}


# likelihood
# parameters = (beta, theta)
# config$transform_theta = TRUE means thetas are log(sd) otherwise theta=sd
result = loglik(parameters, 
             gamma_start, 
             data=fitD$tmb_data, 
    config = config,
    control = list(maxit=100, start.trust.radius = 0.1,
      report.level=4, report.freq=1))

result$minusLogLik
result$gradL

step = 0.001
result2 = loglik(
  parameters + c(step, rep(0, length(parameters)-1)),
  config = result$config,
  gamma_start = result$solution,
  data = fitD$tmb_data,
    control = list(maxit=100, start.trust.radius = 0.1,
      report.level=4, report.freq=1)
  )

(dpart = c(result$fval, result2$fval))
diff(dpart)/step
result$extra$grad[1]

(ldet=c(result$logDetHessian, result2$logDetHessian))
diff(ldet)/step
result$extra$d1[1]
result$extra$d2[1]

(llik=c(result$minusLogLik, result2$minusLogLik))
diff(llik)/step
result$gradL[1]

```


```{r checkDerivs}



fitD$config$maxDeriv = 2
fitD$config$verobse=FALSE
parametersFull = c(
  result$parameters[1:nrow(fitD$tmb_data$XTp)], # beta
  result$solution,  # gamma
  result$parameters[-(1:nrow(fitD$tmb_data$XTp))] # theta
  )


wrappers <- hpolcc::make_trustoptim_wrappers(
  data=fitD$tmb_data,
  config=list(verbose=FALSE, transform_theta = TRUE, 
    dirichelet = fitD$config$dirichelet, 
    num_threads = fitD$config$num_threads),
  debug=FALSE
)

newmat = Matrix::Matrix(0, 
  nrow=nrow(result$extra$denseHessian), ncol=nrow(result$extra$denseHessian),
  sparse=TRUE)
newmat[seq(from=Nbeta+1, len=Ngamma), seq(from=Nbeta+1, len=Ngamma)] = result$hessian
newmat[1:8,1:8]
result$extra$denseHessian[1:8,1:8] 
wrappers$hs(parametersFull)[1:8,1:8]

thirdDiag = 2*(result$extra$denseDiag - result$extra$denseHessian)


# thirdDiag[a,b] = T_abb
result$extra$denseDiag[1:4, 1:9]
thirdDiag[1:4, 1:9]


# test third

indexForThird = length(parametersFull)
forThirdGamma1 = forThirdGamma2 = 5

stepsize = 0.0001
SforDeriv = seq(-3,3)*stepsize + parametersFull[indexForThird]

Sparams = matrix(parametersFull, length(parametersFull), length(SforDeriv))
Sparams[indexForThird,] = SforDeriv
Sparams = as.data.frame(Sparams)

hsEval = mapply(
  wrappers$hs,
  Sparams
  )

# 2nd deriv a v H_bb(a) 
Shs = unlist(lapply(hsEval, function(xx) xx[forThirdGamma1, forThirdGamma2]))
plot(SforDeriv, Shs, type='o')


newx = seq(SforDeriv[1]+stepsize, max(SforDeriv)-stepsize, len=length(SforDeriv)-1)
numthird = diff(Shs)/stepsize

plot(newx, numthird, type='o')


abline(h=thirdDiag[indexForThird, forThirdGamma1])


(diff(Shs)/stepsize)/2 + approx(SforDeriv, Shs, newx)$y
result$extra$denseDiag[indexForThird, forThirdGamma1]
result$extra$denseDiag[forThirdGamma1, indexForThird]
result$extra$denseDiag[indexForThird, forThirdGamma1] 


(diff(Shs)/stepsize)/2 
approx(SforDeriv, Shs, newx)$y

result$extra$denseDiag[indexForThird, forThirdGamma1]-result$extra$denseHessian[indexForThird, forThirdGamma1] 

2*(result$extra$denseDiag[indexForThird, forThirdGamma1] - result$extra$denseHessian[indexForThird, forThirdGamma1]) 


start_val = parametersFull
# checking derivatives
fnSub = function(xx) wrappers$fn(c(
  start_val[
    seq(1, length(start_val)-length(xx))
    ], xx
    )
  )

StoDiff = seq(to=length(start_val), len=7)
testGrad = numDeriv::grad(fnSub, 
  start_val[StoDiff]
  )
(xx = rbind(
  autodiff=wrappers$gr(start_val)[StoDiff],
  num = testGrad
))

autoHessian = wrappers$hs(start_val)

grSub = function(xx) wrappers$gr(
  c(
  start_val[
    seq(1, length(start_val)-length(xx))
    ], xx
    )
    )

testHessian = numDeriv::jacobian(grSub, start_val[StoDiff])[StoDiff,]
options(width=200)
rbind(
  testHessian, 
  NA, 
  autoHessian[StoDiff, StoDiff]
)



plot(SforDeriv, fnEval)

plot(seq(SforDeriv[1]+stepsize, max(SforDeriv)-stepsize, len=length(SforDeriv)-1), 
  diff(fnEval)/stepsize, type='o')
(theG = wrappers$gr(parametersFull)[indexForThird])  
abline(h=theG, lty=2)


plot(SforDeriv[-c(1, length(SforDeriv))], 
  diff(diff(fnEval))/stepsize^2, type='o')
abline(h=wrappers$hs(parametersFull)[indexForThird,indexForThird], lty=2)



hsEval = mapply(
  wrappers$hs,
  Sparams
  )

forThirdGamma1 = 1+Nbeta
forThirdGamma2 = 1+Nbeta




str(result$extra$third)

thirdHere = result$third[
  ( (result$third$paramInParams+1) == indexForThird),
  ]


diagHere = result$extra$diag[,unlist(thirdHere[1,'param'])+1 ]

# Tijj + T_ikk
diagHere = Matrix::forceSymmetric(outer(diagHere, diagHere, FUN='+'))

# 2 T_{ijk} + T_ijj + T_ikk
taylor3mat = Matrix::sparseMatrix(
  i = thirdHere$iInParams, j = thirdHere$gammaInParams,
  x = thirdHere$taylor3, index1=FALSE, symmetric=TRUE,
  dims = dim(diagHere)
  )

diag_masked <- diagHere * (taylor3mat != 0)
Tderiv = (taylor3mat - diag_masked)/2

Shs = unlist(lapply(hsEval, function(xx) xx[forThirdGamma1, forThirdGamma2]))
plot(SforDeriv, Shs, type='o')

plot(seq(SforDeriv[1]+stepsize, max(SforDeriv)-stepsize, len=length(SforDeriv)-1), 
  diff(Shs)/stepsize, type='o')

abline(h=thirdDiag[forThirdGamma1,indexForThird])




```


```{r mapEstimator, eval=FALSE}
if(FALSE) {
# map estimator  
library('trustOptim')

mapOptim <- trust.optim(
  x = start_val,
  fn = wrappers$fn,
  gr = wrappers$gr,
  hs = wrappers$hs,
  method = "Sparse",
  control = list(
    start.trust.radius = 50, maxit=150, trust.iter = 5000,
    preconditioner=1, report.level=4, report.freq=1)
)

mapOptim$solution[1:Nbeta]
exp(mapOptim$solution[-seq(1, Nbeta + Ngamma)])

}


Nsim = 200
gammaSim =  res$gamma_hat + 
  cholVariance %*% matrix(
  rnorm(Nsim*nrow(cholVariance)),
  nrow(cholVariance), Nsim)


# only plot the global effects
rowsToKeep = grep("temp.*global", rownames(fitD$tmb_data$ATp))

Stemp = seq(0,41,by=0.5)
SobsTemp = na.omit(match(Stemp, round(fitD$data$tempClean, 1)))

par(mar=c(3,3,0,0), bty='n')
matplot(
  Stemp,
  crossprod(
    fitD$tmb_data$XTp[, SobsTemp],
    parameters[1:Nbeta]
    ) + 
  crossprod(
    fitD$tmb_data$ATp[rowsToKeep, SobsTemp], 
    gammaSim[rowsToKeep,]),
  type='l', col='#00000020', lty=1, xaxs='i',
  yaxs='i'
)

```




```{r third_deriv_stuff}
# note wrappers$hs must have been run 
parametersGamma = hpolcc::sparsityForThird(
  hessian = environment(wrappers$fn)$cache_env$config1$sparsity, 
    data = fitD$tmb_data)

config = c(
  parametersGamma,
  fitD$config[setdiff(
      names(fitD$config), 
      names(parametersGamma)
    )]
)
config$verbose=TRUE

resThird = hpolcc:::derivForLaplace(
  parameters=start_val, data=fitD$tmb_data, config
  ) 
str(resThird)

resThird$ddw = matrix(resThird$ddw, ncol=2, byrow=TRUE)
resThird$taylor3 = matrix(resThird$taylor3, ncol=3, byrow=TRUE)

head(resThird$ddw)
head(resThird$taylor3)
testHessian[,3]
testGrad


resThird$forDiag[1:150]




```



```{r theResults}
result$parameters = hpolcc::formatParameters(
  result$solution, fitD, TRUE)

result$parameters$theta

# parameters are (beta, gamma, theta)

randomEffects = seq(from=length(result$parameters$beta)+1,
  length = nrow(result$parameters$gamma))

result$Htilde = result$hessian[randomEffects, randomEffects]
result$HtildeChol = Matrix::solve(Matrix::chol(result$Htilde))

Nsim = 200
simInd = matrix(rnorm(Nsim * nrow(result$Htilde)),
 nrow(result$Htilde), Nsim)

# sample from [U|Y, theta_map]
simGamma = Matrix::crossprod(result$HtildeChol,  simInd) 
rownames(simGamma) = rownames(fitD$tmb_data$ATp)

simGamma = simGamma + unlist(result$parameters$gamma[
  rownames(simGamma), 'value'])


Stemp = seq(0,45,by=0.1)
SobsTemp = na.omit(match(Stemp, round(fitD$data$tempClean, 1)))
Spm = seq(1,1000, by=1)
SobsPm = na.omit(match(Spm, round(fitD$data$pm)))

Sobs = SobsTemp
nameX = 'tempClean'

Bmat = fitD$tmb_data$ATp[, Sobs]
# only want the global effects
Bmat = Bmat[grep(paste0(nameX,".*global"), rownames(Bmat)),]


simF = crossprod(Bmat, simGamma[rownames(Bmat),]) 

covHere = grep(nameX, names(result$parameters$beta), value=TRUE)
fixedPart = drop(crossprod(fitD$tmb_data$XTp[covHere,Sobs, drop=FALSE], 
                           result$parameters$beta[covHere]))

simSum = simF + fixedPart 



theX= fitD$data[[covHere[1] ]][Sobs]
if(grepl("^sqrt", nameX)) theX = theX^2

par(mar=c(3,3,0,0), bty='l')
matplot(theX, simSum, type='l',
  col='#00000010', lty=1, xaxs='i')


```



```{r otherstuff22, eval=FALSE}

Sstart = 1:(nrow(fitD$tmb_data$XTp)+1)
Send = seq(to = length(fitD$start_parameters), by=1, 
  len=length(fitD$theta_info$name)+1)

bob = function(xx) {
  hpolcc:::objectiveFunctionC( 
c(xx[Sstart], fitD$start_parameters[-c(Sstart, Send)], 
  xx[-Sstart]),
 fitD$tmb_data, 
 c(
 fitD$config[setdiff(names(fitD$config), c('verbose','maxDeriv'))], 
  list(maxDeriv = 0, verbose=FALSE))
  )$value
}

round(numDeriv::hessian(bob, fitD$start_parameters[c(Sstart, Send)]), 3)

res$hessian[c(Sstart, Send), c(Sstart, Send)]



beta = fitD$start_parameters[1:nrow(fitD$tmb_data$XTp)]
gamma =fitD$start_parameters[seq(from=length(beta)+1, 
  len=nrow(fitD$tmb_data$ATp)
  )]
theta = fitD$start_parameters[seq(from = length(beta) + length(gamma)+1,
  to=length(fitD$start_parameters))]


resFun = function(xx)  hpolcc:::objectiveFunctionC(xx, fitD$tmb_data, 
  c(fitD$config[
    setdiff(names(fitD$config), 'maxDeriv')], 
  list(maxDeriv = 0)))$value


resFun(fitD$start_parameters)  

Matrix::determinant(res$hessian)



fitD$config$verbose=FALSE
res$grad
theGradAgain = numDeriv::grad(resFun, fitD$start_parameters)  

rbind(res$grad, theGradAgain)



theHessAgain = numDeriv::hessian(resFun, fitD$start_parameters)


```

```{r originalPackage, eval=FALSE}
# using the hpoltest original package (and TMB)

fitM <- hnlm(
  formula= y ~ f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 22, p=3, knots = seq(0, 42, by=2), group_var = ecoRegion), 
  data=dataMedSimple,
  dirichlet=FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  tmb_parameters = list(theta = thetaStartM),
  optim_parameters = list(maxit = 20),
  verbose=TRUE, optimizer='optim', intern=FALSE, inner.method = 'BFGS', 
  inner.control = list(maxit = 500), for_dev=FALSE)
```

```{r plotFit2, fig.height=4, fig.width=4, eval=FALSE}
simTemp =  hpoltest:::condSim(fitD, 'tempClean') 

theYlim = c(0.2, 5) #exp(quantile(simTemp$y[,'GLOBAL', ], c(0.01, 0.99)))
for(D in dimnames(simTemp$y)[[2]]) {
matplot(simTemp$x, exp(simTemp$y[,D, ]), type='l', col='#00000010', lty=1, 
        xaxs='i', log='y',
        ylim =theYlim,bty='n', 
        xlab='temperature',ylab=D)
lines(simTemp$x, exp(simTemp$fixed_mean), col='yellow')
lines(simTemp$x, exp(apply(simTemp$y[,D, ], 1, mean)), col='red')
}
```

```{r plotDate, eval=FALSE}
simDate = hpoltest:::condSim(fitD, 'date') 

matplot(as.Date(simDate$x), t(apply(simDate$y, 1, quantile, c(0.025, 0.975, 0.5))), col=c('grey','grey','black'), cex=0.3, pch=16,
        xlab='date', ylab='logrr')

```


```{r debug, eval=FALSE}

xx = fitD$obj$env$last.par.best
xx = xx[grep("gamma", names(xx))]
names(xx) = colnames(fitD$obj$env$data$A)
xxDate = xx[grep("date", names(xx))]
xx = xx[grep("date", names(xx), invert=TRUE)]

```


```{r digitPreference, eval=FALSE}
checkForDigitPreference1 = aggregate(deaths[,'medical', drop=FALSE],
  deaths[,c('dom','month')], sum)
checkForDigitPreference = dcast(as.data.table(checkForDigitPreference1), dom ~ month)
image(as.numeric(checkForDigitPreference$dom),
 as.numeric(colnames(checkForDigitPreference)[-1]),
  as.matrix(checkForDigitPreference[,-1]))
```

```{r indiaDataAlex, eval=FALSE}
load("/home/patrick/research/hpoltest/applications/india/combinedindianmorttempdata_meannofilter.Rdata")
x = indianmorttempdata

table(x$death_cause, x$case)
quantile(x$final_temp)
tempSdByStrata = tapply(x$final_temp, x$riskset, sd)
countByStrata = tapply(x$case, x$riskset, sum)
# nuber cases per day
table(tapply(x$case, x$date, sum))


tempRef = 32

x$tempHigh = pmax(x$final_temp - tempRef, 0)
x$tempLow = pmin(x$final_temp - tempRef, 0)

x$tempC = x$final_temp - tempRef
x$tempSq = (x$final_temp - tempRef)^2
x$temp3 = (x$final_temp - tempRef)^3

x$dateInt = as.integer(x$date)

xSub = x[x$death_cause == 'stroke', ]
Slow = unique(unlist(xSub[xSub$final_temp < 10, 'riskset']))

Shigh = unique(unlist(xSub[xSub$final_temp >37, 'riskset']))



xSub2 = xSub[! xSub$riskset %in% c(Slow, Shigh), ]

library('survival')
fitClogit = clogit(case ~ tempHigh + tempLow + strata(riskset),
       data = x[x$death_cause == 'stroke',])
summary(fitClogit)$coef


#library('hpoltest')
fitLinear <- hnlm(
  formula = case ~  tempHigh + tempLow + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = 20, map = list(theta=factor(NA))),
  data = xSub2,
  verbose=FALSE
  )

fitLinear$est$param$est

randomDate = as.Date(
  as.integer(gsub("factor.dateInt.", "", names(fitLinear$est$random$est))),
  as.integer(as.Date('1970/1/1')))
  
badDates = randomDate[order(abs(fitLinear$est$random$est), decreasing=TRUE)[1:2]]
stuff = as.data.frame(xSub[xSub$date %in% badDates, c('riskset', 'date','final_temp','case')])
stuff = stuff[order(stuff$date,stuff$riskset), ]
stuff


plot(randomDate, fitLinear$est$random$est)

fitLinear2 <- hnlm(
  formula = case ~  tempHigh + tempLow + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = xSub2,
  verbose=FALSE
  )

fitLinear$est$param$est
fitLinear2$est$param$est


fitQuad  <- hnlm(
  formula = case ~  tempC + tempSq + temp3 + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = -5, map = list(theta=factor(NA))),
  data = xSub2,
  verbose=FALSE
  )
fitLinear$est$param$est
fitLinear2$est$param$est
fitQuad$est$param$est
```

```{r comparePlots, eval=FALSE}
tempSeq = seq(10, 40)
tempSeqC = tempSeq - tempRef

plot(tempSeq,
     tempSeqC*fitQuad$est$param$est['tempC'] + 
     tempSeqC^2*fitQuad$est$param$est['tempSq'] +
     tempSeqC^3*fitQuad$est$param$est['temp3'],
     type='l'
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitClogit$coef['tempLow'] + 
      pmax(tempSeqC, 0) * fitClogit$coef['tempHigh'],
      col='green', lwd=3, lty=1
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitLinear$est$param$est['tempLow'] + 
      pmax(tempSeqC, 0) * fitLinear$est$param$est['tempHigh'],
      col='red', lwd=3, lty=2
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitLinear2$est$param$est['tempLow'] + 
      pmax(tempSeqC, 0) * fitLinear2$est$param$est['tempHigh'],
      col='orange', lwd=3, lty=2
)

legend('bottomleft', bty='n', lty=1, col=c('black','red','orange','green'),
       legend = c('cubic', 'linear no OD','linear','clogit'))

```

```{r fitSimple, eval=FALSE}
tempKnots = seq(0,40, by=0.5)
fitSimple <- hnlm(
  formula = case ~  f(final_temp, model='iwp', ref_value = tempRef, p=4, knots = tempKnots) + 
    f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = c(-1,-5), map = list(theta=factor(c(NA,1)))),
  data = xSub2,
  verbose=FALSE
  )

fitSimple$est$param$est 
fitQuad$est$param$est
```






```{r fitTemperatureAlex, include=FALSE, eval=FALSE}

tempKnots = seq(-5,40,by=2.5)
library('hpoltest')
fitAsAlex <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
fitIwp3NoOd <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=3, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )

fitIwp3 <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=3, knots = tempKnots) +
     f(date, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=FALSE),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5),
  verbose=FALSE
  )
fitIwp2 <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots) +
     f(date, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
fitAsAlex <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
```




```{r moreSimlpeModels, eval=FALSE}
  
dataMedSimplePm = removeUnusedStrata(
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01' &
    !deaths$ecoRegion %in% c(7, 8,10), ],
  Sstrata, outcome)
formulaMedSimplePm <- y ~ dom + f(date, model='iid') +
    f(sqrtPm, model='iwp',ref_value = 25, p=3, knots = pmKnots)

fitPm<- hnlm(
  formula=formulaMedSimplePm, 
  data=dataMedSimplePm,
#  tmb_parameters = list(theta = c(5.25,31)),
  cc_design =  ccDesign(strat_vars = Sstrata),
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE, for_dev=FALSE 
  )

#  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 
#  f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var = ecoRegion)

formulaCvd = cvd ~ dom + f(date, model='iid') + f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var = ecoRegion)

formulaMedSenior <- medical_senior ~ dom + f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var=ecoRegion)
#  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 


formulaCvdSimple = cvd ~ dom + f(date, model='iid') + 
  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 
#f(tempClean, model='iwp',ref_value = 25, p=3, knots = tempKnots)

formulaMedSeniorSimple = medical_senior ~ dom + f(date, model='iid') + 
  f(tempClean, model='iwp',ref_value = 25, p=3, knots = tempKnots)

    dataMed = deaths[which(deaths$N_Week3_medical > 0 & deaths$date > as.Date('2009/1/1')),]

        dataMedSenior = deaths[which(deaths$N_Week3_medical_senior > 0 & 
                                   deaths$date > as.Date('2006/1/1')),]
dataCvd = deaths[which(deaths$N_Week3_cvd > 0 & deaths$date > as.Date('2006/1/1')), ]

if(FALSE) {

fitMedSeniorSimple = hnlm(formulaMedSeniorSimple, dataMedSenior, cc_design, 
                            optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
                          verbose=TRUE)

fitCvdSimple<- hnlm(
  formula=formulaCvdSimple, 
  data=dataCvd,
#  tmb_parameters = list(theta = c(5.25,31)),
  cc_design = cc_design,
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE, for_dev=FALSE 
  )
saveRDS(fitCvdSimple, file.path(storeDir, 'cvdSimple.rds'))

fitCvdSetup <- hnlm(
  formula=formulaCvd, 
  data=dataCvd,
  cc_design = cc_design,
  verbose=TRUE, for_dev=TRUE 
  )

fitCvdSetup$tmb_parameters$beta = fitCvdSimple$est$param$est[1:length(fitCvdSetup$tmb_parameters$beta)]
thetaHat = fitCvdSimple$est$param$est[grep("theta", names(fitCvdSimple$est$param$est))]

fitCvdSetup$tmb_parameters$theta[1:length(thetaHat)] = thetaHat
fitCvdSetup$tmb_parameters$theta[-(1:length(thetaHat))] = 0
fitCvdSetup$tmb_parameters$gamma[1:length(fitCvdSimple$est$random$est)] = fitCvdSimple$est$random$est

  obj <- MakeADFun(data = fitCvdSetup$tmb_data,
                   parameters = fitCvdSetup$tmb_parameters,
#                   random = 'gamma',
                   map = list(
                    beta = factor(rep(NA, length(fitCvdSetup$tmb_parameters$beta))),
                    theta=factor(rep(NA, length(fitCvdSetup$map$theta)))),
                   intern=FALSE, type='ADFun',
                   DLL = "hpoltest")

  parCat = rep(NA, ncol(obj$env$.data$A))
  parCat[grep("date", colnames(obj$env$.data$A))]='date'
  parCat[grep("hrpoly", colnames(obj$env$.data$A))]='poly'
  parCat[grep("hiwp", colnames(obj$env$.data$A))]='hiwp'
  parCat[grep("global", colnames(obj$env$.data$A))]='global'

  the19 = grep('hiwp_19', colnames(obj$env$.data$A), value=TRUE)

  obj$fn(obj$par)
    opt <- optim(
      par = rnorm(length(obj$par), obj$par, 1e-2),
      fn = obj$fn, 
      gr = obj$gr,
      method = 'BFGS',
      control = list(
        trace=TRUE, 
        maxit=1e5,
        parscale = c(date=1,global=1e-5, hiwp=1e-2, poly=1e-2)[parCat]
    )
  )


            names(opt$par) = colnames(obj$env$.data$A)

                opt$par1 = opt$par[grep("date", names(opt$par))]
    opt$par2 = opt$par[!names(opt$par)%in% names(opt$par1)]

    quantile(obj$gr(opt$par))

        
fitMedSenior <- hnlm( 
  formula=formulaMedSenior, data=dataMedSenior, #[which(deaths$date > as.Date('2009/1/1')), ], 
  cc_design = cc_design,verbose=TRUE, for_dev=FALSE,
    optim_parameters = list(eval.max=1e4, iter.max=1e4),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE)
  )


  obj <- MakeADFun(data = fitMed1$tmb_data,
                   parameters = fitMed1$tmb_parameters[c('beta','gamma','theta')],
                   random = 'gamma',
                   map = fitMed1$map,
                   intern=FALSE, type='ADFun',
                   DLL = "hpoltest")
  obj$fn(obj$par)
    opt <- optim(par = obj$par, 
      fn = obj$fn, gr = obj$gr,
      control = list(trace=TRUE)
  )
saveRDS(fitMed, file.path(storeDir, 'ccMed.rds'))
fitCvd <- hnlm(
  formula=formulaCvd, 
  data=dataCvd,
  cc_design = cc_design,
#  optim_parameters = list(eval.max=100, iter.max=5),
  verbose=TRUE, for_dev=FALSE 
  )
saveRDS(fitCvd, file.path(storeDir, 'ccCvd.rds'))
}
fitCvd = readRDS(file.path(storeDir, 'ccCvd.rds'))
```

```{r temperatureDataFromIndia, eval=FALSE}
deathsTemp = as.data.table(readRDS( file.path(storeDir, "indiaTemperature.rds")))
deathsTemp$dom = as.numeric(format(deathsTemp$date, '%d'))
deathsTemp[deathsTemp$dom == Hmisc::monthDays(deathsTemp$date), 'dom'] = 31
deathsTemp[format(deathsTemp$date, '%m%d') == '0101', 'dom'] = 0
deathsTemp$dom = relevel(factor(deathsTemp$dom), '1')
deathsTemp$month = format(deathsTemp$date, '%m')


(theEtable = sort(table(deathsTemp$ecoRegion), decreasing=TRUE))
# 10, 17, 20 looks dodgy



Sstrata = c("cell","ecoRegion", "yearWeek3Dow")
knotsTemperature = seq(0,50, by=2.5)  

outcome = 'medical'

deathsTempMed = removeUnusedStrata(deathsTemp[
  deathsTemp$date >= as.Date('2006/1/1') &
  format(deathsTemp$date, '%b%d') != 'Jan01', ],
  Sstrata, outcome)


deathsTempMed$tempCut = 2*round(deathsTempMed$maxtemp/2)
ttable1 = dcast(deathsTempMed, tempCut ~ ecoRegion, value.var=outcome, fun.aggregate=sum)
ttable2 = dcast(deathsTempMed, tempCut ~ ecoRegion, value.var=outcome, fun.aggregate=length)

matplot(ttable1[,1], as.matrix(ttable1[,-1]/ttable2[,-1])[,names(theEtable)[10+1:9]])
merge(
  deathsTempMed, 
    as.data.frame(deathsTempMed)[order(deathsTempMed$y, decreasing=TRUE)[1:3], Sstrata],
    all.x=FALSE, all.y=TRUE)

deathsTempMed$tempLow = pmin(25, deathsTempMed$maxtempLag01)
deathsTempMed$tempHigh = pmax(25, deathsTempMed$maxtempLag01)

SexcludeRegions  = c(7, 10,13)
SincludeRegions = c(1:6, 8,9)
library(hpoltest)
fit <- hnlm(
  formula= y ~  dom + 
    f(date, model='iid') +
    f(maxtempLag01, model='hiwp',ref_value = 25, p=3, knots = knotsTemperature, group_var=ecoRegion),
  data=deathsTempMed[deathsTempMed$ecoRegion %in% SincludeRegions,],
  cc_design = ccDesign(strat_vars = Sstrata),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    smartsearch=FALSE, 
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)



formulaMedSeniorIwp = medical_senior ~ dom + f(date, model='iid') + 
  f(maxtemp, model='iwp',ref_value = 25, p=3, knots = knotsTemperature)

fitTempMedSeniorIwp <- hnlm(
  formula=formulaMedSeniorIwp, 
  data=deathsTempSeniorMed3,
  cc_design = ccDesign(strat_vars = c("cell","ecoRegion", "yearWeek3Dow")),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)

fitTempMedIwp <- hnlm(
  formula=formulaMedIwp, 
  data=deathsTempMed,
  cc_design = ccDesign(strat_vars = Sstrata),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)
```


```
# darjeeling
# INTEL oneMKL ERROR: /usr/lib/x86_64-linux-gnu/libmkl_core.so.2: cannot open shared object file: No such file or directory.
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_core.so.2  /usr/lib/x86_64-linux-gnu/libmkl_core.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_gnu_thread.so.2  /usr/lib/x86_64-linux-gnu/libmkl_gnu_thread.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_tbb_thread.so.2  /usr/lib/x86_64-linux-gnu/libmkl_tbb_thread.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_intel_lp64.so.2 /lib/x86_64-linux-gnu/libmkl_intel_lp64.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_intel_lp64.so.2 /lib/x86_64-linux-gnu/libmkl_intel_lp64.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_def.so.2 /lib/x86_64-linux-gnu/libmkl_def.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_avx512.so.2 /lib/x86_64-linux-gnu/libmkl_avx512.so.2
```
