```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')


dataMedSimple =
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01'
    & deaths$tempClean >=0
#    & deaths$ecoRegion %in% 2:6 #setdiff(1:20, c(1,7,19)) # regions 1, 7, 19 have some very low temperatures.
    , ]
save(dataMedSimple, file='dataMedSimple.RData')
# load("dataMedSimple.RData")
```



```{r fit}
# R CMD INSTALL --clean --preclean hpoltest_0.0.1.tar.gz   --configure-args="--enable-openmp" --configure-vars="PKG_CPPFLAGS='-fopenmp' PKG_LIBS='-fopenmp'"
# devtools::install_github("samperochkin/hpoltest")
# ulimit -s unlimited 


library('hpolcc')

fitD <- hpolcc::hnlm(
  formula= medical ~ #f(date, model='iid') +
#    f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
#      knots = seq(0,32, by=2),
#      group_var = ecoRegion) + 
    f(tempClean, model='hiwp',ref_value = 20, p=3, 
      knots = seq(0, 45, by=5), 
    group_var = ecoRegion), 
  data=dataMedSimple[dataMedSimple$ecoRegion < 7,],
  dirichlet= FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  for_dev=TRUE, verbose=FALSE
  )

  parameters = c(
    0.0, 0.01, # beta (slope, quadratic) 
    log(1), # sd global
    log(0.1), # sd local
    log(0.001), log(0.001), # sd local slope, quad
    log(0.1)[fitD$config$dirichlet] # dirichelet OD
    )


fitD$config$transform_theta = TRUE
  fitD$config$num_threads = parallel::detectCores()


  Nbeta = nrow(fitD$tmb_data$XTp)
  Ngamma = nrow(fitD$tmb_data$ATp)

  Ntheta = length(fitD$theta_info$name)


 gamma_start = runif(Ngamma, -0.01, 0.01) #mapOptim$solution[seq(Nbeta+1, len=Ngamma)]

    config = fitD$config
    config$dense = TRUE
    config$transform_theta = TRUE

# likelihood
# parameters = (beta, theta)R
# config$transform_theta = TRUE means thetas are log(sd) otherwise theta=sd

if(exists("result")) gamma_start = result$solution

library('hpolcc')

result = loglik(parameters, 
             gamma_start, 
             data=fitD$tmb_data, 
    config = config,
      control=list(maxit=1000, start.trust.radius = 0.1, prec=1e-12, stop.trust.radius = 1e-12,
        cg.tol = 1e-12,
      report.level=4, report.freq=1)
      )

result$extra
```








```{r checkDeriv}

  DparL = 2
  Sp1L = seq(-0.0002, 0.0002, len=5)+result$parameters[DparL]

  bobL =  mapply(function(xx) {
    parameters2 = result$parameters
    parameters2[DparL] = xx
    loglik(parameters2, gamma_start = result$solution, data=fitD$tmb_data, config=result$config,
      control=list(maxit=1000, start.trust.radius = 0.1, prec=1e-12, stop.trust.radius = 1e-12,
        cg.tol = 1e-12,
      report.level=4, report.freq=1))
    }, xx =Sp1L, SIMPLIFY=FALSE)

bobH = mapply(function(xx) {
    parameters2 = result$fullParameters
    parameters2[DparL] = xx  
    hpolcc:::objectiveFunctionC(parameters2, fitD$tmb_data, result$config)
  }, xx =Sp1L, SIMPLIFY=FALSE)

bobT = mapply(function(xx) {
    parameters2 = result$fullParameters
    parameters2[DparL] = xx  
    thirdDeriv(x=parameters2, data=fitD$tmb_data, config=result$config)
  }, xx =Sp1L, SIMPLIFY=FALSE)


plot(Sp1L, unlist(lapply(bobH, function(xx) xx$value)))

plot(Sp1L, unlist(lapply(bobH, function(xx) xx$grad[DparL])), col='red', type='o', lwd=4, pch=16)
points(Sp1L, unlist(lapply(bobT, function(xx) xx$first[DparL])), col='blue', pch=3)
points(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(unlist(lapply(bobH, function(xx) xx$value)))/diff(Sp1L))


DparL2 = 9;DparL3 = 3 # this one is very different

plot(Sp1L, unlist(lapply(bobH, function(xx) xx$denseHessian[DparL, DparL2])), col='red', type='o', lwd=4, pch=16)
points(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(unlist(lapply(bobH, function(xx) xx$grad[DparL2])))/diff(Sp1L))
points(Sp1L, unlist(lapply(bobT, function(xx) xx$fullHessian[DparL, DparL2])), pch=4, col='blue')




bob0 = unlist(lapply(bobL, function(xx) xx$minusLogLik))
Sdet = unlist(lapply(bobL, function(xx) xx$logDetHessian))

uHat = do.call(cbind, lapply(bobL, function(xx) xx$solution))
dUhat = t(apply(uHat, 1, diff))/diff(Sp1L[1:2])

  Nbeta = nrow(fitD$tmb_data$XTp)
  Ngamma = nrow(fitD$tmb_data$ATp)
Sgamma1 = seq(Nbeta+1, len=Ngamma)

dUhatAd = do.call(cbind,lapply(bobL, function(xx) {
  invH = solve(xx$fullHessian[Sgamma1,Sgamma1])
  - invH %*% xx$fullHessian[Sgamma1,DparL]
  }))


Dgamma=7
plot(Sp1L, dUhatAd[Dgamma,], col='red', type='o', lwd=4, pch=16)
points(
  Sp1L[-1] - diff(Sp1L)/2,
  dUhat[Dgamma,])

DH = DHtheta = list()
for(D in 1:(length(bobL)-1)) {
  DH[[D]] = (bobL[[D+1]]$fullHessian[Sgamma1, Sgamma1] - bobL[[D]]$fullHessian[Sgamma1, Sgamma1]) / diff(Sp1L[seq(D, len=2)])
  DHtheta[[D]] = Matrix::forceSymmetric(bobH[[D+1]]$denseHessian[Sgamma1, Sgamma1] - bobH[[D]]$denseHessian[Sgamma1, Sgamma1] ) / diff(Sp1L[seq(D, len=2)])
}

DHadTheta = DHad = list()
for(Dx in 1:length(Sp1L)) {
  DHad[[Dx]] = 0
  DHadTheta[[Dx]] = as.matrix(bobL[[Dx]]$third[[DparL]])[Sgamma1, Sgamma1]
for(Dgamma in 1:length(Sgamma1)) {
  There = bobL[[Dx]]$third[[ Sgamma1[Dgamma] ]][Sgamma1, Sgamma1]
  incHere =  There * dUhatAd[Dgamma,Dx]
  DHad[[Dx]] = DHad[[Dx]] + incHere
}
DHad[[Dx]] = DHad[[Dx]] +  DHadTheta[[Dx]] 
}



stuff = DHtheta[[2]] - (DHadTheta[[2]]+DHadTheta[[3]])/2
options(width=140)
DHtheta[[2]][1:8,1:8]
DHadTheta[[2]][1:8,1:8]
DHadTheta[[3]][1:8,1:8]
stuff[1:8,1:8]


DparL2 = 9;DparL3 = 3 # this one is very different

bobH[[1]]$denseHessian[DparL2, DparL3]
bobH[[2]]$denseHessian[DparL3, DparL2]
bobT[[3]]$third[[DparL]][DparL2, DparL3]
bobT[[3]]$third[[DparL3]][DparL2, DparL]
bobT[[3]]$third[[DparL2]][DparL, DparL3]

# U constant v remaxed hessian
plot(Sp1L, unlist(lapply(bobH, function(xx) xx$denseHessian[DparL2, DparL])), col='red', type='o', lwd=4, pch=16)
#lines(Sp1L, unlist(lapply(bobL, function(xx) xx$fullHessian[DparL2, DparL3])), col='blue')
lines(Sp1L, unlist(lapply(bobT, function(xx) xx$fullHessian[DparL2, DparL])), col='green')
points(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(unlist(lapply(bobT, function(xx) xx$first[DparL2])))/diff(Sp1L), 
  col='blue')

# U constant v remaxed hessian
plot(Sp1L, unlist(lapply(bobH, function(xx) xx$denseHessian[DparL3, DparL])), col='red', type='o', lwd=4, pch=16)
#lines(Sp1L, unlist(lapply(bobL, function(xx) xx$fullHessian[DparL2, DparL3])), col='blue')
lines(Sp1L, unlist(lapply(bobT, function(xx) xx$fullHessian[DparL3, DparL])), col='green')
points(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(unlist(lapply(bobT, function(xx) xx$first[DparL3])))/diff(Sp1L), 
  col='blue')




# deriv of u constant hessian
plot(Sp1L, unlist(lapply(bobT, function(xx) xx$fullHessian[DparL2, DparL3])), pch=4, col='blue')



plot(Sp1L, unlist(lapply(bobT, function(xx) xx$thirdList[[DparL]][DparL2, DparL3])), pch=4, col='blue')


plot(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(unlist(lapply(bobH, function(xx) xx$denseHessian[DparL2, DparL3])))/diff(Sp1L), 
  col='red', type='o', lwd=4, pch=16
)
points(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(unlist(lapply(bobT, function(xx) xx$fullHessian[DparL2, DparL3])))/diff(Sp1L), 
  col='blue')


 unlist(lapply(bobT, function(xx) xx$thirdList[[DparL]][DparL2, DparL3]))
  unlist(lapply(bobT, function(xx) xx$thirdList[[DparL2]][DparL, DparL3]))
   unlist(lapply(bobT, function(xx) xx$thirdList[[DparL3]][DparL, DparL2]))


abline(h=c(
result$third[[DparL]][DparL2, DparL3],
result$third[[DparL3]][DparL2, DparL],
result$third[[DparL2]][DparL3, DparL]), 
lwd=c(3:1), col=3:1)



# hessian wrt theta, u remax
plot(Sp1L, unlist(lapply(bobL, function(xx) xx$fullHessian[DparL2, DparL3])), col='blue')

plot(Sp1L, unlist(lapply(DHad, function(xx) xx[which(Sgamma1 ==DparL2), which(Sgamma1==DparL3)])), 
  col='red', pch=16)
points(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(unlist(lapply(bobL, function(xx) xx$fullHessian[DparL2, DparL3])))/diff(Sp1L), 
  col='blue'
)


plot(Sp1L, unlist(lapply(DHadTheta, function(xx) xx[DparL2, DparL23])), type='o', col='red', pch=16,
  ylim = c(32, 35)*1000)
points(
  Sp1L[-1] - diff(Sp1L)/2,
  unlist(lapply(DHtheta, function(xx) xx[DparL2, DparL3]))
)


stuff = DH[[1]] - (DHad[[1]] + DHad[[2]])/2
stuff[1:5,1:5]

plot(Sp1L, unlist(lapply(DHad, function(xx) xx[DparL2, DparL3])), type='o', col='red', pch=16)
points(
  Sp1L[-1] - diff(Sp1L)/2,
  unlist(lapply(DH, function(xx) xx[DparL2, DparL3]))
)

Ddet = c()
for(Dx in 1:length(Sp1L)) {
  Ddet[Dx] = sum(Matrix::diag(
    solve(bobL[[Dx]]$fullHessian[Sgamma1, Sgamma1]) %*% DHad[[Dx]][Sgamma1, Sgamma1]
    ))
}

plot(Sp1L, Sdet)

plot(Sp1L, Ddet, pch=16, col='red')
points(
  Sp1L[-1] - diff(Sp1L)/2,
  diff(Sdet)/diff(Sp1L))



  invH = solve(result$fullHessian[Sgamma1,Sgamma1])
  dUhat = - invH %*% result$fullHessian[Sgamma1, -Sgamma1]
hhstuff = 0

for(D in 1:length(Sgamma1)) {
  hhstuff = hhstuff + result$third[[Sgamma1[D] ]][Sgamma1, Sgamma1] * dUhat[D, DparL]
}

sum(Matrix::diag(
  invH %*% (
  result$third[[DparL]][Sgamma1, Sgamma1] + hhstuff)
  ))
  

abline(h=result$extra[DparL,'grad'])


plot(
  Sp1L[-1] + diff(Sp1L)/2,
  diff(bob1)/diff(Sp1L))
abline(h=apply(result$extra[DparL,c("W", "VHH")],1, sum))


theT1 = result$third[[DparL]]
theT2 = (bobL[[3]]$fullHessian - bobL[[2]]$fullHessian)/(diff(Sp1L)[3])

theT1[1:7,1:7]
theT2[1:7,1:7]

  Nbeta = nrow(fitD$tmb_data$XTp)
  Ngamma = nrow(fitD$tmb_data$ATp)
Sgamma1 = seq(Nbeta+1, len=Ngamma)
invH = solve(bobL[[3]]$fullHessian[Sgamma1,Sgamma1])

Strace = mapply(function(xx) sum(Matrix::diag(invH %*% bobL[[3]]$third[[xx]][Sgamma1, Sgamma1])),
  xx = 1:length(result$fullParameters))
VhInv = Strace[Sgamma1] %*% invH
# VHH
VhInv %*% bobL[[3]]$fullHessian[Sgamma1,-Sgamma1]
#W
Strace[DparL]
sum(Matrix::diag(invH %*% theT2[Sgamma1,Sgamma1]))

```




```{r checkSimple}
library(hpolcc)
config1 = fitD$config
config1$dense = TRUE
config1$transform_theta = TRUE
config1$maxDeriv = 2
config1$verbose=FALSE

paramsAll = result$fullParameters

#    pairs1 = expand.grid(i=Sparams, j=seq(from=Nbeta, len=Ngamma), p=0, pEnd=0)
Nparams = length(result$fullParameters)
    pairs1 = expand.grid(i=seq(0, len=Nparams), j=seq(0, len=Nparams), p=0, pEnd=0)
    pairs1 = pairs1[pairs1$i != pairs1$j, ]

    pairs1$index = 1:nrow(pairs1)
config1$sparsity = list(third=list(pairs=pairs1))

bobD = thirdDiagonals(
      result$fullParameters, fitD$tmb_data, config1
    ) 

bob3 = thirdOffDiagonals(
      result$fullParameters, fitD$tmb_data, config1
    ) 

  Dpar = 2
Dpar2=3;Dpar3 = 9
Shere = c( Dpar-1 ,Dpar2-1, Dpar3-1)    

pairs3 = pairs1[which(apply(pairs1[,c('i','j')], 1, function(xx) all(xx %in% Shere))),]
config3 = config1
config3$verbose=TRUE
config3$sparsity = list(third=list(pairs=pairs3))

stuff = thirdOffDiagonals(
      result$fullParameters, fitD$tmb_data, config3
    ) 
pairs3a = cbind(pairs3[,c('i','j')], t(stuff))
pairs3b = reshape2::melt(pairs3a, id.vars=c('i','j'), value.name='taylor3', variable.name='k')
pairs3b = pairs3b[(as.numeric(as.character(pairs3b$k))-1) %in% Shere, ]
pairs3b$taylor3

  Sp1 = seq(-0.0002, 0.0002, len=11)+result$fullParameters[Dpar]


  bob =  mapply(function(xx) {
    parameters2 = result$fullParameters
    parameters2[Dpar] = xx
    hpolcc:::objectiveFunctionC(parameters2, fitD$tmb_data, config)
    }, xx =Sp1, SIMPLIFY=FALSE)

bobC = bob[[6]]

bobC$hessian[1:5,1:5]
result$fullHessian[1:5,1:5]
bobL[[3]]$fullHessian[1:5,1:5] 
bobL[[2]]$fullHessian[1:5,1:5] 
bob[[5]]$hessian[1:5,1:5]

bob0 = unlist(lapply(bob, function(xx) xx$value))
bob1 = unlist(lapply(bob, function(xx) xx$grad[Dpar]))


bob1a = unlist(lapply(bob, function(xx) xx$grad[Dpar2]))
bob1b = unlist(lapply(bob, function(xx) xx$grad[Dpar3]))
bob2 = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar]))
bob2a = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar2]))
bob2c = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar3]))
bob2b = unlist(lapply(bob, function(xx) xx$hessian[Dpar2,Dpar3]))

diffH = list()
for(D in 1:(length(bob)-1)) {
  diffH[[D]] = (bob[[D+1]]$hessian -bob[[D]]$hessian) / (diff(Sp1)[D])
}



pairsHere = which(apply(pairs1[,c('i','j')], 1, function(xx) all(xx %in% Shere)))

Sk = apply(pairs1[pairsHere, c('i','j')], 1, function(xx) setdiff(Shere, xx))
pairsHere2 = pairs1[rep(pairsHere, unlist(lapply(Sk, length))), ]
pairsHere2$k = unlist(Sk)
pairsHere2 = pairsHere2[pairsHere2$i != pairsHere2$j, ]
pairsHere2$taylor3 = pairsHere2$Tiik = pairsHere2$Tjjk =pairsHere2$Tkki = pairsHere2$Tkkj = NA
for(D in 1:nrow(pairsHere2)) {
  pairsHere2[D, 'taylor3'] = bob3[pairsHere2[D, 'k']+1, pairsHere2[D, 'index']]  
  pairsHere2[D, 'Tiik'] = 2*bobD$diag[pairsHere2[D, 'k']+1, pairsHere2[D, 'i']+1]  
  pairsHere2[D, 'Tjjk'] = 2*bobD$diag[pairsHere2[D, 'k']+1, pairsHere2[D, 'j']+1]  
  pairsHere2[D, 'Tkkj'] = 2*bobD$diag[pairsHere2[D, 'j']+1, pairsHere2[D, 'k']+1]  
  pairsHere2[D, 'Tkki'] = 2*bobD$diag[pairsHere2[D, 'i']+1, pairsHere2[D, 'k']+1]  
}
  # 3rd taylor is   T_iik/2 + T_jjk/2 + T_ijk
pairsHere2$Tijk = pairsHere2$taylor3 - (pairsHere2$Tjjk + pairsHere2$Tiik)/2
pairsHere2$taylorReconstruct = (pairsHere2$Tjjk + pairsHere2$Tiik)/2

pairsHere2

pairsHere2[apply(pairsHere2[,c('i','j','k')], 1, function(xx) all(xx %in% Shere)), ]

config3 = config1
config3$sparsity$third$pairs = pairsHere2
bob3 = thirdOffDiagonals(
      result$fullParameters, fitD$tmb_data, config3
    ) 


Shere1 = Shere+1
2*bobD$diag[Shere+1, Shere+1]
bobD$second[Shere+1, Shere+1]



# grad 
plot(Sp1, bob1, type='l')
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob0)/diff(Sp1))
abline(h=bobD$first[Dpar])
abline(v=paramsAll[Dpar])


# 2nd Dpar Dpar
plot(Sp1, bob2, type='l', lwd=3)
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob1)/diff(Sp1))
abline(h=bobD$second[Dpar,Dpar], col='red')
abline(v=paramsAll[Dpar])


# 2nd Dpar Dpar2
plot(Sp1, bob2a, type='l', lwd=3)
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob1a)/diff(Sp1))
abline(h=bobD$second[Dpar,Dpar2], col='red')
abline(v=paramsAll[Dpar])

# 2nd Dpar Dpar3
plot(Sp1, bob2c, type='l', lwd=3)
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob1b)/diff(Sp1))
abline(h=bobD$second[Dpar,Dpar3], col='red')
abline(v=paramsAll[Dpar])


# Dpar2 Dpar Dpar
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2a)/diff(Sp1), type='o'
  )
abline(h=pairsHere2[pairsHere2$i==(Dpar-1) & pairsHere2$k == (Dpar2-1), 'Tiik'], lwd=3, col='green')
abline(h=pairsHere2[pairsHere2$j==(Dpar-1) & pairsHere2$k == (Dpar2-1), 'Tjjk'], lty=3, lwd=3, col='yellow')
abline(h=2*bobD$diag[Dpar2, Dpar])
abline(v=paramsAll[Dpar])

pairsHere2[pairsHere2$j==(Dpar2-1) & pairsHere2$k == (Dpar-1), 'Tkkj']
pairsHere2[pairsHere2$i==(Dpar2-1) & pairsHere2$k == (Dpar-1), 'Tkki']
abline(h=c(theT1[Dpar, Dpar2], theT2[Dpar, Dpar2]), col=c('red','blue'))

theT2[Dpar, Dpar2]

# Dpar3 Dpar Dpar
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2c)/diff(Sp1)
  )
abline(h=pairsHere2[pairsHere2$i==(Dpar-1) & pairsHere2$k == (Dpar3-1), 'Tiik'], lwd=3, col='green')
abline(h=pairsHere2[pairsHere2$j==(Dpar-1) & pairsHere2$k == (Dpar3-1), 'Tjjk'], lty=3, lwd=3, col='yellow')
abline(h=2*bobD$diag[Dpar3, Dpar])


# Dpar Dpar Dpar
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2 )/diff(Sp1))
abline(h=2*bobD$diag[Dpar, Dpar])

theT1[Dpar, Dpar]
theT2[Dpar, Dpar]

# 2nd Dpar2 Dpar3
plot(Sp1, bob2b, type='l', lwd=3)
abline(h=bobD$second[Dpar2,Dpar3], col='red')
abline(v=paramsAll[Dpar])

# Dpar Dpar2 Dpar3
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2b)/diff(Sp1)
  )
abline(h=pairsHere2$Tijk, lwd=nrow(pairsHere2):1, lty=1:nrow(pairsHere2),col=1:nrow(pairsHere2))





```


```{r theResults}
result$parameters = hpolcc::formatParameters(
  result$solution, fitD, TRUE)

result$parameters$theta

# parameters are (beta, gamma, theta)

randomEffects = seq(from=length(result$parameters$beta)+1,
  length = nrow(result$parameters$gamma))

result$Htilde = result$hessian[randomEffects, randomEffects]
result$HtildeChol = Matrix::solve(Matrix::chol(result$Htilde))

Nsim = 200
simInd = matrix(rnorm(Nsim * nrow(result$Htilde)),
 nrow(result$Htilde), Nsim)

# sample from [U|Y, theta_map]
simGamma = Matrix::crossprod(result$HtildeChol,  simInd) 
rownames(simGamma) = rownames(fitD$tmb_data$ATp)

simGamma = simGamma + unlist(result$parameters$gamma[
  rownames(simGamma), 'value'])


Stemp = seq(0,45,by=0.1)
SobsTemp = na.omit(match(Stemp, round(fitD$data$tempClean, 1)))
Spm = seq(1,1000, by=1)
SobsPm = na.omit(match(Spm, round(fitD$data$pm)))

Sobs = SobsTemp
nameX = 'tempClean'

Bmat = fitD$tmb_data$ATp[, Sobs]
# only want the global effects
Bmat = Bmat[grep(paste0(nameX,".*global"), rownames(Bmat)),]


simF = crossprod(Bmat, simGamma[rownames(Bmat),]) 

covHere = grep(nameX, names(result$parameters$beta), value=TRUE)
fixedPart = drop(crossprod(fitD$tmb_data$XTp[covHere,Sobs, drop=FALSE], 
                           result$parameters$beta[covHere]))

simSum = simF + fixedPart 



theX= fitD$data[[covHere[1] ]][Sobs]
if(grepl("^sqrt", nameX)) theX = theX^2

par(mar=c(3,3,0,0), bty='l')
matplot(theX, simSum, type='l',
  col='#00000010', lty=1, xaxs='i')


```



```{r otherstuff22, eval=FALSE}

Sstart = 1:(nrow(fitD$tmb_data$XTp)+1)
Send = seq(to = length(fitD$start_parameters), by=1, 
  len=length(fitD$theta_info$name)+1)

bob = function(xx) {
  hpolcc:::objectiveFunctionC( 
c(xx[Sstart], fitD$start_parameters[-c(Sstart, Send)], 
  xx[-Sstart]),
 fitD$tmb_data, 
 c(
 fitD$config[setdiff(names(fitD$config), c('verbose','maxDeriv'))], 
  list(maxDeriv = 0, verbose=FALSE))
  )$value
}

round(numDeriv::hessian(bob, fitD$start_parameters[c(Sstart, Send)]), 3)

res$hessian[c(Sstart, Send), c(Sstart, Send)]



beta = fitD$start_parameters[1:nrow(fitD$tmb_data$XTp)]
gamma =fitD$start_parameters[seq(from=length(beta)+1, 
  len=nrow(fitD$tmb_data$ATp)
  )]
theta = fitD$start_parameters[seq(from = length(beta) + length(gamma)+1,
  to=length(fitD$start_parameters))]


resFun = function(xx)  hpolcc:::objectiveFunctionC(xx, fitD$tmb_data, 
  c(fitD$config[
    setdiff(names(fitD$config), 'maxDeriv')], 
  list(maxDeriv = 0)))$value


resFun(fitD$start_parameters)  

Matrix::determinant(res$hessian)



fitD$config$verbose=FALSE
res$grad
theGradAgain = numDeriv::grad(resFun, fitD$start_parameters)  

rbind(res$grad, theGradAgain)



theHessAgain = numDeriv::hessian(resFun, fitD$start_parameters)


```

```{r originalPackage, eval=FALSE}
# using the hpoltest original package (and TMB)

fitM <- hnlm(
  formula= y ~ f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 22, p=3, knots = seq(0, 42, by=2), group_var = ecoRegion), 
  data=dataMedSimple,
  dirichlet=FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  tmb_parameters = list(theta = thetaStartM),
  optim_parameters = list(maxit = 20),
  verbose=TRUE, optimizer='optim', intern=FALSE, inner.method = 'BFGS', 
  inner.control = list(maxit = 500), for_dev=FALSE)
```

```{r plotFit2, fig.height=4, fig.width=4, eval=FALSE}
simTemp =  hpoltest:::condSim(fitD, 'tempClean') 

theYlim = c(0.2, 5) #exp(quantile(simTemp$y[,'GLOBAL', ], c(0.01, 0.99)))
for(D in dimnames(simTemp$y)[[2]]) {
matplot(simTemp$x, exp(simTemp$y[,D, ]), type='l', col='#00000010', lty=1, 
        xaxs='i', log='y',
        ylim =theYlim,bty='n', 
        xlab='temperature',ylab=D)
lines(simTemp$x, exp(simTemp$fixed_mean), col='yellow')
lines(simTemp$x, exp(apply(simTemp$y[,D, ], 1, mean)), col='red')
}
```

```{r plotDate, eval=FALSE}
simDate = hpoltest:::condSim(fitD, 'date') 

matplot(as.Date(simDate$x), t(apply(simDate$y, 1, quantile, c(0.025, 0.975, 0.5))), col=c('grey','grey','black'), cex=0.3, pch=16,
        xlab='date', ylab='logrr')

```


```{r debug, eval=FALSE}

xx = fitD$obj$env$last.par.best
xx = xx[grep("gamma", names(xx))]
names(xx) = colnames(fitD$obj$env$data$A)
xxDate = xx[grep("date", names(xx))]
xx = xx[grep("date", names(xx), invert=TRUE)]

```


```{r digitPreference, eval=FALSE}
checkForDigitPreference1 = aggregate(deaths[,'medical', drop=FALSE],
  deaths[,c('dom','month')], sum)
checkForDigitPreference = dcast(as.data.table(checkForDigitPreference1), dom ~ month)
image(as.numeric(checkForDigitPreference$dom),
 as.numeric(colnames(checkForDigitPreference)[-1]),
  as.matrix(checkForDigitPreference[,-1]))
```

```{r indiaDataAlex, eval=FALSE}
load("/home/patrick/research/hpoltest/applications/india/combinedindianmorttempdata_meannofilter.Rdata")
x = indianmorttempdata

table(x$death_cause, x$case)
quantile(x$final_temp)
tempSdByStrata = tapply(x$final_temp, x$riskset, sd)
countByStrata = tapply(x$case, x$riskset, sum)
# nuber cases per day
table(tapply(x$case, x$date, sum))


tempRef = 32

x$tempHigh = pmax(x$final_temp - tempRef, 0)
x$tempLow = pmin(x$final_temp - tempRef, 0)

x$tempC = x$final_temp - tempRef
x$tempSq = (x$final_temp - tempRef)^2
x$temp3 = (x$final_temp - tempRef)^3

x$dateInt = as.integer(x$date)

xSub = x[x$death_cause == 'stroke', ]
Slow = unique(unlist(xSub[xSub$final_temp < 10, 'riskset']))

Shigh = unique(unlist(xSub[xSub$final_temp >37, 'riskset']))



xSub2 = xSub[! xSub$riskset %in% c(Slow, Shigh), ]

library('survival')
fitClogit = clogit(case ~ tempHigh + tempLow + strata(riskset),
       data = x[x$death_cause == 'stroke',])
summary(fitClogit)$coef


#library('hpoltest')
fitLinear <- hnlm(
  formula = case ~  tempHigh + tempLow + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = 20, map = list(theta=factor(NA))),
  data = xSub2,
  verbose=FALSE
  )

fitLinear$est$param$est

randomDate = as.Date(
  as.integer(gsub("factor.dateInt.", "", names(fitLinear$est$random$est))),
  as.integer(as.Date('1970/1/1')))
  
badDates = randomDate[order(abs(fitLinear$est$random$est), decreasing=TRUE)[1:2]]
stuff = as.data.frame(xSub[xSub$date %in% badDates, c('riskset', 'date','final_temp','case')])
stuff = stuff[order(stuff$date,stuff$riskset), ]
stuff


plot(randomDate, fitLinear$est$random$est)

fitLinear2 <- hnlm(
  formula = case ~  tempHigh + tempLow + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = xSub2,
  verbose=FALSE
  )

fitLinear$est$param$est
fitLinear2$est$param$est


fitQuad  <- hnlm(
  formula = case ~  tempC + tempSq + temp3 + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = -5, map = list(theta=factor(NA))),
  data = xSub2,
  verbose=FALSE
  )
fitLinear$est$param$est
fitLinear2$est$param$est
fitQuad$est$param$est
```

```{r comparePlots, eval=FALSE}
tempSeq = seq(10, 40)
tempSeqC = tempSeq - tempRef

plot(tempSeq,
     tempSeqC*fitQuad$est$param$est['tempC'] + 
     tempSeqC^2*fitQuad$est$param$est['tempSq'] +
     tempSeqC^3*fitQuad$est$param$est['temp3'],
     type='l'
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitClogit$coef['tempLow'] + 
      pmax(tempSeqC, 0) * fitClogit$coef['tempHigh'],
      col='green', lwd=3, lty=1
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitLinear$est$param$est['tempLow'] + 
      pmax(tempSeqC, 0) * fitLinear$est$param$est['tempHigh'],
      col='red', lwd=3, lty=2
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitLinear2$est$param$est['tempLow'] + 
      pmax(tempSeqC, 0) * fitLinear2$est$param$est['tempHigh'],
      col='orange', lwd=3, lty=2
)

legend('bottomleft', bty='n', lty=1, col=c('black','red','orange','green'),
       legend = c('cubic', 'linear no OD','linear','clogit'))

```

```{r fitSimple, eval=FALSE}
tempKnots = seq(0,40, by=0.5)
fitSimple <- hnlm(
  formula = case ~  f(final_temp, model='iwp', ref_value = tempRef, p=4, knots = tempKnots) + 
    f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = c(-1,-5), map = list(theta=factor(c(NA,1)))),
  data = xSub2,
  verbose=FALSE
  )

fitSimple$est$param$est 
fitQuad$est$param$est
```






```{r fitTemperatureAlex, include=FALSE, eval=FALSE}

tempKnots = seq(-5,40,by=2.5)
library('hpoltest')
fitAsAlex <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
fitIwp3NoOd <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=3, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )

fitIwp3 <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=3, knots = tempKnots) +
     f(date, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=FALSE),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5),
  verbose=FALSE
  )
fitIwp2 <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots) +
     f(date, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
fitAsAlex <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
```




```{r moreSimlpeModels, eval=FALSE}
  
dataMedSimplePm = removeUnusedStrata(
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01' &
    !deaths$ecoRegion %in% c(7, 8,10), ],
  Sstrata, outcome)
formulaMedSimplePm <- y ~ dom + f(date, model='iid') +
    f(sqrtPm, model='iwp',ref_value = 25, p=3, knots = pmKnots)

fitPm<- hnlm(
  formula=formulaMedSimplePm, 
  data=dataMedSimplePm,
#  tmb_parameters = list(theta = c(5.25,31)),
  cc_design =  ccDesign(strat_vars = Sstrata),
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE, for_dev=FALSE 
  )

#  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 
#  f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var = ecoRegion)

formulaCvd = cvd ~ dom + f(date, model='iid') + f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var = ecoRegion)

formulaMedSenior <- medical_senior ~ dom + f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var=ecoRegion)
#  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 


formulaCvdSimple = cvd ~ dom + f(date, model='iid') + 
  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 
#f(tempClean, model='iwp',ref_value = 25, p=3, knots = tempKnots)

formulaMedSeniorSimple = medical_senior ~ dom + f(date, model='iid') + 
  f(tempClean, model='iwp',ref_value = 25, p=3, knots = tempKnots)

    dataMed = deaths[which(deaths$N_Week3_medical > 0 & deaths$date > as.Date('2009/1/1')),]

        dataMedSenior = deaths[which(deaths$N_Week3_medical_senior > 0 & 
                                   deaths$date > as.Date('2006/1/1')),]
dataCvd = deaths[which(deaths$N_Week3_cvd > 0 & deaths$date > as.Date('2006/1/1')), ]

if(FALSE) {

fitMedSeniorSimple = hnlm(formulaMedSeniorSimple, dataMedSenior, cc_design, 
                            optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
                          verbose=TRUE)

fitCvdSimple<- hnlm(
  formula=formulaCvdSimple, 
  data=dataCvd,
#  tmb_parameters = list(theta = c(5.25,31)),
  cc_design = cc_design,
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE, for_dev=FALSE 
  )
saveRDS(fitCvdSimple, file.path(storeDir, 'cvdSimple.rds'))

fitCvdSetup <- hnlm(
  formula=formulaCvd, 
  data=dataCvd,
  cc_design = cc_design,
  verbose=TRUE, for_dev=TRUE 
  )

fitCvdSetup$tmb_parameters$beta = fitCvdSimple$est$param$est[1:length(fitCvdSetup$tmb_parameters$beta)]
thetaHat = fitCvdSimple$est$param$est[grep("theta", names(fitCvdSimple$est$param$est))]

fitCvdSetup$tmb_parameters$theta[1:length(thetaHat)] = thetaHat
fitCvdSetup$tmb_parameters$theta[-(1:length(thetaHat))] = 0
fitCvdSetup$tmb_parameters$gamma[1:length(fitCvdSimple$est$random$est)] = fitCvdSimple$est$random$est

  obj <- MakeADFun(data = fitCvdSetup$tmb_data,
                   parameters = fitCvdSetup$tmb_parameters,
#                   random = 'gamma',
                   map = list(
                    beta = factor(rep(NA, length(fitCvdSetup$tmb_parameters$beta))),
                    theta=factor(rep(NA, length(fitCvdSetup$map$theta)))),
                   intern=FALSE, type='ADFun',
                   DLL = "hpoltest")

  parCat = rep(NA, ncol(obj$env$.data$A))
  parCat[grep("date", colnames(obj$env$.data$A))]='date'
  parCat[grep("hrpoly", colnames(obj$env$.data$A))]='poly'
  parCat[grep("hiwp", colnames(obj$env$.data$A))]='hiwp'
  parCat[grep("global", colnames(obj$env$.data$A))]='global'

  the19 = grep('hiwp_19', colnames(obj$env$.data$A), value=TRUE)

  obj$fn(obj$par)
    opt <- optim(
      par = rnorm(length(obj$par), obj$par, 1e-2),
      fn = obj$fn, 
      gr = obj$gr,
      method = 'BFGS',
      control = list(
        trace=TRUE, 
        maxit=1e5,
        parscale = c(date=1,global=1e-5, hiwp=1e-2, poly=1e-2)[parCat]
    )
  )


            names(opt$par) = colnames(obj$env$.data$A)

                opt$par1 = opt$par[grep("date", names(opt$par))]
    opt$par2 = opt$par[!names(opt$par)%in% names(opt$par1)]

    quantile(obj$gr(opt$par))

        
fitMedSenior <- hnlm( 
  formula=formulaMedSenior, data=dataMedSenior, #[which(deaths$date > as.Date('2009/1/1')), ], 
  cc_design = cc_design,verbose=TRUE, for_dev=FALSE,
    optim_parameters = list(eval.max=1e4, iter.max=1e4),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE)
  )


  obj <- MakeADFun(data = fitMed1$tmb_data,
                   parameters = fitMed1$tmb_parameters[c('beta','gamma','theta')],
                   random = 'gamma',
                   map = fitMed1$map,
                   intern=FALSE, type='ADFun',
                   DLL = "hpoltest")
  obj$fn(obj$par)
    opt <- optim(par = obj$par, 
      fn = obj$fn, gr = obj$gr,
      control = list(trace=TRUE)
  )
saveRDS(fitMed, file.path(storeDir, 'ccMed.rds'))
fitCvd <- hnlm(
  formula=formulaCvd, 
  data=dataCvd,
  cc_design = cc_design,
#  optim_parameters = list(eval.max=100, iter.max=5),
  verbose=TRUE, for_dev=FALSE 
  )
saveRDS(fitCvd, file.path(storeDir, 'ccCvd.rds'))
}
fitCvd = readRDS(file.path(storeDir, 'ccCvd.rds'))
```

```{r temperatureDataFromIndia, eval=FALSE}
deathsTemp = as.data.table(readRDS( file.path(storeDir, "indiaTemperature.rds")))
deathsTemp$dom = as.numeric(format(deathsTemp$date, '%d'))
deathsTemp[deathsTemp$dom == Hmisc::monthDays(deathsTemp$date), 'dom'] = 31
deathsTemp[format(deathsTemp$date, '%m%d') == '0101', 'dom'] = 0
deathsTemp$dom = relevel(factor(deathsTemp$dom), '1')
deathsTemp$month = format(deathsTemp$date, '%m')


(theEtable = sort(table(deathsTemp$ecoRegion), decreasing=TRUE))
# 10, 17, 20 looks dodgy



Sstrata = c("cell","ecoRegion", "yearWeek3Dow")
knotsTemperature = seq(0,50, by=2.5)  

outcome = 'medical'

deathsTempMed = removeUnusedStrata(deathsTemp[
  deathsTemp$date >= as.Date('2006/1/1') &
  format(deathsTemp$date, '%b%d') != 'Jan01', ],
  Sstrata, outcome)


deathsTempMed$tempCut = 2*round(deathsTempMed$maxtemp/2)
ttable1 = dcast(deathsTempMed, tempCut ~ ecoRegion, value.var=outcome, fun.aggregate=sum)
ttable2 = dcast(deathsTempMed, tempCut ~ ecoRegion, value.var=outcome, fun.aggregate=length)

matplot(ttable1[,1], as.matrix(ttable1[,-1]/ttable2[,-1])[,names(theEtable)[10+1:9]])
merge(
  deathsTempMed, 
    as.data.frame(deathsTempMed)[order(deathsTempMed$y, decreasing=TRUE)[1:3], Sstrata],
    all.x=FALSE, all.y=TRUE)

deathsTempMed$tempLow = pmin(25, deathsTempMed$maxtempLag01)
deathsTempMed$tempHigh = pmax(25, deathsTempMed$maxtempLag01)

SexcludeRegions  = c(7, 10,13)
SincludeRegions = c(1:6, 8,9)
library(hpoltest)
fit <- hnlm(
  formula= y ~  dom + 
    f(date, model='iid') +
    f(maxtempLag01, model='hiwp',ref_value = 25, p=3, knots = knotsTemperature, group_var=ecoRegion),
  data=deathsTempMed[deathsTempMed$ecoRegion %in% SincludeRegions,],
  cc_design = ccDesign(strat_vars = Sstrata),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    smartsearch=FALSE, 
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)



formulaMedSeniorIwp = medical_senior ~ dom + f(date, model='iid') + 
  f(maxtemp, model='iwp',ref_value = 25, p=3, knots = knotsTemperature)

fitTempMedSeniorIwp <- hnlm(
  formula=formulaMedSeniorIwp, 
  data=deathsTempSeniorMed3,
  cc_design = ccDesign(strat_vars = c("cell","ecoRegion", "yearWeek3Dow")),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)

fitTempMedIwp <- hnlm(
  formula=formulaMedIwp, 
  data=deathsTempMed,
  cc_design = ccDesign(strat_vars = Sstrata),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)
```


```
# darjeeling
# INTEL oneMKL ERROR: /usr/lib/x86_64-linux-gnu/libmkl_core.so.2: cannot open shared object file: No such file or directory.
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_core.so.2  /usr/lib/x86_64-linux-gnu/libmkl_core.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_gnu_thread.so.2  /usr/lib/x86_64-linux-gnu/libmkl_gnu_thread.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_tbb_thread.so.2  /usr/lib/x86_64-linux-gnu/libmkl_tbb_thread.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_intel_lp64.so.2 /lib/x86_64-linux-gnu/libmkl_intel_lp64.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_intel_lp64.so.2 /lib/x86_64-linux-gnu/libmkl_intel_lp64.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_def.so.2 /lib/x86_64-linux-gnu/libmkl_def.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_avx512.so.2 /lib/x86_64-linux-gnu/libmkl_avx512.so.2
```
