---
title: "India"
geometry: margin=1in
output: pdf_document
header-includes:
  - |
    ```{=latex}
    \usepackage[table]{xcolor}
    \expandafter\def\csname ver@subfig.sty\endcsname{}
    \usepackage{graphicx,subcaption}
    \makeatletter\@ifundefined{subfloat}{\newcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}{\renewcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}\makeatother
    ```
---

Time-stratified case-crossover for Indian grid-cell mortality with
population-weighted PM$_{2.5}$ and temperature. Strata are cell $\times$
eco-region $\times$ year–week3–day-of-week. The conditional likelihood is a
Dirichlet-multinomial with estimated overdispersion. Exposure–response uses
hierarchical IWP (`hiwp`) smooths of order 2 on $\sqrt{\mathrm{PM}_{2.5}}$ and
order 3 on temperature/10, each grouped by eco-region.

Joint-density derivative compute is knitr-cached; the plot chunk is not.
All chunks after the plot are `eval=FALSE`.

```{r setup, include=FALSE}
knitr::knit_hooks$set(margins = Pmisc::hook_plot_margins)
knitr::knit_hooks$set(plot = knitr::hook_plot_tex)

knitr::opts_chunk$set(
  echo = FALSE, prompt = TRUE, comment = NA,
  dev = "pdf", margins = 1, fig.cap = " ", fig.pos = "hp",
  fig.path = file.path("figure", "moreBayes"),
  fig.width = 5, fig.height = 5, dpi = NULL,
  tidy = TRUE, tidy.opts = list(indent = 2, width.cutoff = 45)
)

options(width = 80)
dir.create("analysis_cache", showWarnings = FALSE)
fit_rds <- "analysis_cache/res_hiwp_pm_temp.rds"
deaths_rds <- "deaths.RData"

# Placeholders so knitr still resolves fig.subcap / fig.cap on eval=FALSE chunks.
for_global <- data.frame(
  var = character(0),
  type = character(0),
  var_name = character(0),
  subcap = character(0),
  stringsAsFactors = FALSE
)
d_var <- character(0)
s_group <- character(0)
s_var_names <- c(temp_div10 = "temperature", sqrt_pm = "PM2.5")
```

```{r dirs, include=FALSE}
s_store_dir <- c("/store/patrick/copernicus/agg", "~/Downloads/agg")
store_dir <- s_store_dir[file.exists(path.expand(s_store_dir))][1]
has_store <- !is.na(store_dir) && nzchar(store_dir)
has_deaths <- file.exists(deaths_rds)
has_fit <- file.exists(fit_rds)
```

```{r ecoRegionData, eval=isTRUE(has_store)}
india_eco_regions <- terra::unwrap(
  readRDS(file.path(store_dir, "indiaEcoRegions.rds"))
)
india_eco_regions$name <- gsub(
  " AND ", ", ",
  gsub(
    " E.O.REGION|[(].*[)]|TO PERHUMID.*|PENINSULA ", "",
    india_eco_regions$physio_reg
  )
)
india_eco_regions$id_name <- paste(
  india_eco_regions$objectid, india_eco_regions$name
)
india_eco_regions$col <- rep_len(
  c(
    RColorBrewer::brewer.pal(12, "Set3"),
    RColorBrewer::brewer.pal(8, "Set2")
  ),
  length(india_eco_regions)
)
```

```{r worldPopByRegion, cache=TRUE, eval=isTRUE(has_store)}
year <- 2020
dir.create("data/gpw", recursive = TRUE, showWarnings = FALSE)

pop_density <- geodata::population(year, res = 5, path = "data/gpw")
eco <- terra::project(india_eco_regions, terra::crs(pop_density))
pop_density_india_eco <- terra::mask(terra::crop(pop_density, eco), eco)
cell_area_km2 <- terra::cellSize(pop_density_india_eco, unit = "km")
pop_count <- pop_density_india_eco * cell_area_km2
eco_pop <- terra::extract(
  pop_count, eco,
  fun = sum, na.rm = TRUE, exact = TRUE
)
eco$gpw_2020 <- eco_pop[, 2]
eco_pop_df <- data.frame(
  objectid = eco$objectid,
  physio_reg = eco$physio_reg,
  gpw_2020 = eco$gpw_2020
)
eco_pop_df$weight <- eco_pop_df$gpw_2020 / sum(eco_pop_df$gpw_2020)
eco_weights <- stats::setNames(
  eco_pop_df$weight,
  as.character(eco_pop_df$objectid)
)
```

```{r deaths_data, eval=FALSE}
# Manually execute to rebuild deaths.RData from the store RDS.
stopifnot(isTRUE(has_store))
deaths <- data.table::as.data.table(
  readRDS(file.path(store_dir, "indiaCC.rds"))
)
data.table::setnames(
  deaths,
  old = c("ecoRegion", "yearWeek3Dow"),
  new = c("eco_region", "year_week3_dow"),
  skip_absent = TRUE
)
deaths$sqrt_pm <- sqrt(pmin(deaths$pm, 1000))
deaths$temp_div10 <- pmax(deaths$temp, -5) / 10
deaths$dom <- as.numeric(format(deaths$date, "%d"))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), "dom"] <- 31
deaths[format(deaths$date, "%m%d") == "0101", "dom"] <- 0
deaths$dom <- droplevels(relevel(factor(deaths$dom), "1"))
deaths$eco_region <- factor(as.character(deaths$eco_region))
save(deaths, file = deaths_rds)
```

```{r formula_factory, include=FALSE}
formula_lhs <- adlaplace::dirichlet_multinom(
  "medical",
  by = c("cell", "eco_region", "year_week3_dow"),
  init = 0.1,
  parscale = 10
)
formula_rhs <- c(
  adlaplace::linear("dom"),
  adlaplaceHgp::hiwp(
    "sqrt_pm",
    p = 2,
    ref_value = 8,
    knots = seq(0, 32, by = 2),
    by = "eco_region",
    init = 1e-1,
    lower = 1e-12
  ),
  adlaplaceHgp::hiwp(
    "temp_div10",
    p = 3,
    ref_value = 25 / 10,
    knots = seq(-5, 42.5, by = 2.5) / 10,
    by = "eco_region",
    init = 1e-1,
    lower = 1e-12
  )
)
cc_formula <- c(formula_lhs, formula_rhs)
```

```{r check_joint_deriv, cache=TRUE}
# Cached: outer_fn warm-start, 5 derivative-free outer steps, then joint dens scan.
# Subset to year >= 2010 to keep the probe cheap.
stopifnot(file.exists(deaths_rds))
load(deaths_rds)
deaths <- data.table::as.data.table(deaths)
deaths <- deaths[as.integer(format(date, "%Y")) >= 2010L]
deaths[, eco_region := factor(as.character(eco_region))]
deaths[, dom := droplevels(dom)]
message(
  sprintf(
    "check_joint_deriv subset: N=%d (years >= 2010)",
    nrow(deaths)
  )
)
utils::flush.console()

md <- adlaplace::model_data(cc_formula, data = deaths, verbose = FALSE)
n_shards_dev <- as.integer(
  min(200L, max(50L, data.table::uniqueN(deaths$cell)))
)
config_dev <- list(
  transform_theta = TRUE,
  num_threads = 1L,
  verbose = FALSE,
  obs_groups = adlaplace::obs_groups(
    A = md$term_data$A,
    elgm_matrix = md$term_data$elgm_matrix,
    num_shards = n_shards_dev
  )
)

message("building ad_pack (shards=", n_shards_dev, ")...")
utils::flush.console()
ad_plain <- adlaplace::ad_pack(md, config_dev, num_threads = 1L)
dens_ptr <- adlaplace::clone_ad_pack_ptr(ad_plain@ptr)

sz <- ad_plain@sizes
n_beta <- as.integer(sz["beta"])
n_gamma <- as.integer(sz["gamma"])
n_theta <- as.integer(sz["theta"])
x0_start <- c(
  md$term_data$info$beta$init,
  adlaplace::apply_theta_log(md$term_data$info$theta, cols = "init")$init
)
ctrl_inner <- list(
  maxit = 5L, report.level = 3, report.freq = 1,
  restart.gamma.clamp = Inf
)
cache_j <- new.env(parent = emptyenv())
cache_j$gamma <- rep(0, n_gamma)

message(
  "outer_fn(x0): inner opt maxit=", ctrl_inner$maxit,
  " (gamma length=", n_gamma, ")..."
)
utils::flush.console()
t_fn <- proc.time()[["elapsed"]]
fn0 <- tryCatch(
  adlaplace::outer_fn(
    x0_start,
    config = config_dev, ad_pack = ad_plain, cache = cache_j,
    control_inner = ctrl_inner
  ),
  error = function(e) e
)
fn0_sec <- proc.time()[["elapsed"]] - t_fn
fn0_err <- if (inherits(fn0, "error")) conditionMessage(fn0) else NA_character_
fn0_val <- if (inherits(fn0, "error")) NA_real_ else as.numeric(fn0)
message(
  sprintf(
    "outer_fn done in %.1fs: %s",
    fn0_sec,
    if (!is.na(fn0_err)) fn0_err else format(fn0_val)
  )
)
utils::flush.console()

gamma_after_fn0 <- as.numeric(cache_j$gamma)
gamma_fn0_sum <- list(
  n = length(gamma_after_fn0),
  n_finite = sum(is.finite(gamma_after_fn0)),
  range = if (any(is.finite(gamma_after_fn0))) {
    range(gamma_after_fn0, na.rm = TRUE)
  } else {
    c(NA_real_, NA_real_)
  },
  max_abs = if (any(is.finite(gamma_after_fn0))) {
    max(abs(gamma_after_fn0), na.rm = TRUE)
  } else {
    NA_real_
  }
)

# 5 Nelder-Mead outer steps (no derivatives); record each outer_fn eval.
ctrl_nm_inner <- list(
  maxit = 5L, report.level = 0, report.freq = 0,
  restart.gamma.clamp = Inf
)
cfg_nm <- utils::modifyList(config_dev, list(verbose = FALSE))
fn_hist <- data.frame(
  eval = integer(),
  neg_log_lik = numeric(),
  log_lik = numeric(),
  finite = logical()
)
fn_nm <- function(par) {
  v <- tryCatch(
    adlaplace::outer_fn(
      par,
      config = cfg_nm, ad_pack = ad_plain, cache = cache_j,
      control_inner = ctrl_nm_inner
    ),
    error = function(e) NA_real_
  )
  i <- nrow(fn_hist) + 1L
  fn_hist <<- rbind(
    fn_hist,
    data.frame(
      eval = i,
      neg_log_lik = v,
      log_lik = -v,
      finite = is.finite(v)
    )
  )
  if (i == 1L || i %% 5L == 0L || !is.finite(v)) {
    message(
      sprintf(
        "  NM eval %d: neg_log_lik=%s finite=%s",
        i, format(v), is.finite(v)
      )
    )
    utils::flush.console()
  }
  if (!is.finite(v)) {
    return(.Machine$double.xmax)
  }
  v
}
message("Nelder-Mead outer: maxit=5 (no derivatives)...")
utils::flush.console()
t_nm <- proc.time()[["elapsed"]]
opt_nm <- tryCatch(
  stats::optim(
    par = x0_start,
    fn = fn_nm,
    method = "Nelder-Mead",
    control = list(maxit = 5L, reltol = 1e-8)
  ),
  error = function(e) e
)
nm_sec <- proc.time()[["elapsed"]] - t_nm
if (!inherits(opt_nm, "error") && nrow(fn_hist)) {
  fn_hist$delta_ll <- fn_hist$log_lik - fn_hist$log_lik[[1L]]
}
message(
  sprintf(
    "Nelder-Mead done in %.1fs: %s",
    nm_sec,
    if (inherits(opt_nm, "error")) {
      conditionMessage(opt_nm)
    } else {
      sprintf(
        "convergence=%d n_evals=%d value=%s",
        opt_nm$convergence, nrow(fn_hist), format(opt_nm$value)
      )
    }
  )
)
utils::flush.console()

x_nm <- if (inherits(opt_nm, "error")) x0_start else opt_nm$par
gamma_after_nm <- as.numeric(cache_j$gamma)
nm_err <- if (inherits(opt_nm, "error")) conditionMessage(opt_nm) else NA_character_

x_here <- c(
  x_nm[seq_len(n_beta)],
  gamma_after_nm,
  x_nm[seq(n_beta + 1L, length.out = n_theta)]
)
Dpar_dens <- n_beta + n_gamma + 1L
message("deriv_scan_joint: n=5 around full index ", Dpar_dens, "...")
utils::flush.console()
t_scan <- proc.time()[["elapsed"]]
scan_j <- adlaplace::deriv_scan_joint(
  dens_ptr,
  x = x_here,
  index = Dpar_dens,
  width = 0.1,
  n = 5L,
  ad_shards = NULL,
  negative = FALSE
)
message(
  sprintf("deriv_scan_joint done in %.1fs", proc.time()[["elapsed"]] - t_scan)
)
# Drop AD handles before cache serialize (external pointers).
rm(ad_plain, dens_ptr, cache_j, md, fn_nm)
```

```{r check_outer_fn, dependson="check_joint_deriv"}
# Diagnose outer_fn using cached fn0 + Nelder-Mead path (no recomputation).
cat("=== outer_fn at start (fn0) ===\n")
cat(
  sprintf(
    "finite=%s  value=%s  error=%s  time=%.1fs\n",
    is.finite(fn0_val), format(fn0_val),
    if (is.na(fn0_err)) "none" else fn0_err,
    fn0_sec
  )
)
cat(
  sprintf(
    "gamma after fn0: n=%d finite=%d/%d range=[%s, %s] max|g|=%s\n",
    gamma_fn0_sum$n, gamma_fn0_sum$n_finite, gamma_fn0_sum$n,
    format(gamma_fn0_sum$range[[1L]]), format(gamma_fn0_sum$range[[2L]]),
    format(gamma_fn0_sum$max_abs)
  )
)

cat("\n=== Nelder-Mead outer (5 iters, no derivatives) ===\n")
if (!is.na(nm_err)) {
  cat("optim error: ", nm_err, "\n", sep = "")
} else {
  cat(
    sprintf(
      "convergence=%d  value=%s  time=%.1fs  n_evals=%d\n",
      opt_nm$convergence, format(opt_nm$value), nm_sec, nrow(fn_hist)
    )
  )
}
if (nrow(fn_hist)) {
  print(fn_hist)
  cat(
    sprintf(
      "log_lik change (last - first): %s\n",
      format(fn_hist$log_lik[[nrow(fn_hist)]] - fn_hist$log_lik[[1L]])
    )
  )
}

# Consistency: first NM eval should match fn0 (same x0; gamma warm-started).
if (nrow(fn_hist) && is.finite(fn0_val) && is.finite(fn_hist$neg_log_lik[[1L]])) {
  d_fn <- fn_hist$neg_log_lik[[1L]] - fn0_val
  cat(
    sprintf(
      "\nfn0 vs first NM eval: diff=%s (should be ~0 if warm-start reused)\n",
      format(d_fn)
    )
  )
}

dx <- x_nm - x0_start
cat(
  sprintf(
    "\nouter par move: max|dx|=%s  ||dx||2=%s  n_moved(>1e-8)=%d / %d\n",
    format(max(abs(dx))), format(sqrt(sum(dx * dx))),
    sum(abs(dx) > 1e-8), length(dx)
  )
)

cat("\n=== flags (problems with outer_fn / start) ===\n")
flags <- c(
  fn0_not_finite = !is.finite(fn0_val),
  fn0_error = !is.na(fn0_err),
  gamma_not_all_finite = gamma_fn0_sum$n_finite < gamma_fn0_sum$n,
  nm_any_nonfinite_eval = nrow(fn_hist) > 0L && !all(fn_hist$finite),
  nm_no_improvement = nrow(fn_hist) > 1L &&
    is.finite(fn_hist$log_lik[[1L]]) &&
    is.finite(fn_hist$log_lik[[nrow(fn_hist)]]) &&
    (fn_hist$log_lik[[nrow(fn_hist)]] <= fn_hist$log_lik[[1L]] + 1e-8),
  nm_error = !is.na(nm_err)
)
print(flags)
if (isTRUE(flags[["fn0_not_finite"]]) || isTRUE(flags[["fn0_error"]])) {
  message(
    "Likely cause of L-BFGS-B 'needs finite values of fn': outer_fn(x0) is not finite."
  )
}
if (isTRUE(flags[["nm_any_nonfinite_eval"]])) {
  message("Nelder-Mead hit non-finite outer_fn along its path.")
}
if (isTRUE(flags[["nm_no_improvement"]]) && !isTRUE(flags[["fn0_not_finite"]])) {
  message(
    "Derivative-free steps did not improve log_lik from fn0 (flat / worse / stuck)."
  )
}
if (!any(unlist(flags))) {
  message("No obvious outer_fn failure flags at this cheap probe.")
}
```

```{r check_joint_deriv_plot, fig.height=5, dependson="check_joint_deriv"}
par(mfrow = c(3, 2), mar = c(2, 2, 2, 0), mgp = c(1, 0.5, 0))
plot(scan_j$grid, scan_j$value, type = "l", ylab = "joint log dens")
plot(scan_j$grid, scan_j$grad[, Dpar_dens], type = "l", ylab = "AD gradient")
points(scan_j$mid, scan_j$fd_value, pch = 16)
legend(
  "topright",
  legend = c("AD", "finite diff"),
  lty = c(1, NA), pch = c(NA, 16), bty = "n"
)

s_par2 <- sort(unique(c(
  Dpar_dens, 1:3, seq(to = length(x_here), length.out = 3)
)))
s_par2 <- s_par2[seq_len(min(prod(par("mfrow")) - 2L, length(s_par2)))]
for (Dpar2 in s_par2) {
  plot(
    scan_j$grid, scan_j$hess_row[, Dpar2],
    type = "l",
    ylab = paste0("H[", Dpar_dens, ",", Dpar2, "]"),
    ylim = range(
      c(scan_j$hess_row[, Dpar2], scan_j$fd_grad[, Dpar2]),
      na.rm = TRUE
    )
  )
  points(scan_j$mid, scan_j$fd_grad[, Dpar2], pch = 16)
}
```

```{r fit, eval=FALSE}
# Manually execute to fit / refresh analysis_cache/res_hiwp_pm_temp.rds.
stopifnot(file.exists(deaths_rds))
load(deaths_rds)
deaths <- data.table::as.data.table(deaths)
deaths[, eco_region := factor(as.character(eco_region))]

n_threads <- parallel::detectCores()
if (is.na(n_threads) || n_threads < 1L) {
  n_threads <- 1L
}
n_shards <- as.integer(
  min(1e4L, max(100L, data.table::uniqueN(deaths$cell)))
)
message(
  sprintf(
    "fitting medical N=%d threads=%d shards=%d -> %s",
    nrow(deaths), n_threads, n_shards, fit_rds
  )
)

fit <- adlaplace::adlaplace(
  formula = cc_formula,
  data = deaths,
  config = list(
    transform_theta = TRUE,
    reorder_shards = "none",
    num_threads = as.integer(n_threads),
    num_shards = n_shards
  ),
  control = list(
    maxit = 80,
    trace = 3,
    REPORT = 1,
    pgtol = 1e-5,
    factr = 1e7,
    lmm = 8
  ),
  verbose = TRUE
)

if (!is.null(fit$details) && !is.null(fit$model_data)) {
  wts <- if (exists("eco_weights")) eco_weights else NULL
  fit$sample <- tryCatch(
    adlaplaceHgp::cond_sim_iwp(
      fit = fit$details,
      model_data = fit$model_data,
      n = 500L,
      weights = wts
    ),
    error = function(e) {
      message("cond_sim_iwp failed: ", conditionMessage(e))
      NULL
    }
  )
}

# Lean cache for knit; keep full `fit` (ad_pack) for check_lik_deriv.
fit_ok <- is.null(fit$details$error) &&
  !is.null(fit$par_info) &&
  all(is.finite(fit$par_info$mle_internal))
if (!fit_ok) {
  message(
    "WARNING: outer fit did not produce finite MLEs",
    if (!is.null(fit$details$error)) paste0(": ", fit$details$error) else "",
    "; not overwriting ", fit_rds
  )
} else {
  fit_save <- fit
  fit_save$model_data <- NULL
  fit_save$ad_pack <- NULL
  dir.create("analysis_cache", showWarnings = FALSE)
  saveRDS(fit_save, fit_rds)
  message("saved: ", fit_rds)
  rm(fit_save)
}
```

```{r check_lik_deriv, fig.height=3.5, eval=FALSE}
# Fast debug: joint dens + AD grad vs 2-point FD along one outer theta.
# No Laplace / no inner opt / no Hessian (those were too slow and crashed).
# Reuses ad_pack/dens_ptr from this session when present.
fit_chk <- if (exists("fit") && !is.null(fit$par_info)) {
  fit
} else if (file.exists(fit_rds)) {
  message("loading estimates from ", fit_rds)
  readRDS(fit_rds)
} else {
  stop("no fit in memory and missing ", fit_rds, "; run the fit chunk first")
}

need_pack <- !(exists("dens_ptr") && !is.null(dens_ptr)) &&
  !(exists("ad_pack") && !is.null(ad_pack))
if (!need_pack && exists("ad_pack") && !exists("dens_ptr")) {
  dens_ptr <- adlaplace::clone_ad_pack_ptr(ad_pack@ptr)
}
if (need_pack) {
  if (!is.null(fit_chk$ad_pack)) {
    ad_pack <- fit_chk$ad_pack
    md <- fit_chk$model_data
  } else {
    message("rebuilding ad_pack (once)...")
    stopifnot(file.exists(deaths_rds))
    load(deaths_rds)
    deaths <- data.table::as.data.table(deaths)
    deaths[, eco_region := factor(as.character(eco_region))]
    md <- adlaplace::model_data(cc_formula, data = deaths, verbose = FALSE)
    n_threads <- parallel::detectCores()
    if (is.na(n_threads) || n_threads < 1L) {
      n_threads <- 1L
    }
    n_shards <- as.integer(
      min(800L, max(50L, data.table::uniqueN(deaths$cell)))
    )
    config_dev <- list(
      transform_theta = TRUE,
      num_threads = 1L,
      verbose = FALSE,
      reorder_shards = "none",
      obs_groups = adlaplace::obs_groups(
        A = md$term_data$A,
        elgm_matrix = md$term_data$elgm_matrix,
        num_shards = n_shards
      )
    )
    ad_pack <- adlaplace::ad_pack(md, config_dev, num_threads = 1L)
  }
  dens_ptr <- adlaplace::clone_ad_pack_ptr(ad_pack@ptr)
}

sz <- ad_pack@sizes
n_beta <- as.integer(sz["beta"])
n_theta <- as.integer(sz["theta"])
n_gamma <- as.integer(sz["gamma"])
par_lab_md <- if (!is.null(fit_chk$model_data)) {
  fit_chk$model_data$term_data$info$parameters$label
} else if (exists("md")) {
  md$term_data$info$parameters$label
} else {
  fit_chk$par_info$label
}

x_named <- stats::setNames(
  as.numeric(fit_chk$par_info$mle_internal),
  as.character(fit_chk$par_info$label)
)
x_outer <- as.numeric(x_named[par_lab_md])
if (!all(is.finite(x_outer))) {
  x_outer <- as.numeric(fit_chk$cache$last_par_fn)
  if (!all(is.finite(x_outer)) || length(x_outer) != length(par_lab_md)) {
    x_outer <- as.numeric(fit_chk$par_info$init)
  }
  message("using non-MLE outer start (fit MLEs not finite)")
}
gamma0 <- if (!is.null(fit_chk$gamma) && length(fit_chk$gamma) == n_gamma) {
  as.numeric(fit_chk$gamma)
} else {
  rep(0, n_gamma)
}

# Full parameter index for first/last theta (outer theta block after gamma).
theta_labs <- grep("_iwp$|_hiwp$|dirichlet", par_lab_md, value = TRUE)
par_lab <- if (length(theta_labs)) {
  theta_labs[length(theta_labs)]
} else {
  par_lab_md[length(par_lab_md)]
}
j_outer <- match(par_lab, par_lab_md)
# full x = c(beta, gamma, theta); theta starts after beta+gamma
j_full <- n_beta + n_gamma + (j_outer - n_beta)
stopifnot(is.finite(j_full), j_full > n_beta + n_gamma)

x0 <- c(x_outer[seq_len(n_beta)], gamma0, x_outer[seq(n_beta + 1L, length.out = n_theta)])
eps <- 1e-4
xp <- xm <- x0
xp[j_full] <- x0[j_full] + eps
xm[j_full] <- x0[j_full] - eps

message("joint dens + grad FD for ", par_lab, " (full index ", j_full, ")")
d0 <- adlaplace::joint_log_dens(dens_ptr, x0, negative = FALSE)
dp <- adlaplace::joint_log_dens(dens_ptr, xp, negative = FALSE)
dm <- adlaplace::joint_log_dens(dens_ptr, xm, negative = FALSE)
g0 <- adlaplace::grad(dens_ptr, x0, inner = FALSE, negative = FALSE)
fd <- (dp - dm) / (2 * eps)

cmp <- data.frame(
  label = par_lab,
  joint_dens = d0,
  ad_grad = unname(g0[j_full]),
  fd_grad = fd,
  abs_err = abs(unname(g0[j_full]) - fd)
)
print(cmp)

plot(
  c(xm[j_full], x0[j_full], xp[j_full]),
  c(dm, d0, dp),
  type = "b", xlab = par_lab, ylab = "joint log dens",
  main = "3 dens evals (no Laplace)"
)
abline(
  a = d0 - g0[j_full] * x0[j_full],
  b = g0[j_full],
  col = 2, lty = 2
)
legend("topright", c("dens", "AD slope"), lty = c(1, 2), col = c(1, 2), bty = "n")
```

```{r explore_na_mle, fig.height=5, eval=FALSE}
# Why are mle / mle_internal NA? Pick one NA slot and probe outer_fn / joint dens.
# Outer optim failure sets *all* MLEs to NA (see adlaplace:::.adlaplace_fit_optim_failed).
fit_na <- if (exists("fit") && !is.null(fit$par_info)) {
  fit
} else if (file.exists(fit_rds)) {
  message("loading: ", fit_rds)
  readRDS(fit_rds)
} else {
  stop("no fit available; run the fit chunk first")
}

pi <- fit_na$par_info
na_rows <- which(!is.finite(pi$mle_internal))
message(
  "outer error: ",
  if (!is.null(fit_na$details$error)) fit_na$details$error else "(none)",
  "\nNA mle_internal: ", length(na_rows), " / ", nrow(pi)
)
print(pi[na_rows, c("label", "mle", "mle_internal", "init", "lower", "upper", "log")])

# Pick one NA parameter (prefer a theta/IWP scale if present).
pick <- na_rows[grep("_iwp$|_hiwp$|dirichlet", pi$label[na_rows])[1]]
if (is.na(pick)) {
  pick <- na_rows[1L]
}
stopifnot(length(pick) == 1L, is.finite(pick))
message(
  sprintf(
    "probing parameter [%d] %s (init=%g, log=%s)",
    pick, pi$label[pick], pi$init[pick], pi$log[pick]
  )
)

# Rebuild AD handle if the cached fit is lean.
if (!is.null(fit_na$ad_pack) && !is.null(fit_na$model_data)) {
  ad_na <- fit_na$ad_pack
  md_na <- fit_na$model_data
  cfg_na <- utils::modifyList(fit_na$config, list(verbose = FALSE))
} else {
  stopifnot(file.exists(deaths_rds))
  load(deaths_rds)
  deaths <- data.table::as.data.table(deaths)
  deaths[, eco_region := factor(as.character(eco_region))]
  md_na <- adlaplace::model_data(cc_formula, data = deaths, verbose = FALSE)
  n_threads <- parallel::detectCores()
  if (is.na(n_threads) || n_threads < 1L) {
    n_threads <- 1L
  }
  n_shards <- as.integer(
    min(1e4L, max(100L, data.table::uniqueN(deaths$cell)))
  )
  cfg_na <- list(
    transform_theta = TRUE,
    num_threads = as.integer(n_threads),
    verbose = FALSE,
    reorder_shards = "none",
    obs_groups = adlaplace::obs_groups(
      A = md_na$term_data$A,
      elgm_matrix = md_na$term_data$elgm_matrix,
      num_shards = n_shards
    )
  )
  message("building ad_pack for NA probe...")
  ad_na <- adlaplace::ad_pack(md_na, cfg_na, num_threads = as.integer(n_threads))
}
dens_na <- adlaplace::clone_ad_pack_ptr(ad_na@ptr)
sz <- ad_na@sizes
n_beta <- as.integer(sz["beta"])
n_theta <- as.integer(sz["theta"])
n_gamma <- as.integer(sz["gamma"])

# Outer start: last_par_fn if present, else inits.
x0 <- if (!is.null(fit_na$cache$last_par_fn) &&
  length(fit_na$cache$last_par_fn) == nrow(pi)) {
  as.numeric(fit_na$cache$last_par_fn)
} else {
  as.numeric(pi$init)
}
names(x0) <- pi$label
g0 <- if (!is.null(fit_na$gamma) && length(fit_na$gamma) == n_gamma) {
  as.numeric(fit_na$gamma)
} else {
  rep(0, n_gamma)
}
ctrl_inner <- list(maxit = 100L, report.level = 3, report.freq = 1)
cfg_na$verbose <- TRUE
cache_na <- new.env(parent = emptyenv())
cache_na$gamma <- g0

# 1) Outer objective / gradient at the start point used by optim.
message(
  "outer_fn(x0): starting inner opt (gamma length=", length(g0),
  ", maxit=", ctrl_inner$maxit, ")..."
)
utils::flush.console()
t_fn <- proc.time()[["elapsed"]]
fn0 <- tryCatch(
  adlaplace::outer_fn(
    x0,
    config = cfg_na, ad_pack = ad_na, cache = cache_na,
    control_inner = ctrl_inner
  ),
  error = function(e) e
)
message(
  sprintf(
    "outer_fn done in %.1fs: %s",
    proc.time()[["elapsed"]] - t_fn,
    if (inherits(fn0, "error")) conditionMessage(fn0) else format(fn0)
  )
)
utils::flush.console()

message("outer_gr(x0): starting...")
utils::flush.console()
t_gr <- proc.time()[["elapsed"]]
gr0 <- tryCatch(
  adlaplace::outer_gr(
    x0,
    config = cfg_na, ad_pack = ad_na, cache = cache_na,
    control_inner = ctrl_inner
  ),
  error = function(e) e
)
message(
  sprintf("outer_gr done in %.1fs", proc.time()[["elapsed"]] - t_gr)
)
utils::flush.console()
if (!inherits(gr0, "error")) {
  message(
    "outer_gr(x0): finite=", all(is.finite(gr0)),
    " range=", paste(range(gr0, na.rm = TRUE), collapse = " .. ")
  )
  print(utils::head(data.frame(label = pi$label, grad = as.numeric(gr0)), 12))
} else {
  message("outer_gr(x0) error: ", conditionMessage(gr0))
}

# 2) Joint dens at full (beta, gamma, theta); then profile lik.
x_full0 <- c(x0[seq_len(n_beta)], g0, x0[seq(n_beta + 1L, length.out = n_theta)])
jd0 <- tryCatch(
  adlaplace::joint_log_dens(dens_na, x_full0, negative = FALSE),
  error = function(e) e
)
message("joint_log_dens(full): ", if (inherits(jd0, "error")) {
  conditionMessage(jd0)
} else {
  jd0
})

message("log_lik_laplace(deriv=TRUE): starting...")
utils::flush.console()
t_ll <- proc.time()[["elapsed"]]
lik_na <- tryCatch(
  adlaplace::log_lik_laplace(
    x = x0,
    gamma = g0,
    ad_pack = ad_na,
    config = cfg_na,
    control = ctrl_inner,
    deriv = TRUE
  ),
  error = function(e) e
)
message(
  sprintf("log_lik_laplace done in %.1fs", proc.time()[["elapsed"]] - t_ll)
)
utils::flush.console()
if (inherits(lik_na, "error")) {
  message("log_lik_laplace error: ", conditionMessage(lik_na))
} else {
  message(
    "log_lik_laplace: neg_log_lik=", lik_na$neg_log_lik,
    " finite=", is.finite(lik_na$neg_log_lik)
  )
  if (!is.null(lik_na$deriv$d_neg_log_lik)) {
    message(
      "d_neg_log_lik[", pick, "] (", pi$label[pick], ") = ",
      lik_na$deriv$d_neg_log_lik[pick]
    )
  }
  g0 <- as.numeric(lik_na$inner_opt$solution)
  cache_na$gamma <- g0
}

# 3) Per-shard dens: find shards that are non-finite at x_full0.
n_grp <- adlaplace:::n_groups(dens_na)
message("per-shard joint dens: evaluating ", n_grp, " shards...")
utils::flush.console()
shard_vals <- rep(NA_real_, n_grp)
for (i in seq_len(n_grp)) {
  if (i == 1L || i == n_grp || i %% 500L == 0L) {
    message("  shard ", i, "/", n_grp)
    utils::flush.console()
  }
  shard_vals[[i]] <- tryCatch(
    adlaplace::joint_log_dens(
      dens_na, x_full0,
      ad_shards = i - 1L, negative = FALSE
    ),
    error = function(e) NA_real_
  )
}
bad_shards <- which(!is.finite(shard_vals))
message(
  "shards: n=", n_grp,
  " non-finite=", length(bad_shards),
  if (length(bad_shards)) {
    paste0(" (first: ", paste(utils::head(bad_shards - 1L, 10), collapse = ", "), ")")
  } else {
    ""
  }
)
print(utils::head(sort(shard_vals), 10))
print(utils::tail(sort(shard_vals), 5))

# 4) Short 1-D path in the chosen NA parameter (quiet inner reports).
ctrl_path <- utils::modifyList(ctrl_inner, list(report.level = 0, report.freq = 0))
cfg_path <- utils::modifyList(cfg_na, list(verbose = FALSE))
Sx <- x0[pick] + c(-0.5, 0, 0.5)
fn_path <- rep(NA_real_, length(Sx))
for (i in seq_along(Sx)) {
  message(sprintf("outer_fn path %d/%d at %s=%g ...", i, length(Sx), pi$label[pick], Sx[[i]]))
  utils::flush.console()
  xv <- x0
  xv[pick] <- Sx[[i]]
  t_i <- proc.time()[["elapsed"]]
  fn_path[[i]] <- tryCatch(
    adlaplace::outer_fn(
      xv,
      config = cfg_path, ad_pack = ad_na, cache = cache_na,
      control_inner = ctrl_path
    ),
    error = function(e) NA_real_
  )
  message(
    sprintf(
      "  done in %.1fs: %s",
      proc.time()[["elapsed"]] - t_i,
      format(fn_path[[i]])
    )
  )
  utils::flush.console()
}
path_tbl <- data.frame(
  x = Sx,
  outer_fn = fn_path,
  finite = is.finite(fn_path)
)
print(path_tbl)

par(mfrow = c(1, 2), mar = c(3, 3, 2, 0), mgp = c(1.5, 0.5, 0))
plot(
  path_tbl$x, path_tbl$outer_fn,
  type = "b", xlab = pi$label[pick], ylab = "outer_fn",
  main = "outer_fn along NA parameter"
)
abline(v = x0[pick], lty = 2, col = "grey")
plot(
  seq_along(shard_vals) - 1L, shard_vals,
  pch = 16, cex = 0.4, xlab = "shard", ylab = "joint dens",
  main = "per-shard joint dens at x0"
)
if (length(bad_shards)) {
  points(bad_shards - 1L, shard_vals[bad_shards], col = 2, pch = 16)
}
```

```{r load_cache, include=FALSE, eval=FALSE}
res <- NULL
if (has_fit) {
  message("loading: ", fit_rds)
  res <- readRDS(fit_rds)
} else {
  message("missing: ", fit_rds, "; figures/tables will leave gaps")
}
has_sample <- !is.null(res$sample$sim)
s_var <- if (has_sample) names(res$sample$sim) else character(0)
s_var_names <- c(temp_div10 = "temperature", sqrt_pm = "PM2.5")
# Placeholders so knitr fig.subcap / fig.cap options always resolve
# even when sample caches are missing (eval=FALSE still evaluates options).
for_global <- data.frame(
  var = character(0),
  type = character(0),
  var_name = character(0),
  subcap = character(0),
  stringsAsFactors = FALSE
)
d_var <- character(0)
s_group <- character(0)
the_x_lim <- list(temp_div10 = c(4, 38), sqrt_pm = c(0, 450))
the_y_lim <- list()
```

```{r par_table, eval=FALSE}
th <- res$par_info
keep <- grepl(
  "dirichlet_multinom|_iwp$|_hiwp$|_rpoly|_fpoly",
  th$label
)
theta_table <- th[keep, c("label", "mle", "se", "lower", "upper", "log"), drop = FALSE]
theta_table
```

```{r globalSetup, eval=FALSE}
for_global <- expand.grid(
  var = s_var,
  type = c("Common", "Pop Avg"),
  stringsAsFactors = FALSE
)
for_global$var_name <- unname(s_var_names[for_global$var])
for_global$subcap <- paste(for_global$var_name, for_global$type)

the_y_lim <- list()
for (d in s_var) {
  x_here <- res$sample$x[[d]]
  if (grepl("sqrt", d)) x_here <- x_here^2
  if (grepl("div10", d)) x_here <- 10 * x_here
  use_x <- x_here <= max(the_x_lim[[d]]) & x_here >= min(the_x_lim[[d]])
  range1 <- range(
    res$sample$quantiles$common[[d]][use_x, , drop = FALSE],
    na.rm = TRUE
  )
  env <- res$sample$envelope$common[[d]]
  range2 <- if (!is.null(env)) {
    range(env[use_x, , drop = FALSE], na.rm = TRUE)
  } else {
    NULL
  }
  the_y_lim[[d]] <- range(c(-2, 100 * (c(range1, range2) - 1)), na.rm = TRUE)
}
```

```{r globalRes, fig.cap='Common component', fig.subcap = for_global$subcap, out.width='0.48\\linewidth', fig.height=5, fig.width=5, eval=FALSE}
par(mar = c(3, 3, 0, 0), bty = "l")

for (d in seq_len(nrow(for_global))) {
  d_var <- for_global[d, "var"]
  x_here <- res$sample$x[[d_var]]
  if (grepl("sqrt", d_var)) x_here <- x_here^2
  if (grepl("div10", d_var)) x_here <- x_here * 10

  plot(
    NA,
    xlim = the_x_lim[[d_var]], ylim = the_y_lim[[d_var]],
    xlab = s_var_names[d_var], ylab = "100 x (RR-1)",
    xaxs = "i", yaxs = "i"
  )

  if (for_global[d, "type"] == "Common") {
    sample <- 100 * (res$sample$sim[[d_var]] - 1)
    matlines(x_here, sample, type = "l", col = "#00000010", lty = 1)
    envelope_here <- res$sample$envelope$common[[d_var]]
    quantiles_here <- res$sample$quantiles$common[[d_var]]
  } else {
    envelope_here <- res$sample$envelope$weighted[[d_var]]
    quantiles_here <- res$sample$quantiles$weighted[[d_var]]
  }
  if (!is.null(quantiles_here)) {
    matlines(
      x_here, 100 * (quantiles_here - 1),
      lty = c(2, 1, 2), lwd = c(1, 3, 1), col = "green"
    )
  }
  if (!is.null(envelope_here)) {
    matlines(
      x_here, 100 * (envelope_here - 1),
      lty = c(3, 2, 3), lwd = c(1, 2, 1), col = "red"
    )
  }
}
```

```{r byRegionTemplate, include=FALSE, eval=FALSE}
x_here <- res$sample$x[[d_var]]
s_group <- names(res$sample$envelope$group[[d_var]])
if (grepl("sqrt", d_var)) x_here <- x_here^2
if (grepl("div10", d_var)) x_here <- x_here * 10
sim_global_here_y <- 100 * (res$sample$quantiles$common[[d_var]] - 1)

for (d_group in s_group) {
  group_key <- as.character(d_group)
  the_ye <- 100 * (res$sample$envelope$group[[d_var]][[group_key]] - 1)
  the_yq <- 100 * (res$sample$quantiles$group[[d_var]][[group_key]] - 1)
  if (is.null(the_yq)) {
    warning("missing group quantiles for ", d_var, " / ", d_group)
    next
  }
  plot(
    x_here, sim_global_here_y[, 2],
    type = "l",
    xlab = d_var, ylab = "100 x (RR-1)",
    col = "grey", xaxs = "i", lwd = 4,
    ylim = the_y_lim[[d_var]], xlim = the_x_lim[[d_var]]
  )
  matlines(
    x_here, the_yq,
    lty = c(2, 1, 2), lwd = c(1, 2, 1), col = "green"
  )
  matlines(
    x_here, the_ye,
    lty = c(2, 1, 2), lwd = c(1, 2, 1), col = "red"
  )
}
```

```{r byRegionSetup1, eval=FALSE}
d_var <- grep("temp", names(res$sample$sim), value = TRUE)
s_group <- if (length(d_var)) {
  names(res$sample$quantiles$group[[d_var]])
} else {
  character(0)
}
```

```{r byRegion1, fig.cap=if (length(d_var)) s_var_names[d_var] else "temperature", fig.subcap = s_group, fig.height=4, fig.width=4, ref.label = 'byRegionTemplate', out.width='0.23\\linewidth', fig.align='center', eval=FALSE}
```

```{r byRegionSetup2, eval=FALSE}
d_var <- grep("[pP]m$", names(res$sample$sim), value = TRUE)
s_group <- if (length(d_var)) {
  names(res$sample$quantiles$group[[d_var]])
} else {
  character(0)
}
```

```{r byRegion2, fig.cap=if (length(d_var)) s_var_names[d_var] else "PM2.5", fig.subcap = s_group, fig.height=4, fig.width=4, ref.label = 'byRegionTemplate', out.width='0.23\\linewidth', fig.align='center', eval=FALSE}
```

```{r ecoRegionmap, fig.cap = 'Eco regions', fig.subcap = c('regions','legend'), fig.height=5, fig.width=5, out.width = '0.48\\linewidth', eval=FALSE}
mapmisc::map.new(india_eco_regions)
terra::plot(india_eco_regions, add = TRUE, col = india_eco_regions$col)
terra::text(terra::centroids(india_eco_regions), india_eco_regions$objectid)

mapmisc::map.new(india_eco_regions)
mapmisc::legendBreaks(
  "left",
  col = rev(india_eco_regions$col),
  legend = rev(india_eco_regions$id_name),
  cex = 0.7, bty = "n", width = Inf
)
```

```{r ecoTable, results='asis', include=FALSE, eval=FALSE}
knitr::kable(
  mapmisc::legendTable(data.frame(
    label = india_eco_regions$id_name,
    col = india_eco_regions$col
  ), type = "latex"),
  format = "latex", escape = FALSE,
  align = c("c", "l"),
  col.names = c("", "Eco region"),
  caption = "Eco-regions used in the analysis"
) |>
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    font_size = 9
  ) |>
  kableExtra::column_spec(1, latex_column_spec = "p{0.05\\\\linewidth}") |>
  kableExtra::column_spec(2, latex_column_spec = "p{0.7\\\\linewidth}")
```
