```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')


dataMedSimple =
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01'
    & deaths$tempClean >= (0)  & deaths$ecoRegion %in% 1:10 #setdiff(1:20, c(1,7,19)) # regions 1, 7, 19 have some very low temperatures.
    , ]
save(dataMedSimple, file='dataMedSimple.RData')
```




```{r fit}
# R CMD INSTALL --clean --preclean hpoltest_0.0.1.tar.gz   --configure-args="--enable-openmp" --configure-vars="PKG_CPPFLAGS='-fopenmp' PKG_LIBS='-fopenmp'"
# devtools::install_github("samperochkin/hpoltest")
# ulimit -s unlimited 


library('hpolcc')
# load("dataMedSimple.RData")

# add day iid
fitD <- hpolcc::hnlm(
  formula= medical ~ #f(date, model='iid') +
#    f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
#      knots = seq(0,32, by=2),
#      group_var = ecoRegion) + 
    f(tempClean, model='hiwp',ref_value = 20, p=3, 
      knots = seq(0, 45, by=2.5), 
    group_var = ecoRegion), 
  data=dataMedSimple, 
 control=list(maxit=1000, start.trust.radius = 1, prec=1e-6, stop.trust.radius = 1e-9,
        cg.tol = 1e-6, report.level=4, report.freq=1, report.header.freq=10, report.precision=7),
  control_inner=list(report.level=0),
  dirichlet= TRUE, config = list(num_threads=parallel::detectCores()),
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  for_dev=TRUE, verbose=TRUE
  )
#rm(dataMedSimple)
```

```{r sparsePlot}
    image(seq(0, len=nrow(fitD$groups$firstDeriv)), seq(0, len=ncol(fitD$groups$firstDeriv)), 
        as.matrix(fitD$groups$firstDeriv[, 1+fitD$groups$groups$i])
    )
    abline(h=fitD$groups$groups$p, col='blue', lty=1)
```

```{r dostuff}
library(hpolcc)


adfun = fitD$adfun()
grad2 = grad(fitD$start_gamma, data=fitD$tmb_data, config=fitD$config, adfun=adfun)
hes2 = hessian(fitD$start_gamma, data=fitD$tmb_data, config=fitD$config, adfun=adfun)


innerRes = trustOptim::trust.optim(
    x = fitD$start_gamma, 
    fn = jointLogDens,
    gr = grad,
    hs = hessian,
    method = 'Sparse',
    control = list(start.trust.radius = 0.1, report.freq=1, report.level=10),
    data=fitD$tmb_data, config = configHere, adfun = adfun
  )


cache = fitD$cache
assign("file", "india.txt", cache)
if(file.exists(get("file", cache))) file.remove(get("file", cache))

mleX =  trustOptim::trust.optim(
    x = c(0.021726796, -2.118758e-04 ,-4.253132, -7.918471, -5.099570 ,-7.996464, -2.621749), #fitD$parameters, #c(1, 0.3, log(c(0.05, 0.001, 0.05, 0.001))),
    fn = wrappers_outer$fn,
    gr = wrappers_outer$gr,
    method = 'SR1',
    control = list(start.trust.radius = 0.05, REPORT=1, trace=10),
    data=fitD$tmb_data, config = fitD$config, cache =  cache, controlInner = fitD$control_inner
  )
mleX$solution

bob = loglik(
  c(0.021726796, -2.118758e-04 ,-4.253132, -7.918471, -5.099570 ,-7.996464, -2.621749),
  gamma_start = get("gamma_start", cache),
 data=fitD$tmb_data, config = fitD$config, control = fitD$control_inner)


mleB <- BB::spg(
  par = c(0.021726796, -2.118758e-04 ,-4.253132, -7.918471, -5.099570 ,-7.996464, -2.621749),
  fn = wrappers_outer$fn,
  gr = wrappers_outer$gr,
  data=fitD$tmb_data, config = fitD$config, cache =  cache, controlInner = fitD$control_inner,
  control = list(maxit = 1e2, ftol=1e-6, gtol = 1e-4, M = 5, trace=TRUE, checkGrad=TRUE))  # M = nonmonotone history


  parameters = c(
    0.0, 0.01, # beta (slope, quadratic) 
    log(1), # sd global
    log(0.1), # sd local
    log(0.001), log(0.001), # sd local slope, quad
    log(0.1)[fitD$config$dirichlet] # dirichelet OD
    )


fitD$config$transform_theta = TRUE
  fitD$config$num_threads = parallel::detectCores()


  Nbeta = nrow(fitD$tmb_data$XTp)
  Ngamma = nrow(fitD$tmb_data$ATp)

  Ntheta = length(fitD$theta_info$name)



 gamma_start = seq(-0.01, 0.01, len=Ngamma) #mapOptim$solution[seq(Nbeta+1, len=Ngamma)]
    config = fitD$config
    config$transform_theta = TRUE


# likelihood
# parameters = (beta, theta)R
# config$transform_theta = TRUE means thetas are log(sd) otherwise theta=sd

# test when beta and theta in config


library('hpolcc')

config$verbose=FALSE
config$num_threads =  parallel::detectCores()
config$strataPerIter = 500

result <- loglik(parameters, 
             gamma_start, 
             data=fitD$tmb_data, 
    config = config,
      control=list(maxit=1000, start.trust.radius = 1, prec=1e-9, stop.trust.radius = 1e-15,
        cg.tol = 1e-9,
      report.level=4, report.freq=1)
      )
```










```{r mLE}

      control=list(maxit=1000, start.trust.radius = 1, prec=1e-9, stop.trust.radius = 1e-15,
        cg.tol = 1e-9, report.precision=7,
      report.level=4, report.freq=1, report.header.freq=1)
controlInner = control
controlInner$report.level=3
controlInner$report.freq=20
controlInner$report.header.freq=100
controlInner$report.precision = 5

config$verbose=FALSE
config$num_threads =  parallel::detectCores()
config$strataPerIter = 500

cache = new.env()
assign("Nfun", 0, cache)
assign("Ngr", 0, cache)

mle = trustOptim::trust.optim(
    x = parameters,
    fn = wrappers_outer$fn,
    gr = wrappers_outer$gr,
    method = 'BFGS',
    control = control,
    data=fitD$tmb_data, config = config, cache =  cache, controlInner = controlInner
  )
mle$extra = loglik(mle$solution, get("gamma_start", cache), data, config, controlInner)
mle$gamma_hat = mle$extra$solution
saveRDS(mle, file='mle.rds')
```



```{r theResults}
result$parameters = hpolcc::formatParameters(
  result$solution, fitD, TRUE)

result$parameters$theta

# parameters are (beta, gamma, theta)

randomEffects = seq(from=length(result$parameters$beta)+1,
  length = nrow(result$parameters$gamma))

result$Htilde = result$hessian[randomEffects, randomEffects]
result$HtildeChol = Matrix::solve(Matrix::chol(result$Htilde))

Nsim = 200
simInd = matrix(rnorm(Nsim * nrow(result$Htilde)),
 nrow(result$Htilde), Nsim)

# sample from [U|Y, theta_map]
simGamma = Matrix::crossprod(result$HtildeChol,  simInd) 
rownames(simGamma) = rownames(fitD$tmb_data$ATp)

simGamma = simGamma + unlist(result$parameters$gamma[
  rownames(simGamma), 'value'])


Stemp = seq(0,45,by=0.1)
SobsTemp = na.omit(match(Stemp, round(fitD$data$tempClean, 1)))
Spm = seq(1,1000, by=1)
SobsPm = na.omit(match(Spm, round(fitD$data$pm)))

Sobs = SobsTemp
nameX = 'tempClean'

Bmat = fitD$tmb_data$ATp[, Sobs]
# only want the global effects
Bmat = Bmat[grep(paste0(nameX,".*global"), rownames(Bmat)),]


simF = crossprod(Bmat, simGamma[rownames(Bmat),]) 

covHere = grep(nameX, names(result$parameters$beta), value=TRUE)
fixedPart = drop(crossprod(fitD$tmb_data$XTp[covHere,Sobs, drop=FALSE], 
                           result$parameters$beta[covHere]))

simSum = simF + fixedPart 



theX= fitD$data[[covHere[1] ]][Sobs]
if(grepl("^sqrt", nameX)) theX = theX^2

par(mar=c(3,3,0,0), bty='l')
matplot(theX, simSum, type='l',
  col='#00000010', lty=1, xaxs='i')


```



```{r otherstuff22, eval=FALSE}

Sstart = 1:(nrow(fitD$tmb_data$XTp)+1)
Send = seq(to = length(fitD$start_parameters), by=1, 
  len=length(fitD$theta_info$name)+1)

bob = function(xx) {
  hpolcc:::objectiveFunctionC( 
c(xx[Sstart], fitD$start_parameters[-c(Sstart, Send)], 
  xx[-Sstart]),
 fitD$tmb_data, 
 c(
 fitD$config[setdiff(names(fitD$config), c('verbose','maxDeriv'))], 
  list(maxDeriv = 0, verbose=FALSE))
  )$value
}

round(numDeriv::hessian(bob, fitD$start_parameters[c(Sstart, Send)]), 3)

res$hessian[c(Sstart, Send), c(Sstart, Send)]



beta = fitD$start_parameters[1:nrow(fitD$tmb_data$XTp)]
gamma =fitD$start_parameters[seq(from=length(beta)+1, 
  len=nrow(fitD$tmb_data$ATp)
  )]
theta = fitD$start_parameters[seq(from = length(beta) + length(gamma)+1,
  to=length(fitD$start_parameters))]


resFun = function(xx)  hpolcc:::objectiveFunctionC(xx, fitD$tmb_data, 
  c(fitD$config[
    setdiff(names(fitD$config), 'maxDeriv')], 
  list(maxDeriv = 0)))$value


resFun(fitD$start_parameters)  

Matrix::determinant(res$hessian)



fitD$config$verbose=FALSE
res$grad
theGradAgain = numDeriv::grad(resFun, fitD$start_parameters)  

rbind(res$grad, theGradAgain)



theHessAgain = numDeriv::hessian(resFun, fitD$start_parameters)


```

```{r originalPackage, eval=FALSE}
# using the hpoltest original package (and TMB)

fitM <- hnlm(
  formula= y ~ f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 22, p=3, knots = seq(0, 42, by=2), group_var = ecoRegion), 
  data=dataMedSimple,
  dirichlet=FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  tmb_parameters = list(theta = thetaStartM),
  optim_parameters = list(maxit = 20),
  verbose=TRUE, optimizer='optim', intern=FALSE, inner.method = 'BFGS', 
  inner.control = list(maxit = 500), for_dev=FALSE)
```

```{r plotFit2, fig.height=4, fig.width=4, eval=FALSE}
simTemp =  hpoltest:::condSim(fitD, 'tempClean') 

theYlim = c(0.2, 5) #exp(quantile(simTemp$y[,'GLOBAL', ], c(0.01, 0.99)))
for(D in dimnames(simTemp$y)[[2]]) {
matplot(simTemp$x, exp(simTemp$y[,D, ]), type='l', col='#00000010', lty=1, 
        xaxs='i', log='y',
        ylim =theYlim,bty='n', 
        xlab='temperature',ylab=D)
lines(simTemp$x, exp(simTemp$fixed_mean), col='yellow')
lines(simTemp$x, exp(apply(simTemp$y[,D, ], 1, mean)), col='red')
}
```
