```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(
  file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')


dataMedSimple =
  deaths[
    deaths$date >= as.Date('2010/1/1') &
    format(deaths$date, '%b%d') != 'Jan01'
    & deaths$tempClean >= (0) # & deaths$ecoRegion %in% 1:10 #setdiff(1:20, c(1,7,19)) # regions 1, 7, 19 have some very low temperatures.
    , ]
save(dataMedSimple, file='dataMedSimple.RData')
```




```{r fit}
library('hpolcc')
load("dataMedSimple.RData")

res <- hpolcc::hnlm(
  formula= medical ~ #f(date, model='iid') +
    f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
      knots = seq(0,32, by=2),
      group_var = ecoRegion) + 
    f(tempClean, model='hiwp',ref_value = 25, p=3, 
      knots = seq(0, 45, by=2.5), 
    group_var = ecoRegion), 
  data=dataMedSimple, 
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
    dirichlet= TRUE, 
  config = list(transform_theta = TRUE, 
    dense = FALSE,
    num_threads=parallel::detectCores(), 
    Nclusters = 4*parallel::detectCores()),
 control=list(maxit=1000, start.trust.radius = 1, prec=1e-6, stop.trust.radius = 1e-9,
        cg.tol = 1e-6, report.level=4, report.freq=1, report.header.freq=10, report.precision=7),
  control_inner=list(report.level=0),
  for_dev=TRUE, verbose=TRUE
  )
saveRDS(res, file='res.rds')
#rm(dataMedSimple)
```

```{r sparsePlot}
par(mar=c(3,3,0,0), bty='n')
bob= terra::rast(as.matrix(1.1*res$groups$firstDeriv[, 
  1 + res$groups$groups$i]))
terra::plot(bob, asp=FALSE, col=c('white','black'))
abline(v = res$groups$groups$p,
       col = 'blue',
       lty = 1)
```

```{r mle}
res = readRDS("res.rds")
library('hpolcc')
adFunFull = getAdFun(res$parameters_and_gamma, res$tmb_data, res$config)


assign("file", "india.txt", res$cache)
if(file.exists(get("file", res$cache))) file.remove(get("file", res$cache))


mleX1 = optim(
    rep(c(0,-3), 
      c(length(res$config$beta), length(res$config$theta))),
    fn = outer_fn,
    gr = outer_gr,
    method = 'L-BFGS-B',#'SR1',
    lower = rep(c(-0.5, -Inf), c(length(res$config$beta), length(res$config$theta))),
    upper = rep(c(0.5, 1), c(length(res$config$beta), length(res$config$theta))),
    control = list(trace=3, REPORT=1, 
      parscale = rep(c(1e-3, 10), c(length(res$config$beta), length(res$config$theta)))
      ),#start.trust.radius = 0.05, REPORT=1, trace=10),
    data=res$tmb_data, config = res$config, cache =  res$cache, control_inner = res$control_inner,
    adFunFull = adFunFull
  )
saveRDS(mleX1, file='mleX1.rds')  
mleX = trustOptim::trust.optim(
#      rep(c(0,-3, -1),  c(length(res$config$beta), length(res$config$theta)-1), 1),
    mleX1$par,
    fn = outer_fn,
    gr = outer_gr,
    method = 'SR1',
    control = list(start.trust.radius = 0.05, REPORT=1, trace=10),
    data=res$tmb_data, config = res$config, cache =  res$cache, control_inner = res$control_inner,
    adFunFull = adFunFull
  )  
saveRDS(mleX, file='mleX.rds')  
```



```{r theResults}
library('hpolcc')
res = readRDS("resMed.rds")
result = mleX = readRDS("mleMed.rds")
mleXcheck = loglik(
  parameters = mleX$par, 
  gamma_start = get("gamma_start", res$cache),
  config=res$config, data=res$tmb_data, adFunFull=adFunFull, deriv=0)

result$parameters = hpolcc::formatParameters(
  c(mleX$par[1:2], mleXcheck$solution, mleX$par[-(1:2)]), 
  res, TRUE)

result$parameters$theta

# parameters are (beta, gamma, theta)

result$Htilde = mleXcheck$hessian
result$HtildeChol = as(mleXcheck$cholHessian,"Matrix")
result$HtildeCholInv = solve(result$HtildeChol)

Nsim = 500
simInd = matrix(rnorm(Nsim * nrow(result$Htilde)),
 nrow(result$Htilde), Nsim)

# sample from [U|Y, theta_map]
simGamma = Matrix::crossprod(result$HtildeCholInv,  simInd) 
rownames(simGamma) = rownames(res$tmb_data$ATp)

simGamma = simGamma + unlist(result$parameters$gamma[
  rownames(simGamma), 'value'])


Stemp = seq(0,45,by=0.1)
SobsTemp = na.omit(match(Stemp, round(res$data$tempClean, 1)))
newConstr = 25;newConstrIndex = which.min(abs(
  res$data$tempClean[SobsTemp] - newConstr))

#Spm = seq(1,1000, by=1)
#SobsPm = na.omit(match(Spm, round(res$data$pm)))

Sobs = SobsTemp
nameX = 'tempClean'


covHere = grep(nameX, names(result$parameters$beta), value=TRUE)
xHere = res$tmb_data$XTp[covHere,Sobs, drop=FALSE]
fixedPart = drop(crossprod(xHere, 
              result$parameters$beta[covHere]))


BmatAll = res$tmb_data$ATp[, Sobs]
# only want the global effects
Bmat = BmatAll[grep(paste0(nameX,".*global"), rownames(BmatAll)),]


simF = crossprod(Bmat, simGamma[rownames(Bmat),]) 


simSum = simF + fixedPart 

toSubtract = matrix(simSum[newConstrIndex,],
  nrow(simSum), ncol(simSum), byrow=TRUE)
simSum = simSum - toSubtract

theX= res$data[[covHere[1] ]][Sobs]
if(grepl("^sqrt", nameX)) theX = theX^2

par(mar=c(3,3,0,0), bty='l')
matplot(theX, 100*(exp(simSum)-1), type='l',
  col='#00000010', lty=1, xaxs='i',
  ylim = c(-1/3, 31), xlim = c(6, 38))


# eco region 2
D = 3
Bmat2 = BmatAll[grep(paste0(nameX,"_h(iwp|rpoly)_",D,"_"), rownames(BmatAll)),]
simF2 = crossprod(Bmat2, simGamma[rownames(Bmat2),]) 
simSum2 = simSum + simF2 
matplot(theX, 100*(exp(simSum2)-1), type='l',
  col='#00000010', lty=1, xaxs='i',
  ylim = c(-5, 50), xlim = c(6, 38))
```







```{r moreoldstuff, eval=FALSE}
library('hpolcc')
adFunFull = getAdFun(res$parameters_and_gamma, res$tmb_data, res$config)
outer_fn(
  res$parameters,
  data = res$tmb_data, config= modifyList(res$config, list(verbose=TRUE)), 
  adFunFull = adFunFull, control_inner = res$control_inner, cache=res$cache
  )

xx = loglik(res$parameters, get("gamma_start", res$cache), res$tmb_data, res$config, deriv=0, check=TRUE, 
  control=list(prec=1e-9, stop.trust.radius=1e-12, cg.tol=1e-9, report.level=4, report.freq=1))
quantile(xx$solution - xx$oldSolution)

dataNoQ = res$tmb_data
dataNoQ$Qdiag = rep(1, 0)



myh = hessianDense(xx$fullParameters, dataNoQ, res$config)
myh2 = hessian(xx$fullParameters, dataNoQ, modifyList(res$config, list(verbose=TRUE)), adFunFull)
myh2d = hessian(xx$fullParameters, dataNoQ, modifyList(res$config, list(verbose=TRUE, debug=TRUE)), adFunFull)

config1 = config2 = res$config
config1$groups$p = res$config$groups$p[1:2]
config2$groups$p = res$config$groups$p[2:3]
myhG = hessianDense(xx$fullParameters, dataNoQ, config1)
myhG2 = hessianDense(xx$fullParameters, dataNoQ, config2)

myh2d1 = Matrix::sparseMatrix(
  i = res$config$group_sparsity[[1]]$second$full$i,
  j = res$config$group_sparsity[[1]]$second$full$j,
  x = myh2d[[1]], index1=FALSE, symmetric=TRUE
  )
bob = myhG - as.matrix(myh2d1)
quantile(bob)

quantile(myhG *(1-hessianByBlock2[[1]]))
quantile(as.matrix(myh2d1) *(1-hessianByBlock2[[1]]))

quantile(bob)


idx <- min(which(abs(bob)>0.1))
arrayInd(idx, dim(bob))


round(cbind(myh2d1[1:20,71:75],NA, myhG[1:20,71:75]),1) 

bob1 = as.matrix(myh2) - myh
quantile(bob1)


(i1 = which.max(apply(abs(bob1),1,max)))
(i2 = which.max(abs(bob1[i1,])))

sseq = seq(-2,2)
round(cbind(myh2[i1+sseq, i2+sseq], NA, myh[i1+sseq, i2+sseq]),2)


resDense = res
Nparams = length(xx$fullParameters)
newDenseSecondDf = expand.grid(i=seq(0,len=Nparams), j=seq(0,len=Nparams))
newDenseSecondDf = newDenseSecondDf[newDenseSecondDf$j >= newDenseSecondDf$i, ]

theOrig= as.data.frame(res$config$group_sparsity[[1]]$second$full[c('i','j')]))
xxx = rbind(
  newDenseSecondDf[newDenseSecondDf$j < (Nparams-1),],
  theOrig[thrOrig$j >= (Nparams-1), ])
xxx = xxx[!duplicated(xxx),]
xxx = xxx[order(xxx$j, xxx$i),]

newDenseSecondDf = as.list(xxx)
newDenseSecondDf$p = c(0,cumsum(as.vector(table(newDenseSecondDf$j))))
newDenseSecondDf$match = seq(0, len=length(newDenseSecondDf$i))
resDense$config$sparsity$second$full = newDenseSecondDf
resDense$config$group_sparsity[[1]]$second$full = newDenseSecondDf
configDense2 = configDense1 = resDense$config
configDense1$groups$p = res$config$groups$p[1:2]

myh3 = hessian(xx$fullParameters, dataNoQ, modifyList(configDense1, list(debug=TRUE,verbose=TRUE)))
myh3a = Matrix::sparseMatrix(
  i = resDense$config$group_sparsity[[1]]$second$full$i,
  j = resDense$config$group_sparsity[[1]]$second$full$j,
  x = myh3[[1]], index1=FALSE, symmetric=TRUE
  )

quantile(as.matrix(myh3a) *(1-hessianByBlock2[[1]]))
quantile(as.matrix(myh3a) - myhG)


quantile(as.matrix(myh2d1) - as.matrix(myh3a))



myh3 = hessian(xx$fullParameters, dataNoQ, modifyList(resDense$config, list(debug=TRUE,verbose=TRUE)))

myh3d1 = Matrix::sparseMatrix(
  i = resDense$config$group_sparsity[[1]]$second$full$i,
  j = resDense$config$group_sparsity[[1]]$second$full$j,
  x = myh3[[1]], index1=FALSE, symmetric=TRUE
  )
myh3d2 = Matrix::sparseMatrix(
  i = resDense$config$group_sparsity[[2]]$second$full$i,
  j = resDense$config$group_sparsity[[2]]$second$full$j,
  x = myh3[[2]], index1=FALSE, symmetric=TRUE
  )
bob = myhG - as.matrix(myh3d1)
bob2 = myhG2 - as.matrix(myh3d2)

myhG2a = Matrix::forceSymmetric(Matrix::Matrix(myhG2))

bob1 = as.matrix(myh3) - myh
quantile(bob1)
(i1 = which.max(apply(abs(bob1),1,max)))
(i2 = which.max(abs(bob1[i1,])))
round(cbind(myh3[i1+sseq, i2+sseq], NA, myh[i1+sseq, i2+sseq]),2)
idx <- which.min(abs(myh - 24112585))
arrayInd(idx, dim(myh))


myh1 = hessianDense(xx$solution, res$tmb_data, res$config)
myh2a = hessian(xx$fullParameters, res$tmb_data, modifyList(res$config, list(verbose=TRUE)))
myh3 = hessian(xx$solution, res$tmb_data, res$config)





which(myh2@x == myh2[i1, i2])
myh2[i1, i2] # should be 0.2308968 not -19317.62

Sgamma1 = seq(length(res$config$beta)+1, len=legnth(res$gamma_start))
bob2 = as.matrix(myh2[Sgamma1, Sgamma1]) - as.matrix(myh3) #xx$hessian)


quantile(myh[Sgamma1, Sgamma1] - myh1)

outer_gr(
  res$parameters,
  data = res$tmb_data, config= modifyList(res$config, list(verbose=TRUE)), 
  adFunFull = adFunFull, control_inner = res$control_inner, cache=res$cache
  )

jointLogDens(res$parameters_and_gamma, res$tmb_data, modifyList(res$config, list(verbose=TRUE)))
gD = grad(res$parameters_and_gamma, res$tmb_data, 
  modifyList(res$config, list(verbose=TRUE, dense=TRUE)), 
  adFun = adFunFull)
gS = grad(res$parameters_and_gamma, res$tmb_data, 
  modifyList(res$config, list(verbose=TRUE, dense=FALSE)), 
  adFun = adFunFull)
quantile(gD - gS)  


bob = function(xx) {
  jointLogDens(c(
    res$parameters_and_gamma[seq(1, length(res$parameters_and_gamma) - length(xx))], xx), 
    res$tmb_data, modifyList(res$config, list(verbose=FALSE)))
}
theSeq = seq(to=length(res$parameters_and_gamma), len=5)
bob2 = numDeriv::grad(bob, res$parameters_and_gamma[theSeq])
rbind(bob2, gD[theSeq], gS[theSeq])

adFunHere = getAdFun(
  res$gamma_start, 
  res$tmb_data, 
  modifyList(res$config, list(verbose=TRUE)))

gDr = grad(res$gamma_start, res$tmb_data, 
  modifyList(res$config, list(verbose=TRUE, dense=TRUE)), 
  adFun = adFunHere)
gDr1 = grad(res$gamma_start, res$tmb_data, 
  modifyList(res$config, list(verbose=TRUE, dense=TRUE)))
gSr = grad(res$gamma_start, res$tmb_data, 
  modifyList(res$config, list(verbose=TRUE, dense=FALSE)), 
  adFun = adFunHere)
quantile(gDr - gSr)  


mys = seq(-3,3) + which.max(abs(gDr - gSr))
rbind(mys,rbind(gDr, gSr,gD=gD[Sgamma1], gS=gS[Sgamma1])[,mys] )

rbind(gD[Sgamma1], gS[Sgamma1], gDr, gDr1, gSr)[,1:5]



x0= hessian(res$gamma_start, res$tmb_data, modifyList(res$config, list(verbose=TRUE)), 
  adFun = adFunHere)
system.time(x1 <- 
  hessian(res$parameters_and_gamma, res$tmb_data, modifyList(res$config, list(verbose=TRUE, dense=FALSE)), 
  adFun = adFunFull))
system.time(x2 <- 
  hessianDense(res$parameters_and_gamma, res$tmb_data, modifyList(res$config, list(verbose=TRUE))))


```
