```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')


dataMedSimple =
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01'
    & deaths$tempClean >=0
#    & deaths$ecoRegion %in% 2:6 #setdiff(1:20, c(1,7,19)) # regions 1, 7, 19 have some very low temperatures.
    , ]
save(dataMedSimple, file='dataMedSimple.RData')
# load("dataMedSimple.RData")
```




```{r tmbpackage, eval=FALSE}
library(hpoltest)
formula <- count ~ f(date, model = 'iid')  + 
  f(tempClean, model = 'hiwp', ref_value = 20, p=3, 
      knots = seq(0, 45, by=2.5), 
    group_var = ecoRegion)
  fit <- hnlm(formula, dataMedSimple, 
    cc_design =  ccDesign(time_var = 'date',
    strat_vars=c("cell","ecoRegion", "yearWeek3Dow"))
  )
  sdr <- sdreport(fit$obj, getJointPrecision=TRUE)
  mu <- fit$obj$env$last.par

```


```{r fit}
# R CMD INSTALL --clean --preclean hpoltest_0.0.1.tar.gz   --configure-args="--enable-openmp" --configure-vars="PKG_CPPFLAGS='-fopenmp' PKG_LIBS='-fopenmp'"
# devtools::install_github("samperochkin/hpoltest")
# ulimit -s unlimited 


library('hpolcc')

fitD <- hpolcc::hnlm(
  formula= medical ~ #f(date, model='iid') +
#    f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
#      knots = seq(0,32, by=2),
#      group_var = ecoRegion) + 
    f(tempClean, model='hiwp',ref_value = 20, p=3, 
      knots = seq(0, 45, by=5), 
    group_var = ecoRegion), 
  data=dataMedSimple[dataMedSimple$ecoRegion < 7,],
  dirichlet= FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  for_dev=TRUE, verbose=TRUE
  )

  parameters = c(
    0.0, 0.01, # beta (slope, quadratic) 
    log(1), # sd global
    log(0.1), # sd local
    log(0.001), log(0.001), # sd local slope, quad
    log(0.1)[fitD$config$dirichlet] # dirichelet OD
    )


fitD$config$transform_theta = TRUE
if(ps::ps_system_memory()$total > 5e10) {
  fitD$config$num_threads = parallel::detectCores()
} else {
  fitD$config$num_threads = 8
}


  Nbeta = nrow(fitD$tmb_data$XTp)
  Ngamma = nrow(fitD$tmb_data$ATp)

  Ntheta = length(fitD$theta_info$name)


 gamma_start = runif(Ngamma, -0.01, 0.01) #mapOptim$solution[seq(Nbeta+1, len=Ngamma)]

    config = fitD$config
    config$verbose=FALSE
    config$dense = TRUE
    config$transform_theta = TRUE

# likelihood
# parameters = (beta, theta)R
# config$transform_theta = TRUE means thetas are log(sd) otherwise theta=sd



if(exists("result")) gamma_start = result$solution



library('hpolcc')

result = loglik(parameters, 
             gamma_start, 
             data=fitD$tmb_data, 
    config = config,
    control = list(maxit=1000, start.trust.radius = 0.1,
      report.level=4, report.freq=1))


result$extra
```

```{r checkSimple}
library(hpolcc)
config1 = fitD$config
config1$dense = TRUE
config1$transform_theta = TRUE
config1$maxDeriv = 2
config1$verbose=FALSE

  Dpar = 2;Dpar2=4;Dpar3 = 5

  Sp1 = seq(-0.0005, 0.0005, len=11)+paramsAll[Dpar]

  bob =  mapply(function(xx) {
    parameters2 = paramsAll
    parameters2[Dpar] = xx
    hpolcc:::objectiveFunctionC(parameters2, fitD$tmb_data, config1)
    }, xx =Sp1, SIMPLIFY=FALSE)

bobC = bob[[6]]
bob0 = unlist(lapply(bob, function(xx) xx$value))
bob1 = unlist(lapply(bob, function(xx) xx$grad[Dpar]))
bob1a = unlist(lapply(bob, function(xx) xx$grad[Dpar2]))
bob2 = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar]))
bob2a = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar2]))
bob2b = unlist(lapply(bob, function(xx) xx$hessian[Dpar2,Dpar3]))


bobD = thirdDiagonals(
      paramsAll, fitD$tmb_data, config1
    ) 


    pairs1 = expand.grid(i=Sparams, j=seq(from=Nbeta, len=Ngamma), p=0, pEnd=0)
    pairs1$index = 1:nrow(pairs1)
config1$sparsity = list(third=list(pairs=pairs1))


bob3 = thirdOffDiagonals(
      paramsAll, fitD$tmb_data, config1
    ) 
bob4 = matrix(
  bob3[[3]][,pairs1[pairs1$i == (Dpar-1) & pairs1$j == (Dpar2-1), 'index']],
  ncol=3, byrow=TRUE)


pairs1[apply(pairs1, 1, function(xx) all(xx[c('i','j')] %in% (c(Dpar, Dpar2, Dpar3)-1))), ]

# grad 
plot(Sp1, bob1, type='l')
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob0)/diff(Sp1))
abline(h=bobD$first[Dpar])
abline(v=paramsAll[Dpar])

abline(h=
  bob3[[1]][1, pairs1[pairs1$i == (Dpar-1) & pairs1$j == (Dpar2-1), 'index']],
  col='green', lty=3
  )
c(bobD$first[Dpar,1], bob4[Dpar, 3])



# 2nd Dpar Dpar
plot(Sp1, bob2, type='l', lwd=3)
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob1)/diff(Sp1))
abline(h=bobD$second[Dpar,Dpar], col='red')
abline(v=paramsAll[Dpar])

c(bobD$second[Dpar,Dpar], bob4[Dpar, 2], bobC$hessian[Dpar, Dpar])


# 2nd Dpar Dpar2
plot(Sp1, bob2a, type='l', lwd=3)
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob1a)/diff(Sp1))
abline(h=bobD$second[Dpar,Dpar2], col='red')
abline(v=paramsAll[Dpar])

# Dpar2 Dpar Dpar
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2a)/diff(Sp1)
  )
abline(h=2*bobD$diag[Dpar2, Dpar])


# 2nd Dpar2 Dpar3
plot(Sp1, bob2b, type='l', lwd=3)
abline(h=bobD$second[Dpar2,Dpar3], col='red')
abline(v=paramsAll[Dpar])

# Dpar Dpar2 Dpar3
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2b)/diff(Sp1)
  )
abline(h=(bob4[Dpar3, 1] - bobD$diag[Dpar3, Dpar] - bobD$diag[Dpar3, Dpar2]))


pairs1[pairs1$i == (Dpar-1) & pairs1$j == (Dpar2-1),  ]
dim(bob4)



c(  
  bob3[[2]][1, pairs1[pairs1$i == (Dpar-1) & pairs1$j == (Dpar2-1), 'index']],
  bobC$hessian[Dpar, Dpar]/2 + bobC$grad[Dpar2])



```

```{r checkDeriv}
order(apply(abs(resThirdDiag$diag[1:70,1:70]), 1, max), decreasing=TRUE)[1:6]
order(apply(abs(resThirdDiag$diag), 2, max), decreasing=TRUE)[1:6]
resThirdDiag$diag[,34]

  Dpar = 2
  Sp1 = seq(-0.0005, 0.0005, len=11)+result$fullParameters[Dpar]

  bob =  mapply(function(xx) {
    parameters2 = result$fullParameters
    parameters2[Dpar] = xx
    hpolcc:::objectiveFunctionC( 
      parameters2,
   fitD$tmb_data, 
   c(
   fitD$config[setdiff(names(fitD$config), c('verbose','maxDeriv'))], 
    list(maxDeriv = 2, verbose=FALSE, transform_theta=TRUE))
    )
    }, xx =Sp1, SIMPLIFY=FALSE)

  bob0 = unlist(lapply(bob, function(xx) xx$value))
  plot(Sp1, bob0)

bob1 = unlist(lapply(bob, function(xx) xx$grad[Dpar]))
bob2 = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar]))

plot(Sp1, bob1, type='l')
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob0)/diff(Sp1))


abline(h=resThirdDiag$first[Dpar,1])

plot(Sp1, bob2, type='l')
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob1)/diff(Sp1))
abline(h=resThirdDiag$second[Dpar,Dpar], col='red', lty=2)

plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2)/diff(Sp1)) 
abline(h=2*resThirdDiag$diag[Dpar,Dpar], col='red', lty=2)


Dpar2 = 4
bob2 = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar2]))
bob1 = unlist(lapply(bob, function(xx) xx$grad[Dpar2]))
plot(Sp1, bob2, type='l', ylim = range(c(bob2, bob2[1]+ c(-1, 1))))
points(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob1)/diff(Sp1))
abline(h=resThirdDiag$second[Dpar2,Dpar], col='red', lty=2)


plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2)/diff(Sp1))
abline(h=2*resThirdDiag$diag[Dpar2,Dpar], col='red', lty=2)

Dpar3 = 5
bob2 = unlist(lapply(bob, function(xx) xx$hessian[Dpar,Dpar3]))
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(bob2)/diff(Sp1))
abline(h=2*resThirdDiag$diag[Dpar3,Dpar], col='red', lty=2)



bob2 = unlist(lapply(bob, function(xx) xx$hessian[Dpar3,Dpar2]))
plot(Sp1, bob2)

stuff = thirdNonDiag[thirdNonDiag$i == (Dpar-1) & thirdNonDiag$k %in% Sgamma0, ]
parhere = c(Dpar2, Dpar3)-1
parhere2 = c(Dpar, Dpar2, Dpar3)-1

stuff[stuff$k %in% parhere & stuff$j %in% parhere,]
thirdDiag[thirdDiag$k %in% parhere2 & thirdDiag$i %in% parhere2, ]


tijk = data.frame(
  x=Sp1[-1] - diff(Sp1)/2, 
  y=diff(bob2)/diff(Sp1)
  )
plot(tijk, ylim = range(outer(tijk$y, c(-10,10), "+")))

abline(h=stuff[apply(
  stuff[,c('i','j','k')],
  1, 
  function(xx) all(xx %in% parhere2)
  ), 'x']
  )

abline(h=thirdList[[Dpar]][Dpar2, Dpar3], col='red', lwd=3)
abline(h=thirdList[[Dpar2]][Dpar, Dpar3], col='blue', lty=3, lwd=3)
abline(h=thirdList[[Dpar3]][Dpar2, Dpar], col='orange')

thirdNonDiag[apply(thirdNonDiag[,c('i','j','k')], 1,
function(xx)  all(xx %in% (c(Dpar, Dpar2, Dpar3)-1)) 
), ]

thirdOffDiag1[apply(thirdOffDiag1[,c('i','j','k')], 1,
function(xx)  all(xx %in% (c(Dpar, Dpar2, Dpar3)-1)) 
), ]
thirdDiag[apply(thirdDiag[,c('i','j','k')], 1,
function(xx)  all(xx %in% (c(Dpar, Dpar2, Dpar3)-1)) 
), ]

Sdet = unlist(lapply(bob, function(xx) Matrix::determinant(Matrix::forceSymmetric(xx$hessian[Sgamma1, Sgamma1]),
    log = TRUE, sqrt=FALSE)$modulus))
plot(Sp1, Sdet)

Slik = unlist(lapply(bob, function(xx) xx$value)) + Sdet/2
plot(Sp1, Slik)


plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(Slik)/diff(Sp1))

abline(h=sum(result$extra[Dpar,c('theta','VHH')]))

Sthird = list()
for(Dthird in seq(1, length(bob)-1)) {
  Sthird[[Dthird]] = (bob[[Dthird+1]]$denseHessian - bob[[Dthird]]$denseHessian)/(diff(Sp1)[Dthird])
}

Sthird[[5]][1:4,1:4]

thirdList[[Dpar]][2:5,3:5]


bob2 = (Sthird[[5]] - thirdList[[Dpar]]


lapply(bob, function(xx) xx$hessian[c(Dpar, Dpar2, Dpar3),c(Dpar, Dpar2, Dpar3)]    )
plot(
  Sp1[-1] - diff(Sp1)/2, 
  diff(Sdet)/diff(Sp1),
  ylim = range(outer(diff(Sdet)/diff(Sp1),c(0.5, 2))))
abline(h=sum(Matrix::diag(invHessianRandom %*% Sthird[[5]][Sgamma1, Sgamma1])))

(stuff = thirdNonDiag[apply(thirdNonDiag[,c('i','j','k')], 1, 
  function(xx) all(sort(xx) == (sort(c(Dpar, Dpar2, Dpar3)-1)))), ])

resThirdDiag$second[c(Dpar, Dpar2, Dpar3),c(Dpar, Dpar2, Dpar3)]

```

```{r stuff}
Sgamma = seq(from=Nbeta+1, len=Ngamma)
Sparams = setdiff(1:length(result$fullParameters), Sgamma)
Sk = 1:length(Sparams)

step = 0.0001
result2 = list()
for(Dk in Sk) {
result2 = c(result2, list(loglik(
  parameters = result$parameters + step * (1:length(parameters) == Dk),
  config = result$config,
  gamma_start = result$solution,
  data = fitD$tmb_data,
    control = list(maxit=100, start.trust.radius = 0.1,
      report.level=4, report.freq=1)
  )))
}


# numeric differentiation of hessians to get third deriv
diffH = lapply(result2, function(xx) {
  (xx$fullHessian - result$fullHessian)/step 
  })

unlist(lapply(diffH, function(Hinv, Tuux, Sgamma1) {
      sum(Matrix::diag(Hinv %*% Tuux[Sgamma1,Sgamma1]))
  },  
  Hinv=invHessianRandom, Sgamma1 = Sgamma1))

# check diagonals
thirdDiagDiff = do.call(cbind, lapply(diffH, function(xx) {
    Matrix::diag(xx)
  })) # T_iik, columns are k

thirdDiagDiff[1:9,]
resThirdDiag$diag[1:9, Sparams]




diffTaylorDiag = 0.5*diffTaylorDiag1 + result$dense$hessian[,Sparams]
diffTaylorDiag[1:12,]

# in derivLogDet.cpp, columns are i.  transpose
t(result$dense$diagTaylor)[1:12,Sparams]


# 3rd parameter, 78 in full hessian
Dk = 79;Dk0 = Dk-1;Dpar = which(Sparams == Dk)

Tarray[1,2,Dpar]
Tarray[Dk,2,1]
Tarray[2,Dk,1]
Tarray[1,Dk,2]


# should have 
bob=cbind(
  (result2[[1]]$dense$hessian - result$dense$hessian)[,Dk],
  (result2[[Dpar]]$dense$hessian - result$dense$hessian)[,1]
)



Dik = which(apply(
  as.matrix(result$config$sparsity$third$ijk[,c('i','j','k')])==Dk0,
  1, any))

newijk = do.call(rbind,apply(
  result$config$sparsity$third$ijk[Dik,c('i','j','k')], 
  1,
  function(xx) sort(setdiff(xx, Dk0)),
  simplify=FALSE
  ))
newijk = newijk[order(newijk[,1], newijk[,2]),]  
bob = Matrix::sparseMatrix(
  i = newijk[,1],
  j=newijk[,2],
  x=result$third$third[Dik],
  dims = rep(Nparams, 2),
  index1=FALSE
  )
bob2 = (bob + Matrix::t(bob))
diffH[[which(Sparams == Dk)]][1:5,1:5]


result$dense$hessian[1:5, 75:82]
result2[[3]]$dense$hessian[1:5, 75:82]
diffH[[3]][1:5, 75:82]
result$third$thirdList[[78]][1:5, 75:82]

theDiff = diffH[[1]] - as.matrix(result$dense$third[[1]])

(t(theDiag)/2 + result$dense$hessian[-Sgamma1,])[,1:5]
result$dense$diagTaylor[-Sgamma1, 1:5]


theDiag[1:5,]
result$dense$diag[1:5, -Sgamma1]


quantile(theDiff[Sgamma1,Sgamma1])

Strace = unlist(lapply(diffH, function(Hinv, Tuux, Sgamma) {
      sum(Matrix::diag(Hinv %*% Tuux[Sgamma,Sgamma]))
  },  
  Hinv=result$dense$invHessianRandom, Sgamma = Sgamma1))

rbind(
  num=Strace, 
  dense=result$dense$trace[-Sgamma1],
unlist(lapply(result2, function(xx) xx$dense$logDet- result$dense$logDet))/step
)

bob1 = diffH[[1]][Sgamma1,Sgamma1]
bob2 = result$dense$third[[1]][Sgamma1, Sgamma1]

# strange one is
c(0,0,4)

llik=rbind(result$minusLogLik, unlist(lapply(result2, function(xx) xx$minusLogLik)))
rbind(
  num=apply(llik, 2, diff)/step,
  sparse=result$gradL, 
  dense=result$dense$extra$grad)



diffH = (result2$third$denseHessian - result$third$denseHessian)/step
diffHad = result$third$thirdList[[82]][Sgamma1,Sgamma1]
diffH[1:9,1:9]
diffHad[1:9,1:9]


Sindex = 2+1:6

result$hessian[Sindex,Sindex]
result$third$denseHessian[Sindex,Sindex]

  # taylor is   T_iik/2 + T_jjk/2 +  T_ijk 
Tiik= Matrix::diag(diffH)
taylor2 = Tiik/2 + result$third$denseHessian[,82]
toAdd = outer(diagT, diagT, FUN="+")
taylor = diffH + toAdd/2


denseTaylor = Matrix::sparseMatrix(
  i=result$config$sparsity$third$ij$i,
  j=result$config$sparsity$third$ij$j,
  x=result$third$denseThird[,82],
  index1=0, dims=dim(result$third$denseHessian), symmetric=TRUE)

 result$third$denseDiag[82, 1:5]
 result$third$denseDiag[1:5, 82]
 taylor2[1:5]

range( (taylor2 -  result$third$denseDiag[82,])/taylor2, na.rm=TRUE)

plot(abs(taylor2), abs(result$third$denseDiag[82,]), log='xy');abline(c(0,1))

denseThirdDiag = 2*(result$third$denseDiag[82,] - result$third$denseHessian[,82])


plot(abs(denseThirdDiag), abs(Tiik), log='xy');abline(c(0,1))


plot(abs(as.vector(taylor)), abs(as.vector(denseTaylor)), log='xy');abline(c(0,1))

# should be the same as taylor

denseTaylor[Sindex,Sindex]
taylor[Sindex,Sindex]
diffH[Sindex,Sindex]
result$third$denseHessian[Sindex,Sindex]

TijkAd = denseTaylor - outer(denseThirdDiag, denseThirdDiag, FUN='+')

TijkAd[Sindex,Sindex]
diffH[Sindex,Sindex]

(dpart = c(result$fval, result2$fval))
diff(dpart)/step
(ldet=c(result$logDetHessian, result2$logDetHessian))
diff(ldet)/step
sum(Matrix::diag(solve(result$third$denseHessian)%*% diffH))
sum(Matrix::diag(solve(result$hessian)%*% thirdList[[82]][Sgamma1,Sgamma1]))


cbind(result$extra, result2$extra)





result$gradL[1]

```

```{r map}
library('hpolcc')

  parameters = c(
    0.0, 0.01, # beta (slope, quadratic) 
    rep(0, nrow(fitD$tmb_data$ATp)),
    log(1), # sd global
    log(0.1), # sd local
    log(0.001), log(0.001), # sd local slope, quad
    log(0.1)[fitD$config$dirichelet] # dirichelet OD
    )

map = map_estimator(
  parameters, 
  data = fitD$tmb_data,
  config = list(verbose=FALSE, transform_theta=TRUE, dirichelet=TRUE, num_threads=4),
    control = list(
      maxit=1000, start.trust.radius = 0.1,trust.iter= 1e5,
      report.level=4, report.freq=1)
  )
map$map$parameters  
```




```{r doSomeSim, eval=FALSE}

Nsim = 200
gammaSim =  res$gamma_hat + 
  cholVariance %*% matrix(
  rnorm(Nsim*nrow(cholVariance)),
  nrow(cholVariance), Nsim)


# only plot the global effects
rowsToKeep = grep("temp.*global", rownames(fitD$tmb_data$ATp))

Stemp = seq(0,41,by=0.5)
SobsTemp = na.omit(match(Stemp, round(fitD$data$tempClean, 1)))

par(mar=c(3,3,0,0), bty='n')
matplot(
  Stemp,
  crossprod(
    fitD$tmb_data$XTp[, SobsTemp],
    parameters[1:Nbeta]
    ) + 
  crossprod(
    fitD$tmb_data$ATp[rowsToKeep, SobsTemp], 
    gammaSim[rowsToKeep,]),
  type='l', col='#00000020', lty=1, xaxs='i',
  yaxs='i'
)

```



```{r theResults}
result$parameters = hpolcc::formatParameters(
  result$solution, fitD, TRUE)

result$parameters$theta

# parameters are (beta, gamma, theta)

randomEffects = seq(from=length(result$parameters$beta)+1,
  length = nrow(result$parameters$gamma))

result$Htilde = result$hessian[randomEffects, randomEffects]
result$HtildeChol = Matrix::solve(Matrix::chol(result$Htilde))

Nsim = 200
simInd = matrix(rnorm(Nsim * nrow(result$Htilde)),
 nrow(result$Htilde), Nsim)

# sample from [U|Y, theta_map]
simGamma = Matrix::crossprod(result$HtildeChol,  simInd) 
rownames(simGamma) = rownames(fitD$tmb_data$ATp)

simGamma = simGamma + unlist(result$parameters$gamma[
  rownames(simGamma), 'value'])


Stemp = seq(0,45,by=0.1)
SobsTemp = na.omit(match(Stemp, round(fitD$data$tempClean, 1)))
Spm = seq(1,1000, by=1)
SobsPm = na.omit(match(Spm, round(fitD$data$pm)))

Sobs = SobsTemp
nameX = 'tempClean'

Bmat = fitD$tmb_data$ATp[, Sobs]
# only want the global effects
Bmat = Bmat[grep(paste0(nameX,".*global"), rownames(Bmat)),]


simF = crossprod(Bmat, simGamma[rownames(Bmat),]) 

covHere = grep(nameX, names(result$parameters$beta), value=TRUE)
fixedPart = drop(crossprod(fitD$tmb_data$XTp[covHere,Sobs, drop=FALSE], 
                           result$parameters$beta[covHere]))

simSum = simF + fixedPart 



theX= fitD$data[[covHere[1] ]][Sobs]
if(grepl("^sqrt", nameX)) theX = theX^2

par(mar=c(3,3,0,0), bty='l')
matplot(theX, simSum, type='l',
  col='#00000010', lty=1, xaxs='i')


```



```{r otherstuff22, eval=FALSE}

Sstart = 1:(nrow(fitD$tmb_data$XTp)+1)
Send = seq(to = length(fitD$start_parameters), by=1, 
  len=length(fitD$theta_info$name)+1)

bob = function(xx) {
  hpolcc:::objectiveFunctionC( 
c(xx[Sstart], fitD$start_parameters[-c(Sstart, Send)], 
  xx[-Sstart]),
 fitD$tmb_data, 
 c(
 fitD$config[setdiff(names(fitD$config), c('verbose','maxDeriv'))], 
  list(maxDeriv = 0, verbose=FALSE))
  )$value
}

round(numDeriv::hessian(bob, fitD$start_parameters[c(Sstart, Send)]), 3)

res$hessian[c(Sstart, Send), c(Sstart, Send)]



beta = fitD$start_parameters[1:nrow(fitD$tmb_data$XTp)]
gamma =fitD$start_parameters[seq(from=length(beta)+1, 
  len=nrow(fitD$tmb_data$ATp)
  )]
theta = fitD$start_parameters[seq(from = length(beta) + length(gamma)+1,
  to=length(fitD$start_parameters))]


resFun = function(xx)  hpolcc:::objectiveFunctionC(xx, fitD$tmb_data, 
  c(fitD$config[
    setdiff(names(fitD$config), 'maxDeriv')], 
  list(maxDeriv = 0)))$value


resFun(fitD$start_parameters)  

Matrix::determinant(res$hessian)



fitD$config$verbose=FALSE
res$grad
theGradAgain = numDeriv::grad(resFun, fitD$start_parameters)  

rbind(res$grad, theGradAgain)



theHessAgain = numDeriv::hessian(resFun, fitD$start_parameters)


```

```{r originalPackage, eval=FALSE}
# using the hpoltest original package (and TMB)

fitM <- hnlm(
  formula= y ~ f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 22, p=3, knots = seq(0, 42, by=2), group_var = ecoRegion), 
  data=dataMedSimple,
  dirichlet=FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  tmb_parameters = list(theta = thetaStartM),
  optim_parameters = list(maxit = 20),
  verbose=TRUE, optimizer='optim', intern=FALSE, inner.method = 'BFGS', 
  inner.control = list(maxit = 500), for_dev=FALSE)
```

```{r plotFit2, fig.height=4, fig.width=4, eval=FALSE}
simTemp =  hpoltest:::condSim(fitD, 'tempClean') 

theYlim = c(0.2, 5) #exp(quantile(simTemp$y[,'GLOBAL', ], c(0.01, 0.99)))
for(D in dimnames(simTemp$y)[[2]]) {
matplot(simTemp$x, exp(simTemp$y[,D, ]), type='l', col='#00000010', lty=1, 
        xaxs='i', log='y',
        ylim =theYlim,bty='n', 
        xlab='temperature',ylab=D)
lines(simTemp$x, exp(simTemp$fixed_mean), col='yellow')
lines(simTemp$x, exp(apply(simTemp$y[,D, ], 1, mean)), col='red')
}
```

```{r plotDate, eval=FALSE}
simDate = hpoltest:::condSim(fitD, 'date') 

matplot(as.Date(simDate$x), t(apply(simDate$y, 1, quantile, c(0.025, 0.975, 0.5))), col=c('grey','grey','black'), cex=0.3, pch=16,
        xlab='date', ylab='logrr')

```


```{r debug, eval=FALSE}

xx = fitD$obj$env$last.par.best
xx = xx[grep("gamma", names(xx))]
names(xx) = colnames(fitD$obj$env$data$A)
xxDate = xx[grep("date", names(xx))]
xx = xx[grep("date", names(xx), invert=TRUE)]

```


```{r digitPreference, eval=FALSE}
checkForDigitPreference1 = aggregate(deaths[,'medical', drop=FALSE],
  deaths[,c('dom','month')], sum)
checkForDigitPreference = dcast(as.data.table(checkForDigitPreference1), dom ~ month)
image(as.numeric(checkForDigitPreference$dom),
 as.numeric(colnames(checkForDigitPreference)[-1]),
  as.matrix(checkForDigitPreference[,-1]))
```

```{r indiaDataAlex, eval=FALSE}
load("/home/patrick/research/hpoltest/applications/india/combinedindianmorttempdata_meannofilter.Rdata")
x = indianmorttempdata

table(x$death_cause, x$case)
quantile(x$final_temp)
tempSdByStrata = tapply(x$final_temp, x$riskset, sd)
countByStrata = tapply(x$case, x$riskset, sum)
# nuber cases per day
table(tapply(x$case, x$date, sum))


tempRef = 32

x$tempHigh = pmax(x$final_temp - tempRef, 0)
x$tempLow = pmin(x$final_temp - tempRef, 0)

x$tempC = x$final_temp - tempRef
x$tempSq = (x$final_temp - tempRef)^2
x$temp3 = (x$final_temp - tempRef)^3

x$dateInt = as.integer(x$date)

xSub = x[x$death_cause == 'stroke', ]
Slow = unique(unlist(xSub[xSub$final_temp < 10, 'riskset']))

Shigh = unique(unlist(xSub[xSub$final_temp >37, 'riskset']))



xSub2 = xSub[! xSub$riskset %in% c(Slow, Shigh), ]

library('survival')
fitClogit = clogit(case ~ tempHigh + tempLow + strata(riskset),
       data = x[x$death_cause == 'stroke',])
summary(fitClogit)$coef


#library('hpoltest')
fitLinear <- hnlm(
  formula = case ~  tempHigh + tempLow + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = 20, map = list(theta=factor(NA))),
  data = xSub2,
  verbose=FALSE
  )

fitLinear$est$param$est

randomDate = as.Date(
  as.integer(gsub("factor.dateInt.", "", names(fitLinear$est$random$est))),
  as.integer(as.Date('1970/1/1')))
  
badDates = randomDate[order(abs(fitLinear$est$random$est), decreasing=TRUE)[1:2]]
stuff = as.data.frame(xSub[xSub$date %in% badDates, c('riskset', 'date','final_temp','case')])
stuff = stuff[order(stuff$date,stuff$riskset), ]
stuff


plot(randomDate, fitLinear$est$random$est)

fitLinear2 <- hnlm(
  formula = case ~  tempHigh + tempLow + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = xSub2,
  verbose=FALSE
  )

fitLinear$est$param$est
fitLinear2$est$param$est


fitQuad  <- hnlm(
  formula = case ~  tempC + tempSq + temp3 + f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = -5, map = list(theta=factor(NA))),
  data = xSub2,
  verbose=FALSE
  )
fitLinear$est$param$est
fitLinear2$est$param$est
fitQuad$est$param$est
```

```{r comparePlots, eval=FALSE}
tempSeq = seq(10, 40)
tempSeqC = tempSeq - tempRef

plot(tempSeq,
     tempSeqC*fitQuad$est$param$est['tempC'] + 
     tempSeqC^2*fitQuad$est$param$est['tempSq'] +
     tempSeqC^3*fitQuad$est$param$est['temp3'],
     type='l'
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitClogit$coef['tempLow'] + 
      pmax(tempSeqC, 0) * fitClogit$coef['tempHigh'],
      col='green', lwd=3, lty=1
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitLinear$est$param$est['tempLow'] + 
      pmax(tempSeqC, 0) * fitLinear$est$param$est['tempHigh'],
      col='red', lwd=3, lty=2
)
lines(tempSeq,
      pmin(tempSeqC, 0) * fitLinear2$est$param$est['tempLow'] + 
      pmax(tempSeqC, 0) * fitLinear2$est$param$est['tempHigh'],
      col='orange', lwd=3, lty=2
)

legend('bottomleft', bty='n', lty=1, col=c('black','red','orange','green'),
       legend = c('cubic', 'linear no OD','linear','clogit'))

```

```{r fitSimple, eval=FALSE}
tempKnots = seq(0,40, by=0.5)
fitSimple <- hnlm(
  formula = case ~  f(final_temp, model='iwp', ref_value = tempRef, p=4, knots = tempKnots) + 
    f(dateInt, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  tmb_parameters = list(theta = c(-1,-5), map = list(theta=factor(c(NA,1)))),
  data = xSub2,
  verbose=FALSE
  )

fitSimple$est$param$est 
fitQuad$est$param$est
```






```{r fitTemperatureAlex, include=FALSE, eval=FALSE}

tempKnots = seq(-5,40,by=2.5)
library('hpoltest')
fitAsAlex <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
fitIwp3NoOd <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=3, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )

fitIwp3 <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=3, knots = tempKnots) +
     f(date, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=FALSE),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5),
  verbose=FALSE
  )
fitIwp2 <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots) +
     f(date, model='iid'),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
fitAsAlex <- hnlm(
  case ~  f(final_temp, model='iwp', ref_value = 30, p=2, knots = tempKnots),
  cc_design = ccDesign(strat_vars = c('riskset')),
  data = x[x$death_cause == 'stroke',],
  verbose=FALSE
  )
```




```{r moreSimlpeModels, eval=FALSE}
  
dataMedSimplePm = removeUnusedStrata(
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01' &
    !deaths$ecoRegion %in% c(7, 8,10), ],
  Sstrata, outcome)
formulaMedSimplePm <- y ~ dom + f(date, model='iid') +
    f(sqrtPm, model='iwp',ref_value = 25, p=3, knots = pmKnots)

fitPm<- hnlm(
  formula=formulaMedSimplePm, 
  data=dataMedSimplePm,
#  tmb_parameters = list(theta = c(5.25,31)),
  cc_design =  ccDesign(strat_vars = Sstrata),
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE, for_dev=FALSE 
  )

#  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 
#  f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var = ecoRegion)

formulaCvd = cvd ~ dom + f(date, model='iid') + f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var = ecoRegion)

formulaMedSenior <- medical_senior ~ dom + f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 25, p=3, knots = tempKnots, group_var=ecoRegion)
#  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 


formulaCvdSimple = cvd ~ dom + f(date, model='iid') + 
  f(sqrtPm, model='hiwp', ref_value = 7, p=2, knots = pmKnots, group_var = ecoRegion) 
#f(tempClean, model='iwp',ref_value = 25, p=3, knots = tempKnots)

formulaMedSeniorSimple = medical_senior ~ dom + f(date, model='iid') + 
  f(tempClean, model='iwp',ref_value = 25, p=3, knots = tempKnots)

    dataMed = deaths[which(deaths$N_Week3_medical > 0 & deaths$date > as.Date('2009/1/1')),]

        dataMedSenior = deaths[which(deaths$N_Week3_medical_senior > 0 & 
                                   deaths$date > as.Date('2006/1/1')),]
dataCvd = deaths[which(deaths$N_Week3_cvd > 0 & deaths$date > as.Date('2006/1/1')), ]

if(FALSE) {

fitMedSeniorSimple = hnlm(formulaMedSeniorSimple, dataMedSenior, cc_design, 
                            optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
                          verbose=TRUE)

fitCvdSimple<- hnlm(
  formula=formulaCvdSimple, 
  data=dataCvd,
#  tmb_parameters = list(theta = c(5.25,31)),
  cc_design = cc_design,
  optim_parameters = list(eval.max=1e4, iter.max=1e4, trace=1),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE, for_dev=FALSE 
  )
saveRDS(fitCvdSimple, file.path(storeDir, 'cvdSimple.rds'))

fitCvdSetup <- hnlm(
  formula=formulaCvd, 
  data=dataCvd,
  cc_design = cc_design,
  verbose=TRUE, for_dev=TRUE 
  )

fitCvdSetup$tmb_parameters$beta = fitCvdSimple$est$param$est[1:length(fitCvdSetup$tmb_parameters$beta)]
thetaHat = fitCvdSimple$est$param$est[grep("theta", names(fitCvdSimple$est$param$est))]

fitCvdSetup$tmb_parameters$theta[1:length(thetaHat)] = thetaHat
fitCvdSetup$tmb_parameters$theta[-(1:length(thetaHat))] = 0
fitCvdSetup$tmb_parameters$gamma[1:length(fitCvdSimple$est$random$est)] = fitCvdSimple$est$random$est

  obj <- MakeADFun(data = fitCvdSetup$tmb_data,
                   parameters = fitCvdSetup$tmb_parameters,
#                   random = 'gamma',
                   map = list(
                    beta = factor(rep(NA, length(fitCvdSetup$tmb_parameters$beta))),
                    theta=factor(rep(NA, length(fitCvdSetup$map$theta)))),
                   intern=FALSE, type='ADFun',
                   DLL = "hpoltest")

  parCat = rep(NA, ncol(obj$env$.data$A))
  parCat[grep("date", colnames(obj$env$.data$A))]='date'
  parCat[grep("hrpoly", colnames(obj$env$.data$A))]='poly'
  parCat[grep("hiwp", colnames(obj$env$.data$A))]='hiwp'
  parCat[grep("global", colnames(obj$env$.data$A))]='global'

  the19 = grep('hiwp_19', colnames(obj$env$.data$A), value=TRUE)

  obj$fn(obj$par)
    opt <- optim(
      par = rnorm(length(obj$par), obj$par, 1e-2),
      fn = obj$fn, 
      gr = obj$gr,
      method = 'BFGS',
      control = list(
        trace=TRUE, 
        maxit=1e5,
        parscale = c(date=1,global=1e-5, hiwp=1e-2, poly=1e-2)[parCat]
    )
  )


            names(opt$par) = colnames(obj$env$.data$A)

                opt$par1 = opt$par[grep("date", names(opt$par))]
    opt$par2 = opt$par[!names(opt$par)%in% names(opt$par1)]

    quantile(obj$gr(opt$par))

        
fitMedSenior <- hnlm( 
  formula=formulaMedSenior, data=dataMedSenior, #[which(deaths$date > as.Date('2009/1/1')), ], 
  cc_design = cc_design,verbose=TRUE, for_dev=FALSE,
    optim_parameters = list(eval.max=1e4, iter.max=1e4),
  inner.control = list(maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE)
  )


  obj <- MakeADFun(data = fitMed1$tmb_data,
                   parameters = fitMed1$tmb_parameters[c('beta','gamma','theta')],
                   random = 'gamma',
                   map = fitMed1$map,
                   intern=FALSE, type='ADFun',
                   DLL = "hpoltest")
  obj$fn(obj$par)
    opt <- optim(par = obj$par, 
      fn = obj$fn, gr = obj$gr,
      control = list(trace=TRUE)
  )
saveRDS(fitMed, file.path(storeDir, 'ccMed.rds'))
fitCvd <- hnlm(
  formula=formulaCvd, 
  data=dataCvd,
  cc_design = cc_design,
#  optim_parameters = list(eval.max=100, iter.max=5),
  verbose=TRUE, for_dev=FALSE 
  )
saveRDS(fitCvd, file.path(storeDir, 'ccCvd.rds'))
}
fitCvd = readRDS(file.path(storeDir, 'ccCvd.rds'))
```

```{r temperatureDataFromIndia, eval=FALSE}
deathsTemp = as.data.table(readRDS( file.path(storeDir, "indiaTemperature.rds")))
deathsTemp$dom = as.numeric(format(deathsTemp$date, '%d'))
deathsTemp[deathsTemp$dom == Hmisc::monthDays(deathsTemp$date), 'dom'] = 31
deathsTemp[format(deathsTemp$date, '%m%d') == '0101', 'dom'] = 0
deathsTemp$dom = relevel(factor(deathsTemp$dom), '1')
deathsTemp$month = format(deathsTemp$date, '%m')


(theEtable = sort(table(deathsTemp$ecoRegion), decreasing=TRUE))
# 10, 17, 20 looks dodgy



Sstrata = c("cell","ecoRegion", "yearWeek3Dow")
knotsTemperature = seq(0,50, by=2.5)  

outcome = 'medical'

deathsTempMed = removeUnusedStrata(deathsTemp[
  deathsTemp$date >= as.Date('2006/1/1') &
  format(deathsTemp$date, '%b%d') != 'Jan01', ],
  Sstrata, outcome)


deathsTempMed$tempCut = 2*round(deathsTempMed$maxtemp/2)
ttable1 = dcast(deathsTempMed, tempCut ~ ecoRegion, value.var=outcome, fun.aggregate=sum)
ttable2 = dcast(deathsTempMed, tempCut ~ ecoRegion, value.var=outcome, fun.aggregate=length)

matplot(ttable1[,1], as.matrix(ttable1[,-1]/ttable2[,-1])[,names(theEtable)[10+1:9]])
merge(
  deathsTempMed, 
    as.data.frame(deathsTempMed)[order(deathsTempMed$y, decreasing=TRUE)[1:3], Sstrata],
    all.x=FALSE, all.y=TRUE)

deathsTempMed$tempLow = pmin(25, deathsTempMed$maxtempLag01)
deathsTempMed$tempHigh = pmax(25, deathsTempMed$maxtempLag01)

SexcludeRegions  = c(7, 10,13)
SincludeRegions = c(1:6, 8,9)
library(hpoltest)
fit <- hnlm(
  formula= y ~  dom + 
    f(date, model='iid') +
    f(maxtempLag01, model='hiwp',ref_value = 25, p=3, knots = knotsTemperature, group_var=ecoRegion),
  data=deathsTempMed[deathsTempMed$ecoRegion %in% SincludeRegions,],
  cc_design = ccDesign(strat_vars = Sstrata),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    smartsearch=FALSE, 
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)



formulaMedSeniorIwp = medical_senior ~ dom + f(date, model='iid') + 
  f(maxtemp, model='iwp',ref_value = 25, p=3, knots = knotsTemperature)

fitTempMedSeniorIwp <- hnlm(
  formula=formulaMedSeniorIwp, 
  data=deathsTempSeniorMed3,
  cc_design = ccDesign(strat_vars = c("cell","ecoRegion", "yearWeek3Dow")),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)

fitTempMedIwp <- hnlm(
  formula=formulaMedIwp, 
  data=deathsTempMed,
  cc_design = ccDesign(strat_vars = Sstrata),
  optimizer='optim',
  optim_parameters = list(
    method = 'BFGS',
    maxit=1e5, eval.max=1e4, iter.max=1e4),
  inner.control = list(
    maxit = 1e4, tol10=1e-4, tol=1e-9, ustep=0.5, intern=TRUE),
  verbose=TRUE)
```


```
# darjeeling
# INTEL oneMKL ERROR: /usr/lib/x86_64-linux-gnu/libmkl_core.so.2: cannot open shared object file: No such file or directory.
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_core.so.2  /usr/lib/x86_64-linux-gnu/libmkl_core.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_gnu_thread.so.2  /usr/lib/x86_64-linux-gnu/libmkl_gnu_thread.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_tbb_thread.so.2  /usr/lib/x86_64-linux-gnu/libmkl_tbb_thread.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_intel_lp64.so.2 /lib/x86_64-linux-gnu/libmkl_intel_lp64.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_intel_lp64.so.2 /lib/x86_64-linux-gnu/libmkl_intel_lp64.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_def.so.2 /lib/x86_64-linux-gnu/libmkl_def.so.2
sudo ln -s /opt/intel/oneapi/mkl/latest/lib/libmkl_avx512.so.2 /lib/x86_64-linux-gnu/libmkl_avx512.so.2
```
