```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')


dataMedSimple =
  deaths[
    deaths$date >= as.Date('2006/1/1') &
    format(deaths$date, '%b%d') != 'Jan01'
    & deaths$tempClean >= (0) # & deaths$ecoRegion %in% 1:10 #setdiff(1:20, c(1,7,19)) # regions 1, 7, 19 have some very low temperatures.
    , ]
save(dataMedSimple, file='dataMedSimple.RData')
```




```{r fit}
library('hpolcc')
load("dataMedSimple.RData")

res <- hpolcc::hnlm(
  formula= medical ~ #f(date, model='iid') +
#    f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
#      knots = seq(0,32, by=2),
#      group_var = ecoRegion) + 
    f(tempClean, model='hiwp',ref_value = 20, p=3, 
      knots = seq(0, 45, by=2.5), 
    group_var = ecoRegion), 
  data=dataMedSimple, 
 control=list(maxit=1000, start.trust.radius = 1, prec=1e-6, stop.trust.radius = 1e-9,
        cg.tol = 1e-6, report.level=4, report.freq=1, report.header.freq=10, report.precision=7),
  control_inner=list(report.level=0),
  dirichlet= TRUE, 
  config = list(transform_theta = TRUE, num_threads=parallel::detectCores(),Nclusters = 30),
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  for_dev=TRUE, verbose=TRUE
  )
saveRDS(res, file='res.rds')
#rm(dataMedSimple)
```

```{r sparsePlot}
par(mar=c(3,3,0,0), bty='n')
image(seq(0, len = nrow(res$groups$firstDeriv)),
      seq(0, len = ncol(res$groups$firstDeriv)),
      as.matrix(res$groups$firstDeriv[, 1 + res$groups$groups$i]))
abline(h = res$groups$groups$p,
       col = 'blue',
       lty = 1)
```

```{r mle}
library('hpolcc')
adFunFull = getAdFun(res$parameters_and_gamma, res$tmb_data, res$config)
adFunFull1 = getAdFun(res1$parameters_and_gamma, res1$tmb_data, res1$config)


mleX =  trustOptim::trust.optim(
    x = res$parameters,
    fn = wrappers_outer$fn,
    gr = wrappers_outer$gr,
    method = 'SR1',
    control = list(start.trust.radius = 0.05, REPORT=1, trace=10),
    data=res$tmb_data, config = res$config, cache =  cache, control_inner = res$control_inner,
    adFunFull = adFunFull
  )

mleX$check = loglik(mleX$solution, config=res$config, cache=cache, deriv=0, check=TRUE)

Sall = 1:length(res$parameters_and_gamma)
Sgamma = seq(length(res$config$beta)+1, len=length(res$gamma_start))

Dpar = 2;Dpar0 = Dpar-1
Sall = 1:length(res$parameters_and_gamma)
Sgamma = seq(length(res$config$beta)+1, len=length(res$gamma_start))
Spar = setdiff(Sall, Sgamma)
DparInAll = which(Sall == Spar[Dpar])
parMat = data.frame(res$parameters)[,rep(1, 6)]
Sx = seq(-0.5,0.5, len=ncol(parMat)) + parMat[Dpar,1]
Sx1 = Sx[-1] - diff(Sx)/2

parMatFull = data.frame(par=estHere$fullParameters)[,rep(1, 6)]
parMatFull[DparInAll,] = parMat[Dpar,] = Sx

rx = mapply(
  loglik,
  parameters = as.list(parMat),
  MoreArgs = list(
    gamma_start = get('gamma_start', cache)+ rnorm(length(Sgamma), 0, 1e-4),
    control = list(
  maxit = 5000L,
  gtol  = 1e-13,   # gradient-norm tolerance
  xtol  = 1e-13,   # parameter step tolerance
  ftol  = 1e-12,   # objective change tolerance
  stop.trust.radius = 1e-15,
  cg.tol = 1e-15,
  prec   = 1e-15,
  report.level = 8L, report.freq = 1L
),
    data=res$tmb_data, config = res$config, adFun=adFun, adFunFull = adFunFull),
  SIMPLIFY=FALSE
  )

lapply(rx, '[[', 'parameters')

the30 = hpolcc:::thirdDeriv(x=estHere$fullParameters, data=res$tmb_data, config = res$config, adFun=adFunFull, extra=TRUE)
config2 = res$config;config2$dense=TRUE
the3b = hpolcc:::thirdDeriv(x=estHere$fullParameters, data=res$tmb_data, config = config2, adFun=adFunFull, extra=TRUE)

bob = as.matrix(the30$thirdList[[1]][Sgamma,Sgamma]) - as.matrix(the3b$thirdList[[1]][Sgamma,Sgamma])

p1 = estHere$fullParameters
p1[1] = estHere$fullParameters + 0.01
h1 = hessian(parameters=p1, data=res$tmb_data, config = res$config, adFun=adFunFull)
dh = (as.matrix(h1)[Sgamma,Sgamma]- as.matrix(estHere$hessian))/0.01
dh2 = the30$thirdList[[Dpar]][Sgamma,Sgamma]
dh3 = the3b$thirdList[[Dpar]][Sgamma,Sgamma]

sum(Matrix::diag(the30$invHessianRandom %*% dh2))
sum(the30$invHessianRandom * dh2)

tracehu = mapply(function(ih, tuuu,Sgamma) {
  sum(Matrix::diag(ih %*% tuuu[Sgamma, Sgamma]))
  },
  tuuu = the30$thirdList,
  MoreArgs = list(ih = the30$invHessianRandom, Sgamma=Sgamma))
cbind(tracehu[-Sgamma],
as.vector(crossprod(the30$dUhat, tracehu[Sgamma])))

bob1 = Matrix::sparseMatrix(i=res$config$sparsity$second$nonSymmetric$i, j=res$config$sparsity$second$nonSymmetric$j, 
  x=the30$raw$diag, 
  index1=0, dims = c(125, 125))
bob2 = matrix(the3b$raw$diag, ncol=125)
quantile(as.matrix(bob1) - bob2)

bob3 = merge(the30$third, the3b$third, by =c('i','j','k'), suffixes = c('s','f'), all=TRUE)
bob3$diff = abs(bob3$xf - bob3$xs)
bob3[which(is.na(bob3$diff))[1:12],]

plot(Sx1, diff(unlist(lapply(rx, "[[", "minusLogLik")))/mean(diff(Sx)),  pch=16, xlim = range(Sx))
points(Sx, unlist(lapply(rx, function(xx) xx$deriv[Dpar,'dL'])), col='red', type='o', cex=2, pch=15)


theU = lapply(rx, "[[", "solution")
theU = do.call(cbind, theU)
(themax <- order(apply(theU, 1, sd), decreasing=TRUE)[1:10])
dU = apply(theU, 1, diff)/mean(diff(Sx))
matplot(Sx,t(theU[themax,]), type='o')


myDu = as.matrix(the30$invHessianRandom %*% the30$hessianFull[Sgamma, -Sgamma])

matplot(Sx1, dU[,themax], xlim = range(Sx), type='o')
abline(h=the30$dUhat[themax, Dpar])


theH = lapply(rx, "[[", "hessian")
theH = lapply(theH, as.matrix)
theH = do.call(abind::abind, c(theH, list(along=3)))
dH = apply(theH, 1:2, diff)/mean(diff(Sx))

theSd = apply(dH, 2:3, sd)
themax = order(diag(theSd), decreasing=TRUE)[1:8]


matplot(Sx1, t(apply(dH, 1, diag)[themax,]), col=1:length(themax))
abline(h=Matrix::diag(the30$dH[[Dpar]])[themax], col=1:length(themax))




Dpar1=14;Dpar2=16
plot(Sx1, dH[,Dpar1, Dpar2])
abline(h=the30$dH[[Dpar]][Dpar1, Dpar2])


plot(Sx1, diff(unlist(lapply(rx, "[[", "fval")))/mean(diff(Sx)), type='o')
points(Sx, unlist(lapply(rx, function(xx) xx$deriv[Dpar,'theta'])), col='red')



theD = unlist(lapply(rx, "[[", "halfLogDet"))
dD = diff(theD)/mean(diff(Sx))
plot(Sx1, dD, xlim = range(Sx))
abline(h=the30$dDet[Dpar])

dhub = mapply(
  function(tuuu, hinv, Sgamma) {
    if(is.null(tuuu)) return(0)
    if(nrow(tuuu) == 0) return(0)
    tsub = tuuu[Sgamma, Sgamma]
    res = try(sum(as.vector(hinv * as.matrix(tsub))))
    if('try-error' %in% class(res)) res = list(hinv, tuuu, Sgamma, tsub)
    res
  },
  tuuu = the3b$thirdList,
  MoreArgs = list(hinv = Matrix::as.matrix(the3b$invHessianRandom), Sgamma = Sgamma)
  )

ddet =  as.vector(unlist(dhu)[Sgamma] %*% the30$dUhat) + unlist(dhu)[-Sgamma]

as.vector(unlist(dhub)[Sgamma] %*% the3b$dUhat) + unlist(dhub)[-Sgamma]

abline(h=ddet[Dpar], col='red')


the3b = hpolcc:::thirdStrata(parameters=estHere$fullParameters, data=res$tmb_data, config = res1$config, adFun=adFunFull1)



Shere = -(1:117)
round(bob1[Shere, Shere],2)
round(bob2[Shere, Shere],2)

cbind(Matrix::diag(the30$thirdList[[122]]), Matrix::diag(the3b$thirdList[[122]]))

theH = parallel::mcmapply(hessianDense,
  parameters = parMatFull,
  MoreArgs = list(data=res$tmb_data, config = res$config), SIMPLIFY=FALSE,
  mc.cores=2
)

Nparams = length(rx[[3]]$fullParameters)
theH2 = array(unlist(theH), c(Nparams, Nparams, length(theH)))
dH = apply(theH2, 1:2, diff)/mean(diff(Sx))



config3 = res$config;config3$num_threads = 1
config2 = res$config;config2$dense=TRUE
config4 = res1$config;config4$dense = FALSE
the30 = hpolcc:::thirdDeriv(x=estHere$fullParameters, data=res$tmb_data, config = res$config, adFun=adFunFull, extra=TRUE)
the31 = hpolcc:::thirdDeriv(x=estHere$fullParameters, data=res1$tmb_data, config = res1$config, adFun=adFunFull1, extra=TRUE)



# dense, multi groups
the32 = hpolcc:::thirdDeriv(x=estHere$fullParameters, data=res$tmb_data, config = config2, adFun=adFunFull, extra=TRUE)
the33 = hpolcc:::thirdDeriv(x=estHere$fullParameters, data=res$tmb_data, config = config3, adFun=adFunFull, extra=TRUE)
the34 = hpolcc:::thirdDeriv(x=estHere$fullParameters, data=res1$tmb_data, config = config4, adFun=adFunFull1, extra=TRUE)


the30$thirdList[[DparInAll]][1:5,1:5]

Dpar1=3;Dpar2=4
plot(Sx1, dH[,Dpar1,Dpar2])
abline(h=the30$thirdList[[DparInAll]][Dpar1, Dpar2], col='red', lwd=2) # bad

abline(h=the34$thirdList[[Dpar]][Dpar1, Dpar2], col='orange', lty=3,lwd=6) # bad
abline(h=the31$thirdList[[Dpar]][Dpar1, Dpar2], lwd=5, col='grey')
abline(h=the33$thirdList[[Dpar]][Dpar1, Dpar2], col='yellow', lty=2,lwd=1)
abline(h=the32$thirdList[[Dpar]][Dpar1, Dpar2], col='blue', lty=3,lwd=2)

quantile(the30$fullHessian - the33$fullHessian)
quantile(the31$fullHessian - the33$fullHessian)

quantile((bob1 <- the31$fullHessian - theH2[,,3]))
quantile((bob2 <- the33$fullHessian - theH2[,,3]))
quantile(the31$fullHessian[Sgamma,Sgamma] - rx[[3]]$hessian)







theUhat = apply(do.call(cbind, lapply(rx, '[[', "solution")),1,diff)/mean(diff(Sx))
Dpar2 = 15
plot(Sx1, theUhat[,Dpar2])

the30$dUhat[Dpar2, Dpar]
the31$dUhat[Dpar2, Dpar]
the33$dUhat[Dpar2, Dpar]


theD = unlist(lapply(rx, function(xx) xx$halfLogDet))
plot(Sx, theD)

plot(Sx1, diff(theD)/mean(diff(Sx)), xlim = range(Sx), type='o',)
points(Sx, unlist(lapply(rx, function(xx) xx$deriv[Dpar,'det'])), col='red')


plot(Sx, unlist(lapply(rx, "[[", "minusLogLik")))






```

```{r checkDeriv}
library('hpolcc')
adFun = getAdFun(res$gamma_start, res$tmb_data, res$config)
adFunFull = getAdFun(res$parameters_and_gamma, res$tmb_data, res$config)

res$config$dense = FALSE
estHere= loglik(res$parameters, res$gamma_start, res$tmb_data, res$config, adFun, adFunFull, check=FALSE)

bob2 = thirdDeriv(
      estHere$fullParameters, data=res$tmb_data, config=res$config, adFunFull,
      extra=TRUE)

SparAll = 1:length(res$parameters_and_gamma)
Sgamma = seq(length(res$config$beta)+1, len=nrow(res$tmb_data$ATp)   )
SparOnly = setdiff(SparAll, Sgamma)
DparL = 1;DparL2 = SparOnly[DparL]
bnd = 10^(floor(log10(abs(res$parameters[DparL])))-1)
SxL = seq(-bnd,bnd, len=6) + res$parameters[DparL]
SxL1 = SxL[-1] - diff(SxL)/2
parMat2 = as.data.frame(matrix(res$parameters, ncol=length(SxL), nrow=length(res$parameters), byrow=FALSE))
parMat2[DparL,] = SxL

config2 = res$config;config2$dense=TRUE
theL = mapply(
  loglik,
  parameters = parMat2,
  MoreArgs = list(
    gamma_start =estHere$solution, data = res$tmb_data, config=config2, adFun = adFun,
    adFunFull = adFunFull, control = list(report.level=10, report.freq=5), check=FALSE),
  SIMPLIFY=FALSE
)

estHere$deriv[DparL,]

plot(SxL, unlist(lapply(theL, "[[", "fval")))

plot(SxL1, diff(unlist(lapply(theL, "[[", "fval")))/mean(diff(SxL)), type='o')
points(SxL, unlist(lapply(theL, function(xx) xx$deriv[DparL,'theta'])), col='red')
abline(h=estHere$deriv[DparL,'theta'])



theD = unlist(lapply(theL, function(xx) xx$halfLogDet))
plot(SxL, theD)

plot(SxL1, diff(theD)/mean(diff(SxL)), xlim = range(SxL), type='o')
points(SxL, unlist(lapply(theL, function(xx) xx$deriv[DparL,'det'])), col='red')

abline(h=theL[[3]]$deriv[DparL, 'det'])

xx = solve(theL[[3]]$hessian) %*% Matrix::as.matrix(bob2$dH[[DparL]]) 
sum(Matrix::diag(xx))



theU = do.call(cbind, lapply(theL, "[[", "solution"))
dU = apply(theU, 1, diff)
DuIndex=1

which.max(abs(dU[3,] - bob2$dUhat[,DparL]))

plot(SxL1, dU[,DuIndex]/mean(diff(SxL)))
abline(h=bob2$dUhat[DuIndex, DparL]);abline(v=res$parameters[DparL])





theH = lapply(theL, '[[', 'hessian')
theH = do.call(abind::abind, c(lapply(theH, Matrix::as.matrix), list(along=3)))
dH = apply(theH, 1:2, diff)/mean(diff(SxL))

round(bob2$dH[[DparL]][1:35,1:16], 1)
round(dH[3,1:35,1:16], 1)

D1 = 14;D2 = 27
plot(SxL1, dH[,D1, D2])
abline(h=bob2$dH[[DparL]][D1, D2]);abline(v=res$parameters[DparL])

xx = as.matrix(bob2$dH[[DparL]]) - dH[3,,] 
which.max(apply(abs(xx), 1, sum))
which.max(abs(xx[,14]))

cbind(round(bob2$dH[[DparL]][10:30,12:17],3),
NA,
round(dH[3,10:30,12:17],3))

theD = unlist(lapply(theL, function(xx) Matrix::determinant(xx$hessian, log=TRUE)$modulus))
plot(SxL, theD)

plot(SxL1, diff(theD)/mean(diff(SxL)), xlim = range(SxL))
abline(h=estHere$deriv[DparL,'det'], lwd=3, col='yellow')
abline(h=bob2$dDet[DparL]);abline(v=res$parameters[DparL])
points(SxL, unlist(lapply(theL, function(xx) xx$deriv[DparL, 'det'])), col='red')

xx = bob2$invHessianRandom %*% dH[3,,] 
sum(Matrix::diag(xx))
xx = bob2$invHessianRandom %*% Matrix::as.matrix(bob2$dH[[DparL]]) 
sum(Matrix::diag(xx))
estHere$deriv[DparL,'det']

theF = unlist(lapply(theL, '[[', 'fval'))
plot(SxL, theF)

plot(SxL1, diff(theF)/mean(diff(SxL)))

estHere$deriv[DparL, ]


  multGammaParam = bob2$invHessianRandom %*% bob2$fullHessian[Sgamma, -Sgamma]
as.vector(bob2$first[Sgamma] %*% multGammaParam )

```

```{r checkGrad}

mygrad = grad(estHere$solution, data=res$tmb_data, config=res$config, adFun = adFun)

mygrad2 = numDeriv::grad(
  jointLogDens, estHere$solution, data=res$tmb_data, config=res$config, adFun = adFun,
    method.args = list(
    d        = 1e-4,                         # relative step for |x_i| > zero.tol
    eps      = 1e-6,                         # absolute step for near-zero x_i
    zero.tol = sqrt(.Machine$double.eps),    # what counts as “near zero”
    r        = 6,                            # Richardson iterations (↑ accuracy, ↑ evals)
    v        = 2                             # step shrink factor
  )
)
which.max(abs(mygrad - mygrad2))


Dpar = 9
parMat = data.frame(estHere$solution)[,rep(1, 6)]
Sx = seq(-0.0001,0.0001, len=ncol(parMat)) + parMat[Dpar,1]
Sx1 = Sx[-1] - diff(Sx)/2

parMat[Dpar,] = Sx

rx = unlist(mapply(
  jointLogDens,
  x = as.list(parMat),
  MoreArgs = list(data=res$tmb_data, config = res$config, adFun=adFun),
  SIMPLIFY=FALSE
  ))
dx = do.call(cbind, mapply(
  grad,
  parameters = as.list(parMat),
  MoreArgs = list(data=res$tmb_data, config = res$config, adFun=adFun),
  SIMPLIFY=FALSE
  ))

plot(Sx, rx)

plot(Sx1, diff(rx)/mean(diff(Sx)), type='o', xlim = range(Sx))
points(Sx, dx[Dpar,], col='red')
abline(h=mygrad2[Dpar], col='yellow', lwd=3)
abline(h=mygrad[Dpar])

```

```{r dostuff}
library(hpolcc)

adFun = getAdFun(res$gamma_start, res$tmb_data, res$config)


grad2 = grad(res$start_gamma, data=res$tmb_data, config=res$config, adfun=adfun)
hes2 = hessian(res$start_gamma, data=res$tmb_data, config=res$config, adfun=adfun)

innerRes = trustOptim::trust.optim(
    x = fitD$start_gamma, 
    fn = jointLogDens,
    gr = grad,
    hs = hessian,
    method = 'Sparse',
    control = list(start.trust.radius = 0.1, report.freq=5, report.level=10),
    data=fitD$tmb_data, config = fitD$config, adfun = fitD$adfun()
  )

```

```{r morestuff}
library('hpolcc')
adFun = getAdFun(res$gamma_start, res$tmb_data, res$config)
adFunFull = getAdFun(res$parameters_and_gamma, res$tmb_data, res$config)

cache = res$cache
assign("file", "india.txt", cache)
assign("gamma_start", res$gamma_start, cache)
if(file.exists(get("file", cache))) file.remove(get("file", cache))

wrappers_outer$fn(res$parameters,    data=res$tmb_data, config = res$config, cache =  cache, control_inner = res$control_inner,
    adFun = adFun, adFunFull = adFunFull)


mleX =  trustOptim::trust.optim(
    par = c(0.009,  0.012, -3, -7, -5, -6, -1),
    fn = wrappers_outer$fn,
    fn = wrappers_outer$gr,
    method = 'Sparse',
    data=res$tmb_data, config = res$config, cache =  cache, control_inner = res$control_inner,
    adFun = adFun, adFunFull = adFunFull
  )
mleX$solution

bob = loglik(
  c(0.021726796, -2.118758e-04 ,-4.253132, -7.918471, -5.099570 ,-7.996464, -2.621749),
  gamma_start = get("gamma_start", cache),
 data=fitD$tmb_data, config = fitD$config, control = fitD$control_inner)


mleB <- BB::spg(
  par = c(0.021726796, -2.118758e-04 ,-4.253132, -7.918471, -5.099570 ,-7.996464, -2.621749),
  fn = wrappers_outer$fn,
  gr = wrappers_outer$gr,
  data=fitD$tmb_data, config = fitD$config, cache =  cache, controlInner = fitD$control_inner,
  control = list(maxit = 1e2, ftol=1e-6, gtol = 1e-4, M = 5, trace=TRUE, checkGrad=TRUE))  # M = nonmonotone history


  parameters = c(
    0.0, 0.01, # beta (slope, quadratic) 
    log(1), # sd global
    log(0.1), # sd local
    log(0.001), log(0.001), # sd local slope, quad
    log(0.1)[fitD$config$dirichlet] # dirichelet OD
    )


fitD$config$transform_theta = TRUE
  fitD$config$num_threads = parallel::detectCores()


  Nbeta = nrow(fitD$tmb_data$XTp)
  Ngamma = nrow(fitD$tmb_data$ATp)

  Ntheta = length(fitD$theta_info$name)



 gamma_start = seq(-0.01, 0.01, len=Ngamma) #mapOptim$solution[seq(Nbeta+1, len=Ngamma)]
    config = fitD$config
    config$transform_theta = TRUE


# likelihood
# parameters = (beta, theta)R
# config$transform_theta = TRUE means thetas are log(sd) otherwise theta=sd

# test when beta and theta in config


library('hpolcc')

config$verbose=FALSE
config$num_threads =  parallel::detectCores()
config$strataPerIter = 500

result <- loglik(parameters, 
             gamma_start, 
             data=fitD$tmb_data, 
    config = config,
      control=list(maxit=1000, start.trust.radius = 1, prec=1e-9, stop.trust.radius = 1e-15,
        cg.tol = 1e-9,
      report.level=4, report.freq=1)
      )
```










```{r mLE}


mle = optim(
    c(0.00858558945462489, 0.0100467550976791, -3.1175149461754 ,-7.02350405776971, -5.01580155255101, -6.04704512936413, -1.16549925013542),
    fn = wrappers_outer$fn,
#    gr = wrappers_outer$gr,
    method = 'L-BFGS-B',
    control = res$control,
lower = c(-0.2, -0.1, -8,-8,-8,-8,-8),
upper = c(0.2, 0.1, 8,8,8, 8, 8),
    data=res$tmb_data, config = res$config, cache =  res$cache, control_inner = control_inner,
    adFun=adFun, adFunFull = adFunFull
  )

mle$extra = loglik(mle$solution, get("gamma_start", cache), data, config, controlInner)
mle$gamma_hat = mle$extra$solution
saveRDS(mle, file='mle.rds')
```



```{r theResults}
result$parameters = hpolcc::formatParameters(
  result$solution, fitD, TRUE)

result$parameters$theta

# parameters are (beta, gamma, theta)

randomEffects = seq(from=length(result$parameters$beta)+1,
  length = nrow(result$parameters$gamma))

result$Htilde = result$hessian[randomEffects, randomEffects]
result$HtildeChol = Matrix::solve(Matrix::chol(result$Htilde))

Nsim = 200
simInd = matrix(rnorm(Nsim * nrow(result$Htilde)),
 nrow(result$Htilde), Nsim)

# sample from [U|Y, theta_map]
simGamma = Matrix::crossprod(result$HtildeChol,  simInd) 
rownames(simGamma) = rownames(fitD$tmb_data$ATp)

simGamma = simGamma + unlist(result$parameters$gamma[
  rownames(simGamma), 'value'])


Stemp = seq(0,45,by=0.1)
SobsTemp = na.omit(match(Stemp, round(fitD$data$tempClean, 1)))
Spm = seq(1,1000, by=1)
SobsPm = na.omit(match(Spm, round(fitD$data$pm)))

Sobs = SobsTemp
nameX = 'tempClean'

Bmat = fitD$tmb_data$ATp[, Sobs]
# only want the global effects
Bmat = Bmat[grep(paste0(nameX,".*global"), rownames(Bmat)),]


simF = crossprod(Bmat, simGamma[rownames(Bmat),]) 

covHere = grep(nameX, names(result$parameters$beta), value=TRUE)
fixedPart = drop(crossprod(fitD$tmb_data$XTp[covHere,Sobs, drop=FALSE], 
                           result$parameters$beta[covHere]))

simSum = simF + fixedPart 



theX= fitD$data[[covHere[1] ]][Sobs]
if(grepl("^sqrt", nameX)) theX = theX^2

par(mar=c(3,3,0,0), bty='l')
matplot(theX, simSum, type='l',
  col='#00000010', lty=1, xaxs='i')


```



```{r otherstuff22, eval=FALSE}

Sstart = 1:(nrow(fitD$tmb_data$XTp)+1)
Send = seq(to = length(fitD$start_parameters), by=1, 
  len=length(fitD$theta_info$name)+1)

bob = function(xx) {
  hpolcc:::objectiveFunctionC( 
c(xx[Sstart], fitD$start_parameters[-c(Sstart, Send)], 
  xx[-Sstart]),
 fitD$tmb_data, 
 c(
 fitD$config[setdiff(names(fitD$config), c('verbose','maxDeriv'))], 
  list(maxDeriv = 0, verbose=FALSE))
  )$value
}

round(numDeriv::hessian(bob, fitD$start_parameters[c(Sstart, Send)]), 3)

res$hessian[c(Sstart, Send), c(Sstart, Send)]



beta = fitD$start_parameters[1:nrow(fitD$tmb_data$XTp)]
gamma =fitD$start_parameters[seq(from=length(beta)+1, 
  len=nrow(fitD$tmb_data$ATp)
  )]
theta = fitD$start_parameters[seq(from = length(beta) + length(gamma)+1,
  to=length(fitD$start_parameters))]


resFun = function(xx)  hpolcc:::objectiveFunctionC(xx, fitD$tmb_data, 
  c(fitD$config[
    setdiff(names(fitD$config), 'maxDeriv')], 
  list(maxDeriv = 0)))$value


resFun(fitD$start_parameters)  

Matrix::determinant(res$hessian)



fitD$config$verbose=FALSE
res$grad
theGradAgain = numDeriv::grad(resFun, fitD$start_parameters)  

rbind(res$grad, theGradAgain)



theHessAgain = numDeriv::hessian(resFun, fitD$start_parameters)


```

```{r originalPackage, eval=FALSE}
# using the hpoltest original package (and TMB)

fitM <- hnlm(
  formula= y ~ f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 22, p=3, knots = seq(0, 42, by=2), group_var = ecoRegion), 
  data=dataMedSimple,
  dirichlet=FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  tmb_parameters = list(theta = thetaStartM),
  optim_parameters = list(maxit = 20),
  verbose=TRUE, optimizer='optim', intern=FALSE, inner.method = 'BFGS', 
  inner.control = list(maxit = 500), for_dev=FALSE)
```

```{r plotFit2, fig.height=4, fig.width=4, eval=FALSE}
simTemp =  hpoltest:::condSim(fitD, 'tempClean') 

theYlim = c(0.2, 5) #exp(quantile(simTemp$y[,'GLOBAL', ], c(0.01, 0.99)))
for(D in dimnames(simTemp$y)[[2]]) {
matplot(simTemp$x, exp(simTemp$y[,D, ]), type='l', col='#00000010', lty=1, 
        xaxs='i', log='y',
        ylim =theYlim,bty='n', 
        xlab='temperature',ylab=D)
lines(simTemp$x, exp(simTemp$fixed_mean), col='yellow')
lines(simTemp$x, exp(apply(simTemp$y[,D, ], 1, mean)), col='red')
}
```
