```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')


dataMedSimple =
  deaths[
    format(deaths$date, '%b%d') != 'Jan01'
    & deaths$tempClean >= (0) # & deaths$ecoRegion %in% 1:10 #setdiff(1:20, c(1,7,19)) # regions 1, 7, 19 have some very low temperatures.
    , ]
save(dataMedSimple, file='dataMedSimple.RData')
```




```{r fit}
library('hpolcc')
load("dataMedSimple.RData")

# add day iid
res <- hpolcc::hnlm(
  formula= cvd ~ #f(date, model='iid') +
    f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
      knots = seq(0,32, by=2),
      group_var = ecoRegion) + 
    f(tempClean, model='hiwp',ref_value = 20, p=3, 
      knots = seq(0, 45, by=2.5), 
    group_var = ecoRegion), 
  data=dataMedSimple, 
  dirichlet= TRUE, 
  config = list(transform_theta = TRUE, 
                num_threads=parallel::detectCores(),
                Nclusters = 4*parallel::detectCores()),
  control_inner=list(report.level=0),
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  for_dev=TRUE, verbose=TRUE
  )
saveRDS(res, file='resCvd.rds')
#rm(dataMedSimple)
# res = readRDS('res.rds')
```


```{r sparsePlot}
par(mar=c(3,3,0,0), bty='n')
bob= terra::rast(as.matrix(1.1*res$groups$firstDeriv[, 
  1 + res$groups$groups$i]))
terra::plot(bob, asp=FALSE, col=c('white','black'))
abline(v = res$groups$groups$p,
       col = 'blue',
       lty = 1)
```

```{r checkDeriv}
res = readRDS('resCvd.rds')
library('hpolcc')
adFunFull = getAdFun(res$parameters_and_gamma, res$tmb_data, res$config)

assign("file", "indiaCvd.txt", res$cache)
if(file.exists(get("file", res$cache))) file.remove(get("file", res$cache))

forRep = c(length(res$config$beta), length(res$config$theta)-1, 1)
mleX = optim(
    c(0.002, -0.004, 0.0008,
      rep(-4, forRep[2]), 
      -1.2),
    fn = wrappers_outer$fn,
    gr = wrappers_outer$gr,
    method = 'L-BFGS-B', 
    lower = rep(c(-0.5, -15,-5), forRep),
    upper = rep(c(0.5, 0, 0), forRep),
    control = list(trace=3, REPORT=1, 
      parscale = rep(c(1e-3, 10, 1), forRep)
      ),#start.trust.radius = 0.05, REPORT=1, trace=10),
    data=res$tmb_data, config = res$config, cache =  res$cache, 
    control_inner = res$control_inner,
    adFunFull = adFunFull
  )

mleX$extra = loglik(
  mleX$par,
  gamma_start = get("gamma_start", res$cache),
 data=res$tmb_data, config = res$config, control = res$control_inner,
 deriv = 0)
saveRDS(mleX, file='mleCvd.rds')
```




```{r theResults}
library('hpolcc')
theData = readRDS("resCvd.rds")
result = readRDS("mleCvd.rds")

result$parameters = hpolcc::formatParameters(
  result$extra$fullParameters, 
  theData, TRUE)
result$parameters$theta

HtildeChol = Matrix::Cholesky(result$extra$hessian, LDL=TRUE)
HtildeCholEx = list(Lt = Matrix::expand1(HtildeChol, "L1."),
                    D = Matrix::expand1(HtildeChol, "D"),
                    Pt = Matrix::expand1(HtildeChol, "P1.")
)

quantile(result$extra$hessian - HtildeCholEx$Pt %*% Matrix::t(HtildeCholEx$Lt) %*% HtildeCholEx$D %*% HtildeCholEx$L %*% Matrix::t(HtildeCholEx$P))


# H = Pt L D Lt P 
# V = Pt Ltinv Dinv Linv P

Nsim = 500
predSeq = list(
  tempClean = seq(0,45,by=0.2),
  sqrtPm = sqrt(seq(0,1000, by=2))
)
newConstr = list(tempClean = 25, sqrtPm = sqrt(25))


simInd = matrix(
  rnorm(Nsim * nrow(result$extra$hessian)),
  nrow(result$extra$hessian), Nsim)
simInd1 = simInd/sqrt(HtildeCholEx$D@x)
simGamma = HtildeCholEx$Pt %*% Matrix::solve(HtildeCholEx$Lt, simInd1) +
  result$extra$solution
rownames(simGamma) = rownames(theData$tmb_data$ATp)

if(FALSE) {
Hinv = Matrix::solve(result$extra$hessian)
varHere =var(as.matrix(Matrix::t(simGamma)))
quantile(Hinv - varHere)

plot(
  sqrt(diag(varHere)),
  sqrt(Matrix::diag(Hinv)));abline(0,1)
}

Sref = unlist(lapply(theData$terms, '[[', "ref_value"))
Svar = unlist(lapply(theData$terms, '[[', "var"))

simGlobal = simF = fixedPart = predX = list()
for(D in names(predSeq)) {
  refHere = Sref[match(D, Svar)]
  Sobs = na.omit(match(predSeq[[D]], round(refHere + theData$tmb_data$XTp[D,], 1)))
  predX[[D]] = refHere + theData$tmb_data$XTp[D,Sobs]
  
  colsA = grep(paste0(D, "_hiwp_global_"),rownames(theData$tmb_data$ATp), value=TRUE)
  simF[[D]] = Matrix::crossprod(theData$tmb_data$ATp[colsA,Sobs], simGamma[colsA,])
  
  betaHere = grep(paste0("^", D, "([[:digit:]]+)?$"), names(result$parameters$beta), value=TRUE)
  fixedPart[[D]] =  Matrix::crossprod(
    theData$tmb_data$XTp[betaHere, Sobs,drop=FALSE], 
    result$parameters$beta[betaHere])
  simGlobalHere= simF[[D]] + fixedPart[[D]]
  newConstrIndex = which.min(abs(predX[[D]] - newConstr[[D]]))

  toSubtract = matrix(simGlobalHere[newConstrIndex,],
    nrow(simGlobalHere), ncol(simGlobalHere), byrow=TRUE)
  simGlobal[[D]] = simGlobalHere - toSubtract
}

#  matplot(simF[[D]], type='l')
    
if(FALSE) {
   newDf = data.frame(x = predSeq[[D]], ecoRegion = NA)
  colnames(newDf)[1] = D
  newXA[[D]] = hpolcc:::getNewXA(
  terms = theData$terms, 
  df= newDf
  )
  xColId = as.numeric(gsub( paste0("^", D, "([[:digit:]]+)$"), "\\1", colnames(newXA[[D]]$X)))
  newColNames = paste0(D, c("", seq(1, len=length(xColId)-1)[xColId])[xColId])
  fixedPart[[D]] = newXA[[D]]$X %*% result$parameters$beta[newColNames]
}
  


```


```{r globalRes, fig.cap='Common component', fig.subcap = c("temperature", "PM 2.5)}

par(mar=c(3,3,0,0), bty='l')
matplot(predX[["tempClean"]], 
        100*(exp(simGlobal[["tempClean"]])-1), type='l',
  col='#00000010', lty=1, xaxs='i',
  ylim = c(-1/3, 31), xlim = c(6, 38))

theY =  100*(exp(simGlobal[["sqrtPm"]])-1)
theX = predX[["sqrtPm"]]^2
matplot(theX, theY , type='l',
  col='#00000010', lty=1, xaxs='i',
  xlim = c(0,600), ylim = c(-4, 12),
  xlab = 'PM 2.5', ylab = '100 (RR-1)')
theYq = t(apply(theY, 1, quantile, probs=c(0.025, 0.975, 0.5)))
matlines(theX, theYq, lty=c(2,2,1), lwd=c(1,1,2), col='blue')

```


```{r otherstuff22, eval=FALSE}

Sstart = 1:(nrow(fitD$tmb_data$XTp)+1)
Send = seq(to = length(fitD$start_parameters), by=1, 
  len=length(fitD$theta_info$name)+1)

bob = function(xx) {
  hpolcc:::objectiveFunctionC( 
c(xx[Sstart], fitD$start_parameters[-c(Sstart, Send)], 
  xx[-Sstart]),
 fitD$tmb_data, 
 c(
 fitD$config[setdiff(names(fitD$config), c('verbose','maxDeriv'))], 
  list(maxDeriv = 0, verbose=FALSE))
  )$value
}

round(numDeriv::hessian(bob, fitD$start_parameters[c(Sstart, Send)]), 3)

res$hessian[c(Sstart, Send), c(Sstart, Send)]



beta = fitD$start_parameters[1:nrow(fitD$tmb_data$XTp)]
gamma =fitD$start_parameters[seq(from=length(beta)+1, 
  len=nrow(fitD$tmb_data$ATp)
  )]
theta = fitD$start_parameters[seq(from = length(beta) + length(gamma)+1,
  to=length(fitD$start_parameters))]


resFun = function(xx)  hpolcc:::objectiveFunctionC(xx, fitD$tmb_data, 
  c(fitD$config[
    setdiff(names(fitD$config), 'maxDeriv')], 
  list(maxDeriv = 0)))$value


resFun(fitD$start_parameters)  

Matrix::determinant(res$hessian)



fitD$config$verbose=FALSE
res$grad
theGradAgain = numDeriv::grad(resFun, fitD$start_parameters)  

rbind(res$grad, theGradAgain)



theHessAgain = numDeriv::hessian(resFun, fitD$start_parameters)


```

```{r originalPackage, eval=FALSE}
# using the hpoltest original package (and TMB)

fitM <- hnlm(
  formula= y ~ f(date, model='iid') +
    f(tempClean, model='hiwp',ref_value = 22, p=3, knots = seq(0, 42, by=2), group_var = ecoRegion), 
  data=dataMedSimple,
  dirichlet=FALSE,
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
  tmb_parameters = list(theta = thetaStartM),
  optim_parameters = list(maxit = 20),
  verbose=TRUE, optimizer='optim', intern=FALSE, inner.method = 'BFGS', 
  inner.control = list(maxit = 500), for_dev=FALSE)
```

```{r plotFit2, fig.height=4, fig.width=4, eval=FALSE}
simTemp =  hpoltest:::condSim(fitD, 'tempClean') 

theYlim = c(0.2, 5) #exp(quantile(simTemp$y[,'GLOBAL', ], c(0.01, 0.99)))
for(D in dimnames(simTemp$y)[[2]]) {
matplot(simTemp$x, exp(simTemp$y[,D, ]), type='l', col='#00000010', lty=1, 
        xaxs='i', log='y',
        ylim =theYlim,bty='n', 
        xlab='temperature',ylab=D)
lines(simTemp$x, exp(simTemp$fixed_mean), col='yellow')
lines(simTemp$x, exp(apply(simTemp$y[,D, ], 1, mean)), col='red')
}
```
