```{r setup, include=FALSE}
#devtools::install_github("samperochkin/hpoltest")
```



on darjeeling
```{r loadpackage, eval=FALSE}

library('hpolcc', lib.loc = 
  "/home/patrick/R/x86_64-pc-linux-gnu-library/4.5")

```

for new package
Need cppad, openmp

brew install cppad
brew install openmp (?)

might need 

~/.R/Makevars with

# Use Homebrew LLVM clang++ for OpenMP support
CXX = /opt/homebrew/opt/llvm/bin/clang++
CC  = /opt/homebrew/opt/llvm/bin/clang

CXXFLAGS += -std=c++17 -I/opt/homebrew/include -fopenmp
LDFLAGS  += -L/opt/homebrew/lib -lcppad_lib -fopenmp

# Tell clang where to find OpenMP
CPPFLAGS += -I/opt/homebrew/opt/libomp/include
LDFLAGS  += -L/opt/homebrew/opt/libomp/lib -lomp



```{r setup2}


library('data.table')
# /home/patrick/Admin/globalHealth/indiaSpatial/daily.Rmd
SstoreDir = c('/store/patrick/copernicus/agg', '~/Downloads/agg')
storeDir = SstoreDir[1]
if(!dir.exists(storeDir)) storeDir = SstoreDir[2]

deaths = as.data.table(readRDS( file.path(storeDir, "indiaCC.rds")))

indiaEcoRegions = terra::unwrap(readRDS(
  file.path(storeDir, "indiaEcoRegions.rds")))


# trransform variables
deaths$sqrtPm = sqrt(pmin(deaths$pm,1000))
deaths$timeIid = as.numeric(deaths$date)

# table(deaths$ecoRegion, deaths$temp < -10)

deaths$tempClean = pmax(deaths$temp, -5)


deaths$dom = as.numeric(format(deaths$date, '%d'))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), 'dom'] = 31
deaths[format(deaths$date, '%m%d') == '0101', 'dom'] = 0
deaths$dom = relevel(factor(deaths$dom), '1')
deaths$month = format(deaths$date, '%m')

deaths = deaths[format(deaths$date, '%b%d') != 'Jan01', ]
save(deaths, file='deaths.RData')
```




```{r fit}
library('hpolcc')
load("deaths.RData")

res <- hpolcc::hnlm(
  formula= medical ~ dom + 
  f(temp, model='hiwp',ref_value = 25, p=3, 
      knots = seq(-5, 42.5, by=2.5), 
    group_var = ecoRegion)
  , 
  data=deaths[
    which(
      deaths$date > as.Date("2008/1/1") & 
      deaths$temp >= -5
      ), ],
  cc_design =  c("cell","ecoRegion", "yearMonthDow"),
  config = list(
    transform_theta = TRUE, 
    boundary_is_random= TRUE,
    dirichlet= TRUE, verbose=FALSE,
    num_threads=parallel::detectCores(), 
    Nclusters = 100*parallel::detectCores()
    ),
  for_dev=TRUE
  )
saveRDS(res, file='resTemp.rds')

resPm <- hpolcc::hnlm(
  formula= medical ~ date5 + #humidity + #f(date, model='iid') +
   f(sqrtPm, model='hiwp', ref_value = 8, p=2, 
     knots = seq(0,25, by=1),
     group_var = ecoRegion, init = 0.1) 
  #    +
#   f(tempClean, model='hiwp',ref_value = 25, p=3, 
#      knots = seq(0, 45, by=2.5), 
#    group_var = ecoRegion)
  , 
  data=dataMedSimple[dataMedSimple$date > as.Date('2010/1/1'),], 
  cc_design =  c("cell","ecoRegion", "yearWeek3Dow"),
   control=list(
    maxit= 2000,
    start.trust.radius = 0.1, report.level=4, report.freq=10, report.header.freq=10, report.precision=5),
  config = list(
    transform_theta = TRUE, 
    boundary_is_random= TRUE,
    dirichlet= TRUE, verbose=FALSE,
    num_threads=parallel::detectCores(), 
    Nclusters = 50*parallel::detectCores()
    ),
  for_dev=FALSE
  )
saveRDS(resPm, file='resPm.rds')


```



```{r theResults}
library('hpolcc')
res = readRDS("resPm.rds")
res = readRDS("resTemp.rds")
simGlobal = res$sample$global

predSeq = res$sample$x
simGlobalSumToZero = simGlobal[[1]] - 
  matrix(apply(simGlobal[[1]][predSeq[[1]]>15 & predSeq[[1]]<30 ,], 2, mean), nrow(simGlobal[[1]]), ncol(simGlobal[[1]]), byrow=TRUE)
```





```{r globalRes, fig.cap='Common component', fig.subcap = c("temperature", "PM 2.5")}
par(mar=c(3,3,0,0), bty='l', oma=c(0,0,0,0))
if("temp" %in% names(simGlobal)) {
theY = 100*(exp(simGlobal[["temp"]])-1)
matplot(predSeq[["temp"]], theY, type='l',
  col='#00000010', lty=1, xaxs='i',
  xlab='degrees C', ylab = '100 x (RR-1)',
  ylim = c(-40, 86), xlim = c(6, 41))
theYq = t(apply(theY, 1, quantile, probs=c(0.025, 0.975, 0.5)))
matlines(predSeq[["temp"]], theYq, lty=c(2,2,1), lwd=c(1,1,2), col='blue')

theYZero = 100*(exp(simGlobalSumToZero)-1)
matplot(predSeq[["tempClean"]], theYZero, type='l',
  col='#00000010', lty=1, xaxs='i',
  xlab='degrees C', ylab = '100 x (RR-1)',
  ylim = c(-40, 86), xlim = c(6, 41))
theYq = t(apply(theYZero, 1, quantile, probs=c(0.025, 0.975, 0.5)))
matlines(predSeq[["tempClean"]], theYq, lty=c(2,2,1), lwd=c(1,1,2), col='blue')


}

if("sqrtPm" %in% names(simGlobal)) {

theY =  100*(exp(simGlobal[["sqrtPm"]])-1)
theX = predSeq[["sqrtPm"]]^2
matplot(theX, theY , type='l',
  col='#00000010', lty=1, xaxs='i',
  xlim = c(0,600), ylim = c(-4, 12),
  xlab = 'PM 2.5', ylab = '100 (RR-1)')
theYq = t(apply(theY, 1, quantile, probs=c(0.025, 0.975, 0.5)))
matlines(theX, theYq, lty=c(2,2,1), lwd=c(1,1,2), col='blue')
}
```

```{r check}
res$extra$parameters$beta
xx = c(rep(0,nrow(res$objects$gamma_info)), res$extra$fullParameters[
  seq(nrow(res$objects$gamma_info), length(res$extra$fullParameters))
  ])
#xx = res$extra$fullParameters  
xx[1:3]
res$objects$gamma_info[which(res$objects$gamma_info$global)[1:12],]


Dpar = 30


matplot(res$objects$tmb_data$ATp[1,], 
  res$objects$tmb_data$ATp[Dpar,],
  pch='.', col='red'
  )

Sx = seq(-0.01, 0.01, len=201) + xx[Dpar]
parDf = data.frame(xx)[,rep(1, length(Sx))]
parDf[Dpar,] = Sx
tmb_data_no_q = res$objects$tmb_data
tmb_data_no_q$Qdiag = rep(1, 0)

Sy = mapply(jointLogDens, parameters = as.list(parDf), 
  MoreArgs = list(data=
    tmb_data_no_q,
    #res$objects$tmb_data, 
    config=res$objects$config))
plot(Sx, Sy, type='o')
abline(v= res$extra$fullParameters[Dpar])
```





```{r checkThree, eval=FALSE, include=FALSE}
res = readRDS("res.rds")
library('hpolcc')
res$config$verbose=FALSE

adFun = getAdFun(res$parameters_and_gamma, res$tmb_data, res$config)
adFunInner = getAdFun(res$gamma_start, res$tmb_data, res$config)

system.time(resM <- loglik(res$parameters, res$gamma_start, res$tmb_data, res$config, adFun, res$control, deriv=1))
system.time(resM2 <- loglik(res$parameters, res$gamma_start, res$tmb_data, 
  res$config[setdiff(names(res$config), "first")], 
  adFun, res$control, deriv=1))

bob = Matrix::expand2(resM2$extra$hessian)

Dpar=length(resM$parameters)
Sx = seq(-0.1, 0.1, len=5) + resM$parameters[Dpar]
SxD = Sx[-1] - diff(Sx)/2
parDf = data.frame(resM$parameters)[,rep(1, length(Sx))]
parDf[Dpar,] = Sx

allLik = mapply(
  loglik,
  parameters = as.list(parDf),
  MoreArgs = list(gamma_start = resM$solution, data=res$tmb_data, config=res$config, control=res$control, deriv=0),
  SIMPLIFY=FALSE
  )

Sdet = unlist(lapply(allLik, function(xx) Matrix::determinant(xx$cholHessian, sqrt=FALSE)$mod/2))
plot(Sx, Sdet)

DsDet = diff(Sdet)/diff(Sx)

plot(SxD, DsDet, type='o')
abline(h=resM$deriv[Dpar,'dDet'],col='grey',lwd=4)
abline(h=resM2$deriv[Dpar,'dDet'])
abline(v=resM$parameters[Dpar])



Sdens = unlist(lapply(allLik, '[[', 'fval'))
Ddens = diff(Sdens)/diff(Sx)
plot(SxD, Ddens,type='o')
abline(v=resM$parameters[Dpar])
abline(h=resM$deriv[Dpar,'gradTheta'],col='grey',lwd=4)
abline(h=resM2$deriv[Dpar,'gradTheta'])

Sl = unlist(lapply(allLik, '[[', 'minusLogLik'))
Dl = diff(Sl)/diff(Sx)
plot(SxD, Dl, type='o')
abline(v=resM$parameters[Dpar])
abline(h=resM$deriv[Dpar,'dL'],col='grey',lwd=4)
abline(h=resM2$deriv[Dpar,'dL'])



```
