---
title: "India"
geometry: margin=1in
output: pdf_document
header-includes:
  - |
    ```{=latex}
    \usepackage[table]{xcolor}
    \expandafter\def\csname ver@subfig.sty\endcsname{}
    \usepackage{graphicx,subcaption}
    \makeatletter\@ifundefined{subfloat}{\newcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}{\renewcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2} }}\makeatother
    ```
---

Time-stratified case-crossover for Indian grid-cell mortality with
population-weighted PM$_{2.5}$ and temperature. Strata are cell $\times$
eco-region $\times$ year–week3–day-of-week. The conditional likelihood is a
Dirichlet-multinomial with estimated overdispersion. Exposure–response uses
hierarchical IWP (`hiwp`) smooths of order 2 on $\sqrt{\mathrm{PM}_{2.5}}$ and
order 3 on temperature/10, each grouped by eco-region.

Fitting never runs during knit: the `fit` chunk is `eval=FALSE`. Rendering
loads a successful fit cache when present (`analysis_cache/res_hiwp_pm_temp.rds`).

```{r setup, include=FALSE}
knitr::knit_hooks$set(margins = Pmisc::hook_plot_margins)
knitr::knit_hooks$set(plot = knitr::hook_plot_tex)

knitr::opts_chunk$set(
  echo = FALSE, prompt = TRUE, comment = NA,
  dev = "pdf", margins = 1, fig.cap = " ", fig.pos = "hp",
  fig.path = file.path("figure", "moreBayes", ""),
  fig.width = 5, fig.height = 5,
  tidy = TRUE, tidy.opts = list(indent = 2, width.cutoff = 45)
)

options(width = 80)
dir.create("analysis_cache", showWarnings = FALSE)
fit_rds <- "analysis_cache/res_hiwp_pm_temp.rds"
deaths_rds <- "deaths.RData"
```

```{r dirs, include=FALSE}
s_store_dir <- c("/store/patrick/copernicus/agg", "~/Downloads/agg")
store_dir <- s_store_dir[file.exists(path.expand(s_store_dir))][1]
has_store <- !is.na(store_dir) && nzchar(store_dir)
has_deaths <- file.exists(deaths_rds)
has_fit <- file.exists(fit_rds)
```

```{r ecoRegionData, eval=isTRUE(has_store)}
india_eco_regions <- terra::unwrap(
  readRDS(file.path(store_dir, "indiaEcoRegions.rds"))
)
india_eco_regions$name <- gsub(
  " AND ", ", ",
  gsub(
    " E.O.REGION|[(].*[)]|TO PERHUMID.*|PENINSULA ", "",
    india_eco_regions$physio_reg
  )
)
india_eco_regions$id_name <- paste(
  india_eco_regions$objectid, india_eco_regions$name
)
india_eco_regions$col <- rep_len(
  c(
    RColorBrewer::brewer.pal(12, "Set3"),
    RColorBrewer::brewer.pal(8, "Set2")
  ),
  length(india_eco_regions)
)
```

```{r worldPopByRegion, cache=TRUE, eval=isTRUE(has_store)}
year <- 2020
dir.create("data/gpw", recursive = TRUE, showWarnings = FALSE)

pop_density <- geodata::population(year, res = 5, path = "data/gpw")
eco <- terra::project(india_eco_regions, terra::crs(pop_density))
pop_density_india_eco <- terra::mask(terra::crop(pop_density, eco), eco)
cell_area_km2 <- terra::cellSize(pop_density_india_eco, unit = "km")
pop_count <- pop_density_india_eco * cell_area_km2
eco_pop <- terra::extract(
  pop_count, eco,
  fun = sum, na.rm = TRUE, exact = TRUE
)
eco$gpw_2020 <- eco_pop[, 2]
eco_pop_df <- data.frame(
  objectid = eco$objectid,
  physio_reg = eco$physio_reg,
  gpw_2020 = eco$gpw_2020
)
eco_pop_df$weight <- eco_pop_df$gpw_2020 / sum(eco_pop_df$gpw_2020)
eco_weights <- stats::setNames(
  eco_pop_df$weight,
  as.character(eco_pop_df$objectid)
)
```

```{r deaths_data, eval=FALSE}
# Manually execute to rebuild deaths.RData from the store RDS.
stopifnot(isTRUE(has_store))
deaths <- data.table::as.data.table(
  readRDS(file.path(store_dir, "indiaCC.rds"))
)
data.table::setnames(
  deaths,
  old = c("ecoRegion", "yearWeek3Dow"),
  new = c("eco_region", "year_week3_dow"),
  skip_absent = TRUE
)
deaths$sqrt_pm <- sqrt(pmin(deaths$pm, 1000))
deaths$temp_div10 <- pmax(deaths$temp, -5) / 10
deaths$dom <- as.numeric(format(deaths$date, "%d"))
deaths[deaths$dom == Hmisc::monthDays(deaths$date), "dom"] <- 31
deaths[format(deaths$date, "%m%d") == "0101", "dom"] <- 0
deaths$dom <- droplevels(relevel(factor(deaths$dom), "1"))
deaths$eco_region <- factor(as.character(deaths$eco_region))
save(deaths, file = deaths_rds)
```

```{r formula_factory, include=FALSE}
# Inits/parscale from the first fit (par_info). Theta parscale is on the
# log-SD scale used by optim. Hierarchical SDs that sat at 1e-5/1e-6 are
# restarted at 1e-3 (the start from which temp_div10_iwp did move).
formula_lhs <- adlaplace::dirichlet_multinom(
  "medical",
  by = c("cell", "eco_region", "year_week3_dow"),
  init = 0.35,
  parscale = 0.5
)
# Day-of-month MLEs vs reference day 1 (levels 0, 2, ..., 31).
dom_init <- c(
  0.336, -0.070, -0.095, -0.164, -0.021, -0.116, -0.146, -0.106,
  -0.210, 0.277, -0.193, -0.002, -0.142, -0.110, 0.185, -0.131,
  -0.063, -0.036, -0.193, 0.122, -0.196, -0.209, -0.271, -0.150,
  -0.060, -0.191, -0.201, -0.109, -0.272, -0.185, -0.005
)
formula_rhs <- c(
  adlaplace::linear("dom", init = dom_init, parscale = 0.1),
  adlaplaceHgp::hiwp(
    "sqrt_pm",
    p = 2,
    ref_value = 8,
    knots = seq(0, 32, by = 2),
    by = "eco_region",
    init = c(1e-3, 1.2e-3, 1e-3),
    parscale = c(1, 0.2, 1),
    lower = 1e-12
  ),
  adlaplaceHgp::hiwp(
    "temp_div10",
    p = 3,
    ref_value = 25 / 10,
    knots = seq(-5, 42.5, by = 2.5) / 10,
    by = "eco_region",
    init = c(1e-3, 0.84, 1e-3, 1e-3),
    parscale = c(1, 0.5, 1, 1),
    lower = 1e-12
  )
)
cc_formula <- c(formula_lhs, formula_rhs)
```


```{r fit, eval=FALSE}
# Manually execute to fit / refresh analysis_cache/res_hiwp_pm_temp.rds.
stopifnot(file.exists(deaths_rds))
load(deaths_rds)
deaths <- data.table::as.data.table(deaths)
deaths[, eco_region := factor(as.character(eco_region))]

n_threads <- parallel::detectCores()
if (is.na(n_threads) || n_threads < 1L) {
  n_threads <- 1L
}
n_shards <- as.integer(
  min(1e4L, max(100L, data.table::uniqueN(deaths$cell)))
)
message(
  sprintf(
    "fitting medical N=%d threads=%d shards=%d -> %s",
    nrow(deaths), n_threads, n_shards, fit_rds
  )
)

fit <- adlaplace::adlaplace(
  formula = cc_formula,
  data = deaths,
  config = list(
    reorder_shards = "none",
    num_threads = n_threads,
    num_shards = n_shards
  ),
  control = list(
    maxit = 2000,
    trace = 3,
    REPORT = 5,
    factr = 1e3,
    pgtol = 1e-8
  ),
  hessian = FALSE,
  verbose = FALSE
)

if (!is.null(fit$details) && !is.null(fit$model_data)) {
  wts <- if (exists("eco_weights")) eco_weights else NULL
  fit$sample <- tryCatch(
    adlaplaceHgp::cond_sim_iwp(
      fit = fit$details,
      model_data = fit$model_data,
      n = 500L,
      weights = wts
    ),
    error = function(e) {
      message("cond_sim_iwp failed: ", conditionMessage(e))
      NULL
    }
  )
}
fit_ok <- is.null(fit$details$error) &&
  !is.null(fit$par_info) &&
  all(is.finite(fit$par_info$mle_internal))
if (!fit_ok) {
  message(
    "WARNING: outer fit did not produce finite MLEs",
    if (!is.null(fit$details$error)) paste0(": ", fit$details$error) else "",
    "; not overwriting ", fit_rds
  )
} else {
  fit_save <- fit
  fit_save$ad_pack <- NULL
  dir.create("analysis_cache", showWarnings = FALSE)
  saveRDS(fit_save, fit_rds)
  message(
    "saved: ", fit_rds,
    if (is.null(fit$sample$sim)) " (no sample yet)" else " (with sample)"
  )
  rm(fit_save)
}
```



```{r load_cache, include=FALSE}
res <- NULL
if (file.exists(fit_rds)) {
  message("loading: ", fit_rds)
  obj <- readRDS(fit_rds)
  mle_ok <- !is.null(obj$par_info) && (
    any(is.finite(obj$par_info$mle_internal)) ||
      any(is.finite(obj$par_info$mle))
  )
  if (isTRUE(mle_ok)) {
    res <- obj
    message("accepted: ", fit_rds)
  } else {
    message("skipping unusable cache (no finite MLE): ", fit_rds)
  }
}
if (is.null(res)) {
  message("no usable fit cache; figures/tables will leave gaps")
}

# Knit does not refit, but exposure-response figures need cond_sim draws.
if (
  !is.null(res) && is.null(res$sample$sim) &&
    !is.null(res$details) && !is.null(res$model_data)
) {
  wts <- if (exists("eco_weights")) eco_weights else NULL
  message("cache has no sample; running cond_sim_iwp")
  res$sample <- tryCatch(
    adlaplaceHgp::cond_sim_iwp(
      fit = res$details,
      model_data = res$model_data,
      n = 500L,
      weights = wts
    ),
    error = function(e) {
      message("cond_sim_iwp failed: ", conditionMessage(e))
      NULL
    }
  )
  if (!is.null(res$sample$sim)) {
    saveRDS(res, fit_rds)
    message("saved sample into ", fit_rds)
  }
}

has_sample <- !is.null(res$sample$sim)
s_var <- if (has_sample) names(res$sample$sim) else character(0)
s_var_names <- c(temp_div10 = "temperature", sqrt_pm = "PM2.5")
# Placeholders so knitr fig.subcap / fig.cap options always resolve
# even when sample caches are missing (eval=FALSE still evaluates options).
for_global <- data.frame(
  var = character(0),
  type = character(0),
  var_name = character(0),
  subcap = character(0),
  stringsAsFactors = FALSE
)
d_var <- character(0)
s_group <- character(0)
the_x_lim <- list(temp_div10 = c(4, 38), sqrt_pm = c(0, 440))
the_y_lim <- list(temp_div10 = c(-4, 45), sqrt_pm = c(-4, 11))

sample_x <- function(d_var) {
  x_here <- res$sample$x[[d_var]]
  if (is.data.frame(x_here) || (is.list(x_here) && !is.numeric(x_here))) {
    x_here <- x_here[[1]]
  }
  x_here <- as.numeric(x_here)
  if (grepl("sqrt", d_var)) {
    x_here <- x_here^2
  }
  if (grepl("div10", d_var)) {
    x_here <- x_here * 10
  }
  x_here
}
```

```{r par_table, eval=(!is.null(res)), results='asis'}
# Parameter MLEs (Mexico-style Pmisc::mdTable via canexposure::theta_table logic).
theta_table <- res$par_info[, intersect(names(res$par_info), c("label", "mle", "mle_internal", "init"))]
theta_table <- theta_table[grep("dom", theta_table$label, invert = TRUE), ]
rownames(theta_table) <- NULL
colnames(theta_table) <- gsub("_", " ", colnames(theta_table))
theta_table$label <- gsub("_", " ", theta_table$label)
theta_table$label <- gsub("div10|sqrt", "", theta_table$label)
Pmisc::mdTable(
  theta_table,
  guessGroup = FALSE,
  mdToTex = any(commandArgs() == "mdToTex"),
  dec = 8,
  caption = "Parameter MLEs: Dirichlet-multinomial overdispersion and hierarchical IWP scale parameters."
)
```

```{r dom_plot, fig.cap = "Day-of-month fixed effects (log relative risk vs day 1). Points: MLE; segments: 95\\% Wald CI from par\\_info standard errors.", fig.height = 4, fig.width = 6, eval = !is.null(res) && any(grepl("dom", res$par_info$label))}
pi <- res$par_info
dom <- pi[grepl("dom", pi$label), , drop = FALSE]
est <- if ("mle" %in% names(dom) && any(is.finite(dom$mle))) {
  as.numeric(dom$mle)
} else {
  as.numeric(dom$mle_internal)
}
se <- if ("se" %in% names(dom) && any(is.finite(dom$se))) {
  as.numeric(dom$se)
} else if ("se_internal" %in% names(dom)) {
  as.numeric(dom$se_internal)
} else {
  rep(NA_real_, nrow(dom))
}

# Day labels: prefer model_data / legacy beta_label; else contrast vs reference day 1.
day <- NULL
if (!is.null(res$model_data$term_data$info$beta)) {
  bi <- res$model_data$term_data$info$beta
  bi <- bi[grepl("dom", bi$beta_label), , drop = FALSE]
  if (nrow(bi) == nrow(dom)) {
    day <- as.integer(sub("^.*dom", "", bi$beta_label))
  }
}
if (is.null(day) && !is.null(res$parameters$beta)) {
  bi <- res$parameters$beta
  bi <- bi[grepl("dom", bi$beta_label), , drop = FALSE]
  if (nrow(bi) == nrow(dom)) {
    day <- as.integer(sub("^.*dom", "", bi$beta_label))
  }
}
if (is.null(day)) {
  # linear("dom") drops reference level "1" -> coeffs for 0,2,3,...,31
  day <- as.integer(c(0L, 2:31))
  if (length(day) != nrow(dom)) {
    day <- seq_len(nrow(dom))
  }
}

z <- stats::qnorm(0.975)
dom_df <- data.frame(
  day = day,
  est = est,
  lo = est - z * se,
  hi = est + z * se
)
# Reference day 1 (omitted contrast) at log RR = 0.
dom_df <- rbind(
  data.frame(day = 1L, est = 0, lo = NA_real_, hi = NA_real_),
  dom_df
)
dom_df <- dom_df[order(dom_df$day), , drop = FALSE]

par(mar = c(3, 3, 0.5, 0.5), bty = "l", mgp = c(1.8, 0.5, 0))
ylim <- range(c(dom_df$est, dom_df$lo, dom_df$hi), na.rm = TRUE)
plot(
  dom_df$day, dom_df$est,
  type = "n",
  xlim = range(dom_df$day),
  ylim = ylim,
  xlab = "Day of month",
  ylab = "Log relative risk",
  xaxs = "i"
)
abline(h = 0, col = "grey70", lty = 2)
ok_ci <- is.finite(dom_df$lo) & is.finite(dom_df$hi)
if (any(ok_ci)) {
  arrows(
    dom_df$day[ok_ci], dom_df$lo[ok_ci],
    dom_df$day[ok_ci], dom_df$hi[ok_ci],
    length = 0.03, angle = 90, code = 3, col = "steelblue"
  )
}
points(dom_df$day, dom_df$est, pch = 16, col = "steelblue")
```

```{r globalSetup, eval=isTRUE(has_sample)}
for_global <- expand.grid(
  var = s_var,
  type = c("Common", "Pop Avg"),
  stringsAsFactors = FALSE
)
for_global$var_name <- unname(s_var_names[for_global$var])
for_global$subcap <- paste(for_global$var_name, for_global$type)
```

```{r globalRes, fig.cap='Exposure-response: common component (left column of each pair) and population-weighted average. Grey: conditional draws. Green: pointwise 2.5/50/97.5\\% quantiles. Red: 80\\% global envelope.', fig.subcap = for_global$subcap, out.width='0.48\\linewidth', fig.height=5, fig.width=5, eval=isTRUE(has_sample) && nrow(for_global) > 0}
par(mar = c(3, 3, 0, 0), bty = "l")

for (d in seq_len(nrow(for_global))) {
  d_var <- for_global[d, "var"]
  x_here <- sample_x(d_var)

  plot(
    NA,
    xlim = the_x_lim[[d_var]], ylim = the_y_lim[[d_var]],
    xlab = s_var_names[d_var], ylab = "100 x (RR-1)",
    xaxs = "i", yaxs = "i"
  )

  if (for_global[d, "type"] == "Common") {
    sample <- 100 * (res$sample$sim[[d_var]] - 1)
    matlines(x_here, sample, type = "l", col = "#00000010", lty = 1)
    envelope_here <- res$sample$envelope$common[[d_var]]
    quantiles_here <- res$sample$quantiles$common[[d_var]]
  } else {
    envelope_here <- res$sample$envelope$weighted[[d_var]]
    quantiles_here <- res$sample$quantiles$weighted[[d_var]]
  }
  if (!is.null(quantiles_here) && !inherits(quantiles_here, "try-error")) {
    matlines(
      x_here, 100 * (quantiles_here - 1),
      lty = c(2, 1, 2), lwd = c(1, 3, 1), col = "green"
    )
  }
  if (!is.null(envelope_here) && !inherits(envelope_here, "try-error")) {
    matlines(
      x_here, 100 * (envelope_here - 1),
      lty = c(3, 2, 3), lwd = c(1, 2, 1), col = "red"
    )
  }
}
```

```{r byRegionTemplate, eval=FALSE, include=FALSE}
x_here <- sample_x(d_var)
s_group <- names(res$sample$quantiles$group[[d_var]])
if (!length(s_group)) {
  s_group <- names(res$sample$envelope$group[[d_var]])
}
sim_global_here_y <- 100 * (res$sample$quantiles$common[[d_var]] - 1)

for (d_group in s_group) {
  group_key <- as.character(d_group)
  the_yq_raw <- res$sample$quantiles$group[[d_var]][[group_key]]
  the_ye_raw <- res$sample$envelope$group[[d_var]][[group_key]]
  if (is.null(the_yq_raw) || inherits(the_yq_raw, "try-error")) {
    warning("missing group quantiles for ", d_var, " / ", d_group)
    next
  }
  the_yq <- 100 * (the_yq_raw - 1)
  plot(
    x_here, sim_global_here_y[, 2],
    type = "l",
    xlab = s_var_names[d_var], ylab = "100 x (RR-1)",
    col = "grey", xaxs = "i", lwd = 4,
    ylim = the_y_lim[[d_var]], xlim = the_x_lim[[d_var]]
  )
  matlines(
    x_here, the_yq,
    lty = c(2, 1, 2), lwd = c(1, 2, 1), col = "green"
  )
  if (!is.null(the_ye_raw) && !inherits(the_ye_raw, "try-error")) {
    matlines(
      x_here, 100 * (the_ye_raw - 1),
      lty = c(2, 1, 2), lwd = c(1, 2, 1), col = "red"
    )
  }
}
```

```{r byRegionSetup1, eval=isTRUE(has_sample)}
d_var <- grep("temp", names(res$sample$sim), value = TRUE)[1]
if (is.na(d_var) || !nzchar(d_var)) {
  d_var <- character(0)
}
s_group <- if (length(d_var)) {
  names(res$sample$quantiles$group[[d_var]])
} else {
  character(0)
}
```

```{r byRegion1, fig.cap=if (length(d_var)) paste0(s_var_names[d_var], " by eco-region. Grey: common-component median. Green: pointwise 2.5/50/97.5\\% quantiles of the region-specific curve. Red: 80\\% global envelope (central curve and bounds).") else "temperature", fig.subcap = s_group, fig.height=4, fig.width=4, ref.label = 'byRegionTemplate', eval=isTRUE(has_sample) && length(d_var) > 0, out.width='0.23\\linewidth', fig.align='center'}
```

```{r byRegionSetup2, eval=isTRUE(has_sample)}
d_var <- grep("sqrt_pm|^pm$|_pm$", names(res$sample$sim), value = TRUE)[1]
if (is.na(d_var) || !nzchar(d_var)) {
  d_var <- character(0)
}
s_group <- if (length(d_var)) {
  names(res$sample$quantiles$group[[d_var]])
} else {
  character(0)
}
```

```{r byRegion2, fig.cap=if (length(d_var)) paste0(s_var_names[d_var], " by eco-region. Grey: common-component median. Green: pointwise 2.5/50/97.5\\% quantiles of the region-specific curve. Red: 80\\% global envelope (central curve and bounds).") else "PM2.5", fig.subcap = s_group, fig.height=4, fig.width=4, ref.label = 'byRegionTemplate', eval=isTRUE(has_sample) && length(d_var) > 0, out.width='0.23\\linewidth', fig.align='center'}
```

```{r ecoRegionmap, fig.cap = 'Eco regions', fig.subcap = c('regions','legend'), fig.height=5, fig.width=5, out.width = '0.48\\linewidth', eval=isTRUE(has_store)}
mapmisc::map.new(india_eco_regions)
terra::plot(india_eco_regions, add = TRUE, col = india_eco_regions$col)
terra::text(terra::centroids(india_eco_regions), india_eco_regions$objectid)

mapmisc::map.new(india_eco_regions)
mapmisc::legendBreaks(
  "left",
  col = rev(india_eco_regions$col),
  legend = rev(india_eco_regions$id_name),
  cex = 0.7, bty = "n", width = Inf
)
```

```{r ecoTable, results='asis', eval=FALSE, include=FALSE}
knitr::kable(
  mapmisc::legendTable(data.frame(
    label = india_eco_regions$id_name,
    col = india_eco_regions$col
  ), type = "latex"),
  format = "latex", escape = FALSE,
  align = c("c", "l"),
  col.names = c("", "Eco region"),
  caption = "Eco-regions used in the analysis"
) |>
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    font_size = 9
  ) |>
  kableExtra::column_spec(1, latex_column_spec = "p{0.05\\\\linewidth}") |>
  kableExtra::column_spec(2, latex_column_spec = "p{0.7\\\\linewidth}")
```
