---
title: "Second derivatives of the multivariate normal probability"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
# Disabled: not registered as a package vignette.
# vignette: >
#   %\VignetteIndexEntry{Second derivatives of the multivariate normal probability}
#   %\VignetteEngine{knitr::rmarkdown}
#   %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(admvn)
```

This vignette shows how to obtain second derivatives of a multivariate normal
probability with respect to the upper integration limits. The function
`pmvn()` returns the probability
\[
  F(\mathbf{x}) = P(\mathbf{l} < \mathbf{X} < \mathbf{x}),
  \qquad \mathbf{X} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma}),
\]
together with the gradient $\nabla F(\mathbf{x})$ and Hessian
$\nabla^2 F(\mathbf{x})$ with respect to the upper limits $\mathbf{x}$.

## Bivariate example

```{r bivariate}
sigma <- matrix(c(1.0, 0.4, 0.4, 1.2), 2, 2)
upper <- c(0.5, 0.3)
out <- pmvn(
  upper,
  lower = c(-Inf, -Inf),
  mean = c(0.1, -0.2),
  sigma = sigma,
  n_points = 2000L,
  n_shifts = 12L
)
out$value
out$gradient
out$hessian
```

The `hessian` component is the matrix of second partial derivatives:

$$
\left(\frac{\partial^2 F}{\partial x_i \partial x_j}\right)_{i,j}.
$$

For example, the mixed partial $\partial^2 F / \partial x_1 \partial x_2$ is:

```{r mixed-partial}
out$hessian[1, 2]
```

## Checking against a numerical Hessian

When `numDeriv` is available, the CppAD Hessian can be compared with a
finite-difference approximation of the same QMC probability:

```{r numderiv, eval = requireNamespace("numDeriv", quietly = TRUE)}
f <- function(x) {
  pmvn(
    x,
    lower = c(-Inf, -Inf),
    mean = c(0.1, -0.2),
    sigma = sigma,
    n_points = 2000L,
    n_shifts = 12L
  )$value
}
numDeriv::hessian(f, upper)
out$hessian
```

The two matrices should agree up to Monte Carlo noise in the underlying Genz
approximation.

## Reusing the CppAD tape

If you need the Hessian at many upper limits with fixed $\boldsymbol{\Sigma}$,
$\boldsymbol{\mu}$, and lower limits, build the tape once with `pmvn_fun()`.
The tape covers upper limits, mean, marginal scales, and the Genz Cholesky
factor; by default `eval(..., inner = TRUE)` returns derivatives with respect
to the upper limits only (sparse inner patterns, as in adlaplace):

```{r tape}
fun <- pmvn_fun(
  lower = c(-Inf, -Inf),
  mean = c(0.1, -0.2),
  sigma = sigma,
  n_points = 2000L,
  n_shifts = 12L,
  seed = 42L
)
fun$eval(c(0.5, 0.3), inner = TRUE)$hessian
fun$eval(c(0.6, 0.1), inner = TRUE)$hessian
```

You can change `mean` or `sigma` at evaluation time without rebuilding the tape
(the Genz permutation is fixed from the seed values used at build time):

```{r tape-vary-params}
fun$eval(c(0.5, 0.3), mean = c(0.2, -0.1))$value
fun$eval(c(0.5, 0.3), sigma = sigma * 1.05)$value
```

Each call reuses the same CppAD tape and only re-evaluates the forward pass and
sparse derivative sweeps at the new parameter values.
