---
title: "Laplace Approximations with adlaplace"
author: "Patrick Brown"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Laplace Approximations with adlaplace}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library("Matrix")
library("adlaplace")
```

This vignette walks through a hierarchical negative-binomial model: simulate data,
build an automatic-differentiation (AD) representation as **shards**, validate
derivatives, solve the inner Laplace problem for random effects, and fit outer
parameters with a profile likelihood.

## Model

Observations follow a negative binomial with mean $\mu = \exp(\eta)$. We use two
equivalent parameterizations: NB **size** $r$, and Gamma-mixing **standard deviation**
$\theta$ (with $E[Z]=1$, $\mathrm{sd}(Z)=\theta$ in the Poisson–Gamma construction).
They are linked by $r = 1/\theta^2$.

**Size $r$.**

$$
\begin{aligned}
f(y;\mu,r) =  & \log\Gamma(y + r) - \log\Gamma(r) - \log\Gamma(y + 1) 
	+ r \log\left(\frac{r}{r + \mu}\right)
	+ y \log\left(\frac{\mu}{r + \mu}\right) \\
	= &  \log\Gamma(y + r) - \log\Gamma(r) - \log\Gamma(y + 1)
	+ r \log r - r \log(r + \mu) + y \log(\mu) - y \log(r + \mu)
\end{aligned}
$$

**SD $\theta$** ($r = 1/\theta^2$).

$$
\begin{aligned}
f(y;\mu,\theta) =  & \log\Gamma\left(y + \theta^{-2}\right) - \log\Gamma\left(\theta^{-2}\right) - \log\Gamma(y + 1) 
	+ \theta^{-2} \log\left(\frac{\theta^{-2}}{\theta^{-2} + \mu}\right)
	+ y \log\left(\frac{\mu}{\theta^{-2} + \mu}\right) \\
	= &  \log\Gamma\left(y + \theta^{-2}\right) - \log\Gamma\left(\theta^{-2}\right) - \log\Gamma(y + 1)
	- 2 \theta^{-2} \log \theta - \theta^{-2} \log\left(\theta^{-2} + \mu\right) + y \log(\mu) - y \log\left(\theta^{-2} + \mu\right)
\end{aligned}
$$

The linear predictor is $\eta = X\beta + A\gamma$, with two independent random-effect
groups in $A$ and log-standard-deviation parameters for each group plus a global
overdispersion parameter. Densities are implemented under `src/densities/` (e.g.
`nbinom.cpp` for observations and hyperparameters, `random_diagonal.cpp` for
random effects).

## Simulated data

We generate 5000 observations, fixed effects, two random-effect groups, and
Poisson–Gamma overdispersion. Outer parameters are stored on the log-SD scale
(`transform_theta = TRUE`).

```{r theData}
Nobs <- 5000
Nrandom1 <- 10
Nrandom2 <- 25

set.seed(0)

X <- Matrix::Matrix(cbind(1, rbinom(Nobs, 1, prob = 0.5)))

Adf <- data.frame(
  r1 = sample(Nrandom1, Nobs, replace = TRUE),
  r2 = sample(Nrandom2, Nobs, replace = TRUE)
)
AmatList <- list(
  r1 = Matrix::sparseMatrix(
    i = seq_len(Nobs),
    j = Adf$r1
  ),
  r2 = Matrix::sparseMatrix(
    i = seq_len(Nobs),
    j = Adf$r2
  )
)
Amat <- do.call(cbind, AmatList)

beta <- rep(1, ncol(X))
thetaOrig <- c(0.1, 0.1, 0.2) # SD for each random group, then overdispersion
theta <- log(thetaOrig)

gamma <- rnorm(
  ncol(Amat),
  sd = rep(thetaOrig[1:2], c(Nrandom1, Nrandom2))
)
eta_true <- as.vector(X %*% beta + Amat %*% gamma)
mu_true <- exp(eta_true)
Z <- rgamma(length(mu_true), thetaOrig[3]^(-2), thetaOrig[3]^(-2))
y <- rpois(length(mu_true), mu_true * Z)

config <- list(
  beta = beta,
  theta = log(thetaOrig),
  transform_theta = TRUE,
  gamma = rep(0, ncol(Amat)),
  shards = adlaplace::ad_shards(Amat, num_shards = 1000),
  verbose = FALSE,
  package = "adlaplace"
)

n_beta <- length(config$beta)
n_gamma <- length(config$gamma)
n_theta <- length(config$theta)
```

## Building AD shards

Each model piece is compiled separately with `ad_fun_ptr()`. The full model has
four shards: observation likelihood, hyperparameter (extra) terms, and one shard
per random-effect group. Shards are merged later with `ad_fun()`.

`beta_map` and `gamma_map` map local design columns of `X` and `A` into the
global `(beta, gamma, theta)` vector: column `j` of `beta_map` has one structural
`1` at row `r` when local `X` column `j` uses global beta `r` (and similarly for
`gamma_map` / `A`). When `ncol(X) > 0`, `ncol(beta_map)` must equal `ncol(X)` with
exactly one nonzero per column.

### Observations and hyperparameters

The observation shard uses the third $\theta$ component (global overdispersion).
The extra shard contributes the prior on $\theta$.

```{r adShards_obs}
data_obs <- adlaplace:::ad_data(
  y = y,
  A = Amat,
  X = X,
  theta_map = Matrix::sparseMatrix(
    i = n_theta,
    j = 1L,
    dims = c(n_theta, 1L)
  ),
  ad_kind = "observations",
  ad_fun = "nbinom_obs"
)

data_extra <- adlaplace::ad_data(
  data_obs,
  ad_kind = "parameters",
  ad_fun = "nbinom_extra"
)

ad_fun_obs <- adlaplace::ad_fun_ptr(data = data_obs, config = config)
ad_fun_extra <- adlaplace::ad_fun_ptr(data = data_extra, config = config)
```

### Random effects

Each group gets a diagonal normal prior with its own $\theta$ index.

```{r adShards_random}
model_r1 <- adlaplace:::ad_data(
  beta_map = n_beta,
  gamma_map = Matrix::sparseMatrix(
    i = seq_len(Nrandom1),
    j = seq_len(Nrandom1),
    dims = c(n_gamma, Nrandom1)
  ),
  theta_map = c(1L, n_theta),
  ad_kind = "random",
  ad_fun = "random_diagonal",
  precision = rep(1, Nrandom1)
)
ad_fun_r1 <- adlaplace::ad_fun_ptr(data = model_r1, config = config)

model_r2 <- adlaplace:::ad_data(
  beta_map = n_beta,
  gamma_map = Matrix::sparseMatrix(
    i = seq.int(Nrandom1 + 1L, length.out = Nrandom2),
    j = seq_len(Nrandom2),
    dims = c(n_gamma, Nrandom2)
  ),
  theta_map = c(2L, n_theta),
  ad_kind = "random",
  ad_fun = "random_diagonal",
  precision = rep(1, Nrandom2)
)
ad_fun_r2 <- adlaplace::ad_fun_ptr(data = model_r2, config = config)
```

## Observation likelihood check

Scan the overdispersion parameter while holding $(\beta, \gamma, \theta_{1:2})$
fixed at the data-generating values. The AD joint log density (observation +
extra shards) should match `stats::dnbinom` on the full sample.

```{r obsLogDens}
xx <- c(beta, gamma, config$theta)
log_theta_true <- config$theta[n_theta]

nb_log_dens <- function(log_theta) {
  r <- exp(-2 * log_theta)
  sum(stats::dnbinom(y, mu = mu_true, size = r, log = TRUE))
}

ad_nb_log_dens <- function(log_theta) {
  xx_scan <- xx
  xx_scan[length(xx_scan)] <- log_theta
  adlaplace::joint_log_dens(ad_fun_obs, xx_scan, negative = FALSE) +
    adlaplace::joint_log_dens(ad_fun_extra, xx_scan, negative = FALSE)
}

seq_log_theta <- log_theta_true + seq(-0.5, 0.5, length.out = 7)
cmp <- data.frame(
  theta = exp(seq_log_theta),
  stats = vapply(seq_log_theta, nb_log_dens, numeric(1)),
  ad = vapply(seq_log_theta, ad_nb_log_dens, numeric(1))
)
cmp$diff <- cmp$ad - cmp$stats
cmp
```

## Combining shards

Shard handles are combined with `c()`. This uses move semantics: after
`ad_fun()` is called, the individual `ad_fun_ptr` objects are cleared and must
not be reused.

```{r adShards_combined}
ad_fun_plain <- c(
  ad_fun_obs, ad_fun_extra,
  ad_fun_r1, ad_fun_r2
)
ad_pack <- adlaplace::ad_fun(ad_fun_plain, num_threads = 12)
```

The per-shard log densities sum to the joint log density at a fixed parameter
vector.

```{r checkByShard}
config$gamma <- rep(0.01, length(config$gamma))
x_full <- c(config$beta, config$gamma, config$theta)
shards <- seq.int(from = 0L, length.out = adlaplace::n_groups(ad_fun_plain))

by_shard <- vapply(
  shards,
  function(s) {
    adlaplace::joint_log_dens(
      ad_fun_plain, x_full,
      shards = s, negative = TRUE
    )
  },
  numeric(1)
)

res_fdf <- fun_obj_fdfh(
  ad_pack,
  c(config$beta, config$theta), config$gamma,
  inner = FALSE, verbose = TRUE
)

by_shard[seq(1, min(c(5, length(by_shard))))]
c(
  sum(by_shard),
  adlaplace::joint_log_dens(ad_fun_plain, x_full, negative = TRUE),
  adlaplace::joint_log_dens(ad_pack, x_full, negative = TRUE),
  res_fdf$f
)
```

## Joint log density derivatives

Gradients and Hessians can be evaluated on all shards or on a subset. Results
should agree when the subset is the full model.

```{r testLogDens}
log_all <- adlaplace::joint_log_dens(ad_fun_plain, x_full, negative = TRUE)
log_sub <- adlaplace::joint_log_dens(
  ad_fun_plain, x_full,
  shards = 0:3, negative = TRUE
)

grad_all <- adlaplace::grad(ad_fun_plain, x_full, negative = TRUE)
grad_sub <- adlaplace::grad(
  ad_fun_plain, x_full,
  shards = 0:3, negative = TRUE
)

hess_all <- adlaplace::hessian(ad_fun_plain, x_full, negative = TRUE)
hess_sub <- adlaplace::hessian(
  ad_fun_plain, x_full,
  shards = 0:3, negative = TRUE
)

c(
  log_diff = log_all - res_fdf$f,
  max_abs_grad_diff = max(abs(grad_all - res_fdf$grad)),
  max_abs_hess_diff = max(abs(
    as.matrix(hess_all) - as.matrix(res_fdf$hessian)
  ))
)
```

## Inner optimization

Given outer parameters $(\beta, \theta)$, `inner_opt()` maximizes over
$\gamma$ (equivalently minimizes $-\log p(y, \gamma \mid \beta, \theta)$).
The returned solution should track the simulated random effects.

```{r testInnerOpt}
x_outer <- c(config$beta, config$theta)
x_full <- c(config$beta, config$gamma, config$theta)


inner_res <- adlaplace::inner_opt(
  parameters = x_outer,
  gamma = config$gamma,
  ad_fun = ad_pack,
  control = list(
    maxit = 2,
    report.level = 3,
    report.freq = 1,
    grad.tol = 1e-8,
    cg.tol = 1e-6
  ),
  deriv = TRUE,
  verbose = TRUE
)
adlaplace::joint_log_dens(ad_pack, x_full, negative = TRUE)


inner_res$opt$fval

plot(
  inner_res$opt$solution, gamma,
  xlab = "estimated", ylab = "true"
)
abline(0, 1, col = "red", lty = 2)
```

## likelihood

`log_lik_laplace()` runs the inner optimizer and returns the Laplace
approximation to the marginal log likelihood. With `deriv = TRUE`, it also
returns the profile gradient and third-order trace terms used in the
determinant correction.

OpenMP shard affinity is fixed when you call `ad_fun(ad_fun_plain, num_threads = …)`
(`owner_thread = shard_index %% num_threads`); `log_lik_laplace()` does not change it.
The same assignment is used for inner optimization, outer derivatives when
`deriv = TRUE`, and the profile trace (`trace_hinv_t` logic, same OpenMP shard
groups). `inner_opt()` uses one OpenMP/CppAD parallel session for trust-region
inner optimization, outer `get_fdfh`, and trace evaluation when `deriv = TRUE`.
`ad_fun_ptr()` and `joint_log_dens()` on a plain
ptr do not use threads. For serial runs only, use `ad_fun(..., num_threads = 1L)`.

For thread-affinity diagnostics, reinstall from a fresh tarball with
`PKG_CXXFLAGS="-DDEBUG" R CMD INSTALL adlaplace_*.tar.gz` (use `R CMD INSTALL --clean`
if installing from source). Configure prints when DEBUG is enabled. With
`config$verbose = TRUE`, `inner_opt` prints CppAD session phase messages (trust
region, outer `get_fdfh`, trace shards). With `deriv = TRUE`, trace runs inside
that session; standalone `trace_hinv_t()` remains for direct use. Each
`CppadParallelScope` end flushes per-thread CppAD pools before the next `.Call`.

```{r testLogLik}
res_noderiv <- adlaplace::log_lik_laplace(
  x = x_outer,
  ad_fun = ad_pack,
  config = config,
  deriv = FALSE
)

res_deriv <- adlaplace::log_lik_laplace(
  x = x_outer,
  ad_fun = ad_pack,
  config = config,
  deriv = TRUE
)

data.frame(
  log_lik = c(res_noderiv$log_lik, res_deriv$log_lik),
  neg_log_lik = c(res_noderiv$neg_log_lik, res_deriv$neg_log_lik),
  inner_fval = c(res_deriv$opt$fval, res_deriv$opt$fval),
  row.names = c("deriv = FALSE", "deriv = TRUE")
)

summary(res_deriv$extra$trace3)
res_deriv$grad
```

## Outer optimization

`outer_fn()` and `outer_gr()` wrap `log_lik_laplace()` for use with
`stats::optim()`. A small cache environment stores the current inner
$\gamma$ start.

```{r trustOptimOuterWrappers}
cache <- new.env(parent = emptyenv())
cache$gamma <- config$gamma

control_inner <- list(
  maxit = 100,
  report.level = 0,
  report.freq = 0
)

optim_control <- list(maxit = 200, trace = 0)
bounds <- list(lower = rep(-5, length(x_outer)), upper = rep(5, length(x_outer)))

common_optim_args <- c(
  list(
    par = x_outer,
    fn = adlaplace::outer_fn,
    method = "L-BFGS-B",
    control = optim_control,
    config = config,
    ad_fun = ad_pack,
    cache = cache,
    control_inner = control_inner
  ),
  bounds
)

outer_fit_no_g <- do.call(stats::optim, common_optim_args)
outer_fit <- do.call(stats::optim, c(common_optim_args, list(gr = adlaplace::outer_gr)))

fit_cmp <- rbind(
  optim = outer_fit$par,
  optim_no_grad = outer_fit_no_g$par,
  true = c(beta, log(thetaOrig))
)
colnames(fit_cmp) <- c(
  paste0("beta[", seq_len(n_beta), "]"),
  paste0("log_theta[", seq_len(n_theta), "]")
)
fit_cmp
exp(outer_fit$par[n_beta + seq_len(n_theta)])
outer_fit$value
```

## Derivative checks

Finite-difference checks along one outer coordinate validate the profile gradient,
the determinant derivative, and the chain rule through the inner mode
$u(\beta, \theta)$.

```{r testLogLgrad, fig.height=7}
if (!requireNamespace("abind", quietly = TRUE)) {
  stop("Install suggested package 'abind' to run derivative checks.")
}
config$verbose <- FALSE
par(mfrow = c(3, 2), mar = c(2, 2, 2, 0), mgp = c(1, 0.5, 0))

x1 <- outer_fit$par
Dpar <- length(x1)
Ngrid <- 7L

par_grid <- matrix(x1, nrow = Ngrid, ncol = Dpar, byrow = TRUE)
Sx <- x1[Dpar] + seq(-0.5, 0.5, length.out = Ngrid)
SxD <- Sx[-1] - diff(Sx) / 2
par_grid[, Dpar] <- Sx

res_scan <- lapply(
  split(par_grid, row(par_grid)),
  adlaplace::log_lik_laplace,
  ad_fun = ad_pack,
  config = config,
  deriv = TRUE,
  gamma = cache$gamma
)

SnegLik <- vapply(res_scan, `[[`, numeric(1), "neg_log_lik")
Sdet <- vapply(res_scan, function(r) r$extra$hessian$half_log_det, numeric(1))
grad_mat <- do.call(rbind, lapply(res_scan, `[[`, "grad"))
extra_df <- do.call(abind::abind, c(lapply(res_scan, `[[`, "deriv"), along = 3))
dU <- do.call(
  abind::abind,
  c(lapply(res_scan, function(r) as.matrix(r$extra$dU)), along = 3)
)
u_hat <- do.call(rbind, lapply(res_scan, function(r) r$opt$solution))

res_mid <- res_scan[[(Ngrid + 1L) %/% 2L]]

plot(Sx, SnegLik, type = "l", xlab = expression(theta[3]), ylab = "-log lik")
plot(Sx, grad_mat[, Dpar], type = "l", xlab = expression(theta[3]), ylab = "grad")
points(SxD, diff(SnegLik) / diff(Sx), pch = 16)
legend("topright", legend = c("AD", "finite diff"), lty = c(1, NA), pch = c(NA, 1), bty = "n")

Du <- 2L
plot(Sx, u_hat[, Du], type = "l", xlab = expression(theta[3]), ylab = expression(u[Du]))
plot(
  Sx, dU[Du, Dpar, ],
  type = "l",
  xlab = expression(theta[3]),
  ylab = expression(d * u[Du] / d * theta[3])
)
points(SxD, diff(u_hat[, Du]) / diff(Sx), pch = 16)

plot(Sx, Sdet, type = "l", xlab = expression(theta[3]), ylab = "half log det")
plot(
  Sx, extra_df[Dpar, "d_det", ],
  type = "l",
  xlab = expression(theta[3]),
  ylab = expression(d / d * theta[3] ~ log ~ det)
)
points(SxD, diff(Sdet) / diff(Sx), pch = 16)

hessian_outer <- Matrix::forceSymmetric(res_mid$extra$hessian$outer)
hessian_plain <- adlaplace::hessian(
  ad_fun_plain,
  res_mid$full_parameters,
  inner = FALSE,
  negative = TRUE
)
max_abs_hess_diff <- max(abs((hessian_outer - hessian_plain)@x), na.rm = TRUE)
max_abs_hess_diff
```


```{r checkDu}  
seq_gamma1 <- seq.int(
  ad_pack@sizes["beta"] + 1L,
  length.out = ad_pack@sizes["gamma"]
)
#  dU <- -Hstuff$H_inv %*% hessian_pack$outer[seq_gamma1, -seq_gamma1]
dU_check <- -solve(
  res_mid$extra$hessian$inner,
  res_mid$extra$hessian$outer[seq_gamma1, -seq_gamma1]
)
dU_check2 <- -
  res_mid$extra$hessian$H_inv %*%
  res_mid$extra$hessian$outer[seq_gamma1, -seq_gamma1]

quantile(abs(dU_check - res_mid$extra$dU))
quantile(abs(dU_check - dU_check2))
quantile(abs(dU_check2 - res_mid$extra$dU))

quantile(as.matrix(solve(res_mid$extra$hessian$inner)) - as.matrix(res_mid$extra$hessian$H_inv))
quantile(as.matrix(solve(res_mid$extra$hessian$inner)) - Matrix::tcrossprod(res_mid$extra$hessian$half_H_inv))
quantile(as.matrix(res_mid$extra$hessian$H_inv) - Matrix::tcrossprod(res_mid$extra$hessian$half_H_inv))


(as.matrix(solve(res_mid$extra$hessian$inner)) - as.matrix(res_mid$extra$hessian$H_inv))[1:5, 1:5]


par(mfrow = c(1, 1))

plot(
  Sx, dU[Du, Dpar, ],
  type = "l",
  xlab = expression(theta[3]),
  ylab = expression(d * u[Du] / d * theta[3])
)
points(SxD, diff(u_hat[, Du]) / diff(Sx), pch = 16)
points(rep(res_mid$parameters[Dpar], 2),
  c(dU_check[Du, Dpar], dU_check2[Du, Dpar]),
  col = c("red", "blue"), cex = c(5, 3)
)
```

## Joint density derivative checks

The same finite-difference idea applies to the full joint log density
$-\log p(\beta, \gamma, \theta \mid y)$ and its AD gradient and Hessian.

```{r derivJointDens, fig.height=5}
par(mfrow = c(4, 2), mar = c(2, 2, 2, 0), mgp = c(1, 0.5, 0))

x_here <- res_mid$full_parameters
Dpar_dens <- length(x_here) - 1
Ngrid <- 11L
shards <- seq.int(from = 0L, length.out = adlaplace::n_groups(ad_fun_plain))

par_grid <- matrix(x_here, nrow = Ngrid, ncol = length(x_here), byrow = TRUE)
Sx <- x_here[Dpar_dens] + seq(-0.1, 0.1, length.out = Ngrid)
SxD <- Sx[-1] - diff(Sx) / 2
par_grid[, Dpar_dens] <- Sx
x_list <- split(par_grid, row(par_grid))

dens <- vapply(
  x_list,
  adlaplace::joint_log_dens,
  numeric(1),
  ad_fun = ad_fun_plain,
  shards = shards,
  negative = FALSE
)

rbind(
  dens[1:5],
  dens[seq(to = length(dens), length.out = 5)]
)

grad <- do.call(
  cbind,
  lapply(
    x_list,
    adlaplace::grad,
    ad_fun = ad_fun_plain,
    inner = FALSE,
    shards = shards,
    negative = FALSE
  )
)

plot(Sx, grad[Dpar_dens, ], type = "l", ylab = "AD gradient")
points(SxD, diff(dens) / diff(Sx), pch = 16)
legend("topright", legend = c("AD", "finite diff"), lty = c(1, NA), pch = c(NA, 1), bty = "n")

hes <- array(
  dim = c(length(x_here), length(x_here), Ngrid),
  dimnames = list(NULL, NULL, NULL)
)
for (i in seq_len(Ngrid)) {
  hes[, , i] <- as.matrix(adlaplace::hessian(
    ad_fun_plain,
    x_list[[i]],
    inner = FALSE,
    shards = shards,
    negative = FALSE
  ))
}

grad_slope <- apply(grad, 1, diff) / mean(diff(Sx))
s_par2 <- sort(unique(c(Dpar_dens, 1:3, seq(to = length(x_here), length.out = 5))))
s_par2 <- s_par2[seq(1, min(c(prod(par("mfrow")) - 1, length(s_par2))))]
for (Dpar2 in s_par2) {
  plot(
    Sx, hes[Dpar_dens, Dpar2, ],
    type = "l",
    ylab = paste0("H[", Dpar_dens, ",", Dpar2, "]"),
    ylim = range(c(hes[Dpar_dens, Dpar2, ], grad_slope[, Dpar2]), na.rm = TRUE)
  )
  points(SxD, grad_slope[, Dpar2], pch = 16)
}
```


##  nicer interface

Code verifying the omnibus interface gives the same results as barebones implementation.

```{r checkNicerInterface}
dat <- cbind(
  data.frame(y = y),
  Adf,
  x = X[, 2]
)
md <- model_data(
  data = dat,
  formula =
    nbinom(y, lower = 1e-9, init = 0.15) ~
      x + iid(r1, init = 0.1) + iid(r2, init = 0.1)
)
config2 <- list(
  transform_theta = TRUE,
  shards = ad_shards(
    md$data$A,
    num_shards = 100
  )
)
ad_fun2 <- ad_fun(md, config2, num_threads = 4L)

md$observations$y@XTp[, 1:5]
data_obs@XTp[, 1:5]

md$data$info$parameters[, c("term", "model", "label", "init", "transform")]
data_obs@theta_map

param_remap <- c(2, 1, 5, 3, 4)
log_lik2 <- log_lik_laplace(
  x = res_deriv$parameters[param_remap],
  config = config2,
  ad_fun = ad_fun2,
  deriv = TRUE
)
c(log_lik2$log_lik, res_deriv$log_lik)
rbind(log_lik2$grad, res_deriv$grad[param_remap])
```


# mean estimation

Posterior mean of the random effects can be approximated with sparse nested
Gauss–Hermite quadrature (`mvQuad`): nodes are drawn from the Laplace proposal
`N(hat(gamma), H^{-1})` and reweighted by `joint_log_dens` relative to that
Gaussian.

```{r meanEstimation}
library(mvQuad)
mean_estimate <- mean_mvquad(
  parameters = log_lik2$parameters,
  mode = log_lik2$opt$solution,
  cov = log_lik2$extra$hessian$H_inv,
  ad_fun = ad_fun2,
  n = 3
)
quantile(mean_estimate - log_lik2$opt$solution)
plot(log_lik2$opt$solution, mean_estimate)
abline(0, 1, col = "red", lty = 2)
```