---
title: "Laplace Approximations with adlaplace"
author: "Patrick Brown"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Laplace Approximations with adlaplace}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette is a low-level test and demonstration of the adlaplace package.


Negative binomial with $E(y) = \mu$. We use two equivalent parameterizations: NB **size** $r$, and Gamma-mixing **standard deviation** $\theta$ (with $E[Z]=1$, $\mathrm{sd}(Z)=\theta$ in the Poisson–Gamma construction). They are linked by $r = 1/\theta^2$.

**Size $r$.**

$$
\begin{aligned}
f(y;\mu,r) =  & \log\Gamma(y + r) - \log\Gamma(r) - \log\Gamma(y + 1) 
	+ r \log\left(\frac{r}{r + \mu}\right)
	+ y \log\left(\frac{\mu}{r + \mu}\right) \\
	= &  \log\Gamma(y + r) - \log\Gamma(r) - \log\Gamma(y + 1)
	+ r \log r - r \log(r + \mu) + y \log(\mu) - y \log(r + \mu)
\end{aligned}
$$

**SD $\theta$** ($r = 1/\theta^2$).

$$
\begin{aligned}
f(y;\mu,\theta) =  & \log\Gamma\left(y + \theta^{-2}\right) - \log\Gamma\left(\theta^{-2}\right) - \log\Gamma(y + 1) 
	+ \theta^{-2} \log\left(\frac{\theta^{-2}}{\theta^{-2} + \mu}\right)
	+ y \log\left(\frac{\mu}{\theta^{-2} + \mu}\right) \\
	= &  \log\Gamma\left(y + \theta^{-2}\right) - \log\Gamma\left(\theta^{-2}\right) - \log\Gamma(y + 1)
	- 2 \theta^{-2} \log \theta - \theta^{-2} \log\left(\theta^{-2} + \mu\right) + y \log(\mu) - y \log\left(\theta^{-2} + \mu\right)
\end{aligned}
$$


Generate some data.

```{r theData}
Nobs <- 5000
Nrandom1 <- 10
Nrandom2 <- 25

set.seed(0)
X <- Matrix::Matrix(cbind(1, rbinom(Nobs, 1, prob = 0.5)))
AmatList <- list(
  r1 = Matrix::sparseMatrix(
    i = 1:Nobs,
    j = sample(Nrandom1, Nobs, replace = TRUE)
  ),
  r2 = Matrix::sparseMatrix(
    i = 1:Nobs,
    j = sample(Nrandom2, Nobs, replace = TRUE)
  )
)
Amat <- do.call(cbind, AmatList)

beta <- rep(1, ncol(X))

thetaOrig <- c(0.1, 0.1, 0.2)
theta <- log(thetaOrig)

gamma <- rnorm(ncol(Amat), sd = rep(thetaOrig[1:2], c(Nrandom1, Nrandom2)))
eta_true <- as.vector(X %*% beta + Amat %*% gamma)
etaT <- exp(eta_true)
Z <- rgamma(length(etaT), thetaOrig[3]^(-2), thetaOrig[3]^(-2))
c(mean(Z), sd(Z))
y <- rpois(length(etaT), etaT * Z)


config <- list(
  beta = beta,
  theta = log(thetaOrig), # sd for each Amat, overdisp parameter
  transform_theta = TRUE,
  gamma = rep(0, ncol(Amat)),
  shards = adlaplace::ad_shards(Amat, num_shards = 1000),
  num_threads = 10L,
  verbose = TRUE,
  package = "adlaplace"
)

n_beta <- length(config$beta)
n_gamma <- length(config$gamma)
n_theta <- length(config$theta)
```


Model densities are defined in `src/nbinom.cpp` (observation and extra terms)
and `random.cpp` (random effects).

Each shard is built with `ad_fun_ptr()`. Shards are merged with `c()` before
attaching Hessian templates via `ad_fun()`.

```{r adShards_obs}
data_obs <- adlaplace:::ad_data(
  y = y,
  A = Amat,
  X = X,
  theta_map =
    Matrix::sparseMatrix(
      i = n_theta, # observations uses last theta
      j = 1,
      dims = c(n_theta, 1L)
    ),
  ad_kind = "observations",
  ad_fun = "nbinom_obs"
)

data_extra <- data_obs
data_extra@ad_kind <- "parameters"
data_extra@ad_fun <- "nbinom_extra"

ad_fun_obs <- adlaplace::ad_fun_ptr(data = data_obs, config = config)
ad_fun_extra <- adlaplace::ad_fun_ptr(data = data_extra, config = config)
```

```{r adShards_random_1}
nr1 <- Nrandom1

model_r1 <- adlaplace:::ad_data(
  beta_map = n_beta,
  gamma_map = Matrix::sparseMatrix(
    i = 1:nr1,
    j = 1:nr1,
    dims = c(n_gamma, nr1)
  ),
  theta_map = c(1, n_theta), # sd is first theta
  ad_kind = "random",
  ad_fun = "random_diagonal",
  precision = rep(1, nr1)
)

ad_fun_r1 <- adlaplace::ad_fun_ptr(data = model_r1, config = config)
```



```{r adShards_random_2}
nr2 <- ncol(AmatList$r2)

model_r2 <- adlaplace:::ad_data(
  beta_map = n_beta,
  gamma_map = Matrix::sparseMatrix(
    i = seq.int(nr1 + 1, length.out = nr2),
    j = 1:nr2,
    dims = c(n_gamma, nr2)
  ),
  theta_map = c(2, n_theta), # sd is 2nd theta
  ad_kind = "random",
  ad_fun = "random_diagonal",
  precision = rep(1, nr2)
)
ad_fun_r2 <- adlaplace::ad_fun_ptr(data = model_r2, config = config)
```




```{r obsLogDens}
# nbinom_obs + nbinom_extra (see src/nbinom.cpp): scan overdispersion
# theta (3rd component, log-SD scale) with beta, gamma, and other thetas fixed
# at values used to generate the data.

nb_log_dens_package <- function(log_theta) {
  # Same parameterization as the vignette (r = 1/theta^2, mu = exp(eta)); see ?dnbinom
  r <- exp(-2 * log_theta)
  sum(stats::dnbinom(y, mu = exp(eta_true), size = r, log = TRUE))
}

xx <- c(beta, gamma, config$theta)
log_theta_true <- config$theta[n_theta]

ad_nb_log_dens <- function(log_theta) {
  xx_scan <- xx
  xx_scan[length(xx)] <- log_theta
  adlaplace::joint_log_dens(ad_fun_obs, xx_scan, negative = FALSE) +
    adlaplace::joint_log_dens(ad_fun_extra, xx_scan, negative = FALSE)
}


seq_log_theta <- log_theta_true + seq(-0.5, 0.5, len = 7)

log_pkg_seq <- vapply(seq_log_theta, nb_log_dens_package, numeric(1))
log_ad_seq <- vapply(seq_log_theta, ad_nb_log_dens, numeric(1))


cbind(theta = exp(seq_log_theta), stats = log_pkg_seq, ad = log_ad_seq)
```

Combining shards inside `ad_fun(...)` uses the same move semantics as `c(ad_fun_ptr, ...)`:
the original shard handles are cleared, and those R objects then point to nothing.
```{r adShards_combined}
ad_fun_plain <- c(ad_fun_obs, ad_fun_extra, ad_fun_r1, ad_fun_r2)
```



```{r checkByShard}
shards <- seq.int(from = 0, length.out = adlaplace::n_groups(ad_fun_plain))

by_shard <- mapply(
  adlaplace::joint_log_dens,
  shards = shards,
  MoreArgs = list(x = xx, ad_fun = ad_fun_plain, negative = FALSE),
  SIMPLIFY = TRUE
)
matrix(by_shard[seq(to = length(by_shard), length.out = 12)], ncol = 4)
```

```{r testLogDens}
x <- c(config$beta, rep(0.1, length(config$gamma)), config$theta)
shards <- 0:3
adlaplace::joint_log_dens(ad_fun_plain, x, shards, negative = FALSE)
my_grad <- adlaplace::grad(ad_fun_plain, x, shards = shards, negative = FALSE)
my_grad[1:5]
my_hess <- adlaplace::hessian(ad_fun_plain, x, shards = shards, negative = FALSE)
my_hess[1:5, 1:5]
adlaplace::joint_log_dens(ad_fun_plain, x, negative = FALSE)
my_grad <- adlaplace::grad(ad_fun_plain, x, negative = FALSE)
my_grad[1:5]
my_hess <- adlaplace::hessian(ad_fun_plain, x, negative = FALSE)
my_hess[1:5, 1:5]
```

Check the derivatives of the joint log density.



Make ad function with sparsity information, evaluate derivates to test the accumulation over shards is done correctly.  



```{r testInnerOpt}
config_v <- modifyList(config, list(verbose = TRUE))
ad_pack <- adlaplace::ad_fun(
  ad_fun_plain,
  num_threads = config$num_threads
)

inner_res <- adlaplace::inner_opt(
  parameters = c(config$beta, config$theta),
  gamma = config$gamma,
  ad_fun = ad_pack,
  control = list(
    maxit = 100,
    report.level = 0,
    report.freq = 0
  ),
  deriv = TRUE,
  verbose = TRUE
)
# plot(inner_res$opt$solution, gamma)
# abline(c(0, 1), col = "red")
```




```{r testLogLik}
x_outer <- c(config$beta, config$theta)
res1 <- adlaplace::log_lik_laplace(
  x = x_outer,
  ad_fun = ad_pack,
  config = modifyList(config, list(verbose = TRUE)),
  deriv = FALSE
)
res2 <- adlaplace::log_lik_laplace(
  x = x_outer,
  ad_fun = ad_pack,
  config = modifyList(config, list(verbose = TRUE)),
  deriv = TRUE
)
x_outer
res1$full_parameters[1:5]
res1$log_lik
res2$log_lik
res1$neg_log_lik
res2$neg_log_lik
res2$opt$fval
res2$grad
plot(res2$opt$solution, gamma)
abline(c(0, 1), col = "red")
```



```{r trustOptimOuterWrappers}
# ad_fun from testInnerOpt

cache <- new.env(parent = emptyenv())
cache$gamma <- config$gamma

x0 <- c(config$beta, config$theta)

adlaplace::outer_fn(x = x0, cache = cache, config = config, ad_fun = ad_pack)
adlaplace::outer_gr(x = x0, cache = cache, config = config, ad_fun = ad_pack)

outer_fit_no_g <- stats::optim(
  par = x0,
  fn = adlaplace::outer_fn,
  method = "L-BFGS-B",
  lower = rep(-5, length(x0)),
  upper = rep(5, length(x0)),
  control = list(
    maxit = 1000,
    trace = 3,
    REPORT = 1
  ),
  config = config,
  ad_fun = ad_pack,
  cache = cache,
  control_inner = list(
    maxit = 100,
    report.level = 0,
    report.freq = 0
  )
)
outer_fit <- stats::optim(
  par = x0,
  fn = adlaplace::outer_fn,
  gr = adlaplace::outer_gr,
  method = "L-BFGS-B",
  lower = rep(-5, length(x0)),
  upper = rep(5, length(x0)),
  control = list(
    maxit = 1000,
    trace = 3,
    REPORT = 1
  ),
  config = config,
  ad_fun = ad_pack,
  cache = cache,
  control_inner = list(
    maxit = 100,
    report.level = 0,
    report.freq = 0
  )
)

rbind(
  opt = outer_fit$par,
  opt_no_g = outer_fit_no_g$par,
  true = c(beta, log(thetaOrig))
)

exp(outer_fit$par)
outer_fit$value
```


```{r testLogLgrad}
# ad_fun from testInnerOpt
par(mfrow = c(3, 2), mar = c(2, 2, 0, 0), mgp = c(1, 0.5, 0))

x1 <- outer_fit$par
Npar <- 7
Dpar <- length(x1)

parDf <- data.frame(x1)[, rep(1, Npar)]
Sx <- seq(-0.5, 0.5, len = Npar) + parDf[Dpar, 1]
SxD <- Sx[-1] - diff(Sx) / 2
parDf[Dpar, ] <- Sx


res <- mapply(
  adlaplace::log_lik_laplace,
  x = as.list(parDf),
  MoreArgs = list(
    ad_fun = ad_fun, config = config, deriv = TRUE,
    gamma = cache$gamma
  ),
  SIMPLIFY = FALSE
)
Slik <- unlist(lapply(res, "[[", "log_lik"))
SnegLik <- unlist(lapply(res, "[[", "neg_log_lik"))
Sdet <- unlist(lapply(res, function(xx) xx$extra$hessian$half_log_det))

dU <- do.call(
  abind::abind,
  c(lapply(res, function(xx) as.matrix(xx$extra$dU)), along = 3)
)
uHat <- do.call(rbind, lapply(res, function(xx) xx$opt$solution))
extraDf <- do.call(abind::abind, c(lapply(res, "[[", "deriv"), along = 3))
gradMat <- do.call(rbind, lapply(res, "[[", "grad"))



plot(Sx, SnegLik)

plot(Sx, gradMat[, Dpar], type = "l")
points(SxD, diff(SnegLik) / diff(Sx))
legend("topright",
  legend = c("autodiff", "numeric"),
  lty = c(1, NA), pch = c(NA, 1), bty = "n"
)

Du <- 2
plot(Sx, uHat[, Du])


res_here <- res[[8]]

hessian_outer <- Matrix::forceSymmetric(res_here$extra$hessian$outer)
hessian_outer2 <- adlaplace::hessian(
  ad_fun_plain, res_here$full_parameters,
  inner = FALSE, negative = TRUE
)
quantile((hessian_outer - hessian_outer2)@x, na.rm = FALSE)


to_plot <- diff(uHat[, Du]) / diff(Sx)
plot(Sx, dU[Du, Dpar, ], ylim = range(c(to_plot, dU[Du, Dpar, ]), na.rm = TRUE), type = "l")
points(SxD, to_plot)

plot(Sx, Sdet)

to_plot <- diff(Sdet) / diff(Sx)
plot(Sx, extraDf[Dpar, "d_det", ],
  ylim = range(c(to_plot, extraDf[Dpar, "d_det", ]), na.rm = TRUE),
  type = "l",
)
points(SxD, to_plot)
```


```{r derivJointDens, eval=TRUE}
par(mfrow = c(3, 2), mar = c(2, 2, 0, 0), mgp = c(1, 0.5, 0))
inner <- FALSE
type <- c("outer", "inner")[1 + inner]
ad_fun_here <- ad_fun_plain

Sgamma1 <- seq.int(length(config$beta) + 1, length.out = length(config$gamma))


x_here <- res_here$full_parameters
Npar <- 13
Dpar_dens <- length(x_here)
if (inner) {
  Dpar_dens <- match(Dpar_dens, Sgamma1)
}
Dpar0_dens <- Dpar_dens - 1L

shards <- seq.int(from = 0, length.out = adlaplace::n_groups(ad_fun_here))

parDf <- matrix(x_here, nrow = Npar, ncol = length(x_here), byrow = TRUE)
Sx <- seq(-0.1, 0.1, len = Npar) + x_here[Dpar_dens]
SxD <- Sx[-1] - diff(Sx) / 2
parDf[, Dpar_dens] <- Sx

x_list <- lapply(seq_len(Npar), function(i) parDf[i, ])

dens1 <- mapply(
  adlaplace::joint_log_dens,
  x = x_list,
  MoreArgs = list(ad_fun = ad_fun_plain, shards = shards, negative = FALSE),
  SIMPLIFY = TRUE
)

# plot(Sx, dens1)


grad1 <- mapply(
  adlaplace::grad,
  x = x_list,
  MoreArgs = list(
    ad_fun = ad_fun_plain, inner = inner, shards = shards, negative = FALSE
  ),
  SIMPLIFY = TRUE
)
plot(Sx, grad1[Dpar_dens, ], ylab = paste(Dpar_dens, "grad"), type = "l")
points(SxD, diff(dens1) / diff(Sx))
legend("topright",
  legend = c("autodiff", "numeric"),
  lty = c(1, NA), pch = c(NA, 1), bty = "n"
)


hes1 <- mapply(
  adlaplace::hessian,
  x = x_list,
  MoreArgs = list(
    ad_fun = ad_fun_plain, inner = inner, shards = shards, negative = FALSE
  ),
  SIMPLIFY = FALSE
)

hes2 <- do.call(abind::abind, c(
  lapply(hes1, as.matrix),
  along = 3
))

gradD <- apply(grad1, 1, diff) / mean(diff(Sx))



for (Dpar2 in c(Dpar_dens, 1, length(x_here) - 2, length(x_here) - 1)) {
  to_plot <- gradD[, Dpar2]
  plot(Sx, hes2[Dpar_dens, Dpar2, ],
    ylab = paste(Dpar_dens, Dpar2),
    ylim = range(c(hes2[Dpar_dens, Dpar2, ], to_plot), na.rm = TRUE),
    type = "l"
  )
  points(SxD, to_plot)
}
```


