---
title: "Laplace Approximations with adlaplace"
author: "Patrick Brown"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Laplace Approximations with adlaplace}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette is a low-level test and demonstration of the adlaplace package.


Neg Binoma, $E(y) = \mu$, $\theta$ is size, $1/\sqrt(\theta)$ is sd of Gamma random effect

$$
\begin{aligned}
f(y;\mu,\theta) =  & \log\Gamma(y + r) - \log\Gamma(r) - \log\Gamma(y + 1) 
	+ r \log(\frac{r}{r + \mu})
	+ y \log(\frac{\mu}{r + \mu}) \\
	= &  \log\Gamma(y + r) - \log\Gamma(r) - \log\Gamma(y + 1)  + r \log r - r \log(r + \mu) + y \log(\mu) - y\log(r + \mu)
\end{aligned}
$$


Generate some data.

```{r theData}
Nobs <- 1000
Nrandom1 <- 10
Nrandom2 <- 25

set.seed(0)
X <- Matrix::Matrix(cbind(1, rbinom(Nobs, 1, prob = 0.5)))
AmatList <- list(
  Matrix::sparseMatrix(
    i = 1:Nobs,
    j = sample(Nrandom1, Nobs, replace = TRUE)
  ),
  Matrix::sparseMatrix(
    i = 1:Nobs,
    j = sample(Nrandom2, Nobs, replace = TRUE)
  )
)
Amat <- do.call(cbind, AmatList)
beta <- rep(1, ncol(X))
thetaOrig <- c(0.1, 0.1, 0.1)
theta <- log(thetaOrig)
gamma <- rnorm(ncol(Amat), sd = rep(thetaOrig[1:2], c(Nrandom1, Nrandom2)))
eta <- as.vector(X %*% beta + Amat %*% gamma)
etaT <- exp(eta)
Z <- rgamma(length(etaT), 0.1^(-2), 0.1^(-2))
c(mean(Z), sd(Z))
y <- rpois(length(etaT), etaT * Z)

NperEffect <- unlist(lapply(AmatList, ncol))
# map = rep(seq(0, len=length(NperEffect)), NperEffect)
map <- Matrix::sparseMatrix(
  i = seq(0L, len = ncol(Amat)),
  j = rep(seq(0L, len = length(NperEffect)), NperEffect),
  x = 1L,
  index1 = FALSE,
  dims = c(ncol(Amat), length(thetaOrig))
)

data <- list(
  y = y,
  ATp = as(Matrix::t(Amat), "dMatrix"),
  XTp = as(Matrix::t(X), "CsparseMatrix"),
  map = map,
  Qdiag = rep(1, ncol(Amat))
)


config <- list(
  beta = rep(1, nrow(data$XTp)),
  theta = c(rep(-1, length(NperEffect)), -1),
  transform_theta = TRUE,
  gamma = rep(0, nrow(data$ATp)),
  groups = adlaplace::adFun_groups(data$ATp, Ngroups = 1000),
  num_threads = 2L,
  verbose = TRUE, package = "adlaplace"
)
```


model is defined in the file `src/objectiveFunction.cpp`


```{r testLogDens}
ad_fun_plain <- adlaplace::build_adfun(data, config)

x <- c(config$beta, rep(0.1, length(config$gamma)), config$theta)
s_groups <- 0:3
adlaplace::jointLogDens(ad_fun_plain, x, s_groups)
my_grad <- adlaplace::grad(ad_fun_plain, x, Sgroups = s_groups)
my_grad[1:5]
my_hess <- adlaplace::hessian(ad_fun_plain, x, Sgroups = s_groups)
my_hess[1:5, 1:5]
adlaplace::jointLogDens(ad_fun_plain, x)
my_grad <- adlaplace::grad(ad_fun_plain, x)
my_grad[1:5]
my_hess <- adlaplace::hessian(ad_fun_plain, x)
my_hess[1:5, 1:5]
```

Check the derivatives of the joint log density.


```{r derivJointDens, eval=TRUE}
ad_fun <- adlaplace::build_adfun(data, modifyList(config, list(verbose = TRUE)))

inner <- FALSE
type <- c("outer", "inner")[1 + inner]

Sgamma1 <- seq.int(length(config$beta) + 1, length.out = length(config$gamma))

Npar <- 13
Dpar <- 3
if (inner) {
  Dpar <- match(Dpar, Sgamma1)
}
Dpar0 <- Dpar - 1L

Sgroups <- seq.int(from = 0, length.out = ncol(config$groups))

parDf <- data.frame(x)[, rep(1, Npar)]
Sx <- seq(-0.1, 0.1, len = 13) + parDf[Dpar, 1]
SxD <- Sx[-1] - diff(Sx) / 2
parDf[Dpar, ] <- Sx


dens1 <- mapply(
  adlaplace::jointLogDens,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, Sgroups = Sgroups),
  SIMPLIFY = TRUE
)

plot(Sx, dens1)


grad1 <- mapply(
  adlaplace::grad,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, inner = inner, Sgroups = Sgroups),
  SIMPLIFY = TRUE
)
plot(Sx, grad1[Dpar, ])
lines(SxD, diff(dens1) / diff(Sx))


hes1 <- mapply(
  adlaplace::hessian,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, inner = inner, Sgroups = Sgroups),
  SIMPLIFY = FALSE
)

hes2 <- do.call(abind::abind, c(
  lapply(hes1, as.matrix),
  along = 3
))

gradD <- apply(grad1, 1, diff) / mean(diff(Sx))

Dpar2 <- 14
plot(Sx, hes2[Dpar, Dpar2, ],
  ylab = paste(Dpar, Dpar2),
  ylim = range(hes2[Dpar, Dpar2, ], na.rm = TRUE)
)
lines(SxD, gradD[, Dpar2])
```


Make ad function with sparsity information, evaluate derivates to test the accumulation over shards is done correctly.  

```{r test_all_derivs}  
ad_fun <- adlaplace::getAdFun(data, config)
x <- c(config$beta, config$gamma, config$theta)
my_derivs <- adlaplace::all_derivs(x, ad_fun, config)

c(my_derivs$fval, -adlaplace::jointLogDens(ad_fun, x))

my_grad <- adlaplace::grad(ad_fun, x)
rbind(my_grad, my_derivs$gradient)[, 1:5]

my_hessian <- adlaplace::hessian(ad_fun, x)
abind::abind(as.matrix(my_hessian), as.matrix(my_derivs$hessian), along = 3)[1:5, 1:5, ]
quantile(my_hessian@x + Matrix::forceSymmetric(my_derivs$hessian)@x)
```

```{r testInnerOpt}
inner_res <- adlaplace::inner_opt(
  parameters = c(config$beta, config$theta),
  gamma = config$gamma,
  config = config,
  ad_fun = ad_fun,
  control = list(
    maxit = 100,
    report.level = 0,
    report.freq = 0
  ),
  deriv = FALSE
)
chol_test <- Matrix::expand2(Matrix::Cholesky(inner_res$hessian, perm = TRUE, LDL = TRUE))
inner_res$chol_inner$L1[1:5, 1:5]
chol_test$L1[1:5, 1:5]
quantile(inner_res$hessian_D - chol_test$D@x)
```

```{r testLogLik}
res <- adlaplace::logLikLaplace(
  x = c(config$beta, config$theta),
  ad_fun = ad_fun,
  config = modifyList(config, list(verbose = FALSE)),
  deriv = FALSE
)
res <- adlaplace::logLikLaplace(
  x = c(config$beta, config$theta),
  ad_fun = ad_fun,
  config = modifyList(config, list(verbose = TRUE)),
  deriv = TRUE
)
res$parameters
res$logLik
res$grad
```



```{r trustOptimOuterWrappers}
ad_fun <- adlaplace::getAdFun(data, config)

cache <- new.env(parent = emptyenv())
cache$gamma <- config$gamma

x0 <- c(config$beta, config$theta)

adlaplace::outer_fn(x = x0, cache = cache, config = config, ad_fun = ad_fun)
adlaplace::outer_gr(x = x0, cache = cache, config = config, ad_fun = ad_fun)

outer_fit <- trustOptim::trust.optim(
  x = x0,
  fn = adlaplace::outer_fn,
  gr = adlaplace::outer_gr,
  method = "SR1",
  config = config,
  ad_fun = ad_fun,
  cache = cache,
  control = list(
    maxit = 1000,
    report.level = 4,
    report.freq = 1
  ),
  control_inner = list(
    maxit = 100,
    report.level = 0,
    report.freq = 0
  )
)


outer_fit$solution
outer_fit$fval
```


```{r testLogLgrad}
ad_fun <- adlaplace::getAdFun(data, config)
x <- outer_fit$solution
Npar <- 13
Dpar <- length(x)

parDf <- data.frame(x)[, rep(1, Npar)]
Sx <- seq(-0.1, 0.1, len = 13) + parDf[Dpar, 1]
SxD <- Sx[-1] - diff(Sx) / 2
parDf[Dpar, ] <- Sx

res <- mapply(
  adlaplace::logLikLaplace,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, config = config, deriv = TRUE),
  SIMPLIFY = FALSE
)
Slik <- unlist(lapply(res, "[[", "logLik"))
Sdet <- unlist(lapply(res, function(xx) xx$opt$half_log_det))

dU <- do.call(
  abind::abind,
  c(lapply(res, function(xx) as.matrix(xx$extra$dU)), along = 3)
)
uHat <- do.call(rbind, lapply(res, function(xx) xx$opt$solution))
extraDf <- do.call(abind::abind, c(lapply(res, "[[", "deriv"), along = 3))
gradMat <- do.call(rbind, lapply(res, "[[", "grad"))
gradVec <- unlist(lapply(res, function(xx) xx$deriv[Dpar, "dL"]))


plot(Sx, Slik)

plot(Sx, gradVec)
lines(SxD, diff(Slik) / diff(Sx), lwd = 2)

Du <- 1
plot(Sx, uHat[, Du])

plot(Sx, dU[Du, Dpar, ])
lines(SxD, diff(uHat[, Du]) / diff(Sx))

plot(Sx, Sdet)

plot(Sx, extraDf[Dpar, "dDet", ])
lines(SxD, diff(Sdet) / diff(Sx))
```





```{r testDeriv}
ad_fun <- adlaplace::getAdFun(data, modifyList(config, list(verbose = TRUE)))

x <- c(config$beta, rep(0.1, length(config$gamma)), config$theta)
adlaplace::jointLogDens(x, ad_fun)


str(adlaplace::grad(ad_fun, x, inner = FALSE))
str(adlaplace::grad(ad_fun, x, inner = TRUE))


Dgroup <- ncol(config$groups) + 1
str(h1 <- adlaplace::hess(x, ad_fun, FALSE, Sgroups = Dgroup))
ad_fun$sparsity[1 + Dgroup]
(Shere <- 1 + seq(
  ad_fun$hessians$map$outer$p[Dgroup + 1],
  ad_fun$hessians$map$outer$p[Dgroup + 2] - 1
))
ad_fun$hessians$map$outer$local[Shere]
ad_fun$hessians$map$outer$global[Shere]
(hseq <- which(h1@x != 0))
h1@x[hseq]


str(h2 <- adlaplace::hess(x, ad_fun, TRUE))
```

```{r trustOptimInterface, eval=FALSE}
ad_fun <- adlaplace::getAdFun(data, config)
inner_res <- adlaplace::inner_opt(
  parameters = c(config$beta, config$theta),
  gamma = config$gamma,
  control = list(
    maxit = 50,
    report.level = 0,
    report.freq = 0
  ),
  config = config,
  ad_fun = ad_fun
)

str(inner_res)

quantile(inner_res$gradient)
```

