---
title: "Bernoulli mixed model for the bacteria data"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Bernoulli mixed model for the bacteria data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(adlaplace)
source("vignette_full.R", local = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
if (!requireNamespace("MASS", quietly = TRUE)) {
  knitr::knit_exit()
}
```

The **adlaplace** formula interface is meant for ordinary mixed models:
write a response family, some covariates, and a random-effect term, then call
`adlaplace()`. This vignette fits a Bernoulli GLMM to the `bacteria` data from
**MASS** — presence of *Haemophilus influenzae* in children — with an
independent random intercept for each child.

## Data

`y` is recorded as `"y"` / `"n"`. The Bernoulli term expects 0/1, so recode
presence once and otherwise use the data as they come.

```{r data}
data(bacteria, package = "MASS")
bacteria$present <- as.integer(bacteria$y == "y")
```

There are 220 swabs on 50 children (`ID`), a three-level treatment
(`placebo`, `drug`, `drug+`), and follow-up `week` at 0, 2, 4, 6, and 11.

```{r data-plot, fig.cap="Observed colonization rate by week and treatment."}
rate <- aggregate(present ~ week + trt, data = bacteria, FUN = mean)
cols <- c(placebo = "#4C78A8", drug = "#F58518", "drug+" = "#54A24B")

plot(
  present ~ week, data = rate, type = "n",
  ylim = c(0, 1), ylab = "proportion present"
)
for (g in levels(bacteria$trt)) {
  d <- rate[rate$trt == g, ]
  lines(d$week, d$present, type = "b", pch = 19, col = cols[[g]])
}
legend("bottomleft", legend = names(cols), col = cols, pch = 19, lty = 1, bty = "n")
```

Colonization starts high and falls over the follow-up; both drug arms sit
below placebo.

## Model

A logistic mixed model with treatment, a linear week effect, and a
child-level independent random intercept is

\[
\begin{aligned}
y_{ij} &\sim \mathrm{Bernoulli}(p_{ij}), \\
\mathrm{logit}(p_{ij}) &= \beta_0 + \mathbf{x}_{ij}^\top\boldsymbol{\beta} + u_i, \\
u_i &\sim N(0, \sigma^2).
\end{aligned}
\]

That is the whole specification. The same model is one `adlaplace()` call.

```{r fit}
fit <- adlaplace(
  binomial(present) ~ trt + week + iid(ID, init = 1),
  data = bacteria
)
```

`binomial(present)` is a Bernoulli likelihood (omit `size` for a single
trial). Bare names such as `trt` and `week` become linear terms; factors
drop the first level, so the two treatment coefficients are `drug` and
`drug+` versus placebo. `iid(ID)` is an independent Gaussian intercept per
child. An intercept is added automatically.

The `init = 1` on `iid()` is a starting value for \(\sigma\) on the logit
scale. A value near 1 is a reasonable default for binary data; a start
that is much too small can sit at a near-zero local mode.

## Estimates

```{r summary}
summary(fit)
confint(fit)
```

The intercept is the logit probability at week 0 on placebo. Both treatment
contrasts are negative (less colonization on drug), and `week_linear` is
negative (colonization declines). `ID_iid` is the child-level standard
deviation \(\sigma\).

## Fitted probabilities

`fitted()` returns the linear predictor, including each child's random
intercept. Subject-specific fitted probabilities are `plogis(fitted(fit))`.

```{r fitted-plot, fig.cap="Mean fitted probability by week and treatment (lines), with observed rates as points."}
bacteria$p_hat <- plogis(fitted(fit))
fit_rate <- aggregate(p_hat ~ week + trt, data = bacteria, FUN = mean)

plot(
  p_hat ~ week, data = fit_rate, type = "n",
  ylim = c(0, 1), ylab = "probability present"
)
for (g in levels(bacteria$trt)) {
  d <- fit_rate[fit_rate$trt == g, ]
  o <- rate[rate$trt == g, ]
  lines(d$week, d$p_hat, col = cols[[g]], lwd = 2)
  points(o$week, o$present, col = cols[[g]], pch = 19)
}
legend("bottomleft", legend = names(cols), col = cols, lty = 1, lwd = 2, pch = 19, bty = "n")
```

## Child random effects

With no smooth terms, `plot(fit)` shows the Laplace modes of the 50 child
intercepts.

```{r re-modes, fig.cap="Laplace modes of the child-level random intercepts."}
plot(fit, main = "Child random intercepts")
```

`sim_iid()` draws from the same Laplace approximation and returns pointwise
intervals, still without assembling Hessians or samplers by hand.

```{r re-intervals, fig.cap="Child random intercepts, ordered, with 80% pointwise intervals."}
set.seed(1)
re <- sim_iid(fit, n = if (vignette_full) 400L else 80L)
pw <- re$ID$pointwise
ord <- order(pw[, "0.5"])
pw <- pw[ord, ]
idx <- seq_along(ord)

plot(
  idx, pw[, "0.5"],
  ylim = range(pw),
  xlab = "child (ordered by median)",
  ylab = "random intercept",
  pch = 19, cex = 0.6
)
arrows(idx, pw[, "0.1"], idx, pw[, "0.9"], length = 0, col = "grey50")
abline(h = 0, lty = 2)
```
