---
title: "Case-crossover models with adlaplace"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Case-crossover models with adlaplace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette fits a time-stratified case-crossover model to daily mortality and
PM10 in London (2002–2006), using the example dataset from
[Tobias et al. (2024)](https://github.com/aureliotobias/casecrossover).

Strata are defined by day-of-week within month and year via
`dirichlet_multinom(..., by = "year_month_dow")`. PM10 enters through an
integrated Wiener process smooth (`iwp(pm10, ...)`), with relative humidity as
a linear fixed effect. Fitting is done with `adlaplace()`.

```{r setup, include=FALSE}
library(adlaplace)
source("_vignette_full.R", local = TRUE)
fit_num_shards <- if (vignette_full) 200L else 8L
fit_maxit <- if (vignette_full) 20L else 5L
```

## Data

```{r data}
data("london", package = "adlaplace")
if (!vignette_full) {
  london <- london[london$year == 2003, ]
}
head(london)
quantile(london$pm10, na.rm = TRUE)
quantile(london$tmean, na.rm = TRUE)
```

## Model specification

Case-crossover strata absorb the intercept, so the formula uses `0 +` on the
right-hand side.

```{r model-spec}
london$temp_div10 <- london$tmean / 10
cc_formula <- dirichlet_multinom(
  all,
  by = c("year", "month", "dow"),
  init = 0.015
) ~ 0 + rh + iwp(
  pm10,
  p = 2,
  knots = seq(0, 100, by = 10),
  ref_value = 40,
  init = 0.0001
) + iwp(
  tmean,
  p = 3, knots = seq(-2, 30, by = 2),
  init = 0.002,
  ref_value = 20
)
```


## Fit

```{r fit}
fit <- adlaplace(
  formula = cc_formula,
  data = london,
  config = list(
    num_threads = 2L,
    num_shards = fit_num_shards
  ),
  control = list(
    maxit = fit_maxit
  ),
  verbose = FALSE
)
```

```{r summary}
fit$par_info
```

The same model with overdispersion optimized on the natural scale
(`log = FALSE`) should recover the same MLE and log-likelihood.

```{r fit-natural-scale, eval = vignette_full}
cc_formula_nat <- dirichlet_multinom(
  all,
  by = c("year", "month", "dow"),
  init = 0.015,
  lower = 0,
  upper = 0.5,
  parscale = 1e-3,
  log = FALSE
) ~ 0 + rh + iwp(
  pm10,
  p = 2,
  knots = seq(0, 100, by = 10),
  ref_value = 40,
  init = 0.0001
) + iwp(
  tmean,
  p = 3, knots = seq(-2, 30, by = 2),
  init = 0.002,
  ref_value = 20
)

fit_nat <- adlaplace(
  formula = cc_formula_nat,
  data = london,
  config = list(
    num_threads = 2L,
    num_shards = 100L
  ),
  control = list(
    maxit = 200, trace = 1, REPORT = 1,
    factr = 1e3,
    pgtol = 1e-8
  ),
  verbose = FALSE
)

all.equal(coef(fit), coef(fit_nat), tolerance = 1e-5)
all.equal(as.numeric(logLik(fit)), as.numeric(logLik(fit_nat)), tolerance = 1e-5)
```


```{r plot_results, eval = vignette_full}
num_sim <- 200L
num_grid <- 101L
gamma_sims <- rmvnldl(n = num_sim, fit = fit)

smooth_vars <- unique(fit$model_data$term_data$info$gamma$term)

par(mfrow = c(1, 2), mar = c(2.25, 2.25, 0, 0), bty = "l", mgp = c(1, 0.5, 0))
for (D in smooth_vars) {
  sim_here <- sim_random(
    D,
    fit = fit,
    gamma_sims = gamma_sims,
    num_grid = num_grid,
    transform = exp
  )
  q <- if (!is.null(sim_here$global)) sim_here$global else sim_here$pointwise

  matplot(
    sim_here$x, sim_here$sim,
    type = "l", lty = 1, col = "#00000010",
    xlab = D, ylab = "RR", xaxs = "i", yaxs = "i",
    ylim = quantile(sim_here$sim, prob = c(0.005, 0.95))
  )
  matlines(sim_here$x, q, lty = c(2, 1, 2), col = "red")
}
```


## AD development check

Build `model_data`, shard map, and `ad_pack` without running outer optimization,
then maximize the Laplace profile likelihood with `optim` directly.

```{r for_dev, eval = vignette_full}
md <- model_data(cc_formula, data = london, verbose = FALSE)
config_dev <- list(
  transform_theta = TRUE,
  num_threads = 2L,
  verbose = FALSE
)
config_dev$obs_groups <- obs_groups(
  A = md$term_data$A,
  elgm_matrix = md$term_data$elgm_matrix,
  num_shards = 50L,
  min_shards = min(50L, 2L * 4L)
)
par_info <- md$term_data$info$parameters
cache <- new.env(parent = emptyenv())
cache$gamma <- rep(0, nrow(md$term_data$info$gamma))
control_dev <- list(
  maxit = 200, trace = 0, REPORT = 1,
  factr = 1e3,
  pgtol = 1e-8,
  parscale = par_info$parscale
)
control_inner <- list(report.level = 0)
af <- ad_pack(md, config_dev, num_threads = 2L)
slotNames(af)
dim(config_dev$obs_groups)
par_info
bob <- stats::optim(
  par = par_info$init,
  fn = outer_fn,
  gr = outer_gr,
  method = "L-BFGS-B",
  control = control_dev,
  lower = par_info$lower,
  upper = par_info$upper,
  config = config_dev,
  ad_pack = af,
  cache = cache,
  control_inner = control_inner,
  hessian = TRUE
)
```

## Joint log density

At starting values, evaluate the full log-density and its gradient over all
parameters (`beta`, inner `gamma`, and `theta`).

```{r joint-log-dens, eval = vignette_full}
sz <- af@sizes
gamma0 <- rep(0, nrow(md$term_data$info$gamma))
x_full <- c(
  par_info$init[seq_len(sz["beta"])],
  gamma0,
  par_info$init[seq(sz["beta"] + 1L, length.out = sz["theta"])]
)

# Serial dens/grad/hess on a dens-safe clone (multi-thread affinity cleared).
dens_ptr <- clone_ad_pack_ptr(af@ptr)
shards <- seq.int(from = 0L, length.out = adlaplace:::n_groups(dens_ptr))
by_shard <- vapply(
  shards,
  function(s) joint_log_dens(dens_ptr, x_full, ad_shards = s),
  numeric(1)
)
by_shard[1:5]
by_shard[seq(to = length(by_shard), length.out = 5)]

sum(by_shard)
joint_log_dens(dens_ptr, x_full)
grad(dens_ptr, x_full)
bob <- hessian(dens_ptr, x_full)
```

## Likelihood

With outer parameters fixed at their starting values, `log_lik_laplace()`
optimizes random effects and returns the log-likelihood.

```{r likelihood, eval = vignette_full}
x_outer <- par_info$init

lik <- log_lik_laplace(
  x = x_outer,
  ad_pack = af,
  config = modifyList(config_dev, list(verbose = TRUE)),
  gamma = gamma0,
  control = list(report.level = 2, report.freq = 1),
  deriv = TRUE
)

lik$neg_log_lik
lik$deriv$d_neg_log_lik
head(lik$inner_opt$solution)
```

`neg_log_lik` is the negative profile log-likelihood after inner optimization
over random effects.
