---
title: "Strata-removal AMIP for case-crossover"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Strata-removal AMIP for case-crossover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

This vignette applies the approximate maximum influence perturbation (AMIP)
of Broderick, Giordano, and Meager to the London case-crossover model. After
fitting, outer parameters (`beta`, `theta`) are held at the MLE and only the
inner random effects `gamma` are allowed to move. The fit-time observation
tapes supply the inner Hessian \(H\); per-stratum scores use streaming
one-unit observation gradients (`grad_obs_units`) so scores do not require
holding an identity shard pack. The most influential strata for flipping the
PM10 effect at exposure 80 are identified, then a second inner optimization
with those strata removed confirms the linear approximation. A full refit on
the reduced data is compared on the PM10 relative-risk curve.

```{r setup, include=FALSE}
library(adlaplace)
library(Matrix)
if (file.exists("vignette_full.R")) {
  source("vignette_full.R", local = TRUE)
} else {
  vignette_full <- TRUE
  vignette_num_threads <- 2L
}
fit_num_shards <- if (vignette_full) 200L else 8L
fit_maxit <- if (vignette_full) 200L else 15L
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Data and model

Same London daily mortality / PM10 specification as the case-crossover vignette.
Abbreviated knits use 2003 only.

```{r data}
data("london", package = "adlaplace")
if (!vignette_full) {
  london <- london[london$year == 2003, ]
}

london$month <- factor(format(london$date, "%b"), levels = month.abb)
london$dow <- factor(format(london$date, "%a"),
  levels = format(as.Date(4:10), "%a")
)

cc_formula <- dirichlet_multinom(
  all,
  by = c("year", "month", "dow"),
  init = 0.02
) ~ 0 + rh + iwp(
  pm10,
  p = 2,
  knots = seq(0, 100, by = 10),
  ref_value = 40,
  init = 0.001
) + iwp(
  tmean,
  p = 3,
  knots = seq(-2, 30, by = 2),
  init = 0.001,
  ref_value = 20
)
```

## Fit (outer MLE)

Outer Hessian is not needed for this inner-only AMIP calculation.

```{r fit}
fit <- adlaplace(
  formula = cc_formula,
  data = london,
  config = list(
    num_threads = vignette_num_threads,
    num_shards = fit_num_shards
  ),
  control = list(
    maxit = fit_maxit,
    trace = 3,
    REPORT = 5
  ),
  verbose = FALSE,
  hessian = FALSE
)
md <- fit$model_data
x_outer <- as.numeric(fit$par_info$mle_internal)
gamma0 <- as.numeric(fit$gamma)
fit$par_info[, c("label", "mle", "mle_internal")]
```



## Inner mode and Hessian at the outer MLE

Freeze `beta` and `theta` at the MLE; re-optimize `gamma` on the **fit** AD
pack (coarse observation shards) and keep the inner Hessian \(H\). Per-stratum
scores below do not need an identity retape.

```{r inner-mode}
n_strata <- ncol(md$term_data$elgm_matrix)
af <- fit$ad_pack
sz <- af@sizes
io <- inner_opt(
  parameters = x_outer,
  gamma = gamma0,
  ad_pack = af,
  control = list(report.level = 0, report.freq = 0),
  deriv = TRUE,
  return_hessians = TRUE,
  verbose = FALSE
)
gamma_hat <- as.numeric(io$inner_opt$solution)
H <- io$hessian$inner
n_strata
as.integer(sz)
```

## Quantity of interest: PM10 effect at 80

The PM10 smooth is centered at `ref_value = 40`, so the effect at 80 is a
nontrivial linear functional \(\phi(\gamma) = a^\top\gamma\) of the IWP and
random-polynomial coefficients.

```{r phi-design}
pm_terms <- Filter(
  function(tt) {
    identical(tt@name, "pm10") &&
      identical(as.character(tt@model_role), "random")
  },
  md$terms
)
new_x <- data.frame(pm10 = 80)
design_list <- lapply(pm_terms, design, data = new_x)
design_list <- Filter(Negate(is.null), design_list)
new_design <- as.matrix(do.call(cbind, design_list))

gamma_lab <- md$term_data$info$gamma$gamma_label
a <- numeric(length(gamma_lab))
names(a) <- gamma_lab
a[colnames(new_design)] <- as.numeric(new_design[1L, ])
stopifnot(any(abs(a) > 0))

phi_hat <- sum(a * gamma_hat)
phi_hat
```

## Per-stratum influence scores

Solve \(H v = a\) once, then form stratum scores
\(\psi_s = -v^\top g_s\) from streaming one-unit observation gradients
(`grad_obs_units`; same negative-log sign as `grad` / `inner_opt`). Temporary
tapes use `hessian_sparsity = FALSE` and are discarded after each batch.

```{r influence-scores}
v <- as.numeric(Matrix::solve(H, a))
x_full <- as.numeric(io$full_parameters)
G <- grad_obs_units(
  md$observations[[1]],
  x_full,
  batch_size = 32L,
  inner = TRUE,
  negative = TRUE
)
stopifnot(nrow(G) == length(v), ncol(G) == n_strata)
psi <- as.numeric(-Matrix::crossprod(v, G))
summary(psi)
```

## Approximate most influential set

Leaving out stratum \(s\) changes \(\phi\) by approximately \(-\psi_s\). To flip
the sign of \(\hat\phi > 0\) we need a negative change, so we drop the strata
with the **largest positive** \(\psi_s\) until \(\sum \psi_s \ge \hat\phi\)
(equivalently Broderick et al. applied to \(-\phi\)). If \(\hat\phi < 0\), drop
the most negative scores instead. If that is impossible under the linear
approximation, report that no small removal set is found.

```{r amip-set}
# Dropping S changes phi by approx -sum(psi[S]). Choose S so that sum(psi[S])
# has the same sign as phi_hat and |sum| >= |phi_hat|.
if (phi_hat > 0) {
  ord <- order(psi, decreasing = TRUE)
} else if (phi_hat < 0) {
  ord <- order(psi)
} else {
  ord <- integer(0)
}
cum_psi <- if (length(ord)) cumsum(psi[ord]) else numeric(0)
m_star <- if (phi_hat == 0 || !length(cum_psi)) {
  NA_integer_
} else {
  which(cum_psi / phi_hat >= 1)[1L]
}
if (is.na(m_star)) {
  drop_idx <- integer(0)
  amip_alpha <- NA_real_
  amip_change <- if (length(cum_psi)) -cum_psi[length(cum_psi)] else 0
} else {
  drop_idx <- ord[seq_len(m_star)]
  amip_alpha <- length(drop_idx) / n_strata
  amip_change <- -cum_psi[m_star]
}

c(
  phi_hat = phi_hat,
  n_drop = length(drop_idx),
  alpha = amip_alpha,
  approx_change = amip_change
)

dropped <- data.frame(
  idx = drop_idx,
  strata = colnames(md$term_data$elgm_matrix)[drop_idx],
  psi = psi[drop_idx]
)

dropped
```

```{r psi-plot, fig.width=6, fig.height=3.5}
par(mar = c(2.5, 2.5, 1, 0.5), bty = "l", mgp = c(1.5, 0.5, 0))

parts <- strsplit(colnames(md$term_data$elgm_matrix), ".", fixed = TRUE)
stratum_info <- data.frame(
  year = as.integer(vapply(parts, `[[`, character(1L), 1L)),
  month = vapply(parts, `[[`, character(1L), 2L),
  dow = vapply(parts, `[[`, character(1L), 3L),
  stringsAsFactors = FALSE
)
stratum_info$month <- factor(stratum_info$month, levels = month.abb)
stratum_info$dow <- factor(stratum_info$dow,
  levels = format(as.Date(4:10), "%a")
)
stratum_info$name <- paste(stratum_info$month, stratum_info$dow, sep = "\n")
stratum_info$num <- stratum_info$year +
  (as.integer(stratum_info$month) - 1) / 12 +
  0.1 * (as.integer(stratum_info$dow) - 1) / 7

Scol <- c("#000000", RColorBrewer::brewer.pal(11, "RdYlGn"))
stratum_info$pch <- as.character(as.integer(stratum_info$dow))
stratum_info$col <- Scol[as.integer(stratum_info$month)]
stratum_info$pch[drop_idx] <- "o"
stratum_info$col[drop_idx] <- "firebrick"

plot(
  stratum_info$num, psi,
  pch = stratum_info$pch,
  cex = 0.55, col = stratum_info$col,
  xlab = "stratum", ylab = expression(psi[s])
)

text(
  stratum_info$num[drop_idx], psi[drop_idx],
  stratum_info$name[drop_idx],
  cex = 0.5
)

legend("bottomright", legend = levels(stratum_info$month), fill = Scol, bty = "n", inset = 0)

abline(h = 0, lty = 2, col = "grey50")
```

## Confirm by dropping nominated strata

Rebuild observation tapes without the Approximate Most Influential Set, keep
the outer MLE fixed, and re-optimize `gamma`. The resulting change in
\(\phi(80)\) is an exact finite-sample lower bound on the worst-case
strata-removal effect (Broderick et al.).

```{r confirm, fig.width=6.5, fig.height=3.5}
phi_from_gamma <- function(gamma) {
  sum(a * as.numeric(gamma))
}

pm_curve <- function(gamma, grid = seq(0, 100, by = 2)) {
  new_grid <- data.frame(pm10 = grid)
  design_list <- lapply(pm_terms, design, data = new_grid)
  design_list <- Filter(Negate(is.null), design_list)
  D <- as.matrix(do.call(cbind, design_list))
  g <- as.numeric(gamma)
  names(g) <- gamma_lab
  list(x = grid, f = as.numeric(D %*% g[colnames(D)]))
}

curve0 <- pm_curve(gamma_hat)

if (!length(drop_idx)) {
  phi_drop <- NA_real_
  curve1 <- NULL
  message("No strata dropped; skipping confirmation retape.")
} else {
  keep <- setdiff(seq.int(0L, n_strata - 1L), as.integer(drop_idx - 1L))
  og_drop <- Matrix::sparseMatrix(
    i = keep,
    j = seq_along(keep) - 1L,
    x = 1,
    dims = c(n_strata, length(keep)),
    index1 = FALSE
  )
  af_drop <- ad_pack(
    md,
    config = modifyList(fit$config, list(obs_groups = og_drop, verbose = FALSE)),
    num_threads = vignette_num_threads,
    reorder_shards = "none"
  )
  io_drop <- inner_opt(
    parameters = x_outer,
    gamma = gamma_hat,
    ad_pack = af_drop,
    control = list(report.level = 0, report.freq = 0),
    deriv = FALSE,
    return_hessians = FALSE,
    verbose = FALSE
  )
  gamma_drop <- as.numeric(io_drop$inner_opt$solution)
  phi_drop <- phi_from_gamma(gamma_drop)
  curve1 <- pm_curve(gamma_drop)
}

c(
  phi_full = phi_hat,
  phi_dropped = phi_drop,
  actual_change = if (is.finite(phi_drop)) phi_drop - phi_hat else NA_real_,
  approx_change = if (length(drop_idx)) amip_change else NA_real_
)
```

## Refit without AMIP strata

Drop the nominated strata from the data and re-estimate the full outer MLE
(smoothness and overdispersion included).

```{r refit-drop}
elgm <- md$term_data$elgm_matrix
if (!length(drop_idx)) {
  fit_drop <- NULL
  phi_refit <- NA_real_
  message("No strata dropped; skipping refit.")
} else {
  in_drop <- Matrix::rowSums(elgm[, drop_idx, drop = FALSE]) > 0
  london_drop <- md$term_data$data[!in_drop, , drop = FALSE]
  c(
    n_full = nrow(md$term_data$data),
    n_drop = nrow(london_drop),
    n_strata_drop = length(drop_idx)
  )
  fit_drop <- adlaplace(
    formula = cc_formula,
    data = london_drop,
    config = list(
      num_threads = vignette_num_threads,
      num_shards = fit_num_shards
    ),
    control = list(
      maxit = fit_maxit,
      trace = 3,
      REPORT = 5
    ),
    verbose = FALSE,
    hessian = FALSE
  )
  phi_refit <- sum(a * fit_drop$gamma[names(a)])
  fit_drop$par_info[, c("label", "mle", "mle_internal")]
}

c(
  phi_full = phi_hat,
  phi_inner_drop = phi_drop,
  phi_refit = phi_refit
)
```

```{r plot_results, fig.width=6.5, fig.height=3.5}
num_sim <- 200L
num_grid <- 101L

sim_full <- sim_random(
  "pm10",
  fit = fit,
  num_grid = num_grid,
  num_sim = num_sim,
  transform = exp
)
q_full <- if (!is.null(sim_full$global)) sim_full$global else sim_full$pointwise

sim_drop <- NULL
q_drop <- NULL
if (!is.null(fit_drop)) {
  sim_drop <- sim_random(
    "pm10",
    fit = fit_drop,
    new_x = data.frame(pm10 = sim_full$x),
    num_sim = num_sim,
    transform = exp
  )
  q_drop <- if (!is.null(sim_drop$global)) sim_drop$global else sim_drop$pointwise
}


par(mar = c(3, 3, 0, 0), mgp = c(1.5, 0.5, 0), bty = "l")
matplot(
  sim_drop$x, q_drop,
  type = "l",
  lty = c(3, 1, 3), col = "darkblue", lwd = 2,
  xlab = "pm10", ylab = "RR", xaxs = "i",
  ylim = c(0.975, 1.065), yaxs = "i", bty = "l"
)

matlines(sim_full$x, q_full, lty = c(3, 1, 3), col = "red", lwd = 2)
abline(h = 1, lty = 3, col = "grey50")
legend(
  "topleft",
  legend = c("full data", "AMIP strata removed"),
  col = c("red", "darkblue"),
  lwd = 2,
  bty = "n"
)
```

## Notes

- Units of removal are **strata** (year–month–day-of-week matched sets), not
  individual calendar days.
- The inner confirmation keeps outer parameters at the original MLE: that is
  sensitivity of \(\hat\gamma(\beta,\theta)\), not a refit of smoothness or
  overdispersion. The refit chunk re-estimates the full outer MLE on the
  reduced data.
- Scores use streaming one-unit observation tapes (`grad_obs_units`); \(H\)
  comes from the fit pack via `inner_opt`. Confirmation still retapes a
  masked likelihood after dropping the nominated strata.
