---
title: "Case-crossover models with hpolcc"
format:
  html:
    toc: true
---

## Introduction

This vignette fits a time-stratified case-crossover model to daily mortality and
PM10 in London (2002–2006), using the example dataset from
[Tobias et al. (2024)](https://github.com/aureliotobias/casecrossover).

Strata are defined by day-of-week within month and year via
`dirichlet_multinom(..., by = "year_month_dow")`. PM10 enters through an
integrated Wiener process smooth (`iwp(pm10, ...)`), with relative humidity as
a linear fixed effect.

```{r setup}
# adlaplace must load before hpolcc/data.table so CppAD OpenMP setup runs first
library(adlaplace)
library(hpolcc)
library(data.table)
```

## Data

```{r data}
london <- fread(system.file("extdata", "london.csv", package = "hpolcc"))
london$date <- as.Date(ISOdate(london$year, london$month, london$day))
london$year_month_dow <- format(london$date, "%Y-%m-%a")
head(london)
quantile(london$pm10, na.rm = TRUE)
quantile(london$tmean, na.rm = TRUE)

```

## Model specification

```{r model-spec}
london$temp_div10 = london$tmean / 10
cc_formula <- dirichlet_multinom(
    all,
    by = "year_month_dow",
    init = 0.2,
    parscale = 5
) ~ rh + iwp(
    pm10,
    p = 2,
    knots = seq(0, 100, by = 10),
    ref_value = 40
) + iwp(
    temp_div10, 
p = 3, knots = seq(-2, 30, by = 2)/10, 
init = 0.01, parscale = 100,
ref_value = 20/10)
```


## Fit

```{r fit}
fit <- hnlm(
    formula = cc_formula,
    data = london,
    config = list(
    transform_theta = TRUE,
    num_threads = 8L,
    num_shards = 100L,
    verbose = FALSE
    ),
    control = list(maxit = 100, trace = 3, REPORT = 1)
)
```

```{r summary}
fit
summary(fit)
fit$coefficients$beta
fit$log_lik
stats::coef(fit)
fit$sample$sim[[1]][1:5, 1:3]
```


```{r plot_results}
matplot(fit$sample$x$pm10, fit$sample$sim$pm10, type='l', lty=1, col='#00000010',
    xlab = "PM10", ylab = "RR", xaxs='i', yaxs='i',
    ylim = quantile(fit$sample$sim$pm10, prob = c(0.005, 0.995)))
matlines(fit$sample$x$pm10, fit$sample$envelope$common$pm10, lty=c(2,1,2), col='red')

```


## AD development check

`for_dev = TRUE` builds `model_data`, the combined `ad_fun`, and `config` without
running outer optimization.

```{r for_dev}
config_dev <- list(
    transform_theta = TRUE,
    num_threads = 8L,
    num_shards = 100L,
    verbose = FALSE
)
forres <- hnlm(
    cc_formula,
    data = london,
    config = config_dev,
    for_dev = TRUE
)
slotNames(forres$ad_fun)
dim(forres$config$shards)
forres$model_data$data$info$parameters
```

## Joint log density

At starting values, evaluate the full log-density and its gradient over all
parameters (`beta`, inner `gamma`, and `theta`).

```{r joint-log-dens}
sz <- forres$ad_fun@sizes
x_full <- c(
    forres$config$opt$init[seq_len(sz["beta"])],
    rep(0, sz["gamma"]),
    forres$config$opt$init[seq(sz["beta"] + 1L, length.out = sz["theta"])]
)

forres$model_data$data$info$parameters[, c("label", "init")]
c(beta = sz["beta"], gamma = sz["gamma"], theta = sz["theta"])

joint_log_dens(forres$ad_fun, x_full)
grad(forres$ad_fun, x_full)
```

## Likelihood

With outer parameters fixed at their starting values, `log_lik_laplace()`
optimizes random effects and returns the  log-likelihood.

```{r likelihood}
x_outer <- forres$model_data$data$info$parameters$init

lik = log_lik_laplace(
    x = x_outer,
    ad_fun = forres$ad_fun,
    config = modifyList(forres$config, list(verbose=TRUE)),
    gamma = forres$cache$gamma,
    control = list(report.level = 2, report.freq = 1),
    deriv = FALSE
)

bob = Matrix::Cholesky(lik$extra$hessian$inner, perm = TRUE, LDL = TRUE)
bob_ex <- Matrix::expand2(bob)

ci <- lik$extra$hessian$chol_inner

quantile(as.numeric(bob_ex$D@x) - ci$D)

table(bob@perm == ci$perm)

diff_L1 <- bob_ex$L1 - ci$L1
c(L1 = max(abs(diff_L1)), D = max(abs(diff_D)), perm_match = all(bob@perm == ci$perm))
which(abs(diff_D) > 0)
which(abs(diff_L1) > 0, arr.ind = TRUE)

H_perm <- ci$L1 %*% Matrix::Diagonal(length(ci$D), ci$D) %*% Matrix::t(ci$L1)
reconstructH <- H_perm[order(ci$perm), order(ci$perm)]
max(abs(reconstructH - lik$extra$hessian$inner))

lik$neg_log_lik
lik$grad
head(lik$opt$solution)
```

`neg_log_lik` is the negative profile log-likelihood after inner optimization
over random effects.

Conditional simulations of IWP smooths use `adlaplaceHgp::cond_sim_iwp()` with
the Laplace output and model-data bundle:

```{r cond-sim}
sim <- adlaplaceHgp::cond_sim_iwp(
  laplace = lik,
  model_data = forres$model_data,
  n = 100
)
names(sim)
# inspect assembled inputs:
# str(adlaplaceHgp::cond_sim_iwp_inputs(lik, forres$model_data))
```


