---
title: "Checking the hiwp Log Density"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Checking the hiwp Log Density}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette simulates data from a small `hiwp()` model, evaluates the joint
log density shard by shard, and checks every `"random"` shard against a manual
evaluation of the same log density in R.

```{r setup, include=FALSE}
library("Matrix")
library("adlaplace")
library("adlaplaceHgp")
knitr::opts_chunk$set(collapse = TRUE)
```

# Simulated data

The model mirrors a typical call, with a coarser knot sequence and a handful of
`by` groups so that everything can be printed.

```{r term}
knots <- c(0, 2, 4, 8, 14, 22)
ref_value <- 4
states <- c("aa", "bb", "cc")

hiwp_terms <- adlaplaceHgp::hiwp("sqrt_pm",
  by = "state",
  ref_value = ref_value, p = 2,
  knots = knots,
  init = 0.5
)
names(hiwp_terms)
```

`hiwp()` returns four terms: the group-specific IWP deviations (`hiwp`), a
global IWP (`iwp`), group-specific random slopes (`hrpoly_1`), and a boundary
polynomial (`rpoly`).

```{r data}
set.seed(2024)
n_per_state <- 60
dat <- data.frame(
  sqrt_pm = rep(seq(min(knots), max(knots), length.out = n_per_state),
    times = length(states)),
  state = rep(states, each = n_per_state)
)
```

To simulate we need the design matrix `A` and the prior precisions, so build
the model data first and then draw the random effects from their own priors.

```{r model_data}
model_terms <- c(
  list(y_gaussian = adlaplace::gaussian("y", init = 0.2)),
  hiwp_terms
)
dat$y <- 0
md <- adlaplace::model_data(model_terms, data = dat)

info <- md$term_data$info
n_beta <- nrow(info$beta)
n_gamma <- nrow(info$gamma)
n_theta <- nrow(info$theta)
c(n_beta = n_beta, n_gamma = n_gamma, n_theta = n_theta)
```

```{r theta_info}
info$theta[, c("term", "model", "label", "init", "log")]
```

Each random effect is indexed by its `gamma_label` and carries the `theta_id`
of the standard deviation that scales it.

```{r gamma_info}
info$gamma[, c("label", "gamma_label", "theta_id")]
```

The `hiwp` term contributes a random-effects shard alongside `iwp` and
`hrpoly` (the boundary `rpoly` has a flat prior and is omitted).

```{r random_shards}
names(md$random)
```

```{r prototype_slots}
slots_for <- function(term) {
  c(class = class(term), density = term@density, ad_kind = term@ad_kind,
    model_role = as.character(term@model_role))
}
as.data.frame(t(vapply(md$terms, slots_for, character(4))))
```

# True parameter values

For a `random_diagonal` prior, coefficient `k` of a term with standard
deviation `sd` has precision `Q[k] / sd^2`, so we draw it with standard
deviation `sd / sqrt(Q[k])`.

```{r simulate}
sd_true <- c(
  sqrt_pm_hiwp = 0.30,
  sqrt_pm_iwp = 0.80,
  sqrt_pm_hrpoly_1 = 0.15,
  y_gaussian_sd = 0.20
)

# diagonal prior precision weights, per term, aligned to A's columns
prec_weights <- rep(NA_real_, n_gamma)
names(prec_weights) <- info$gamma$gamma_label
for (term_here in md$terms) {
  q_here <- adlaplace::precision(term_here, md$term_data$data)
  if (is.null(q_here)) next
  prec_weights[colnames(q_here)] <- Matrix::diag(q_here)
}

gamma_true <- rep(0, n_gamma)
has_prior <- !is.na(prec_weights)
gamma_sd <- sd_true[info$gamma$label[has_prior]] / sqrt(prec_weights[has_prior])
gamma_true[has_prior] <- stats::rnorm(sum(has_prior), sd = gamma_sd)

# the boundary polynomial has a flat prior (sd = Inf); give it a fixed slope
gamma_true[!has_prior] <- 0.1

dat$y <- as.numeric(md$term_data$A %*% gamma_true) +
  stats::rnorm(nrow(dat), sd = sd_true[["y_gaussian_sd"]])
```

Rebuild the model data with the simulated response, and assemble the parameter
vector `x = c(beta, gamma, theta)`. With `transform_theta = TRUE` the theta
block holds log standard deviations.

```{r x_vector}
md <- adlaplace::model_data(model_terms, data = dat)
info <- md$term_data$info

theta_true <- log(sd_true[info$theta$label])
x <- c(gamma_true, theta_true)
length(x)
```

# Building the AD pack

```{r ad_pack}
config <- list(
  transform_theta = TRUE,
  obs_groups = adlaplace::obs_groups(md$term_data$A, num_shards = 3L),
  verbose = FALSE
)
ap <- adlaplace::ad_pack(md, config, num_threads = 1L)
dens_ptr <- adlaplace::clone_ad_pack_ptr(ap@ptr)
n_shard <- adlaplace:::n_groups(dens_ptr)
n_shard
```

`ad_pack()` tapes the shards in the order `observations`, `random`,
`parameters`, and each observation shard expands into one AD group per column
of `config$obs_groups`. That is enough to reconstruct the shard index to name
map.

```{r shard_map}
shard_index_map <- function(model_data, config) {
  shard_list <- c(model_data$observations, model_data$random,
    model_data$parameters)
  ad_kind <- vapply(shard_list, function(s) s@ad_kind, character(1))
  n_groups_each <- ifelse(ad_kind == "observations",
    ncol(config$obs_groups), 1L)
  out <- data.frame(
    name = rep(names(shard_list), times = n_groups_each),
    ad_kind = rep(ad_kind, times = n_groups_each),
    density = rep(vapply(shard_list, function(s) s@density, character(1)),
      times = n_groups_each),
    row.names = NULL
  )
  out$shard <- seq_len(nrow(out)) - 1L
  out[, c("shard", "name", "ad_kind", "density")]
}

shard_map <- shard_index_map(md, config)
stopifnot(nrow(shard_map) == n_shard)
shard_map
```

# Log density per shard

```{r per_shard}
shard_map$log_dens <- vapply(shard_map$shard, function(s) {
  adlaplace::joint_log_dens(dens_ptr, x, ad_shards = s, negative = FALSE)
}, numeric(1))
shard_map

all.equal(
  sum(shard_map$log_dens),
  adlaplace::joint_log_dens(dens_ptr, x, negative = FALSE)
)
```

# Manual evaluation of the random shards

Every random shard here uses the `random_diagonal` density. With `tau =
exp(-2 * log_sd)` and diagonal prior weights `Q`, the C++ code evaluates

$$
\log p(\gamma \mid \sigma) =
  \tfrac{1}{2}\sum_k \log Q_k
  - \tfrac{\tau}{2}\sum_k Q_k \gamma_k^2
  - n \log \sigma
  - \tfrac{n}{2}\log(2\pi).
$$

The shard's own `gamma_map` and `theta_map` say which global entries of `x` it
reads, so the manual version can be written without reference to term names.

```{r manual_fun}
manual_random_diagonal <- function(shard, x, n_beta, n_gamma) {
  gm <- methods::as(shard@gamma_map, "TsparseMatrix")
  gamma_rows <- gm@i[order(gm@j)] + 1L

  tm <- methods::as(shard@theta_map, "TsparseMatrix")
  theta_rows <- tm@i[order(tm@j)] + 1L
  stopifnot(length(theta_rows) == 1L)

  q_weights <- as.numeric(shard@precision)
  log_sd <- x[n_beta + n_gamma + theta_rows]
  tau <- exp(-2 * log_sd)
  gamma_here <- x[n_beta + gamma_rows]
  n_here <- length(q_weights)

  0.5 * sum(log(q_weights)) -
    0.5 * tau * sum(q_weights * gamma_here^2) -
    n_here * log_sd -
    n_here * 0.5 * log(2 * pi)
}
```

```{r manual_compare}
random_rows <- shard_map[shard_map$ad_kind == "random", ]
comparison <- do.call(rbind, lapply(seq_len(nrow(random_rows)), function(i) {
  nm <- random_rows$name[i]
  shard <- md$random[[nm]]
  data.frame(
    name = nm,
    n_gamma = ncol(shard@gamma_map),
    adlaplace = random_rows$log_dens[i],
    manual = manual_random_diagonal(shard, x, n_beta, n_gamma)
  )
}))
comparison$difference <- comparison$adlaplace - comparison$manual
comparison

stopifnot(max(abs(comparison$difference)) < 1e-12)
stopifnot("sqrt_pm_hiwp" %in% comparison$name)
```

The C++ densities agree with the manual formula to machine precision for every
random shard, including `hiwp`.

The `hiwp` log standard deviation is not a flat direction of the joint density:
shifting it changes the value.

```{r theta_not_flat}
theta_hiwp <- which(info$theta$label == "sqrt_pm_hiwp")
x_shifted <- x
x_shifted[n_beta + n_gamma + theta_hiwp] <-
  x_shifted[n_beta + n_gamma + theta_hiwp] + 3

c(
  base = adlaplace::joint_log_dens(dens_ptr, x, negative = FALSE),
  theta_shifted = adlaplace::joint_log_dens(dens_ptr, x_shifted, negative = FALSE)
)
```

# Label ordering and `random_info()`

`design()`, `precision()`, and `random_info()` agree on the column / label
order (basis fastest within each `by` group). `random_info()` also calls
`add_by_levels()` so it works on a freshly constructed term.

```{r ordering}
term_here <- hiwp_terms$sqrt_pm_hiwp
a_cols <- colnames(adlaplace::design(term_here, dat))
q_cols <- colnames(adlaplace::precision(term_here, dat))
r_labels <- adlaplace::random_info(term_here, dat)$gamma_label

data.frame(design = a_cols, precision = q_cols, random_info = r_labels)

c(
  design_vs_precision = identical(a_cols, q_cols),
  design_vs_random_info = identical(a_cols, r_labels)
)
```
