---
documentclass: jss
author:
  - name: Patrick Brown
    affiliation: |
      | University of Toronto
    # TODO: confirm address block before submission
    address: |
      | Department of Statistical Sciences
      | University of Toronto
    email: \email{patrick.brown@utoronto.ca}
title:
  formatted: "\\pkg{adlaplace}: Extensible Laplace Approximations for Hierarchical Models with Automatic Differentiation in \\proglang{R}"
  plain: "adlaplace: Extensible Laplace Approximations for Hierarchical Models with Automatic Differentiation in R"
  short: "\\pkg{adlaplace}: Extensible Laplace Approximations"
abstract: >
  The \pkg{adlaplace} package fits hierarchical models in \proglang{R} by
  maximizing a Laplace approximation to the marginal likelihood, with all
  derivatives computed by automatic differentiation through \pkg{CppAD}.
  The package occupies similar ground to \pkg{TMB} but differs in three
  respects.  First, the gradient of the Laplace-approximate log likelihood
  includes the term arising from the dependence of the log-determinant on
  the mode of the random effects, computed with trace contractions that
  avoid forming any third-order derivative tensor explicitly.  Second, the
  joint log density is decomposed into independent shards which are taped
  and evaluated in parallel with OpenMP, giving parallelism for the density,
  gradient, Hessian, and the determinant-derivative contractions alike.
  Third, model components are organized in two extensible layers: high-level
  \proglang{R} model terms (random intercepts, penalized splines, Gaussian
  Markov random fields) that map onto a small set of low-level C++ density
  kernels, and both layers can be extended by third-party packages without
  modifying \pkg{adlaplace} itself.  We describe the methodology and
  architecture, demonstrate the user-facing interface, and validate the
  package against \pkg{mgcv}, \pkg{TMB}, and \pkg{INLA} on three examples:
  a negative binomial additive mixed model, a logistic mixed model for the
  \code{bacteria} data with the built-in \code{binomial()} observation
  density, and the Besag-York-Mollie disease-mapping model for Germany using a
  non-diagonal intrinsic precision matrix.  A fourth illustration uses the
  companion package \pkg{adlaplaceGrf} to assemble a finite-element
  Mat\'{e}rn Gaussian random field whose hyperparameter-dependent precision
  determinant is evaluated on the AD tape.
keywords:
  formatted: [Laplace approximation, automatic differentiation, hierarchical models, "\\proglang{R}", parallel computing, "\\pkg{CppAD}"]
  plain: [Laplace approximation, automatic differentiation, hierarchical models, R, parallel computing, CppAD]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{booktabs}
output: rticles::jss_article
bibliography: adlaplace.bib
---

```{r setup, include=FALSE}
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  cache = TRUE, cache.path = "cache/", fig.path = "figures/"
)
library("Matrix")
library("adlaplace")
have_inla <- requireNamespace("INLA", quietly = TRUE)
have_tmb <- requireNamespace("TMB", quietly = TRUE)
have_mgcv <- requireNamespace("mgcv", quietly = TRUE)
have_grf <- requireNamespace("adlaplaceGrf", quietly = TRUE)
```

# Introduction {#sec:intro}

Hierarchical models with latent Gaussian random effects are a workhorse of
applied statistics: generalized linear mixed models, penalized regression
splines, and spatial models based on Gaussian Markov random fields all take
the form
\begin{equation}
Y \mid U \sim \pi(y \mid U; \theta), \qquad
U \sim \mathrm{N}\{0, Q(\theta)^{-1}\},
\label{eq:model}
\end{equation}
with observations $Y$, latent effects $U$, and a low-dimensional parameter
vector $\theta$.  Maximum likelihood estimation of $\theta$ requires the
marginal likelihood $\pi(y; \theta) = \int \pi(y \mid u; \theta)\,
\pi(u; \theta)\, du$, an integral that is analytically intractable outside
the Gaussian case.  The Laplace approximation
[@tierney1986laplace; @skaug2006admb] replaces the integral with the
integrand evaluated at the conditional mode, corrected by the determinant of
the negative Hessian, and has proved accurate and fast for a wide range of
models [@rue2009inla; @wood2011fast].

The \pkg{TMB} package [@kristensen2016tmb] made this strategy broadly
accessible in \proglang{R} by combining the Laplace approximation with
automatic differentiation (AD): the user writes the joint negative log
density in \proglang{C++}, and \pkg{TMB} supplies its gradient and Hessian
by AD, locates the conditional mode, and differentiates the approximate
marginal likelihood with respect to $\theta$.  \pkg{TMB} descends from
\pkg{ADMB} [@fournier2012admb] and underlies popular front ends such as
\pkg{glmmTMB} [@brooks2017glmmtmb].

The \pkg{adlaplace} package presented here implements the same
Laplace-plus-AD strategy with a different set of design decisions.

First, the *derivative of the Laplace approximation* is computed
exactly.  The approximate log likelihood contains
$\log |H(\hat U(\theta), \theta)|$, where $H$ is the negative Hessian of the
joint log density in $U$ and $\hat U(\theta)$ is the conditional mode.
Differentiating this term with respect to $\theta$ involves third-order
derivatives, and the chain rule contributes a term through the dependence of
$\hat U(\theta)$ on $\theta$ that \pkg{TMB} omits (see Section
\ref{sec:derivative}).  \pkg{adlaplace} evaluates both terms using trace
contractions against a Cholesky factor of $H^{-1}$, so that no third-order
tensor is ever formed explicitly, and the computation remains compatible
with atomic functions whose higher-order derivatives are provided
analytically.

Second, *parallelization* is built into the representation of the model
rather than added afterwards.  The joint log density is written as a sum of
shards -- subsets of observations, blocks of random effects, and
parameter-only terms -- each of which is taped as a separate \pkg{CppAD}
function.  Density evaluations, gradients, Hessians, and the contractions
required for the determinant derivative all decompose over shards and are
evaluated in parallel with OpenMP threads.  The dependency stack is
deliberately small: \pkg{CppAD} [@bell2023cppad] is the only AD library
used, with \pkg{Rcpp} [@eddelbuettel2011rcpp] and \pkg{Matrix}
[@bates2015matrix] handling the interface between \proglang{R} and
\proglang{C++}.

Third, the package is organized for *extensibility* in two layers.  At the
top, \proglang{R} model terms such as `iid()`, `iwp()`, and `gmrf()`
describe model components in the formula language and produce design
matrices, precision structures, and parameter metadata.  At the bottom, a
small set of \proglang{C++} density kernels -- `gaussian_obs`,
`nbinom_obs`, `poisson_obs`, `binomial_obs`, `random_diagonal`,
`random_mult`, and so on -- evaluate log densities on AD types.  Many top-level terms map to one
bottom-level kernel: random intercepts and spline penalties differ in their
design and precision matrices, not in their density kernel.  Both layers
can be extended from separate packages: \pkg{adlaplaceHgp} adds top-level
terms for hierarchical Gaussian process models using only \proglang{R}
code, while \pkg{adlaplaceExample} is a minimal template for adding a new
\proglang{C++} observation density (a skew-normal) in a backend package.
Custom densities that prove generally useful, such as the built-in
\code{binomial_obs} kernel, can be promoted into the core.

The paper proceeds as follows.  Section \ref{sec:methods} sets out the
Laplace approximation, the exact derivative, and the contraction
identities.  Section \ref{sec:parallel} describes the shard decomposition
and the parallel evaluation model.  Section \ref{sec:using} demonstrates
the user-facing interface.  Section \ref{sec:architecture} details the
two-layer architecture and the extension mechanism.  Section
\ref{sec:illustrations} validates the package against \pkg{mgcv}
[@wood2017gam], \pkg{TMB}, and \pkg{INLA} [@rue2009inla] on three
examples and demonstrates a Mat\'{e}rn SPDE field with
\pkg{adlaplaceGrf}, and Section \ref{sec:discussion} concludes.

# Laplace approximation and its derivative {#sec:methods}

## Notation and the approximate likelihood

Write $f(u, \theta) = \log \pi(y, u; \theta)$ for the joint log density of
model \eqref{eq:model} as a function of the latent effects $u \in
\mathbb{R}^m$ and parameters $\theta \in \mathbb{R}^p$, with the data $y$
fixed.  Here $\theta$ collects both regression coefficients $\beta$ and
variance parameters; \pkg{adlaplace} treats them uniformly as outer
parameters.  Define
\begin{align}
G(u, \theta) &= \frac{\partial}{\partial u} f(u, \theta), &
H(u, \theta) &= -\frac{\partial^2}{\partial u \partial u^\top} f(u, \theta),
\end{align}
and let $\hat U(\theta)$ solve $G\{\hat U(\theta), \theta\} = 0$.  The
Laplace approximation to the marginal log likelihood is
\begin{equation}
\ell(\theta) =
f\{\hat U(\theta), \theta\}
+ \frac{m}{2} \log(2\pi)
- \frac{1}{2} \log \big| H\{\hat U(\theta), \theta\} \big|.
\label{eq:laplace}
\end{equation}
AD of $f$ gives $G$ and $H$ at any point; the difficulty is
differentiating \eqref{eq:laplace} with respect to $\theta$, since
$\hat U(\theta)$ has no closed form and $\log |H|$ evaluated at
$\hat U(\theta)$ hides third-order derivatives of $f$.

Throughout, subscripts denote partial derivative blocks evaluated at
$\{\hat U(\theta), \theta\}$:
$H_{UU} = H$, $H_{U\theta} = -\partial^2 f / \partial u\, \partial
\theta^\top$, and the third-order tensors
\begin{align}
T_{UUi} = \frac{\partial}{\partial u_i} H(u, \theta)
  \bigg|_{u = \hat U(\theta)},
\qquad
T_{UU\theta_r} = \frac{\partial}{\partial \theta_r} H(u, \theta)
  \bigg|_{u = \hat U(\theta)}.
\end{align}

## Derivative of the mode

Implicit differentiation of $G\{\hat U(\theta), \theta\} = 0$ gives the
standard identity
\begin{equation}
\frac{\partial}{\partial \theta} \hat U(\theta)
= -H_{UU}^{-1} H_{U\theta},
\label{eq:dmode}
\end{equation}
which requires only second derivatives of $f$.

## Derivative of the log determinant {#sec:derivative}

By the chain rule, the derivative of the log determinant term in
\eqref{eq:laplace} has two parts: a direct dependence on $\theta$, and an
indirect dependence through the mode,
\begin{equation}
\frac{\partial}{\partial \theta_r}
\log \big| H\{\hat U(\theta), \theta\} \big|
=
\underbrace{\operatorname{tr}\!\big( H_{UU}^{-1}\, T_{UU\theta_r} \big)}_{W_r}
\;+\;
\sum_i
\underbrace{\operatorname{tr}\!\big( H_{UU}^{-1}\, T_{UUi} \big)}_{V_i}
\left[ \frac{\partial \hat U(\theta)}{\partial \theta_r} \right]_i .
\label{eq:dlogdet}
\end{equation}
Substituting \eqref{eq:dmode} and writing $V = (V_1, \ldots, V_m)$,
$W = (W_1, \ldots, W_p)$,
\begin{equation}
\frac{\partial}{\partial \theta}
\log \big| H\{\hat U(\theta), \theta\} \big|
= - V H_{UU}^{-1} H_{U\theta} + W .
\end{equation}
Combining with the derivative of $f\{\hat U(\theta), \theta\}$, in which
the indirect term vanishes because $G\{\hat U(\theta), \theta\} = 0$, the
gradient of \eqref{eq:laplace} is
\begin{equation}
\frac{\partial}{\partial \theta} \ell(\theta)
= f_\theta
+ \tfrac{1}{2} V H_{UU}^{-1} H_{U\theta}
- \tfrac{1}{2} W .
\label{eq:gradient}
\end{equation}
The published description of \pkg{TMB} [@kristensen2016tmb, Section 5]
differentiates $\log|H(u, \theta)|$ with $u$ held fixed at the mode, which
yields only the $W$ term of \eqref{eq:dlogdet}; the mode-dependence term
$V H_{UU}^{-1} H_{U\theta}$ is omitted.  For symmetric joint densities the
omitted term is often small, but it is not zero in general, and
\eqref{eq:gradient} is the exact gradient of the function
\eqref{eq:laplace} being maximized.  Exactness matters in practice: it is
what allows the outer optimizer to converge to tight tolerances and what
makes the numerically differentiated Hessian of \eqref{eq:gradient} at the
optimum a valid observed information matrix for standard errors (Section
\ref{sec:using}).  The finite-difference validation in the package test
suite confirms that \eqref{eq:gradient} matches difference quotients of
\eqref{eq:laplace} to numerical precision.

## Trace contractions avoid third-order tensors {#sec:contractions}

Both $V$ and $W$ contract the third-order tensors against $H_{UU}^{-1}$.
Neither tensor is ever formed.  Let $H_{UU}^{-1} = B B^\top$ with $B$ a
(sparse) Cholesky factor obtained from the same factorization used by the
inner Newton iterations.  Then
\begin{equation}
\operatorname{tr}\big( H_{UU}^{-1}\, T_{\,\cdot} \big)
= \operatorname{tr}\big( B^\top T_{\,\cdot}\, B \big)
= \sum_k B_k^\top\, T_{\,\cdot}\, B_k,
\label{eq:contraction}
\end{equation}
where $B_k$ denotes the $k$th column of $B$.  Each summand
$B_k^\top T_{\,\cdot} B_k$ is a second directional derivative of $G$ (or
first directional derivative of $H$) in the direction $B_k$: in AD terms it
is obtained by a forward sweep in direction $B_k$ followed by a reverse
sweep, at a cost proportional to an ordinary gradient evaluation.  The
number of sweeps is the number of structurally nonzero columns of $B$, and
sparsity of $H_{UU}$ -- guaranteed for the models of Section
\ref{sec:parallel} by construction -- keeps both the factorization and the
contraction cheap.

Because the contractions only require forward and reverse sweeps of first
and second order applied to the taped $H$, they remain valid when the
density involves atomic functions (for example $\log \Gamma$, needed by
the negative binomial density) whose derivatives are supplied analytically
rather than by taping elementary operations.  This is the "compatible with
special functions" property: exact third-order information is extracted
without requiring the AD tool to tape third derivatives of every atomic.

## Inner optimization

For fixed $\theta$, the mode $\hat U(\theta)$ is found by a trust-region
Newton iteration using the sparse $H_{UU}$ and $G$ from AD.  The
factorization computed in the final Newton step is reused for the
determinant in \eqref{eq:laplace} and for the contractions
\eqref{eq:contraction}, and the converged mode is cached and used to
warm-start the inner problem at the next outer iterate.

# Parallel evaluation with shards {#sec:parallel}

## Shard decomposition

The joint log density of model \eqref{eq:model} is a sum of conditionally
independent contributions,
\begin{equation}
f(u, \theta) =
\sum_{\ell} \log \pi\big( y^{(\ell)} \mid u; \theta \big)
+ \sum_{m} \log \pi\big( u^{(m)}; \theta \big)
+ \sum_{s} c_s(\theta),
\label{eq:shards}
\end{equation}
where $y^{(\ell)}$ are blocks of observations, $u^{(m)}$ are blocks of
random effects, and $c_s(\theta)$ collects normalizing terms that depend
only on $\theta$.  \pkg{adlaplace} represents each summand as a *shard*: a
self-contained AD function with its own tape, data, and sparsity
information.  Observation shards are formed by partitioning the rows of
the data so that each shard touches a limited set of random effects --
`ad_shards()` computes such a partition from the sparsity pattern of the
random-effects design matrix.  The split of each observation density into
an $\eta$-dependent shard and a parameter-only shard (for example
`gaussian_obs` and `gaussian_extra`) keeps the per-observation tapes
small: normalizing constants such as $-n \log \sigma$ or
$\sum_i \log \Gamma(y_i + \nu)$ are taped once rather than once per shard.

## Everything decomposes over shards

The value, gradient, and Hessian of \eqref{eq:shards} are sums over
shards, and so is the contraction \eqref{eq:contraction}:
\begin{equation}
\operatorname{tr}\big( H_{UU}^{-1} T_{\,\cdot} \big)
= \sum_k \sum_\ell B_k^\top T^{(\ell)}_{\,\cdot} B_k
+ \sum_k \sum_m B_k^\top T^{(m)}_{\,\cdot} B_k .
\end{equation}
Each term touches one shard's tape and one column of $B$, so the entire
outer-gradient computation \eqref{eq:gradient} -- not just the density
evaluation -- is embarrassingly parallel.  Shards are assigned to OpenMP
threads; each thread owns its work arrays, and results are accumulated
into shared sparse structures with scatter maps computed once at tape
time.  The number of threads is a runtime argument (`num_threads`), and
the tapes themselves are recorded in parallel as well, which matters for
models with many shards.

This contrasts with parallelism that is bolted on at the linear-algebra
level: because the shard structure mirrors the statistical structure of
the model, parallel efficiency does not depend on the internals of the AD
tool, and the same decomposition serves taping, evaluation,
differentiation, and the third-order contractions.

# Using adlaplace {#sec:using}

## A first model

The user-facing entry point is `adlaplace()`, which takes a formula built
from model terms, fits the model, and returns an object with the standard
accessor methods.  A minimal example with a Gaussian response and a random
intercept:

```{r usage-sim, echo=TRUE}
set.seed(1)
n <- 500
grp <- factor(sample(20, n, replace = TRUE))
x <- runif(n)
y <- 1 + 2 * x + rnorm(20, sd = 0.5)[grp] + rnorm(n, sd = 0.3)
dat <- data.frame(y = y, x = x, grp = grp)
```

```{r usage-fit, echo=TRUE}
library("adlaplace")
fit <- adlaplace(y ~ x + iid(grp), data = dat)
summary(fit)
```

A bare response on the left-hand side defaults to a Gaussian observation
density with a residual standard deviation parameter; other observation
densities are requested explicitly, as in
`nbinom(y) ~ x + iid(grp)`.  Variance parameters are optimized on the log
scale, and `summary()`, `coef()`, and `confint()` report them back-transformed
by default, with delta-method standard errors:

```{r usage-methods, echo=TRUE}
coef(fit)
confint(fit, parm = "grp_iid")
```

## Standard errors

The outer optimization uses `stats::optim()` with the exact AD gradient
\eqref{eq:gradient} and `hessian = TRUE`, so the observed information is
the numerically differentiated exact gradient at the optimum -- one layer
of finite differencing applied to an exact quantity, rather than two.
`vcov()` returns the resulting covariance on the optimization scale or the
natural scale, and all reported intervals derive from it.

## The pipeline, exposed

`adlaplace()` is a thin orchestration of four steps that remain available
individually for non-standard models (Section \ref{sec:bym} uses them
directly):

```{r usage-pipeline, echo=TRUE, eval=FALSE}
md <- model_data(nbinom(y) ~ x + iid(grp), data = dat)
config <- list(shards = ad_shards(md$data$A, num_shards = 100))
af <- ad_fun(md, config)
opt <- optim(md$data$info$parameters$init, outer_fn, outer_gr,
  ad_fun = af, config = config, cache = new.env(),
  method = "L-BFGS-B", hessian = TRUE)
```

`model_data()` parses the formula into design matrices and parameter
metadata; `ad_shards()` partitions the observations; `ad_fun()` records
the \pkg{CppAD} tapes; and `outer_fn`/`outer_gr` evaluate
\eqref{eq:laplace} and \eqref{eq:gradient}.

# Architecture and extensibility {#sec:architecture}

## Two layers

Model specification is separated into two layers.

*Top layer: \proglang{R} model terms.*  A model term is an S4 object that
knows how to produce a design matrix (`design()`), a precision structure
(`precision()`), and metadata for its parameters (`theta_info()`,
`beta_info()`, `random_info()`).  Built-in terms include `intercept()`,
`linear()`, `iid()` (random intercepts), and `iwp()` (integrated Wiener
process / penalized spline).  Each term names the low-level density kernel
that evaluates its log density, through its `ad_fun` slot.

*Bottom layer: \proglang{C++} density kernels.*  A kernel is a
\proglang{C++} function that evaluates one shard's log density on AD
types.  The built-in kernels are few: `gaussian_obs`, `nbinom_obs`,
`poisson_obs`, and `binomial_obs` for observations, `random_diagonal` and
`random_mult` for Gaussian random effects with fixed precision patterns,
`random_fem_2` and `random_fem_3` for Mat\'{e}rn SPDE fields whose
precision $Q(\kappa, \tau)$ depends on hyperparameters (Section
\ref{sec:matern}), plus the corresponding parameter-only `_extra` kernels
where an observation family has hyperparameters.  Many terms share one
kernel: `iid()` and `iwp()` both use `random_diagonal`, differing only in
the design and precision matrices they construct, and any Gaussian random
field with precision $\tau Q$ for fixed sparse $Q$ uses `random_mult`.

## Extending the top layer: adlaplaceHgp

New model classes that reduce to existing kernels require only
\proglang{R} code.  The \pkg{adlaplaceHgp} package defines terms for
hierarchical Gaussian process models -- `hiwp()`, `hrpoly()`, `rsiid()`,
and relatives -- by subclassing the term classes and providing `design()`
and `precision()` methods.  It contains no \proglang{C++} and no
knowledge of tapes or shards.  The companion \pkg{adlaplaceGrf} package
similarly lives mostly in \proglang{R}, assembling finite-element Grams
for Mat\'{e}rn SPDE fields that are then evaluated by the core
`random_fem_2` / `random_fem_3` kernels (Section \ref{sec:matern}).

## Extending the bottom layer: adlaplaceExample

New observation densities require a backend package with \proglang{C++}
code, for which \pkg{adlaplaceExample} is the template.  It implements a
skew-normal observation density in two source files: `objectiveFunction.cpp`
holds the model-specific log densities (an $\eta$-dependent observation
shard and a parameter-only shard), and `backend.cpp` holds the boilerplate
-- tape-evaluation amalgam, an exported C API created by one macro, and
two resolver functions mapping density names to function pointers.  The
accompanying \proglang{R} file defines the `skewnormal()` term whose
`package` slot directs \pkg{adlaplace} to build tapes with the backend's
entry points.  Tapes are recorded and evaluated inside the backend's
shared library, while the Laplace machinery of Section \ref{sec:methods}
-- inner optimization, determinant, contractions, parallel scheduling --
is called through the C API and never duplicated.

The template is deliberately small (about 200 lines of commented
\proglang{C++} and 100 of \proglang{R}).  Custom densities that become
widely useful can be promoted into the core, as with the built-in
\code{binomial_obs} kernel used in Section \ref{sec:bacteria}.

# Illustrations {#sec:illustrations}

We validate \pkg{adlaplace} on examples chosen to exercise the exact
gradient (Section \ref{sec:gamm}), a built-in logistic mixed model
(Section \ref{sec:bacteria}), a non-diagonal intrinsic precision matrix
(Section \ref{sec:bym}), and a hyperparameter-dependent Mat\'{e}rn
precision with $\log|Q(\theta)|$ on the AD tape (Section
\ref{sec:matern}).  Full code walkthroughs are in
\texttt{docs/paper/examples/}; the emphasis here is on agreement of
estimates and derivatives, not on speed.

## Negative binomial additive mixed model vs mgcv {#sec:gamm}

```{r gamm-run, include=FALSE}
suppressMessages({
  library(Matrix)
  library(adlaplace)
  library(mgcv)
})
set.seed(0)
dat <- gamSim(6, n = 1000, verbose = FALSE)
dat$f2 <- sqrt(dat$f2)
dat$f2 <- dat$f2 - dat$f2[which.min(abs(dat$x2 - 0.5))]
dat$mu <- exp(-1 + 0 * dat$x1 + dat$f2 +
  rnorm(nlevels(dat$fac), sd = 0.25)[as.integer(dat$fac)])
nbinom_sd <- 0.25
dat$Z <- rgamma(nrow(dat), nbinom_sd^(-2), nbinom_sd^(-2))
dat$y <- rpois(nrow(dat), lambda = dat$mu * dat$Z)

b2 <- gamm(
  y ~ x1 + s(x2),
  family = nb(),
  data = dat,
  random = list(fac = ~1)
)

fit_gamm <- adlaplace(
  nbinom(y, lower = 1e-9, init = 0.15) ~ x1 +
    iwp(x2, ref_value = 0.5, p = 2, knots = seq(0, 1, len = 21), init = 1) +
    iid(fac, init = 0.25),
  data = dat,
  num_threads = 4L,
  num_shards = 100L,
  control = list(maxit = 200L, factr = 1e3, pgtol = 1e-12)
)

co_gamm <- coef(fit_gamm)
comparison_table_gamm <- data.frame(
  Parameter = c("x1 (fixed)", "SD random intercept (fac)"),
  adlaplace = c(
    unname(co_gamm[grep("^x1", names(co_gamm))])[1],
    unname(co_gamm[grep("fac", names(co_gamm))])[1]
  ),
  `mgcv::gamm` = c(
    unname(summary(b2$gam)$p.coeff["x1"]),
    as.numeric(nlme::VarCorr(b2$lme)["(Intercept)", "StdDev"])
  ),
  check.names = FALSE
)

res_gamm <- log_lik_laplace(
  x = fit_gamm$optim$par, gamma = fit_gamm$cache$gamma,
  ad_fun = fit_gamm$ad_fun, config = fit_gamm$config, deriv = FALSE
)
newx_gamm <- data.frame(x2 = seq(0, 1, len = 101))
sim_gamm <- sim_fit(x = newx_gamm, data = fit_gamm$model_data, fit = res_gamm, n = 500L)
sim_centered_gamm <- sim_gamm - matrix(
  colMeans(sim_gamm), nrow(sim_gamm), ncol(sim_gamm), byrow = TRUE
)
```

We simulate a negative binomial additive mixed model following the
`mgcv::gamm()` help files [@wood2017gam]: $n = 1000$ observations with a
linear effect of $x_1$, a smooth effect of $x_2$, and a random intercept
over a 40-level factor, with a gamma-distributed multiplicative
overdispersion giving a negative binomial margin.  The \pkg{adlaplace}
formula uses the `iwp()` spline term with 21 knots:

```{r gamm-formula, echo=TRUE, eval=FALSE}
fit <- adlaplace(
  nbinom(y, init = 0.15) ~ x1 +
    iwp(x2, ref_value = 0.5, p = 2, knots = seq(0, 1, len = 21), init = 1) +
    iid(fac, init = 0.25),
  data = dat, num_threads = 4
)
```

The same model is fitted with `mgcv::gamm(y ~ x1 + s(x2), family = nb,
random = list(fac = ~1))`, which uses penalized quasi-likelihood.
Table \ref{tab:gamm} compares the fixed effect and variance components;
Figure \ref{fig:gamm} overlays the estimated smooth with posterior draws
of the \pkg{adlaplace} `iwp` curve.

```{r gamm-table, results="asis", echo=FALSE}
knitr::kable(
  comparison_table_gamm,
  format = "latex", booktabs = TRUE, digits = 3,
  caption = "\\label{tab:gamm}Estimates from adlaplace and mgcv::gamm
  for the simulated negative binomial additive mixed model.")
```

```{r gamm-figure, echo=FALSE, fig.height=4, fig.cap="\\label{fig:gamm}Estimated smooth effect of $x_2$: mgcv::gamm fit (blue) and adlaplace posterior draws of the iwp component (grey), both centred."}
par(mar = c(4, 4, 1, 1), mgp = c(2.2, 0.7, 0))
plot(b2$gam, select = 1, col = "blue", rug = FALSE,
  xlab = "x2", ylab = "s(x2)", shift = 0)
matlines(newx_gamm$x2, sim_centered_gamm, col = "#00000008", lty = 1)
points(dat$x2, dat$f2, col = "red", cex = 0.15, pch = 16)
legend("topleft", bty = "n",
  legend = c("mgcv::gamm", "adlaplace draws", "truth"),
  col = c("blue", "#00000060", "red"), lty = c(1, 1, NA), pch = c(NA, NA, 16))
```

The two fits agree on the fixed effect, the smooth, and the random-intercept
standard deviation to within Monte Carlo and approximation error, confirming
that the exact gradient of Section \ref{sec:derivative} drives the outer
optimizer to the same optimum reached by the mature \pkg{mgcv}
implementation.

## Logistic mixed model vs TMB and INLA {#sec:bacteria}

The \code{bacteria} data [@venables2002mass] record presence of \emph{H.
influenzae} in repeated observations on 50 children.  We fit a Bernoulli
response with logit link, fixed effects of treatment and a week indicator,
and a child-level random intercept using the built-in \code{binomial()}
term and \code{adlaplace()}:

```{r bacteria-formula, echo=TRUE, eval=FALSE}
fit <- adlaplace(
  binomial(yn) ~ trt + wk + iid(ID, init = 1),
  data = bacteria, num_shards = 20L
)
```

The supplement \texttt{examples/bacteria/bacteria.Rmd} gives the full
analysis and comparison code.  We compare against \code{glmmPQL}
(penalized quasi-likelihood), \pkg{TMB} (exact Laplace, via
\pkg{glmmTMB}), and \pkg{INLA} where available.

```{r bacteria-run, include=FALSE}
source("replication/02-bacteria.R", local = (bact_env <- new.env()))
```

```{r bacteria-table, results="asis", echo=FALSE}
if (exists("comparison_table", envir = bact_env)) {
  knitr::kable(
    get("comparison_table", envir = bact_env),
    format = "latex", booktabs = TRUE, digits = 3,
    caption = "\\label{tab:bacteria}Bacteria data: fixed effects and
    random-intercept standard deviation from adlaplace, glmmPQL, TMB, and INLA.")
}
```

\pkg{adlaplace} agrees with \pkg{glmmTMB} on every fixed effect and on the
random-intercept standard deviation to four decimal places (for example
intercepts of $3.548$ versus $3.548$ and standard deviations of $1.242$
versus $1.242$), as expected because both solve the same Laplace
optimization problem.  Differences from \code{glmmPQL} and \pkg{INLA} are
modest and in the direction anticipated for a binary response and for
\pkg{INLA}'s prior on the random-intercept precision.

## Besag-York-Mollie disease mapping vs INLA {#sec:bym}

The final example exercises `random_mult`: a Gaussian random effect with
precision $\tau Q$ for fixed sparse $Q$, with the generalized
log-determinant and rank of $Q$ passed as constants.  We reproduce the
Germany oral-cavity-cancer example from \pkg{INLA}
[@besag1991bym; @rue2009inla] using the low-level pipeline (no formula
interface).  The supplement \texttt{examples/germany/germany.Rmd} documents
the full assembly of observation and random-effect shards, the ICAR
scaling with \pkg{INLA}'s utilities [@sorbye2014scaling], and the
comparison plots.

Because the ICAR precision is rank-deficient, the intercept is omitted
rather than imposing a sum-to-zero constraint.

```{r bym-run, include=FALSE}
if (have_inla) {
  source("replication/03-bym-germany.R", local = (bym_env <- new.env()))
}
```

```{r bym-table, results="asis", echo=FALSE}
if (have_inla && exists("comparison_hyper", envir = bym_env)) {
  knitr::kable(
    get("comparison_hyper", envir = bym_env),
    format = "latex", booktabs = TRUE, digits = 3,
    caption = "\\label{tab:bym}Germany BYM: posterior standard deviations of
    the structured (ICAR) and unstructured spatial components from adlaplace
    and INLA.")
}
```

```{r bym-figure, echo=FALSE, fig.height=4, fig.cap="\\label{fig:bym}Germany oral cavity cancer: posterior mean spatial relative risk from adlaplace (left) and INLA (right)."}
if (have_inla && exists("plot_maps", envir = bym_env)) {
  get("plot_maps", envir = bym_env)()
}
```

```{r bym-scatter, echo=FALSE, fig.height=4, fig.cap="\\label{fig:bymscatter}District-level total spatial effect (structured plus unstructured): adlaplace posterior mode against INLA posterior mean, with the identity line."}
if (have_inla && exists("plot_scatter", envir = bym_env)) {
  get("plot_scatter", envir = bym_env)()
}
```

The district-level total spatial effects agree closely (correlation
$0.99$ across 544 districts; Figure \ref{fig:bymscatter}), and component
standard deviations are comparable (Table \ref{tab:bym}).

## FEM Mat\'{e}rn field with determinant on the AD tape {#sec:matern}

The `random_mult` kernel of Section \ref{sec:bym} covers precision matrices
of the form $\tau Q$ with $Q$ fixed.  Mat\'{e}rn fields obtained from the
Whittle SPDE [@whittle1954; @lindgren2011spde] have a precision
$Q(\kappa, \tau)$ that depends on the range and marginal-variance
hyperparameters in a non-scalar way, so $\log|Q(\theta)|$ must be
evaluated on the AD tape.  The companion package \pkg{adlaplaceGrf}
assembles tensor-product B-spline finite-element Grams on a rectangle,
and the core kernels `random_fem_2` ($\alpha = 2$, $\nu = 1$ in $d = 2$)
and `random_fem_3` ($\alpha = 3$) record a sparse Cholesky update of
$Q(\theta)$ during taping.

We first check that the FEM field approximates the closed-form Mat\'{e}rn
covariance.  With $\nu = 1$, $\kappa = 3$, and $\tau$ chosen so that the
Mat\'{e}rn marginal variance is one, we evaluate the FEM covariance
$\mathbf{A} Q^{-1} \mathbf{A}^{\top}$ at $100$ uniform sites in the unit
square for a coarse and a fine knot grid, and compare to the exact
Mat\'{e}rn correlation as a function of distance
(Figure~\ref{fig:matern}).

```{r matern-run, include=FALSE}
matern_ok <- FALSE
if (have_grf) {
  suppressMessages({
    library(Matrix)
    library(adlaplace)
    library(adlaplaceGrf)
  })
  set.seed(1)
  nu <- 1
  kappa <- 3
  tau <- 1 / sqrt(4 * pi * kappa^2)
  knots_coarse <- list(
    x = seq(-0.2, 1.2, by = 0.2),
    y = seq(-0.2, 1.2, by = 0.2)
  )
  knots_fine <- list(
    x = seq(-1, 2, by = 0.1),
    y = seq(-1, 2, by = 0.1)
  )
  sites_eval <- data.frame(x = runif(100), y = runif(100))
  matern_cov <- function(h, nu, kappa, sigma2 = 1) {
    h <- pmax(h, 1e-12)
    r <- kappa * h
    sigma2 * (2^(1 - nu) / gamma(nu)) * (r^nu) * besselK(r, nu)
  }
  D <- as.matrix(dist(sites_eval))
  Sigma <- matern_cov(D, nu, kappa, sigma2 = 1)
  diag(Sigma) <- 1
  fem_cov_at_sites <- function(sites_eval, knots_list) {
    fem <- grf_bspline(sites_eval, knots_list, degree = 2L)
    Q <- fem_precision(kappa, tau, fem$C, fem$G, fem$G2, alpha = 2L)
    A <- fem$A
    X <- Matrix::solve(Q, Matrix::t(A))
    as.matrix(A %*% X)
  }
  Sig_coarse <- fem_cov_at_sites(sites_eval, knots_coarse)
  Sig_fine <- fem_cov_at_sites(sites_eval, knots_fine)
  rel_err <- function(S_hat, S) {
    sqrt(sum((S_hat - S)^2)) / sqrt(sum(S^2))
  }
  matern_cov_table <- data.frame(
    grid = c("coarse (by = 0.2)", "fine (by = 0.1)"),
    frobenius_rel = c(rel_err(Sig_coarse, Sigma), rel_err(Sig_fine, Sigma)),
    max_abs = c(max(abs(Sig_coarse - Sigma)), max(abs(Sig_fine - Sigma))),
    mean_diag = c(mean(diag(Sig_coarse)), mean(diag(Sig_fine)))
  )
  upper <- upper.tri(D)
  matern_corr_df <- data.frame(
    dist = D[upper],
    matern = Sigma[upper] /
      sqrt(diag(Sigma)[row(Sigma)[upper]] * diag(Sigma)[col(Sigma)[upper]]),
    coarse = Sig_coarse[upper] /
      sqrt(pmax(diag(Sig_coarse)[row(Sig_coarse)[upper]], 1e-8) *
        pmax(diag(Sig_coarse)[col(Sig_coarse)[upper]], 1e-8)),
    fine = Sig_fine[upper] /
      sqrt(pmax(diag(Sig_fine)[row(Sig_fine)[upper]], 1e-8) *
        pmax(diag(Sig_fine)[col(Sig_fine)[upper]], 1e-8))
  )
  matern_corr_df <- matern_corr_df[order(matern_corr_df$dist), ]

  # Joint density AD check on a small FEM model
  knots_list <- list(
    x = seq(0, 1, length.out = 5),
    y = seq(0, 1, length.out = 5)
  )
  sites_small <- do.call(expand.grid, knots_list)
  fem <- grf_bspline(sites_small, knots_list, degree = 2L)
  prec <- fem_precision_payload(fem, alpha = 2L)
  nr <- nrow(fem$C)
  A <- fem$A
  n <- nrow(A)
  sigma <- 0.4
  log_kappa <- log(2)
  log_tau <- log(1)
  w_true <- rnorm(nr, sd = 0.2)
  y <- as.numeric(A %*% w_true + rnorm(n, sd = sigma))
  obs <- adlaplace:::ad_data(
    y = y, A = A,
    gamma_map = Matrix::Diagonal(nr),
    theta_map = Matrix::sparseMatrix(i = 1L, j = 1L, dims = c(3L, 1L)),
    ad_kind = "observations", ad_fun = "gaussian_obs"
  )
  extra <- adlaplace::ad_data(
    obs, ad_kind = "parameters", ad_fun = "gaussian_extra"
  )
  rand <- adlaplace:::ad_data(
    gamma_map = Matrix::Diagonal(nr),
    theta_map = Matrix::sparseMatrix(
      i = c(2L, 3L), j = c(1L, 2L), x = c(1, 1), dims = c(3L, 2L)
    ),
    ad_kind = "random", ad_fun = "random_fem_2", precision = prec
  )
  config <- list(
    gamma = rep(0, nr),
    theta = c(log(sigma), log_kappa, log_tau),
    transform_theta = TRUE
  )
  ptr <- c(
    ad_fun_ptr(obs, config),
    ad_fun_ptr(extra, config),
    ad_fun_ptr(rand, config)
  )
  w <- rnorm(nr, sd = 0.1)
  x <- c(w, log(sigma), log_kappa, log_tau)
  g_ad <- as.numeric(grad(ptr, x, inner = FALSE, negative = FALSE))
  eps <- 1e-5
  g_fd <- vapply(seq_along(x), function(i) {
    xp <- xm <- x
    xp[i] <- xp[i] + eps
    xm[i] <- xm[i] - eps
    (joint_log_dens(ptr, xp, negative = FALSE) -
      joint_log_dens(ptr, xm, negative = FALSE)) / (2 * eps)
  }, numeric(1))
  matern_grad_diff <- max(abs(g_ad - g_fd))
  matern_ok <- TRUE
}
```

```{r matern-formula, echo=TRUE, eval=FALSE}
fem <- grf_bspline(sites_eval, knots_list, degree = 2L)
prec <- fem_precision_payload(fem, alpha = 2L)
rand <- ad_data(
  gamma_map = Diagonal(nrow(fem$C)),
  theta_map = sparseMatrix(i = c(2L, 3L), j = c(1L, 2L), x = c(1, 1), dims = c(3L, 2L)),
  ad_kind = "random", ad_fun = "random_fem_2", precision = prec
)
```

```{r matern-table, results="asis", echo=FALSE}
if (matern_ok) {
  knitr::kable(
    matern_cov_table,
    format = "latex", booktabs = TRUE, digits = 3,
    caption = "\\label{tab:matern}FEM Mat\\'{e}rn covariance at 100 sites
    versus the closed-form Mat\\'{e}rn ($\nu = 1$, $\\kappa = 3$): relative
    Frobenius error, maximum absolute entrywise error, and mean diagonal.")
}
```

```{r matern-figure, echo=FALSE, fig.height=4, fig.cap="\\label{fig:matern}Exact Mat\\'{e}rn correlation versus distance, with FEM approximations on a coarse and a fine knot grid."}
if (matern_ok) {
  par(mar = c(4, 4, 1, 1), mgp = c(2.2, 0.7, 0))
  plot(matern_corr_df$dist, matern_corr_df$matern,
    type = "l", lwd = 2, ylim = c(0, 1), xlim = c(0, 1),
    xlab = "distance", ylab = "correlation"
  )
  lines(matern_corr_df$dist, matern_corr_df$coarse, col = "tomato", lty = 2, lwd = 2)
  lines(matern_corr_df$dist, matern_corr_df$fine, col = "steelblue", lty = 3, lwd = 2)
  legend("topright", bty = "n",
    legend = c("Matern", "FEM coarse", "FEM fine"),
    col = c("black", "tomato", "steelblue"), lty = c(1, 2, 3), lwd = 2)
}
```

The finer knot grid tracks the Mat\'{e}rn correlation more closely
(Table~\ref{tab:matern}; Figure~\ref{fig:matern}).  Wiring the same Grams
into `random_fem_2` with Gaussian observations, the AD gradient of the
joint log density agrees with central finite differences to machine
precision (maximum absolute difference
`r if (matern_ok) format(matern_grad_diff, digits = 2, scientific = TRUE) else "NA"`
in the small check above), confirming that the sparse Cholesky of
$Q(\kappa, \tau)$ is correctly recorded on the tape.

# Discussion {#sec:discussion}

\pkg{adlaplace} implements Laplace-approximate maximum likelihood for
hierarchical models with three distinguishing features: the exact gradient
of the approximate likelihood, including the mode-dependence term of the
log-determinant derivative computed by trace contractions that avoid any
explicit third-order tensor; a shard decomposition that makes the density,
gradient, Hessian, and determinant-derivative contractions all parallel
over OpenMP threads with a single lightweight AD dependency; and a
two-layer, extensible architecture in which new model terms are added in
\proglang{R} and new density kernels in small backend packages, without
modifying the core.  The illustrations show agreement with \pkg{mgcv},
\pkg{TMB}, and \pkg{INLA} across additive, mixed, and spatial models, and
agreement of a Mat\'{e}rn FEM field with its closed-form covariance when
$\log|Q(\theta)|$ is evaluated on the AD tape.

The package is complementary to existing tools.  Its exact gradient and
extensible kernels make it a convenient foundation for methodological work,
and the Laplace-approximate marginal likelihood is a natural building block
for approximate Bayesian inference, as discussed below.

# Future work {#sec:future}

## General precision matrices with the determinant on the AD tape

The `random_mult` kernel of Section \ref{sec:bym} covers precision matrices
of the form $\tau Q$ with $Q$ fixed: because the single hyperparameter
enters only as a multiplier, the log-determinant reduces to
$r \log \tau + \log|Q|_{+}$, and the generalized determinant $\log|Q|_{+}$
and rank $r$ can be computed once in \proglang{R} and enter the tape as
constants.  Section \ref{sec:matern} shows one important extension already
implemented: the Mat\'{e}rn SPDE precisions $Q(\kappa, \tau)$ for operator
orders $\alpha = 2$ and $3$, with a sparse Cholesky update recorded on the
tape through `random_fem_2` / `random_fem_3` and Grams assembled by
\pkg{adlaplaceGrf}.  Other useful models still fall outside this family --
the BYM2 reparameterization [@riebler2016bym2], the Leroux model,
autoregressive structures, and products of Kronecker factors all have
precision matrices $Q(\theta)$ whose determinant depends on the
hyperparameters in a non-trivial way.  Supporting these requires the same
determinant-on-tape pattern, either operation-by-operation for small
blocks or as an atomic sparse Cholesky with a hand-coded derivative, as
\pkg{TMB} does [@kristensen2016tmb].  Because a `random_*` kernel occupies
a single parameter shard in the decomposition of Section
\ref{sec:parallel}, further determinant-on-tape kernels would drop in
alongside `random_diagonal`, `random_mult`, and `random_fem_*` without
changes to the parallel evaluation machinery; the existing kernels remain
preferable when they apply, since they keep the factorization off the tape
entirely whenever $Q$ is fixed up to a scalar.

## Bayesian inference, priors, and aghq

The outer optimization currently targets the Laplace-approximate marginal
likelihood, with no penalization of the hyperparameters.  A modest
extension is a prior layer: penalized-complexity priors
[@simpson2017pc] or user-supplied log-priors on $\theta$ added as an
additional parameter shard, turning the outer objective into a
log-posterior and the reported estimates into posterior modes.  Beyond
that, the quantities \pkg{adlaplace} already computes -- the
Laplace-approximate log-posterior and its exact gradient -- are precisely
the inputs required by adaptive Gauss-Hermite quadrature as implemented in
the \pkg{aghq} package [@stringer2021aghq; @stringer2022fast], which
adapts its quadrature grid using the mode and curvature of the
approximate posterior.  Wrapping an `adlaplace` fit in the function and
gradient pair that \pkg{aghq} expects would deliver posterior summaries
for the hyperparameters, and posterior distributions for the random
effects by mixing the conditional Gaussians over the quadrature points,
while retaining the shard-parallel evaluation and the exact gradient of
Section \ref{sec:methods}.  The same interface would serve Hamiltonian
Monte Carlo on the marginal posterior [@carpenter2017stan].

Other directions include higher-order corrections using the fourth-order
derivative expressions already derived for the log-determinant, and a
broader library of built-in observation densities.

# Computational details {#sec:details}

```{r sessioninfo, echo=FALSE, results="asis"}
toLatex(sessionInfo(), locale = FALSE)
```

# Acknowledgments {#sec:ack}

TODO: funding and acknowledgments.

# References