

```{r header, results='asis', echo=FALSE}
Pmisc::markdownHeader(
title= "Disease mapping with aggregated data",
author= 'Patrick Brown, University of Toronto',
date= 'Nov 2025',
bibliography= normalizePath(c('~/Admin/teaching/asTwo/astwo.bib', '~/Admin/cv/bibs/patrick.bib', '../refs.bib')),
theme='Singapore',
beamer=TRUE,
classoption=c(
  aspectratio=169,
  xcolor='dvipsnames'),
biblatexoptions = c(
  maxbibnames=20,
  maxcitenames=2,doi='true', isbn='false', url='true'),
headerincludes =c(
  '\\usepackage{tikz}',
  '\\usepackage{longtable}',
  '\\usetikzlibrary{shapes}',
  '\\graphicspath{{~/research/book/docs/geostat2015}}',  
  '\\usecolortheme[named=MidnightBlue]{structure}'),
mathCommands=TRUE
)
```


```{r setup, include=FALSE}
library('knitr')

knitr::opts_chunk$set(echo=FALSE, prompt=TRUE, comment=NA,
		dev='png', marginsp=1, fig.cap=' ',
		fig.width=6.5, fig.height=4, half=NULL,
		tidy=TRUE,tidy.opts=list(indent=2, width.cutoff=55)
)

knitr::knit_hooks$set(
		marginsp = function(before, options, envir){	
			if(!before) return()
# use small margins				
			par(mar=c(1.5+0.9*options$marginsp,
							1.5+0.9*options$marginsp,0.2,0.2),
					mgp=c(1.45, 0.45, 0),cex=1.25)
		}		
)

knitr::opts_hooks$set(half = function(options) {
  options$tidy.opts = list(width.cutoff = 12)
  options
})

options(width=80)

knitr::knit_hooks$set(plot=Pmisc::hook_plot_beamer) 
par(bg='#FFFFFF00')

#$
```


```{r setupPackages,include=FALSE}
# packages which print stuff
requireNamespace('Hmisc')
library('mapmisc')
library('geostatsp')
```

# Today

\bcol

\begin{enumerate}

\item What is geostatistics?

\item Gaussian random fields

\item Generalized linear geostatistical model

\end{enumerate}

\newcol


```{r simData, include=FALSE, cache=TRUE}
myRaster <- squareRaster(ext(0,30,0,20), 60)
myModel =  c(variance=1, range=7, shape=1)
f1 <- geostatsp::RFsimulate(
		x=myRaster, model=myModel, n=2)
f = crop(f1, ext(f1) + c(-1,0,0,-1) )
f = wrap(f)
```
```{r unwrapf, include=FALSE}
f = unwrap(f)
```

```{r simDataPlot, echo=FALSE, fig.height=3, fig.width=6}
simCol = colourScale(f, style='equal', dec=0, breaks=10, 
		col='Spectral', rev=TRUE)
for(D in 1:2) {
mapmisc::map.new(f)
plot(f[[D]], col=simCol$col, breaks=simCol$breaks, legend=FALSE, add=TRUE)
legendBreaks('right', simCol)
}
```

\ecol



# What is Geostatistics

 
\bcol[0.45]
 
```{r inspace3, message=FALSE, fig.height=4, fig.width=8}

mapmisc::map.new(f)
plot(f[['sim1']], col=simCol$col, breaks=simCol$breaks, legend=FALSE, add=TRUE)
legendBreaks('right', simCol)


thePoints = vect(
	cbind(
		runif(15, xmin(f), xmax(f)), 
		runif(15, ymin(f), ymax(f))))
values(thePoints)= extract(f, thePoints)

pointsCol = colourScale(thePoints$sim1, breaks=simCol$breaks, col=simCol$col, style='fixed')
mapmisc::map.new(f)
plot(thePoints, col=pointsCol$plot, add=TRUE, cex=2)
```





\newcol

\begin{itemize}
\item There is an underlying and unobserved spatial surface
\item Data is only observed at a few points 
\item Create a statistical model for the surface
\item Estimate the model parameters
\item Make spatial predictions for the full surface (Kriging)
\end{itemize}

\ecol


 
 
# Example : Rongelap Island 

\bcol

```{r rongelapWorld}

pacificMap = openmap(ext(130,179,-25,25),
			path="http://tile.openstreetmap.org/",
			maxTiles=20)

plot(pacificMap,mar=c(0,0,0,0))
rongelapLoc = 166.84 + 11.153i
points(rongelapLoc, col="red")
text(rongelapLoc,"Rongelap", pos=3)

```


\newcol

\begin{itemize}
\item contaminated by nuclear weapons testing in 1954
    \item Leukemia and thyroid-tumors followed.  
    Greenpeace evacuates residents in 1985
\end{itemize}

\ecol

\begin{itemize}
    \item After some removal of soil, radiation levels have fallen
    \item Is it safe to return?
\end{itemize}
``Model-based geostatistics''
PJ Diggle, JA Tawn, RA Moyeed (1998) \emph{Applied Statistics}
Volume 47, Issue 3, 



# Rongelap Island

\bcol

```{r rongelapData}
	data("rongelapUTM", package='geostatsp')  
	rongelapUTM = unwrap(rongelapUTM)
	rongelapUTM$raw = rongelapUTM$count/rongelapUTM$time

	MarshallIslands = geodata::gadm("MHL", path="cache")	
	MarshallIslands = project(MarshallIslands, crs(rongelapUTM))

	rongelapMap = openmap(rongelapUTM, maxTiles=20,buffer=1*1000)


	theColours= mapmisc::colourScale(rongelapUTM$raw,breaks=10, dec=2)

	map.new(rongelapUTM, buffer=0.5*1000)
	plot(rongelapMap, add=TRUE)
	plot(MarshallIslands,add=T)
	plot(rongelapUTM, col=theColours$plot, add=TRUE, cex=0.6)
	

```

\vspace{-20pt}
\begin{itemize}
  \item Radiation measurements taken at a number of locations
  \item Number of particles entering the machine is recorded
\item Count data: non-Gaussian Geostatistics
\end{itemize}

\newcol[0.2]

\noindent Radiation (particles / second)


```{r rongelapLegend,fig.width=2,fig.height=3}
	plot.new()

	mapmisc::legendBreaks("top", theColours)	
```

\ecol


# Random things

\bcolb{In time}

```{r intime, dev='pdf', fig.height=4, fig.width=6}
toplot=cbind(arima.sim(list(ar=0.98), 100)/3,
arima.sim(list(ar=0.3), 100), 
arima.sim(list(ar=0.9), 100)/2)
matplot(seq(1901, 2000, len=100),-min(toplot)+toplot, xlab="years",ylab="cancer rate",
type="l",lty=1)
```
\begin{itemize}
	\item $U(t)$ is a random time series
	\item indexed by time $t$
\item or written $U_t$
\end{itemize}

\newcolb{In space}


```{r theSeedA}
set.seed(1)
```


```{r inSpace, fig.height=4, fig.width=6}
plot(f[[1]])
```

\begin{itemize}
	\item $U(x,y)$ is a random field 
	\item Could write $U(\mathbf{s}), \mathbf{s} = (x, y)$, or
\item $U_i = U(s_i), s_i = (s_{i1}, s_{i2})$
\end{itemize}

\ecolb


# Exponential covariance functions 

\bcol[0.6]

$$
\text{cov}[U(t), U(t+h)] = \sigma^2 \exp(- h/\psi)
$$
or
$$
\text{cov}[U(t), U(t+h)] = \sigma^2 \exp(- \alpha h)
$$

\begin{itemize}
	\item $\phi$ is a \emph{range} parameter
	\item small $\phi$ means covariance falls quickly
	\item \ldots and $U(t)$ is rough
    \item $\alpha = 1/\phi$ is a \emph{scale} parameter
\item $\sigma^2$ is the variance
\item Note that $U(t)$ can be continuous in time.
\end{itemize}

\newcol[0.4]

```{r covarioARa,echo=F, fig.height=4, fig.width=6, dev='pdf'}
hseq = 0:25

rangeseq = c(50, 10, 1)
alphaseq = 1/rangeseq

arseq = exp(-alphaseq)


corrg = exp(-hseq %o% alphaseq )
matplot(hseq, corrg, type="l", lty=1, xlab="h", ylab="corr[U(t), U(t+h)]")
legend("topright", col=1:3, lty=1, legend=as.character(signif(rangeseq,1)),
		title='range')
```

```{r arPlots2,echo=F, fig.height=4, fig.width=6, dev='pdf'}
margVar = 1/(1-arseq^2)
toplot=cbind(arima.sim(list(ar=arseq[1]), 100)/sqrt(margVar[1]),
arima.sim(list(ar=arseq[2]), 100)/sqrt(margVar[2]), 
arima.sim(list(ar=arseq[3]), 100)/sqrt(margVar[3]))
matplot(seq(1901, 2000, len=100),toplot, 
		xlab="years",ylab="cancer rate",
type="l",lty=1, col=1:3 )

```

\ecol



# Gaussian Random Fields

\bcol[0.45]

\begin{itemize}
  \item Consider a random surface $U(s)$ evaluated at a finite number of
  locations $s_1 \ldots s_N$
  \item What's a sensible way to define the distribution of $[U(s_1)\ldots U(s_N)]$?
\end{itemize}

\newcol[0.5]

$$
\left[
\begin{array}{c}
U(s_1)\\
\vdots \\
U(s_n)
\end{array}
\right]
 \sim 
\text{MVN}\left[ 
\left(\begin{array}{c}
\mu_1\\
\vdots \\
\mu_n
\end{array}\right)
, \Sigma\right]
$$

\ecol

\begin{itemize}
\item Let's restrict the class of spatial processes and try to define a nice
model
\item Gaussian distributions are a natural choice for multivariate, correlated
data
\item By restricting ourselves to {\em Gaussian Random Fields}, we only 
need to define means and a variance matrix
\end{itemize}






```{r Srange,include=FALSE}
shape=1
isoParams = c(variance=1, shape=shape, range=NA)
isoParams = rbind(isoParams, isoParams, isoParams)
isoParams[,'range'] = Srange = c(2,4,8)

myraster = squareRaster(ext(-12,12,-8,8),60)
rasterCor = squareRaster(ext(-3, 3, -2, 2),60)


```

```{r simIso, cache=TRUE, include=FALSE}
set.seed(1)
resM = geostatsp::RFsimulate(isoParams, myraster)
names(resM) = paste('range', as.character(isoParams[,'range']))
resMwrap = wrap(resM)
```
```{r unwrapResM}
resM = unwrap(resMwrap)
simColC = colourScale(
		resM, breaks=9,
		style='equal', dec=0,
		col="RdYlBu", rev=TRUE)


corColC = colourScale(
		breaks=sort(c(
						seq(0,1,by=0.2),
						0.95, 0.1)),
		style='fixed',
		col="Reds")

```



```{r maternSimPlot, echo=FALSE, fig.height=3, fig.width=6.5, fig.subcap=c('correlations',names(resM)), fig.cap='Correlations and simulations', fig.start='#', fig.ncol=2}
par(bg='#FFFFFF00')
useq = seq(0, mean(isoParams[,'range'])*2.25,len=400)

theCor = NULL
for(D in 1:nrow(isoParams)) {
	theCor = cbind(theCor, geostatsp::matern(useq, 
					param=isoParams[D,]))
}


matplot(useq, theCor, type='l', 
		lty=1:3, col='black',
		lwd=2,
    xlab=expression(abs('s'['i'] - 's'['j'])),
    ylab=expression(paste(
            rho, '[ (', 's'['i'] - 's'['j'], 
            ')/', phi, ' ]')))	

legend('right',lty=1:3, col='black', lwd=2,
    legend=as.character(Srange),
	title=expression(paste('range, ', phi)),
    bty='n')


 for(D in names(resM)) {
 	mapmisc::map.new(resM)
    plot(resM[[D]],
      col=simColC$col, breaks=simColC$breaks, legend=FALSE,
      add=TRUE)
  mapmisc::legendBreaks('right', simColC, 
      title= 'U(s)', bty='n')
}

```

# Gaussian random fields  

 

\begin{itemize}
\item $U(s)$ is a the value of a Gaussian Random Field at location $s$
\item Correlation is a function of distance
\item $\text{cor}[U(s+h), U(s)] = \sigma^2 \rho(h/\phi)$
\item A range parameter $\phi$ controls how quickly correlation falls
\item big range $\Rightarrow$ slow decay $\Rightarrow$ smooth surface
\end{itemize}

 
# The Matérn family 

     
$$
\rho(h; \phi, \kappa) = \frac{1}{\Gamma(\kappa)2^{\kappa-1}}\left(\frac{\sqrt{8\kappa}||h||}{\phi}\right)^\kappa
K_\kappa\left(\sqrt{8\kappa}||h||/\phi\right).
$$


    
\bcol[0.5]
      
- $\phi$ is the range
- $\kappa$ is a shape parameter!        
- $K_\kappa(\cdot)$ denotes modified Bessel function of order $\kappa$
- $\kappa=0.5$: *exponential* correlation function




\newcol[0.5]

```{r maternCors1setup, cache=TRUE, include=FALSE}
useq = seq(-3, 3,len=1001)

param=cbind(range = c(0.5,1,2,4,6), 
			shape = rev(c(0.25, 0.5,1, 2, 20)),
			variance=1)
	
theCor = apply(param, 1, matern, x=useq)

set.seed(0)
myraster2 = squareRaster(ext(0,15,0,10), 60)

simsM = geostatsp::RFsimulate(param, myraster2)
simsM = crop(simsM, ext(simsM) + c(-xres(simsM),0,0,-yres(simsM)) )

simsMwrap = wrap(simsM)
```

```{r maternCors1,echo=FALSE,fig.height=3,fig.width=6, dev='pdf'}
simsM = unwrap(simsMwrap)
theCol = RColorBrewer::brewer.pal(nrow(param), 'Set2')

matplot(useq, theCor, type="l", lty=1,xlab="h", ylab="rho(h,0)",col=theCol, lwd=3)
legend("topright", lty=1, col=theCol,
		legend = param[,'shape'],bty='n')	
```

\ecol

- $\kappa \rightarrow \infty$: *Gaussian* correlation function
- $U(s)$ is mean-square $m$ times differentiable if and only if $\kappa>m$

        



```{r maternShape, fig.height=3, fig.width=5, fig.cap='Matern correlations and simulations', fig.subcap=c('correlation', param[,'shape']), marginsp=0, fig.start='#', fig.ncol=3}

matplot(useq, theCor, type="l", lty=1,xlab="h", 
		ylab="rho(h,0)",col=theCol,
		ylim=c(0.9, 1), xlim=c(-1, 1)*0.5)

legend('topright',lty=1,lwd=2,
		col=theCol,legend=param[,'shape'], 
		bty='n')

myCol = colourScale(simsM, breaks=10, col='Spectral', dec=0, rev=TRUE, style='equal')

for(D in names(simsM)) {
	map.new(simsM)
	plot(simsM[[D]], col=myCol$col, breaks=myCol$breaks, legend=FALSE, add=TRUE)
}
```



# A Geostatistical model

Write $s_i$ as the location observation $Y_i$ was made at:
\begin{align*} 
Y_{i} \sim & \text{N}\left[\lambda(s_i), \tau^2\right]\\
\lambda(s) = &\mu + X(s) \beta + U(s)\\
\text{cov}\left[U(s +h), U(s)\right] = &\sigma^2 \rho(h/\phi;\nu)
\end{align*}
\begin{itemize}
    \item $U(s)$ is the {\bf Spatial Random Effect} {\small or {\em Residual
    Spatial Variation} or  \emph{Signal process} or
    \emph{latent spatial process}}
\item The $X(s)$ are {\bf covariates} at $s$ {\small or \emph{explanatory
variables},
\emph{independent variables}, \emph{confounders}, or \emph{fixed effects}}
\item $\tau^2$ is the {\bf observation variance} {\small or
\emph{error variance}, \emph{nugget effect}, \emph{noise variance} or more
correctly the {\em independent observation-level random term}}
\end{itemize}




# Interpretation

\begin{description}
    \item[ $Y_i$:]  the (possibly log-transformed) mercury concentration
    measured at location $s_i$
    \item[ $X(s)$:]  a vector of altitude, nighttime light, and categorical
    variables of land use type at $s$.
    \item[ $\tau$:]  the error associated with mercury measurement, or very
    localized factors influencing mercury.
    \item[ $U(s)$:]  the residual spatial variation, the difference
    between actual mercury concentration and what the covariates predict. Depends on 
\begin{description}
\item[$\sigma$] variability in residual variation   
\item[$\phi$] range parameter
\item[$\nu$] shape parameter    
\end{description}
\end{description}





```{r eval=FALSE,include=FALSE}
cfiles = c('timhortons' =
				"https://upload.wikimedia.org/wikipedia/commons/thumb/d/dd/Tim_Hortons_%288138484955%29.jpg/800px-Tim_Hortons_%288138484955%29.jpg",
		"starbucks" = 
				'https://upload.wikimedia.org/wikipedia/commons/thumb/5/55/Starbucks_Coffee_on_Front_Street%2C_Toronto%2C_between_Church_and_Jarvis_streets_-b.jpg/640px-Starbucks_Coffee_on_Front_Street%2C_Toronto%2C_between_Church_and_Jarvis_streets_-b.jpg',
		fordtim = 'http://insidetimmies.com/wp-content/uploads/2014/08/Screen-Shot-2014-08-08-at-10.05.46-PM.png'
)

for(D in names(cfiles)) {
	Djpg = file.path("Figures","overview",paste(D, ".jpg", sep=""))
	if(!file.exists(Djpg)) {
		download.file(cfiles[D], Djpg, method='curl')
	}
}

download.file('https://upload.wikimedia.org/wikipedia/commons/e/e8/Rob_Ford_Mayor.jpg',
		file.path("Figures", "overview",'ford.jpg'),
		method='curl')
```

```{r loadCoffee, include=FALSE}
load("../data/coffee.RData")
dineUtm = vect(dineUtm)
to2010utm = vect(to2010utm)
toCt2006 = vect(toCt2006)
```

```{r toMmap, include=FALSE, cache=TRUE}
themap1 = openmap(to2010utm, 
		fact=2, path='nrcan')
themap = tonerToTrans(themap1, pattern='blue', threshold=200)
themap1 = wrap(themap1)
themap = wrap(themap)
```
```{r murderDataLoad}
themap = unwrap(themap)
themap1 = unwrap(themap1)
data(murder)
torontoBorder = unwrap(torontoBorder)
allTo = project(torontoBorder, crs(themap))
```

```{r getRid,include=FALSE}
getRidUtm = vect(ext(-1700+500000,-800+500000, -5800,-3800), 
	crs= crs(themap))
values(getRidUtm)= 1
dineUtm$getRid = !is.na(extract(getRidUtm, dineUtm)$value)
```


# Election results, 2010

\bcol[0.6]

```{r fordElectionPlot, echo=FALSE, fig.width=6, fig.height=3}
toPlot = function() map.new(allTo, 0.9, buffer=c(0,0,-1000,-200))
toScale = function() scaleBar(to2010utm,'bottomright', 
			bg='white', seg.len=5, 
			inset=c(0.15, 0), bty='n')

myColFord = colourScale(to2010utm$pctFord,
		style='quantile', col='RdYlBu',
		breaks=12,dec=1,
		opacity=1)

toPlot()
plot(to2010utm, col=myColFord$plot,
		add=TRUE,border='#00000020')
plot(themap,add=TRUE)
plot(allTo,add=TRUE)
legendBreaks("right", 
		breaks=paste(round(100*myColFord$breaks), '%', sep=''),
		col=myColFord$col,
		title="Votes\nfor\nFord",inset=0, 
		bty='n')
toScale()
```

\newcol[0.3]

\includegraphics[width=0.7\textwidth]{~/research/book/docs/geostat2015/staticFigures/overview/ford.jpg}

\ecol

``[...] fashioned his image as the defender of the little guy, the
  suburban strivers, against the downtown elites, with their degrees and their
  symphonies'' Andrew Coyne, National Post 15 Nov 2013

\fullcite{hose}

 
# Coffee shop locations  

Locations of establishments inspected by Toronto
Public Health.  

```{r torontoCoffeeShop, fig.height=5,fig.width=10}
myCol = c(RColorBrewer::brewer.pal(8, "Blues")[8:6],
		RColorBrewer::brewer.pal(8, "Oranges")[8:5])
names(myCol) = c("Tim Horton's", "Coffee time","Country style", "Starbuck's","Second Cup","Timothy's", "Aroma Espresso Bar")
map.new(dineUtm)
plotRGB(themap1,add=TRUE)
plot(getRidUtm,border='green',col='green',add=TRUE)
plot(dineUtm, col=
				paste( myCol[as.character(dineUtm$coffee)], 
						"99", sep=""),
		add=TRUE, 
		pch=15,cex=c(0.7, 0.2)[1+dineUtm$getRid])
plot(allTo,add=TRUE, border='green')
scaleBar(to2010utm,'topright')
legend("bottomright", fill=myCol,legend=names(myCol))
```

\url{http://opendata.toronto.ca/public.health/dinesafe/dinesafe.zip}

\url{https://raw.githubusercontent.com/paulgb/dinesafe-api/master/data.csv}



# The Generalised Linear Geostatistical Model

\begin{align}\nonumber
Y_i | U(s_i) \sim &\mathbf{\pi}[\lambda(s_i), \theta]\\
\label{eq:glgm}
g[\lambda(s_i)] = & \beta X(s_i) + U(s_i)\\ \nonumber
\text{cov}[U(s_i), U(s_j)] =& \sigma^2 \rho[ (s_i - s_j) /\phi,\nu].
\end{align}

\begin{itemize}
    \item $U(s)$ is the {\bf Spatial Random Effect} {\small or {\em Residual
    Spatial Variation} or  \emph{Signal process} or
    \emph{latent spatial process}}
\item The $X(s)$ are {\bf covariates} at $s$ {\small or \emph{explanatory
variables},
\emph{independent variables}, \emph{confounders}, or \emph{fixed effects}}
\item $\mathbf{\pi}$ distribution of the observed $Y_i$ conditional on a mean $\lambda(s_i)$
\item $g$ is a link function.
\item $\theta$ are additional parameters for $\pi$
\end{itemize}



# Interpretation

\begin{description}
    \item[ $Y_i$:]  coffee shop type (1=latte) at location $s_i$
    \item[ $\pi$:]  Bernoulli distribution
    \item[$\lambda(s)$:] probability a caf\'e at $s$ serves lattes
    \item[ $X(s)$:]  a vector of covariates at $s$ (proportion of Ford voters, income).
    \item[ $g$:]  logit transformation
    \item[ $U(s)$:]  the residual spatial variation, the difference
    between actual probability and what the covariates predict. Depends on 
\begin{description}
\item[$\sigma$] variability in residual variation   
\item[$\phi$] range parameter
\item[$\nu$] shape parameter    
\end{description}
\end{description}



# Inference

\begin{itemize}
\item Computationally much more difficult than Gaussian models
\item No closed-form expression for the likelihood
\item Bayesian inference is easier than Maximum Likelihood (!?)
\end{itemize}
\begin{description}
      \item[Prior distributions] $[\beta]$, $[\sigma]$, $[\phi]$
\item[Model] $[Y|U, \beta]$, $[U|\sigma, \phi]$
\item[Marginal Posteriors] $[\beta | Y]$, $[\sigma | Y]$, $[\phi | Y]$, 
$[U(s_0)| Y]$
\end{description}
Integrated Nested Laplace Approximation (INLA) and Gaussian
Markov Random Field approximations


# GLGM in R

```{r fitModel, eval=FALSE, echo=TRUE}
coffeeFit = glgm(formula=latte ~ logIncome + ford,
  data=dineSub, covariates=covariates,
  grid=80, shape=2, buffer=4000,
  prior = list(range=5000, sd=0.5),
  family='binomial')
```
\begin{itemize}
  \item {\tt dineSub} is a {\tt vect} with locations of coffee
  shops and variable {\tt latte}
  \item {\tt covariates} is a list of rasters {\tt logIncome} and {\tt ford}
  \item Cover Toronto with a grid of 80 square cells in the x-direction.
  \item {\tt prior} gives prior medians for $\phi$ and $\sigma$
  \begin{itemize}  \item Exponentials for $\sigma$ and $1/\phi$ \end{itemize}
  \item {\tt glgm} is in the {\tt geostatsp} package, which in turn calls {\tt
  inla} from \url{http://r-inla.org}.
\end{itemize}

```{r loadCoffeeResults, include=FALSE}
#res = readRDS("/store/patrick/spatialData/coffeeFit.rds")
(load("../data/coffeeResults.RData"))
to2010utm = vect(to2010utm)
res$raster = rast(res$raster)
```

```{r toMmap2, cache=FALSE, include=FALSE}
themap2 = themap #tonerToTrans(openmap(res$raster, 
#		fact=2, path='stamen-toner'))
```

# Parameter estimates


\bcol[0.7]

```{r coffeeResultsTable, results='asis'}

hrows = c('(Intercept)','range/1000','sd')
notHrows = !rownames(res$parameters$summary)%in%hrows
toprint = rbind(
  cbind(mean=res$parameters$summary['(Intercept)','meanInvLogit'],
    exp(
        res$parameters$summary['(Intercept)',c('0.025quant','0.975quant')])/
      (1 + 
        exp(res$parameters$summary['(Intercept)',c('0.025quant','0.975quant')])
        )
  ),
  cbind(mean=res$parameters$summary[notHrows,'meanExp'],
    exp(res$parameters$summary[notHrows,c('0.025quant','0.975quant')])		
  ),
  res$parameters$summary['range', c('mean','0.025quant','0.975quant')],
  res$parameters$summary['sd', c('mean','0.025quant','0.975quant')]
)

knitr::kable(toprint, digits=2)
```

Posterior means and 95\% credible intervals for odds ratios of
regression coefficients $\exp(\beta)$ and spatial covariance parameters.


\newcol[0.4]

\begin{itemize}
\item Non-Ford voting people really are latte sipping downtown elites.
\item \ldots even accounting for both latte-drinkers and the anti-Ford camp to
have high incommes.
\end{itemize}

\ecol





```{r coffeePriorPost, fig.cap='Prior and posterior distributions of spatial covariance parameters', fig.subcap=c('range $\\phi$','Std Dev $\\sigma$'), fig.height=3, fig.width=4, fig.ncol=2, dev='pdf', fig.start = '#'}

swissFit=res
matplot(res$parameters$range$posterior[,1],
	res$parameters$range$posterior[,-1],
	type="l", col=c('black','blue'), lty=1,
  xlim=c(0, 18*1000), xlab='range', ylab='dens')

legend("topright", col=c('blue','black'), lty=1, legend=c("prior","post'r"), bty='n')

plot(res$parameters$sd$posterior[,1],
	res$parameters$sd$posterior[,2], xlim = c(0, 1),
	type="l",  lty=1, xlab='sd', ylab='dens')
lines(res$parameters$sd$prior[,1], res$parameters$sd$prior[,2], col='blue')	
```


```{r coffeeResultsPlot, fig.cap='Latte prevalence and residual spatial variation', fig.subcap=c('E(rho g data)','E( exp(random) g data)'), fig.ncol=2, fig.start='#'}

toMap = vect(cbind(c(482429.6, 517016.1),c(-6439.042, 10139.631)),
  crs = crs(res$raster))
colFit = colourScale(res$raster[['predict.invlogit']], style='equal',
  dec=-log10(0.05), breaks=7, col='RdYlBu',opacity=0.5, rev=TRUE)
map.new(res$raster)

plot(res$raster[['predict.invlogit']], col=colFit$colOpacity,
  breaks=colFit$breaks,legend=FALSE,add=TRUE)
plot(themap2,add=TRUE)



legendBreaks('bottomright', colFit)
scaleBar(to2010utm,'topright')


malls = rbind(
  "Sherway Gardens" = c(43.611839, -79.557097),
  "Yorkdale" = c(43.725599, -79.452696),
  "Bayview village" = c(43.768661, -79.385812),
  "Scarborough town centre" = c(43.776354, -79.258279),
  "Fairview mall" = c(43.777778, -79.344444),
  "Yonge/ Eglinton" = c(43.706802, -79.398271),
  "Yonge/ Sheppard" = c(43.761389, -79.410833),
  "Bay/ Bloor" = c(43.6646186,-79.3958636)
)
mallsSp = vect(malls[,2:1],
  crs="epsg:4326")

values(mallsSp) = data.frame(name=rownames(malls),  
    pos=1, 
    offset=1,
    row.names=rownames(malls))

mallsSp$pos[as.character(mallsSp$name)=="Bayview village"] = 3
mallsSp$offset[as.character(mallsSp$name)=="Bayview village"] = 0
mallsSp$pos[as.character(mallsSp$name)=="Bayv/ Bloor"] = 2


mallsSp = project(mallsSp, crs(res$raster))

colRandom = colourScale(res$raster[['random.exp']], style='equal',
  dec=-log10(.005)	, breaks=7, col='RdYlBu',opacity=0.5, rev=TRUE)
map.new(res$raster)
plot(getRidUtm,border='green',col='green',add=TRUE)
plot(res$raster[['random.exp']], col=colRandom$colOpacity,
  breaks=colRandom$breaks,legend=FALSE,add=TRUE)
plot(themap2,add=TRUE)
legendBreaks('bottomright', colRandom)
scaleBar(to2010utm,'topright')
points(mallsSp)
text(mallsSp, labels=gsub(" ", "\n", mallsSp$name), pos=mallsSp$pos,
  offset=mallsSp$offset)


```


```{r coffeeResultsExcPlot, fig.cap='Latte exceedance probabilities', fig.start='#', fig.height=4, fig.width=8, fig.subcap = c('pr($\\rho(s) > 0.6 | Y$)', 'pr($U(s) > 0 | Y$)'), fig.ncol=2}


fitP = excProb(res$inla$marginals.predict, log((0.6) / (1-(0.6))),
  template=res$raster)

colFitP = colourScale(fitP , style='fixed',
  breaks=c(0, 0.2, 0.5, 0.8, 0.9, 0.95, 1), 
  col='RdYlGn', rev=TRUE)
map.new(res$raster)
plot(fitP, col=colFitP$col,
  breaks=colFitP$breaks,legend=FALSE,add=TRUE)
plot(themap2,add=TRUE)
legendBreaks('bottomright', colFitP)
scaleBar(to2010utm,'topright')

fitR = excProb(res$inla$marginals.random$space, 
	0, template=res$raster)

map.new(res$raster)
  plot(fitR, col=colFitP$col,
    breaks=colFitP$breaks,legend=FALSE,add=TRUE)
plot(themap2,add=TRUE)
  legendBreaks('bottomright', colFitP)
  scaleBar(to2010utm,'topright')

```








# Log-Gaussian Cox Process

\begin{itemize}
\item $P_i = $ location of murder $i$
\item murders are a Poisson prrocess with intensity $\lambda(s)O(s)$ 
\item $\lambda(s)$ = risk, $O(s)$ = population at risk
$$
|| \{ P_i \in R \} || \sim \text{Poisson}\left[\int_R \lambda(u)O(\cdot) du  \right]
$$
\item $\lambda$ depends on covariates $X_p(s)$ and a residual spatial random
effect $U(s)$
\begin{align*}
\{ P_i \} | U(\cdot) \sim & \text{Poisson process}[\lambda(\cdot) O(\cdot)] \\
\log[\lambda(s)] = & \mu + X(s)^T\beta + U(s) \\
\text{cov}[U(s+h), U(s)] = & \sigma^2\text{Mat\'ern}(|h|/\phi, \nu)
\end{align*}
$\phi_1$ and $\nu$ are spatial range and shape parameters. 
\end{itemize}






```{r lgcp1, include=FALSE}
mymodel = c(mean=-1, variance=1, 
		range=2, shape=2)
myraster = squareRaster(ext(0,10,0,10), 20)
myCovariate = list(
	a=rast(matrix(1:10,10,10), extent=ext(0,10,0,10)),
		b=rast(matrix(1:10,10,10,byrow=T), extent=ext(0,10,0,10)) )

myLgcp=simLgcp(mymodel, myCovariate,betas=c(a=-0.01, b=0.025), 
		rasterTemplate=myraster)
```





```{r covA, fig.cap='An LGCP', fig.subcap=c('$X_1(s)$','$X_2(s)$','$U(s)$','$\\log(\\lambda(s))$','intensity', 'events'), fig.start='#', fig.ncol=3, fig.height=3, fig.width=5}

par(bg='#FFFFFF00')
plot(myLgcp$r[["a"]])

plot(myLgcp$r[["b"]])

plot(myLgcp$r[["random"]])

plot(myLgcp$r[["linearPredictor"]])

plot(myLgcp$r[["intensity"]])

plot(myLgcp$events);axis(1);axis(2)
```





```{r murderSetup, include=FALSE}
data('murder', package='geostatsp')
data('torontoPop')
torontoPdens = unwrap(torontoPdens)
torontoBorder = unwrap(torontoBorder)
torontoNight = unwrap(torontoNight)
torontoIncome = unwrap(torontoIncome)
murder = unwrap(murder)
murderRot = project(murder, mapmisc::omerc(murder, angle=-17))
```


```{r murderSetupMap, cache=FALSE}
distMat = distance(murderRot, murderRot)
closeMurders = distMat < 10
murderRot$Nclose = apply(closeMurders, 2, sum)
murderClust = murderRot[murderRot$Nclose > 7, ]
murderMap = mapmisc::openmap(murderRot, path='nrcan', zoom=9, fact=2.5)
murderMapZoom = mapmisc::openmap(murderClust, buffer=2*1000, fact=1.5)
```


```{r murderExpl, fig.start='#', fig.ncol=2, fig.cap='Murders', fig.subcap=c('all','zoom')}
mapmisc::map.new(murderRot)
plotRGB(murderMap, add=TRUE, maxpixels=10^9)
points(murderRot, pch=16, col=mapmisc::col2html('blue', 0.25))
text(murderRot[murderRot$Nclose > 4, ], murderRot$Nclose[murderRot$Nclose > 4], col='red', cex=1.5)

mapmisc::map.new(murderClust, buffer=0.5*1000)
plot(murderMapZoom, add=TRUE, maxpixels=10^9)
points(murderRot, pch=16, cex=1.5, col=mapmisc::col2html('blue', 0.25))
text(murderRot[murderRot$Nclose > 4, ], murderRot$Nclose[murderRot$Nclose > 4], col='red', cex=1.5)
```


```{r murderCllean}

murderClean = murderRot[murderRot$Nclose < 15, ]

```







# Fitting the murder data

```{r lgcpFit,echo=TRUE, cache=TRUE}
murderFit = lgcp(
  formula= ~ loginc + logpop + loglight,
  data=murderClean, grid=125,
  covariates= list(loginc = log(torontoIncome),
	logpop = log(torontoPdens), loglight = log(torontoNight)),
 shape=1, buffer=2000,
  border=torontoBorder,
  prior = list(
  	range=c(2000,5000),
  	sd=0.25)
)
murderFit$raster = mask(murderFit$raster, torontoBorder)
murderFit$rasterWrap = wrap(murderFit$raster)
``` 
```{r murderFitUnwrap}
murderFit$raster = unwrap(murderFit$rasterWrap)
```

\begin{itemize}
\item {\tt murderRot} is a {\tt vect} object of murder locations
\item {\tt covList} is a list of rasters of log income, log population, and log
ambient light
\item Cell counts for a grid with 125 cells in the x-direction are computed
\item \ldots and a Poisson model is fit by {\tt glgm} and {\tt inla}.
\end{itemize}


# Parameter posterior distributions

```{r murderParam, results='asis', echo=TRUE}
knitr::kable(murderFit$parameters$summary[,
		c('0.5quant','0.025quant','0.975quant')],
		digits=2)
```


# E$[\lambda(s)\mid Y]$ in cases/km2 (for 30 years)

```{r murderSetupMap2, cache=TRUE, include=FALSE}
murderBgTrans =  wrap(mapmisc::tonerToTrans(mapmisc::openmap(
	murderFit$raster, path='nrcan'), pattern='blue', threshold=200))
```
```{r muderStupMapUnwrap}
murderBgTrans = unwrap(murderBgTrans)
```

```{r murderLambda, fig.height=4, fig.width=8}
murderFitKm2 = murderFit$raster$predict.exp * 10^6
murderCol = mapmisc::colourScale(murderFitKm2, 
	breaks=8, style='equal', transform='log', digits=1, col='Spectral',
	rev=TRUE)
mapmisc::map.new(torontoBorder)
plot(murderFitKm2, col=murderCol$col, breaks=murderCol$breaks, add=TRUE, legend=FALSE)
terra::plot(murderBgTrans, add=TRUE)
mapmisc::legendBreaks('bottomright', murderCol, bty='n')
```





# Residual spatial variation E$(U(s)\mid Y)$



```{r murderLambdaResid, fig.height=4, fig.width=8}
murderColR = mapmisc::colourScale(murderFit$raster$random.mean, breaks=8, style='equal', digits=1, col='Spectral',
	rev=TRUE)
mapmisc::map.new(torontoBorder)
plot(murderFit$raster$random.mean, col=murderColR$col, breaks=murderColR$breaks, add=TRUE, legend=FALSE)
plot(murderBgTrans, add=TRUE)
mapmisc::legendBreaks('bottomright', murderColR, bty='n')
```




# Lung and thyroid cancer


\begin{tabular}{ll}
\includegraphics[width=0.5\textwidth]{staticFigures/plotIncome-2.png} &
\includegraphics[width=0.5\textwidth]{staticFigures/plotIncome-3.png}\tabularnewline
a) Asian & b) African or Caribean \tabularnewline
\end{tabular}


\fullcite{thyroid}


# Parameter estimates 

log-Gaussian Cox process model applied to Lung and Thyroid cancer

\begin{tabular}{r|llllllll}
& mean & 0.025q & 0.975q & E$(e^\beta)$ & mean & 0.025q &
0.975q & E$(e^\beta)$\tabularnewline
\hline
(Intercept) & -0.49 & -0.61 & -0.38 & 0.62 & 0.99 & 0.89 & 1.09 &
2.73\tabularnewline
income & -0.20 & -0.35 & -0.05 & 0.83 & 0.04 & -0.11 & 0.19 &
1.05\tabularnewline
asia & -0.50 & -0.79 & -0.20 & 0.61 & 0.40 & 0.13 & 0.66 &
1.50\tabularnewline
africa & -0.31 & -0.92 & 0.28 & 0.75 & -0.30 & -0.89 & 0.28 &
0.76\tabularnewline
immigrant & 0.05 & -0.17 & 0.26 & 1.06 & 0.12 & -0.09 & 0.33 &
1.14\tabularnewline
range & 4858.19 & 2474.94 & 8819.62 & NA & 2355.21 & 809.02 & 5111.70 &
NA\tabularnewline
sd & 0.22 & 0.16 & 0.31 & NA & 0.19 & 0.11 & 0.33 & NA\tabularnewline
\end{tabular}


# Lung and Thyroid Cancer

\begin{tabular}{ll}
\includegraphics[width=0.4\textwidth]{staticFigures/plotRes-1.png}
&
\includegraphics[width=0.4\textwidth]{staticFigures/plotRes-2.png}\tabularnewline
a) L E$(\lambda(s)|Y)$ & b) L E$(U(s)|Y)$\tabularnewline
\includegraphics[width=0.4\textwidth]{staticFigures/plotRes-3.png}
&
\includegraphics[width=0.4\textwidth]{staticFigures/plotRes-4.png}\tabularnewline
e)  T E$(\lambda(s)|Y)$ & f) T E$(U(s)|Y)$\tabularnewline
\end{tabular}



# Rongelap


```{r rongelapGlgmsetup, echo=TRUE, tidy=FALSE}
rongelapUTM$logTime = log(rongelapUTM$time)
rongelapBorder = crop(MarshallIslands, 
	extend(ext(rongelapUTM), 200)) 
```
```{r rongelapGlgm, echo=TRUE, tidy=FALSE, cache=TRUE}
rongelapRes = glgm(
	count ~ offset(logTime),
	data = rongelapUTM,
	prior = list(sd=0.5, range = 1000),
	family='poisson',
	grid = squareRaster(rongelapBorder, 100), 
	buffer=1000)		
rongelapRes$rasterWrap = 
	wrap(rongelapRes$raster)		
```

#


```{r rongelapPlot, echo=FALSE, fig.height=4, fig.width=8}
rongelapRes$raster = unwrap(rongelapRes$rasterWrap)
toPlot = exp(mask(rongelapRes$raster$predict.0.5quant, rongelapBorder))
rongelapCol = colourScale(toPlot, style='equal', breaks=10,dec=-log10(0.5),
	col='Reds', opacity=c(0.1,1))
mapmisc::map.new(rongelapBorder, buffer=100)
plot(rongelapMap, add=TRUE)
plot(toPlot, col=rongelapCol$colOpacity, breaks=rongelapCol$breaks, legend=FALSE, add=TRUE)
plot(rongelapBorder, add=TRUE)
legendBreaks('topleft', rongelapCol, bty='n')
scalBar(toPlot, 'top', bty='n')
```





# more rongelap

```{r params, echo=TRUE}
knitr::kable(rongelapRes$parameters$summary[,4:6], digits=2)
```


# Conclusions 

- Geostatistics is fun!

