---
title: "Laplace Approximations with adlaplaceExample"
author: "Patrick Brown"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Laplace Approximations with adlaplaceExample}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette mirrors the workflow in `adlaplace`, but uses the
`adlaplaceExample` skew-normal objective defined in
`src/objectiveFunction.cpp`.

Skew-normal contribution for one observation:

$$
\ell(y;\xi,\omega,\alpha)
= -\log\omega - \frac12\log(2\pi)
  -\frac12\left(\frac{y-\xi}{\omega}\right)^2
  + \log\left[\operatorname{erfc}\left(-\frac{\alpha(y-\xi)}{\sqrt{2}\,\omega}\right)\right].
$$

## Simulate data

```{r theData}
set.seed(1)

Nobs <- 20000
Nrandom1 <- 15
Nrandom2 <- 25

X <- Matrix::Matrix(cbind(1, rbinom(Nobs, 1, prob = 0.5)))
beta_true <- c(0.4, -0.3)

A1 <- Matrix::sparseMatrix(i = 1:Nobs, j = sample(Nrandom1, Nobs, replace = TRUE), x = 1)
A2 <- Matrix::sparseMatrix(i = 1:Nobs, j = sample(Nrandom2, Nobs, replace = TRUE), x = 1)
A <- cbind(A1, A2)

NperEffect <- c(Nrandom1, Nrandom2)
map <- Matrix::sparseMatrix(
  i = 0:(ncol(A) - 1L),
  j = rep(0:(length(NperEffect) - 1L), NperEffect),
  x = 1L,
  dims = c(ncol(A), length(NperEffect)),
  index1 = FALSE
)

sd_random <- c(0.25, 0.15)
gamma_true <- rnorm(ncol(A), sd = rep(sd_random, NperEffect))

omega_true <- 0.7
alpha_true <- 0.8


eta <- as.vector(X %*% beta_true + A %*% gamma_true)
if (requireNamespace("sn", quietly = TRUE)) {
  y <- sn::rsn(Nobs, xi = eta, omega = omega_true, alpha = alpha_true)
} else {
  y <- stats::rnorm(Nobs, mean = eta, sd = omega_true)
}
data <- list(
  y = y,
  ATp = as(Matrix::t(A), "dMatrix"),
  XTp = as(Matrix::t(X), "CsparseMatrix"),
  map = map,
  Qdiag = rep(1, ncol(A))
)
```


## Configure model and build AD object

```{r config}
config <- list(
  beta = rep(0, nrow(data$XTp)),
  gamma = rep(0, nrow(data$ATp)),
  theta = c(log(0.2), log(0.2), log(0.5), 2),
  transform_theta = TRUE,
  groups = adlaplace::ad_shards(Matrix::t(data$ATp)),
  num_threads = 8L,
  package = "adlaplaceExample",
  verbose = TRUE
)
```

## Evaluate objective, gradient, and Hessian

```{r derivs}
ad_fun <- adlaplace::get_ad_fun(data, config, package = "adlaplaceExample")
true_theta <- c(log(c(sd_random, omega_true)), alpha_true)
x <- c(beta_true, gamma_true, true_theta)

Sgroups <- seq.int(0L, length.out = adlaplace::n_groups(ad_fun$ad_fun))
# Sgroups = 0L

adlaplace::joint_log_dens(ad_fun, x, Sgroups, negative = FALSE)

g <- adlaplace::grad(ad_fun, x, inner = FALSE)
H <- adlaplace::hessian(ad_fun, x, inner = FALSE)

str(g)
str(H)
quantile(eigen(H)$values)
```


## Trust-optimization and Laplace log-likelihood interface

```{r trustOptimInterface}
ad_fun <- adlaplace::get_ad_fun(data, config, package = "adlaplaceExample")

res <- adlaplace::log_lik_laplace(
  x = c(config$beta, config$theta),
  config = config,
  ad_fun = ad_fun,
  deriv = TRUE
)
-res$logLik
res$grad
```


```{r trustOptimOuterWrappers}
cache <- new.env(parent = emptyenv())
cache$gamma <- config$gamma

x0 <- c(config$beta, config$theta)

adlaplace::outer_fn(x = x0, cache = cache, config = config, ad_fun = ad_fun)
adlaplace::outer_gr(x = x0, cache = cache, config = config, ad_fun = ad_fun)

outer_fit <- trustOptim::trust.optim(
  x = x0,
  fn = adlaplace::outer_fn,
  gr = adlaplace::outer_gr,
  method = "SR1",
  config = modifyList(config, list(verbose = FALSE)),
  ad_fun = ad_fun,
  cache = cache,
  control = list(
    maxit = 1000,
    report.level = 4,
    report.freq = 1
  ),
  control_inner = list(
    maxit = 100,
    report.level = 0,
    report.freq = 0
  )
)

outer_fit$solution
outer_fit$fval
```


```{r testLogLgrad}
x <- c(config$beta, config$theta)
Npar <- 13
Dpar <- length(x)

parDf <- data.frame(x)[, rep(1, Npar)]
Sx <- seq(-0.1, 0.1, len = 13) + parDf[Dpar, 1]
SxD <- Sx[-1] - diff(Sx) / 2
parDf[Dpar, ] <- Sx

res <- mapply(
  adlaplace::log_lik_laplace,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, config = config, deriv = TRUE),
  SIMPLIFY = FALSE
)

Slik <- unlist(lapply(res, "[[", "logLik"))
Sdet <- unlist(lapply(res, function(xx) xx$opt$half_log_det))
dU <- do.call(
  abind::abind, c(lapply(res, function(xx) as.matrix(xx$extra$dU)), along = 3)
)
uHat <- do.call(rbind, lapply(res, function(xx) xx$opt$solution))
extraDf <- do.call(abind::abind, c(lapply(res, "[[", "deriv"), along = 3))
gradMat <- do.call(rbind, lapply(res, "[[", "grad"))
gradVec <- unlist(lapply(res, function(xx) xx$deriv[Dpar, "dL"]))


plot(Sx, gradVec)
lines(SxD, diff(Slik) / diff(Sx), lwd = 2, type = "o", pch = ".")
```

```{r more_plots, eval=FALSE}
Du <- 1
plot(Sx, uHat[, Du])

plot(Sx, dU[Du, Dpar, ])
lines(SxD, diff(uHat[, Du]) / diff(Sx))

plot(Sx, Sdet)

plot(Sx, extraDf[Dpar, "dDet", ])
lines(SxD, diff(Sdet) / diff(Sx))
```


```{r derivJointDens, eval=FALSE}
config$gamma <- rep(1, length(config$gamma))
x <- c(config$beta, config$gamma, config$theta)
ad_fun <- adlaplace::get_ad_fun(data, modifyList(config, list(verbose = TRUE)))

inner <- FALSE
type <- c("outer", "inner")[1 + inner]

Sgamma1 <- seq.int(length(config$beta) + 1, length.out = length(config$gamma))

Npar <- 13
Dpar <- length(x) - 1
if (inner) {
  Dpar <- match(Dpar, Sgamma1)
}
Dpar0 <- Dpar - 1L

bob <- as(ad_fun[[type]], "TsparseMatrix")
(whichIndex <- which(bob@i == Dpar0 & bob@j == Dpar0) - 1L)
map_type <- if (inner) ad_fun$map_inner else ad_fun$map_outer
(whichGlobal <- which(map_type$global == whichIndex) - 1L)
(whichP <- mapply(
  function(xx) min(which(map_type$p >= xx)),
  xx = whichGlobal
) - 1L)

sparsity_shard <- lapply(
  seq_len(adlaplace::n_groups(ad_fun$ad_fun)) - 1L,
  function(g) {
    c(
      adlaplace::get_sparse_sizes(ad_fun$ad_fun, g),
      adlaplace::get_sparse_pattern(ad_fun$ad_fun, g)
    )
  }
)

do.call(rbind, sparsity_shard[[whichP[1] + 1L]][c("row_hess", "col_hess")])
do.call(rbind, sparsity_shard[[whichP[1] + 1L]][c("row_hess_inner", "col_hess_inner")])


# Sgroups = length(sparsity_shard)-12L
# Sgroups = seq.int(from=0, length.out = length(sparsity_shard))

Sgroups <- 376


parDf <- data.frame(x)[, rep(1, Npar)]
Sx <- seq(-0.1, 0.1, len = Npar) + parDf[Dpar, 1]
SxD <- Sx[-1] - diff(Sx) / 2
parDf[Dpar, ] <- Sx


dens1 <- mapply(
  adlaplace::joint_log_dens,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, Sgroups = Sgroups),
  SIMPLIFY = TRUE
)

plot(Sx, dens1)


grad1 <- mapply(
  adlaplace::grad,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, inner = inner, Sgroups = Sgroups),
  SIMPLIFY = TRUE
)
plot(Sx, grad1[Dpar, ])
lines(SxD, diff(dens1) / diff(Sx))


hes1 <- mapply(
  adlaplace::hessian,
  x = as.list(parDf),
  MoreArgs = list(ad_fun = ad_fun, inner = inner, Sgroups = Sgroups),
  SIMPLIFY = FALSE
)

hes2 <- do.call(abind::abind, c(
  lapply(hes1, as.matrix),
  along = 3
))

gradD <- apply(grad1, 1, diff) / mean(diff(Sx))

Dpar2 <- Dpar
plot(Sx, hes2[Dpar, Dpar2, ],
  ylab = paste(Dpar, Dpar2),
  ylim = range(hes2[Dpar, Dpar2, ], na.rm = TRUE)
)
lines(SxD, gradD[, Dpar2])
```


## Notes

- `adlaplaceExample` provides model-specific C++ objective code.
- `adlaplace` provides shared optimization and Laplace tooling.
- For production use, optimize outer parameters with `adlaplace` wrappers and
  call `adlaplaceExample` for objective-specific derivatives.
