---
title: "Laplace Approximations with adlaplaceExample"
author: "Patrick Brown"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Laplace Approximations with adlaplaceExample}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This vignette mirrors the workflow in `adlaplace`, but uses the
`adlaplaceExample` skew-normal objective defined in
`src/objectiveFunction.cpp`.

Skew-normal contribution for one observation:

$$$
\ell(y;\xi,\omega,\alpha)
= -\log\omega - \frac12\log(2\pi)
  -\frac12\left(\frac{y-\xi}{\omega}\right)^2
  + \log\left[\operatorname{erfc}\left(-\frac{\alpha(y-\xi)}{\sqrt{2}\,\omega}\right)\right].
$$$

## Simulate data

```{r theData}
set.seed(1)

Nobs <- 20000


## Configure model and build AD object

```{r config}
config <- list(
  beta = rep(0, nrow(data$XTp)),
  gamma = rep(0, nrow(data$ATp)),
  theta = c(log(0.2), log(0.2), log(0.5), 2),
  transform_theta = TRUE,
  groups = adlaplace::adFun_groups(data$ATp),
  num_threads = 8L,
  package = "adlaplaceExample",
  verbose = TRUE)
```

## Evaluate objective, gradient, and Hessian

```{r derivs}
adFun <- adlaplace::getAdFun(data, config, package = "adlaplaceExample")
true_theta = c(log(c(sd_random, omega_true)), alpha_true)
x <- c(beta_true, gamma_true, true_theta)

Sgroups = seq.int(0L, length.out= length(adFun$sparsity))
#Sgroups = 0L

adlaplace::jointLogDens(x, adFun, Sgroups)

g <- adlaplace::grad(x, adFun, inner = FALSE)
H <- adlaplace::hess(x, adFun, inner = FALSE)

str(g)
str(H)
quantile(eigen(H)$values)
```


## Trust-optimization and Laplace log-likelihood interface

```{r trustOptimInterface, eval=FALSE}
adFun <- adlaplace::getAdFun(data, config, package = "adlaplaceExample")


res <- adlaplace::logLikLaplace(
  x = c(config$beta, config$theta),
  config = config,
  adFun = adFun, 
  deriv=TRUE)
res$fval
res$grad
```


```{r trustOptimOuterWrappers, eval=FALSE}
adFun <- adlaplace::getAdFun(data, config)

cache <- new.env(parent = emptyenv())
cache$gamma <- config$gamma

x0 <- c(config$beta, config$theta)

adlaplace::outer_fn(x=x0, cache=cache, config=config, adFun = adFun)
adlaplace::outer_gr(x=x0, cache=cache, config=config, adFun = adFun)

outer_fit <- trustOptim::trust.optim(
  x = x0,
  fn = adlaplace::outer_fn,
  gr = adlaplace::outer_gr,
  method = "SR1",
  config = modifyList(config,list(verbose=FALSE)),
  adFun = adFun,
  cache = cache,
  control = list(
    maxit = 1000,
    report.level = 4,
    report.freq =1 
  ),
  control_inner = list(
    maxit = 100,
    report.level = 0,
    report.freq = 0
  )
)

outer_fit$solution
outer_fit$fval
```


```{r testLogLgrad}
adFun = adlaplace::getAdFun(data,config)
x = c(config$beta,config$theta)
Npar=13;Dpar = length(x)

parDf = data.frame(x)[,rep(1, Npar)]
Sx = seq(-0.1, 0.1, len=13) + parDf[Dpar,1]
SxD = Sx[-1] - diff(Sx)/2
parDf[Dpar,] = Sx

res = mapply(
  adlaplace::logLikLaplace,
  x = as.list(parDf),
  MoreArgs = list(adFun = adFun, config = config, deriv=TRUE),
  SIMPLIFY=FALSE
  )

Slik = unlist(lapply(res, "[[", 'logLik'))
Sdet = unlist(lapply(res, function(xx) xx$opt$halfLogDet))
dU = do.call(
  abind::abind, c(lapply(res, function(xx) as.matrix(xx$extra$dU)), along=3))
uHat = do.call(rbind, lapply(res, function(xx) xx$opt$solution))
extraDf = do.call(abind::abind, c(lapply(res, '[[', "deriv"), along=3))
gradMat = do.call(rbind, lapply(res, '[[', 'grad'))
gradVec = unlist(lapply(res, function(xx) xx$deriv[Dpar,'dL']))


plot(Sx, Slik)

plot(Sx, gradVec)
lines(SxD, diff(Slik)/diff(Sx), lwd=2, type='o', pch='.')

Du = 1
plot(Sx, uHat[,Du])

plot(Sx, dU[Du,Dpar,])
lines(SxD, diff(uHat[,Du])/diff(Sx))

plot(Sx, Sdet)

plot(Sx, extraDf[Dpar,'dDet',])
lines(SxD, diff(Sdet)/diff(Sx))
```


```{r derivJointDens, eval=FALSE}
config$gamma = rep(1, length(config$gamma))
x = c(config$beta, config$gamma, config$theta)
adFun = adlaplace::getAdFun(data, modifyList(config, list(verbose=TRUE)))

inner = FALSE
type = c('outer','inner')[1+inner]

Sgamma1 = seq.int(length(config$beta)+1, length.out = length(config$gamma))

Npar=13;Dpar = length(x)-1
if(inner) {
  Dpar = match(Dpar, Sgamma1)
}
Dpar0=Dpar-1L

bob = as(adFun$hessians$hessian[[type]], 'TsparseMatrix')
(whichIndex = which(bob@i == Dpar0 & bob@j == Dpar0)-1L)
(whichGlobal = which(adFun$hessians$map[[type]]$global==whichIndex)-1L)

(whichP = mapply(
  function(xx) min(which(adFun$hessian$map[[type]]$p >= xx)),
  xx = whichGlobal) -1L)


do.call(rbind, adFun$sparsity[[
  whichP[1] 
  ]][c('row_hess','col_hess')])
do.call(rbind, adlaplace$sparsity[[
  whichP[1] 
  ]][c('row_hess_inner', 'col_hess_inner')])
  

#Sgroups = length(adFun$sparsity)-12L
#Sgroups = seq.int(from=0, length.out = length(adFun$sparsity))

Sgroups = 376


parDf = data.frame(x)[,rep(1, Npar)]
Sx = seq(-0.1, 0.1, len=Npar) + parDf[Dpar,1]
SxD = Sx[-1] - diff(Sx)/2
parDf[Dpar,] = Sx


dens1 = mapply(
  adlaplace::jointLogDens,
  x = as.list(parDf),
  MoreArgs = list(backendContext = adFun, Sgroups = Sgroups),
  SIMPLIFY=TRUE
  )

plot(Sx, dens1)


grad1 = mapply(
  adlaplace::grad,
  x = as.list(parDf),
  MoreArgs = list(backendContext = adFun, inner=inner, Sgroups = Sgroups),
  SIMPLIFY=TRUE
  )
plot(Sx, grad1[Dpar,])
lines(SxD, diff(dens1)/diff(Sx))


hes1 = mapply(
  adlaplace::hess,
  x = as.list(parDf),
  MoreArgs = list(backendContext = adFun, inner=inner, Sgroups = Sgroups),
  SIMPLIFY=FALSE)

hes2 = do.call(abind::abind, c(
  lapply(hes1, as.matrix), along=3))

gradD = apply(grad1, 1, diff) / mean(diff(Sx))

Dpar2 = Dpar
plot(Sx, hes2[Dpar,Dpar2, ], ylab = paste(Dpar, Dpar2),
  ylim = range(hes2[Dpar, Dpar2,], na.rm=TRUE))
lines(SxD, gradD[,Dpar2])
```


## Notes

- `adlaplaceExample` provides model-specific C++ objective code.
- `adlaplace` provides shared optimization and Laplace tooling.
- For production use, optimize outer parameters with `adlaplace` wrappers and
  call `adlaplaceExample` for objective-specific derivatives.
