---
title: "Laplace Approximations with adlaplaceExample"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Laplace Approximations with adlaplaceExample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette mirrors the workflow in `adlaplace`, but uses the
`adlaplaceExample` skew-normal objective defined in
`src/objectiveFunction.cpp`.

Skew-normal contribution for one observation:

$$
\ell(y;\xi,\omega,\alpha)
= -\log\omega - \frac12\log(2\pi)
  -\frac12\left(\frac{y-\xi}{\omega}\right)^2
  + \log\left[\operatorname{erfc}\left(-\frac{\alpha(y-\xi)}{\sqrt{2}\,\omega}\right)\right].
$$



## Simulate data

```{r theData}
set.seed(1)

Nobs <- 20000
Nrandom1 <- 15
Nrandom2 <- 25

X <- Matrix::Matrix(cbind(1, rbinom(Nobs, 1, prob = 0.5)))
beta_true <- c(0.4, -0.3)

A1 <- Matrix::sparseMatrix(i = 1:Nobs, j = sample(Nrandom1, Nobs, replace = TRUE), x = 1)
A2 <- Matrix::sparseMatrix(i = 1:Nobs, j = sample(Nrandom2, Nobs, replace = TRUE), x = 1)
A <- cbind(A1, A2)

NperEffect <- c(Nrandom1, Nrandom2)
map <- Matrix::sparseMatrix(
  i = 0:(ncol(A) - 1L),
  j = rep(0:(length(NperEffect) - 1L), NperEffect),
  x = 1L,
  dims = c(ncol(A), length(NperEffect)),
  index1 = FALSE
)

sd_random <- c(0.25, 0.15)
gamma_true <- rnorm(ncol(A), sd = rep(sd_random, NperEffect))

omega_true <- 0.7
alpha_true <- 0.8


eta <- as.vector(X %*% beta_true + A %*% gamma_true)
y <- sn::rsn(Nobs, xi = eta, omega = omega_true, alpha=alpha_true)

data <- list(
  y = y,
  ATp = as(Matrix::t(A), "dMatrix"),
  XTp = as(Matrix::t(X), "CsparseMatrix"),
  map = map,
  Qdiag = rep(1, ncol(A))
)
```

## Configure model and build AD object

```{r config}
config <- list(
  beta = rep(0, nrow(data$XTp)),
  gamma = rep(0, nrow(data$ATp)),
  theta = c(log(0.2), log(0.2), log(0.5), 2),
  transform_theta = TRUE,
  groups = adlaplace::adFun_groups(500, data$ATp),
  num_threads = 1L,
  package = "adlaplaceExample",
  verbose = FALSE)
```

## Evaluate objective, gradient, and Hessian

```{r derivs}
adFun <- adlaplaceExample::getAdFun(data, config)
true_theta = c(log(c(sd_random, omega_true)), alpha_true)
x <- c(beta_true, gamma_true, true_theta)
x[length(x)] = -0.5

x2 = x;D=length(x)-3L
x2[D] = x[D]+2

Sgroups = seq.int(0L, length.out= length(adFun$sparsity))
#Sgroups = 0L

(val1 <- adlaplaceExample::jointLogDens(x, adFun))
adlaplaceExample::jointLogDens(x, adFun, Sgroups)

(val2 <- adlaplaceExample::jointLogDens(x2, adFun))
adlaplaceExample::jointLogDens(x2, adFun,Sgroups)

g <- adlaplaceExample::grad(x, adFun, inner = FALSE)
H <- adlaplaceExample::hess(x, adFun, inner = FALSE)

str(g)
str(H)
quantile(eigen(H)$values)
```


## Trust-optimization and Laplace log-likelihood interface

```{r trustOptimInterface, eval=FALSE}
adFun <- adlaplaceExample::getAdFun(data, config)

inner_res <- adlaplace::inner_opt(
  parameters = c(config$beta, config$theta),
  gamma = config$gamma,
  control = list(
    maxit = 5,
    report.level = 10,
    report.freq = 1
  ),
  config = config,
  adPack = adFun
)

str(inner_res)
quantile(inner_res$gradient)

res <- adlaplace::logLik(
  x = c(config$beta, config$theta),
  config = config,
  adPack = adFun)
str(res)
```


## Finite-difference check for one parameter

```{r finiteDiff, eval=FALSE}
config2 = config
#config2$gamma = inner_res$solution  
x <- c(
  config2$beta, 
  config2$gamma, 
  config2$theta)


adFun <- adlaplaceExample::getAdFun(data, config)

Dpar <- 3;Dpar2=15
Sgamma1 = seq.int(length(config$beta)+1, len = length(config$gamma))
DparG = match(Dpar, Sgamma1);DparG2 = match(Dpar2, Sgamma1)
Sx <- seq(-0.1, 0.1, length.out = 21) + x[Dpar]
x_list <- lapply(Sx, function(v) { xx <- x; xx[Dpar] <- v; xx })

fvals <- vapply(x_list, function(xx) adlaplaceExample::jointLogDens(xx, adFun), numeric(1))


stuff = adlaplace::inner_opt_test(adFun, config2)
stuff$hessian2 = Matrix::sparseMatrix(
  i = stuff$hessian$i, p = stuff$hessian$p, x = stuff$hessian$x,
  index1=FALSE, dims = rep(length(Sgamma1),2))


plot(Sx, fvals)
abline(h=-stuff$fval)

grad1 = mapply(
  adlaplaceExample::grad,
  x = x_list,
  MoreArgs = list(backendContext = adFun, inner=FALSE),
  SIMPLIFY=TRUE
  )

hes1 = mapply(
  adlaplace::hess,
  x = x_list,
  MoreArgs = list(backendContext = adFun, inner=FALSE),
  SIMPLIFY=FALSE)
do.call(rbind,lapply(hes1, function(xx) table(cut(eigen(xx)$values, c(-Inf, -1e-10, 1e-10, Inf)))  ))
hes2 = do.call(abind::abind, c(
  lapply(hes1, as.matrix), along=3))
gradD = apply(grad1, 1, diff) / mean(diff(Sx))

Sx_mid <- Sx[-1] - diff(Sx) / 2
plot(Sx, grad1[Dpar,], ylab = "Analytic gradient", xlab = "parameter value")
lines(Sx_mid, diff(fvals) / diff(Sx), col = 2)
abline(h=-stuff$gradient[DparG])
legend("topleft", legend = c("AD gradient", "finite difference"), col = c(1, 2), lty = 1, bty = "n")


plot(Sx, hes2[Dpar,Dpar2, ], ylab = paste(Dpar, Dpar2),
  ylim = range(hes2[Dpar, Dpar2,], na.rm=TRUE))
lines(Sx_mid, gradD[,Dpar2], lwd=3, col='grey')
abline(h=-stuff$hessian2[DparG, DparG2], col='blue')


bob = stuff$hessian2 + hes2[Sgamma1, Sgamma1, 10]
```

## Notes

- `adlaplaceExample` provides model-specific C++ objective code.
- `adlaplace` provides shared optimization and Laplace tooling.
- For production use, optimize outer parameters with `adlaplace` wrappers and
  call `adlaplaceExample` for objective-specific derivatives.




```{r derivJointDens, eval=FALSE}
config$gamma = rep(1, length(config$gamma))
x = c(config$beta, config$gamma, config$theta)
adFun = adlaplace::getAdFun(data, modifyList(config, list(verbose=TRUE)))

inner = FALSE
type = c('outer','inner')[1+inner]


Sgamma1 = seq.int(length(config$beta)+1, length.out = length(config$gamma))

Npar=13;Dpar = length(x)-1
if(inner) {
  Dpar = match(Dpar, Sgamma1)
}
Dpar0=Dpar-1L

bob = as(adFun$hessians$hessian[[type]], 'TsparseMatrix')
(whichIndex = which(bob@i == Dpar0 & bob@j == Dpar0)-1L)
(whichGlobal = which(adFun$hessians$map[[type]]$global==whichIndex)-1L)

(whichP = mapply(
  function(xx) min(which(adFun$hessian$map[[type]]$p >= xx)),
  xx = whichGlobal) -1L)



do.call(rbind, adFun$sparsity[[
  whichP[1] 
  ]][c('row_hess','col_hess')])
do.call(rbind, adFun$sparsity[[
  whichP[1] 
  ]][c('row_hess_inner', 'col_hess_inner')])
  

Sgroups = length(adFun$sparsity)-12L
#Sgroups = seq.int(from=0, length.out = length(adFun$sparsity))

#Sgroups = c()


parDf = data.frame(x)[,rep(1, Npar)]
Sx = seq(-0.1, 0.1, len=13) + parDf[Dpar,1]
SxD = Sx[-1] - diff(Sx)/2
parDf[Dpar,] = Sx


dens1 = mapply(
  adlaplace::jointLogDens,
  x = as.list(parDf),
  MoreArgs = list(backendContext = adFun, Sgroups = Sgroups),
  SIMPLIFY=TRUE
  )

plot(Sx, dens1)


grad1 = mapply(
  adlaplace::grad,
  x = as.list(parDf),
  MoreArgs = list(backendContext = adFun, inner=inner, Sgroups = Sgroups),
  SIMPLIFY=TRUE
  )
plot(Sx, grad1[Dpar,])
lines(SxD, diff(dens1)/diff(Sx))


hes1 = mapply(
  adlaplace::hess,
  x = as.list(parDf),
  MoreArgs = list(backendContext = adFun, inner=inner, Sgroups = Sgroups),
  SIMPLIFY=FALSE)

hes2 = do.call(abind::abind, c(
  lapply(hes1, as.matrix), along=3))

gradD = apply(grad1, 1, diff) / mean(diff(Sx))

Dpar2 = Dpar
plot(Sx, hes2[Dpar,Dpar2, ], ylab = paste(Dpar, Dpar2),
  ylim = range(hes2[Dpar, Dpar2,], na.rm=TRUE))
lines(SxD, gradD[,Dpar2])
```






