---
title: "Testing adlaplace with Skew Normal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing adlaplace with Skew Normal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# To make a package using adlaplace

- copy `configure`, `src/Makevars.in`, and `src/Rinterfaces.cpp` from adlaplace source (or this package)
- copy `R/000.R`, change the package name on the first line
- your `DESCRIPTION` file needs
```
SystemRequirements: CppAD, OpenMP
License: GPL-3
Depends: R (>= 4.1.0)
LinkingTo: Rcpp, adlaplace, RcppEigen, trustOptim
Imports: Rcpp
```
- copy `src/objectiveFunction.cpp` and modify it as necessary
- run `Rcpp::compileAttributes` and `roxygen2::roxygenize(... , load = 'source')`

## System requirements

- cppad
- omp (optional)

# Skew normal model


$$
\begin{aligned}
f(y;\xi,\omega,\alpha)
&= -\log\omega
-\frac12\log(2\pi)
-\frac12\left(\frac{y-\xi}{\omega}\right)^2
+\log\!\left[\operatorname{erfc}\!\left(-\frac{\alpha (y-\xi)}{\sqrt2\,\omega}\right)\right].
\end{aligned}
$$

```{r theData}
Nobs = 10000;Nrandom1=10;Nrandom2=25

set.seed(0)
X = Matrix::Matrix(cbind(1, rbinom(Nobs,1, prob= 0.5)))
beta = rep(1, ncol(X))

AmatList = list(
	Matrix::sparseMatrix(
		i = 1:Nobs,
		j= sample(Nrandom1, Nobs, replace=TRUE)
	),
	Matrix::sparseMatrix(
		i = 1:Nobs,
		j= sample(Nrandom2, Nobs, replace=TRUE)
	)
)
Amat = do.call(cbind, AmatList)
NperEffect = unlist(lapply(AmatList, ncol))
map = Matrix::sparseMatrix(
	i = seq(0L, len=ncol(Amat)),
	j = rep(seq(0L, len=length(NperEffect)), NperEffect),
	x = 1L,
	index1=FALSE,
	dims = c(ncol(Amat), length(NperEffect))
	)

sdRandomOrig = rep(0.1, length(AmatList));sdRandomLog = log(sdRandomOrig)
gamma = rnorm(ncol(Amat), sd=rep(sdRandomOrig, NperEffect))

obsPar = list(omega = 0.1, alpha = 2)
eta = as.vector(X %*% beta + Amat %*% gamma)
y = sn::rsn(length(eta), xi=eta, omega=obsPar$omega, alpha = obsPar$alpha)


data = list(y = y, 
	ATp = as(Matrix::t(Amat), 'dMatrix'),
	XTp = as(Matrix::t(X), 'CsparseMatrix'),
	map = map, 
	Qdiag = rep(1, ncol(Amat)))
```



- starting values for theta.  SD's for random effects come first, parameters for observation distribution after
- starting values for beta
- starting values for the random effects
- is theta on the log scale?
- create groups (for which the likelihoods are differentiated separately) based on the columns of the random effects design matrix
- number of parallel cores.
- name of the R package containing the model

```{r exampleConfig}
config = list(
# theta is sd1 sd2, omega, alpha, first 3 logged
	theta = c(rep(-2.3, length(NperEffect)), -2.3, 0.5), 
	beta = rep(1, nrow(data$XTp)), 
	start_gamma = rep(0, nrow(data$ATp)),
	transform_theta=TRUE, 
	groups = adlaplace::adFun_groups(1000, data$ATp),
	num_threads=1,#parallel::detectCores(),
	package='adlaplaceExample', verbose=FALSE)	
```

find the sparsity structure of each group.

```{r sparsity}
sparsity_list = adlaplace::group_sparsity(data=data, config=config)
names(sparsity_list)
length(sparsity_list$hessian_inner)
config = modifyList(config, sparsity_list)
```

create autodiff functions which include model parameters as well as random effects
create AD function

```{r adFunFull}
adFunFull = adlaplaceExample::getAdFun(
	data, modifyList(config, list(verbose=TRUE)), inner=FALSE)
```

fit

```{r outerOpt}
cache = new.env()
assign('start_gamma', config$start_gamma, cache)

ulim = c(rep(Inf, length(config$beta) + length(config$theta) -1), 5)

mle = stats::optim(
	rep(0, length(c(config$beta, config$theta))),
	adlaplace::outer_fn, 
	adlaplace::outer_gr,
	method='L-BFGS-B',
	lower = -ulim, upper = ulim,
	data=data, config=config, cache=cache, adPack = adFunFull,
	control_inner = list(),
	control = list(trace=1)
	)
mle$par
exp(mle$par[-c(1:length(beta), length(mle$par))])
```



Check derivative of likelihood


```{r checkDerivOuter}
x = mle$par #c(config$beta, config$theta)

Npar=25
Dpar = length(x)
parDf = data.frame(x)[,rep(1, Npar)]
Sx = seq(-0.5, 0.5, len=Npar) + parDf[Dpar,1]
parDf[Dpar,] = Sx
SxD = Sx[-1] - diff(Sx)/2

adFunFull = adlaplaceExample::getAdFun(
	data, config, inner=FALSE)

res = mapply(
  adlaplace::logLik,
  x = as.list(parDf),
  MoreArgs = list(config=config, data=data, adPack = adFunFull),
  SIMPLIFY=FALSE
  )
dens1 = unlist(lapply(res, function(xx) xx$inner$minusLogLik))
grad1 = unlist(lapply(res, function(xx) xx$dLogLik[Dpar]))

plot(Sx, dens1, ylab='-log lik')

plot(Sx, grad1, ylab = 'deriv')
lines(SxD, diff(dens1)/diff(Sx))
```



evaluate the likelihood and derivatives


```{r checkDerivInner}
x =c(mle$par[1:length(config$beta)], cache$start_gamma, mle$par[-(1:length(config$beta))])

Npar=13
Dpar = length(x)
parDf = data.frame(x)[,rep(1, Npar)]
Sx = seq(-0.5, 0.5, len=Npar) + parDf[Dpar,1]
parDf[Dpar,] = Sx
SxD = Sx[-1] - diff(Sx)/2


dens1 = mapply(
  adlaplaceExample::jointLogDens,
  parameters = as.list(parDf),
  MoreArgs = list(config=config, adPack = adFunFull)
  )
grad1 = mapply(
  adlaplaceExample::grad,
  parameters = as.list(parDf),
  MoreArgs = list(config=config, adPack = adFunFull)
  )[Dpar,]
hes1 = mapply(
  adlaplaceExample::hessian,
  parameters = as.list(parDf),
  MoreArgs = list(config=config, adPack = adFunFull),
  SIMPLIFY=FALSE
  )
hes1a = unlist(lapply(hes1, '[', Dpar, Dpar))

plot(Sx, dens1)

plot(Sx, grad1)
lines(SxD, diff(dens1)/diff(Sx))

plot(Sx, hes1a)
lines(SxD, diff(grad1)/diff(Sx))
```

evalute the density in R and compare

```{r checkDens}
omega = exp(config$theta[length(config$theta)-1])
alpha = config$theta[length(config$theta)]
sdVec = as.vector(data$map %*% exp(config$theta[
	seq(from=1, len=ncol(data$map))])) 
x = c(config$beta, gamma, config$theta)

# log density
sum(sn::dsn(
	x = data$y, 
	xi = eta, 
	omega  = omega,
	 alpha = alpha,
	log=TRUE)) + sum(dnorm(
		gamma, 
		sd = sdVec,
		log=TRUE
		))

# negative log density
adlaplaceExample::jointLogDens(
	x,
	config = modifyList(config, list(verbose=TRUE)), 
	adPack = adFunFull)

# parts of log density
toSum = c(
	omegabit = - length(data$y) * log(omega),
	twopibit = -length(data$y) * 0.5*log(2*pi),
	normalbit = sum(-0.5*( (y - eta)/omega)^2),
	erfc = sum(log(gsl::erfc(- alpha * (
	data$y - eta
	) / (sqrt(2) * omega)
	))),
	random = sum(dnorm(
		gamma, 
		sd = sdVec,
		log=TRUE
		)))
# verbose output
adlaplaceExample::jointLogDensNoAdfun(
	x,
	config = modifyList(config, list(verbose=TRUE)), 
	data = data)
toSum
# extra
sum(toSum[c('omegabit','twopibit')])
# data part
sum(toSum[c('normalbit','erfc')])
toSum['random']
```





