---
title: "Besag-York-Mollie disease mapping: Germany oral cavity cancer"
author: "Patrick Brown"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
---

## Introduction

This example reproduces the Germany oral-cavity-cancer data distributed with
**INLA** (Besag, York, and Mollié 1991; Rue, Martino, and Chopin 2009). Counts
$y_i$ in 544 districts follow $y_i \sim \mathrm{Poisson}(E_i e^{\eta_i})$ with a
Besag-York-Mollie (BYM) spatial random effect: a structured intrinsic
conditional autoregressive (ICAR) component plus an unstructured iid component.

The example uses the low-level `adlaplace` pipeline and the `random_mult`
kernel. The generalized log-determinant and rank of the ICAR precision are
computed in R and passed as constants; only the precision multipliers enter
the AD tape (Sørbye and Rue 2014). The intercept is omitted because the ICAR
precision is rank-deficient.


## ICAR precision

```{r precision}
data(Germany, package = "INLA")
g_file <- system.file("demodata/germany.graph", package = "INLA")
graph <- INLA::inla.read.graph(g_file)
adj <- INLA::inla.graph2matrix(graph)
adj <- as(adj, "CsparseMatrix")

deg <- Matrix::rowSums(adj)
Q_icar <- Matrix::Diagonal(nrow(adj), deg) - adj
Q_icar <- as(as(Q_icar, "generalMatrix"), "CsparseMatrix")

Q_scaled <- INLA::inla.scale.model(
  Q_icar,
  constr = list(A = matrix(1, 1, nrow(adj)), e = 0)
)
Q_scaled <- as(as(Q_scaled, "generalMatrix"), "CsparseMatrix")

ev <- eigen(as.matrix(Q_scaled), symmetric = TRUE, only.values = TRUE)$values
ev <- sort(ev, decreasing = TRUE)
log_det_gen <- sum(log(ev[seq_len(nrow(adj) - 1L)]))
prec <- list(Q = Q_scaled, log_det = log_det_gen, rank = nrow(adj) - 1L)
```

## adlaplace shards

Latent vector $\gamma = (u_{\mathrm{struct}}, u_{\mathrm{iid}})$ has length
$2n$. Observations use `poisson_obs` with offset $\log E_i$.

```{r shards}
n <- nrow(Germany)
n_theta <- 2L
n_gamma <- 2L * n

A_obs <- cbind(
  Matrix::sparseMatrix(i = seq_len(n), j = seq_len(n), x = 1, dims = c(n, n)),
  Matrix::sparseMatrix(i = seq_len(n), j = seq_len(n), x = 1, dims = c(n, n))
)
X_obs <- Matrix::Matrix(0, n, 0)

config <- list(
  transform_theta = TRUE,
  gamma = rep(0, n_gamma),
  theta = log(c(0.5, 0.1)),
  offset = log(Germany$E),
  shards = adlaplace:::ad_shards(A_obs, num_shards = 2L),
  verbose = FALSE
)

data_obs <- adlaplace:::ad_data(
  y = as.numeric(Germany$Y),
  A = A_obs,
  X = X_obs,
  theta_map = n_theta,
  ad_kind = "observations",
  ad_fun = "poisson_obs"
)

model_struct <- adlaplace:::ad_data(
  gamma_map = Matrix::sparseMatrix(
    i = seq_len(n), j = seq_len(n), x = 1, dims = c(n_gamma, n)
  ),
  theta_map = c(1L, n_theta),
  ad_kind = "random",
  ad_fun = "random_mult",
  precision = prec
)

model_iid <- adlaplace:::ad_data(
  gamma_map = Matrix::sparseMatrix(
    i = seq.int(n + 1L, length.out = n), j = seq_len(n), x = 1,
    dims = c(n_gamma, n)
  ),
  theta_map = c(2L, n_theta),
  ad_kind = "random",
  ad_fun = "random_diagonal",
  precision = rep(1, n)
)
```

```{r ad_pack}
config$verbose <- TRUE
ptr_obs <- adlaplace:::ad_fun_ptr(data = data_obs, config = config)
# this part takes a few minutes
ptr_iid <- adlaplace:::ad_fun_ptr(data = model_iid, config = config)
ptr_struct <- adlaplace:::ad_fun_ptr(data = model_struct, config = config)
ad_pack <- adlaplace:::ad_fun(c(ptr_obs, ptr_struct, ptr_iid), num_threads = 4L)
```

## Outer optimization

```{r optim}
cache <- new.env(parent = emptyenv())
cache$gamma <- rep(0, n_gamma)

opt <- optim(
  par = config$theta,
  fn = adlaplace:::outer_fn, gr = adlaplace:::outer_gr,
  method = "L-BFGS-B",
  lower = rep(-6, n_theta), upper = rep(4, n_theta),
  control = list(maxit = 200L, factr = 1e3, trace = 3, REPORT = 1),
  config = modifyList(config, list(verbose = FALSE)),
  ad_fun = ad_pack, cache = cache,
  control_inner = list(maxit = 100L, report.level = 0, report.freq = 0)
)

res <- adlaplace:::log_lik_laplace(
  x = opt$par, gamma = cache$gamma,
  ad_fun = ad_pack, config = config, control = control_inner, deriv = FALSE
)
u_hat <- res$opt$solution
eta_adl <- u_hat[seq_len(n)] + u_hat[seq.int(n + 1L, length.out = n)]
sd_adl <- exp(opt$par)
sd_adl
```

## INLA reference

```{r inla}
Germany$region.struct <- Germany$region
Germany$region.iid <- Germany$region
inla.setOption(num.threads = "1:1")
inla_fit <- inla(
  Y ~ -1 +
    f(region.struct,
      model = "besag", graph = g_file, scale.model = TRUE,
      constr = TRUE
    ) +
    f(region.iid, model = "iid"),
  family = "poisson", data = Germany, E = E,
  control.compute = list(return.marginals = FALSE)
)
eta_inla <- inla_fit$summary.random$region.struct$mean +
  inla_fit$summary.random$region.iid$mean
```

## Hyperparameter comparison

```{r hyper}
comparison_hyper <- data.frame(
  Component = c("SD structured (ICAR)", "SD unstructured (iid)"),
  adlaplace = sd_adl,
  INLA = c(
    1 / sqrt(inla_fit$summary.hyperpar[
      "Precision for region.struct", "mean"
    ]),
    1 / sqrt(inla_fit$summary.hyperpar[
      "Precision for region.iid", "mean"
    ])
  )
)
comparison_hyper
cor_eta <- cor(eta_adl, eta_inla)
cor_eta
```

## Maps

```{r maps, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1), mgp = c(2.2, 0.7, 0))
plot(seq_len(n), eta_adl,
  type = "h", xlab = "district", ylab = "effect",
  main = "adlaplace"
)
plot(seq_len(n), eta_inla,
  type = "h", xlab = "district", ylab = "effect",
  main = "INLA"
)
```

```{r scatter, fig.height=4}
par(mfrow = c(1, 1), mar = c(4, 4, 2, 1), mgp = c(2.2, 0.7, 0))
rng <- range(c(eta_adl, eta_inla))
plot(eta_inla, eta_adl,
  xlim = rng, ylim = rng, pch = 16, cex = 0.5,
  col = "#00000080", xlab = "INLA posterior mean",
  ylab = "adlaplace posterior mode",
  main = sprintf("Total spatial effect (cor = %.3f)", cor_eta)
)
abline(0, 1, col = "red", lty = 2)
```

The structured (ICAR) component dominates; the unstructured component is near
zero in both fits. Small differences reflect the distinct inferential targets:
`adlaplace` reports the Laplace-approximate MLE and conditional mode; `INLA`
the posterior mean under its default priors.
