---
title: "Loaloa prevalence with FEM Matérn"
author: "Patrick Brown"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 2
    html-math-method: mathml
vignette: >
  %\VignetteIndexEntry{Loaloa prevalence with FEM Matérn}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
has_geostatsp <- requireNamespace("geostatsp", quietly = TRUE) &&
  requireNamespace("terra", quietly = TRUE)
```

```{r pkgs}
library(adlaplace)
library(adlaplaceGrf)
```

This vignette fits a binomial geostatistical model for loaloa prevalence
using a village-level IID effect and a FEM Matérn spatial field
(`degree = 3`, α = 3), matching the Matérn shape used in geostatsp
examples.

```{r skip-check}
#| echo: false
if (!has_geostatsp) {
  knitr::knit_exit()
}
```

## Data

```{r data}
library(geostatsp)
library(terra)

data(loaloa)
loaloa <- unwrap(loaloa)
elev_r <- unwrap(elevationLoa)
evi_r <- unwrap(eviLoa)

dat <- as.data.frame(loaloa)
dat$elev <- as.numeric(terra::extract(elev_r, loaloa)[, 2L])
dat$evi <- as.numeric(terra::extract(evi_r, loaloa)[, 2L])
dat$villageID <- factor(dat$villageID)
# Scale covariates for numerical stability
dat$elev_s <- as.numeric(scale(dat$elev))
dat$evi_s <- as.numeric(scale(dat$evi))

c(n = nrow(dat), mean_prev = mean(dat$y / dat$N))
```

## Knot lines

Use a coarse knot grid over a buffered region around the villages.
(`geostatsp` and `terra` mask `matern` / `crds`, so call them with `::`.)

```{r knots}
xy <- adlaplaceGrf::crds(loaloa)
pad <- 2e4
xlim <- range(xy[, 1L]) + c(-pad, pad)
ylim <- range(xy[, 2L]) + c(-pad, pad)
knots_list <- list(
  x = seq(xlim[1L], xlim[2L], length.out = 8L),
  y = seq(ylim[1L], ylim[2L], length.out = 8L)
)
```

## Fit

geostatsp Matérn **shape = 2** corresponds to ν = 2 and SPDE α = 3 in
two dimensions, so we use `degree = 3` / `random_fem_3`.

```{r fit}
#| cache: false
fit <- adlaplace(
  binomial(y, size = N) ~
    elev_s + evi_s +
    iid(villageID, init = 0.5) +
    adlaplaceGrf::matern(
      x = loaloa, knots = knots_list, degree = 3L, init = c(1e-4, 1)
    ),
  data = dat,
  num_shards = 10L,
  control = list(maxit = 200L)
)
fit$optim$convergence
coef(fit)
```

```{r summary}
logLik(fit)
```
