---
title: "Loaloa prevalence with FEM Matérn"
author: "Patrick Brown"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 2
    html-math-method: mathml
vignette: >
  %\VignetteIndexEntry{Loaloa prevalence with FEM Matérn}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
has_geostatsp <- requireNamespace("geostatsp", quietly = TRUE) &&
  requireNamespace("terra", quietly = TRUE)
has_INLA <- has_geostatsp && requireNamespace("INLA", quietly = TRUE)
```

```{r pkgs}
library(adlaplace)
library(adlaplaceGrf)
```

This vignette fits a binomial geostatistical model for loaloa prevalence
using a village-level IID effect and a FEM Matérn spatial field
(`shape = 2`, ν = 2), matching the Matérn shape used in geostatsp
examples. When **INLA** is available, the same model is also fit with
`geostatsp::glgm` for comparison.

```{r skip-check}
#| echo: false
if (!has_geostatsp) {
  knitr::knit_exit()
}
```

## Data

```{r data}
data("loaloa", package = "geostatsp")

loaloa <- terra::unwrap(loaloa)
elev_r <- terra::unwrap(elevationLoa)
evi_r <- terra::unwrap(eviLoa)

elev_s_r <- (elev_r - 500) / 100
evi_s_r <- (
  evi_r - mean(terra::values(evi_r), na.rm = TRUE)
) / sd(terra::values(evi_r), na.rm = TRUE)

loaloa_for_fit <- geostatsp:::gm.dataSpatial(
  formula = y ~ elev + evi,
  data = loaloa,
  covariates = list(elev = elev_s_r, evi = evi_s_r),
  grid = geostatsp::squareRaster(loaloa, cells = 200, buffer = 1e4)
)
```

## Knot lines

Use a coarse knot raster over a buffered region around the villages; pass it
directly as `knots` to `matern()` (cell centers become the knot lines).
(`geostatsp` / `terra` may mask `matern`, so call it with `::`.)



## Fit

geostatsp Matérn **shape = 2** is ν = 2 (SPDE α = 3 in 2D), so we pass
`shape = 2` to `matern()` (cubic B-splines / `random_fem_3`). Locations come
from the `geometry` column (WKT) in the model data frame.

```{r model_setup}
#| cache: false
knots_grid <- geostatsp::squareRaster(loaloa, cells = 12, buffer = 500 * 1000)
loaloa_df <- as.data.frame(loaloa_for_fit$data, geom = "WKT")

loaloa_model <- adlaplace::binomial(y, size = N) ~
  elev + evi +
  adlaplace::iid(villageID, init = 0.2) +
  adlaplaceGrf::matern(
    geometry,
    knots = knots_grid,
    shape = 2L,
    # range = practical rho = sqrt(8*nu)/kappa (m); sd = field SD
    init = c(100 * 1000, 0.2),
    lower = c(1 * 1000, 0.001)
  )
```

Compare `inner_opt` (trustOptim, AD derivatives) to a derivative-free
`optim(..., method = "Nelder-Mead")` on $\gamma$ only, with $\beta$ and $\theta$
held fixed at their inits. Nelder-Mead starts from the `inner_opt` mode.
Diagnostic chunks below are `eval: false` by default (slow); set them to
`true` locally when debugging.

```{r inner-vs-nm}
#| cache: false
#| eval: false
md <- adlaplace::model_data(
  formula = loaloa_model,
  data = loaloa_df,
  verbose = TRUE
)
config <- list(
  transform_theta = TRUE,
  verbose = TRUE,
  num_threads = 1L,
  shards = adlaplace::ad_shards(md$data$A, num_shards = 20L)
)
af <- adlaplace::ad_fun(md, config, num_threads = config$num_threads)
adlaplace::n_groups(af@ptr)

n_beta <- length(md$data$info$beta$init)
n_gamma <- nrow(md$data$info$gamma)
n_theta <- nrow(md$data$info$theta)

# outer parameters on the optimization scale (theta already logged in info$parameters)
par_outer <- md$data$info$parameters$init
beta <- par_outer[seq_len(n_beta)]
theta <- par_outer[seq.int(n_beta + 1L, length.out = n_theta)]
gamma0 <- rep(0, n_gamma)

make_x <- function(gamma) {
  c(beta, gamma, theta)
}

inner_res <- adlaplace::inner_opt(
  parameters = par_outer,
  gamma = gamma0,
  ad_fun = af,
  control = list(maxit = 500L, report.level = 1, report.freq = 10),
  deriv = FALSE,
  verbose = TRUE
)

g_hat_inner <- as.numeric(inner_res$opt$solution)

# minimize -log joint dens over gamma; start from inner_opt mode
obj_nm <- function(gamma) {
  adlaplace::joint_log_dens(af, make_x(gamma), negative = TRUE)
}

nm_res <- stats::optim(
  par = g_hat_inner,
  fn = obj_nm,
  method = "Nelder-Mead",
  control = list(maxit = 200L, reltol = 1e-10, trace = 1, REPORT = 50)
)

g_nm <- as.numeric(nm_res$par)

f_inner <- adlaplace::joint_log_dens(af, make_x(g_hat_inner), negative = TRUE)
f_nm <- adlaplace::joint_log_dens(af, make_x(g_nm), negative = TRUE)

data.frame(
  method = c("inner_opt", "Nelder-Mead"),
  neg_log_dens = c(f_inner, f_nm),
  log_dens = c(-f_inner, -f_nm),
  opt_value = c(inner_res$opt$fval, nm_res$value),
  convergence = c(NA_integer_, nm_res$convergence),
  max_abs_gamma_diff = c(NA_real_, max(abs(g_hat_inner - g_nm))),
  cor_gamma = c(NA_real_, stats::cor(g_hat_inner, g_nm))
)

c(
  inner_opt_log_lik = inner_res$log_lik,
  max_abs_gamma_diff = max(abs(g_hat_inner - g_nm)),
  rms_gamma_diff = sqrt(mean((g_hat_inner - g_nm)^2))
)
```

Inner (γ-block) gradient and Hessian FD checks at the **Nelder-Mead** γ estimate
(outer $\beta$, $\theta$ fixed).

```{r debug}
#| cache: false
#| eval: false
#| fig.height: 5

gamma_check <- c(2L, 4L)

# base point: Nelder-Mead <U+03B3> with fixed outer parameters
x_here <- make_x(g_nm)

idx_g <- n_beta + gamma_check
Dpar <- idx_g[length(idx_g)]

Ngrid <- 7L
par_grid <- matrix(x_here, nrow = Ngrid, ncol = length(x_here), byrow = TRUE)
Sx <- x_here[Dpar] + seq(-0.15, 0.15, length.out = Ngrid)
SxD <- Sx[-1] - diff(Sx) / 2
par_grid[, Dpar] <- Sx
x_list <- split(par_grid, row(par_grid))

dens <- vapply(
  x_list,
  adlaplace::joint_log_dens,
  numeric(1),
  ad_fun = af,
  negative = FALSE
)

g_ad_scan <- do.call(
  cbind,
  lapply(
    x_list,
    adlaplace::grad,
    ad_fun = af,
    inner = TRUE,
    negative = FALSE
  )
)

i_mid <- (Ngrid + 1L) %/% 2L
eps <- 1e-5
x_mid <- x_list[[i_mid]]
g_ad <- as.numeric(adlaplace::grad(
  af, x_mid,
  inner = TRUE, negative = FALSE
))
g_fd <- vapply(idx_g, function(i) {
  xp <- x_mid
  xm <- x_mid
  xp[i] <- xp[i] + eps
  xm[i] <- xm[i] - eps
  (adlaplace::joint_log_dens(af, xp, negative = FALSE) -
    adlaplace::joint_log_dens(af, xm, negative = FALSE)) / (2 * eps)
}, numeric(1))
grad_err <- abs(g_ad[idx_g] - g_fd)
c(
  n_gamma = n_gamma,
  dens_mid = dens[i_mid],
  max_abs_inner_grad_err = max(grad_err),
  which_gamma = gamma_check[which.max(grad_err)]
)
rbind(ad = g_ad[idx_g], fd = g_fd)

H_ad <- as.matrix(adlaplace::hessian(
  af, x_mid,
  inner = TRUE, negative = FALSE
))
H_fd <- matrix(NA_real_, length(idx_g), length(idx_g))
for (a in seq_along(idx_g)) {
  for (b in seq_along(idx_g)) {
    ia <- idx_g[a]
    ib <- idx_g[b]
    xpp <- x_mid
    xpm <- x_mid
    xmp <- x_mid
    xmm <- x_mid
    xpp[ia] <- xpp[ia] + eps
    xpp[ib] <- xpp[ib] + eps
    xpm[ia] <- xpm[ia] + eps
    xpm[ib] <- xpm[ib] - eps
    xmp[ia] <- xmp[ia] - eps
    xmp[ib] <- xmp[ib] + eps
    xmm[ia] <- xmm[ia] - eps
    xmm[ib] <- xmm[ib] - eps
    H_fd[a, b] <- (
      adlaplace::joint_log_dens(af, xpp, negative = FALSE) -
        adlaplace::joint_log_dens(af, xpm, negative = FALSE) -
        adlaplace::joint_log_dens(af, xmp, negative = FALSE) +
        adlaplace::joint_log_dens(af, xmm, negative = FALSE)
    ) / (4 * eps * eps)
  }
}
H_ad_sub <- H_ad[idx_g, idx_g, drop = FALSE]
list(
  H_ad = H_ad_sub,
  H_fd = H_fd,
  max_abs_hess_err = max(abs(H_ad_sub - H_fd))
)

par(mfrow = c(2, 2), mar = c(2, 2, 2, 0), mgp = c(1, 0.5, 0))
ylab_g <- paste0("gamma[", gamma_check[length(gamma_check)], "]")
plot(Sx, dens, type = "b", xlab = ylab_g, ylab = "joint log dens")
plot(Sx, g_ad_scan[Dpar, ], type = "l", xlab = ylab_g, ylab = "inner AD grad")
points(SxD, diff(dens) / diff(Sx), pch = 16)
legend("topright", legend = c("AD", "finite diff"), lty = c(1, NA), pch = c(NA, 1), bty = "n")

hes_scan <- vapply(seq_len(Ngrid), function(i) {
  Hi <- as.matrix(adlaplace::hessian(
    af, x_list[[i]],
    inner = TRUE, negative = FALSE
  ))
  Hi[Dpar, Dpar]
}, numeric(1))
g_slope <- diff(g_ad_scan[Dpar, ]) / diff(Sx)
plot(
  Sx, hes_scan,
  type = "l", xlab = ylab_g,
  ylab = paste0("H_inner[", Dpar, ",", Dpar, "]"),
  ylim = range(c(hes_scan, g_slope), na.rm = TRUE)
)
points(SxD, g_slope, pch = 16)

if (length(idx_g) >= 2L) {
  i1 <- idx_g[1L]
  i2 <- Dpar
  hes_cross <- vapply(seq_len(Ngrid), function(i) {
    Hi <- as.matrix(adlaplace::hessian(
      af, x_list[[i]],
      inner = TRUE, negative = FALSE
    ))
    Hi[i1, i2]
  }, numeric(1))
  g1_slope <- diff(g_ad_scan[i1, ]) / diff(Sx)
  plot(
    Sx, hes_cross,
    type = "l", xlab = ylab_g,
    ylab = paste0("H_inner[", i1, ",", i2, "]"),
    ylim = range(c(hes_cross, g1_slope), na.rm = TRUE)
  )
  points(SxD, g1_slope, pch = 16)
}
```

```{r fit}
fit <- adlaplace::adlaplace(
  loaloa_model,
  data = loaloa_df,
  num_shards = 20L,
  control = list(maxit = 200, trace = 3, REPORT = 1),
  config = list(num_threads = 1L),
  verbose = TRUE
)
fit$optim$convergence
coef(fit)
```

```{r fit-glgm, eval=has_INLA, cache=FALSE}
loaFit <- geostatsp::glgm(
  y ~ elev_s + evi_s +
    f(
      villageID,
      prior = "pc.prec",
      param = c(1, 0.5),
      model = "iid"
    ),
  loaloa,
  grid = 100,
  covariates = list(elev_s = elev_s_r, evi_s = evi_s_r),
  family = "binomial",
  Ntrials = loaloa$N,
  shape = 2,
  buffer = 25000,
  prior = list(sd = 1, range = 500 * 1000),
  control.inla = list(strategy = "gaussian")
)
```

## Summary

```{r summary}
logLik(fit)

cf <- coef(fit)
ad_pars <- c(
  intercept = unname(cf[["intercept"]]),
  elev = unname(cf[["elev_linear"]]),
  evi = unname(cf[["evi_linear"]]),
  village_sd = unname(cf[["villageID_iid"]]),
  # matern: practical range rho = sqrt(8*nu)/kappa (m); sd = field SD
  range = unname(cf[["geometry_matern_log_range"]]),
  sd = unname(cf[["geometry_matern_log_sd"]])
)
ad_pars
```

```{r summary-compare, eval=has_INLA}
if (length(loaFit$parameters)) {
  glgm_sum <- loaFit$parameters$summary
  knitr::kable(glgm_sum[, c("mean", "0.025quant", "0.975quant")], digits = 3)

  # Map ad_pars names to rows of loaFit$parameters$summary
  row_for <- function(...) {
    pats <- c(...)
    hits <- unique(unlist(lapply(pats, function(p) {
      grep(p, rownames(glgm_sum), ignore.case = TRUE)
    })))
    if (!length(hits)) {
      return(NA_character_)
    }
    rownames(glgm_sum)[hits[1L]]
  }
  map <- c(
    intercept = row_for("^\\(Intercept\\)$", "^Intercept$"),
    elev = row_for("^elev"),
    evi = row_for("^evi"),
    village_sd = row_for("^sd villageID$", "sd.?villageID", "villageID.*[Ss]d"),
    range = row_for("^range/1000$", "^range$"),
    sd = row_for("^sd$", "SD for space")
  )

  glgm_pars <- setNames(
    vapply(map, function(nm) {
      if (is.na(nm)) {
        return(NA_real_)
      }
      as.numeric(glgm_sum[nm, "mean"])
    }, numeric(1)),
    names(map)
  )
  # summary may store range in km as "range/1000"
  if (!is.na(map[["range"]]) && identical(map[["range"]], "range/1000")) {
    glgm_pars[["range"]] <- glgm_pars[["range"]] * 1000
  }

  compare <- rbind(
    adlaplace = ad_pars[names(map)],
    glgm = glgm_pars
  )
  knitr::kable(compare, digits = 3)
}
```

## Estimated field

```{r est}
my_sim <- adlaplaceGrf:::matern_est(fit, loaloa_for_fit$grid, n = 10)
```

Posterior mean and SD of the spatial field from both fits (glgm when available):

```{r plot-surfaces}
#| fig.width: 8
#| fig.height: 8
my_col <- mapmisc::colourScale(my_sim[[grep("sd", names(my_sim), invert = TRUE)]], style = "equal", digits = 2, breaks = 9, col = "Spectral")
sd_col <- mapmisc::colourScale(my_sim[[grep("sd", names(my_sim))]], style = "equal", digits = 2, breaks = 9)
if (has_INLA && exists("loaFit") && length(loaFit$parameters)) {
  old_par <- graphics::par(mfrow = c(2, 2), mar = c(2, 2, 2, 4))
  on.exit(graphics::par(old_par), add = TRUE)
  mapmisc::map.new(loaloa)
  terra::plot(loaFit$raster[["random.mean"]], main = "glgm random mean", add = TRUE, col = my_col$col, breaks = my_col$breaks, legend = FALSE)
  mapmisc::map.new(loaloa)
  terra::plot(my_sim[["mean"]], main = "adlaplace matern mean", add = TRUE, col = my_col$col, breaks = my_col$breaks, legend = FALSE)
  mapmisc::legendBreaks("right", my_col)
  mapmisc::map.new(loaloa)
  terra::plot(loaFit$raster[["random.sd"]], main = "glgm random sd", add = TRUE, col = sd_col$col, breaks = sd_col$breaks, legend = FALSE)
  mapmisc::map.new(loaloa)
  terra::plot(my_sim[["sd"]], main = "adlaplace matern sd", add = TRUE, col = sd_col$col, breaks = sd_col$breaks, legend = FALSE)
  mapmisc::legendBreaks("right", sd_col)
} else {
  old_par <- graphics::par(mfrow = c(1, 2), mar = c(2, 2, 2, 4))
  on.exit(graphics::par(old_par), add = TRUE)
  terra::plot(my_sim[["mean"]], main = "adlaplace matern mean")
  terra::plot(my_sim[["sd"]], main = "adlaplace matern sd")
}
```

Four posterior draws of the adlaplace Matérn field:

```{r plot-sims}
#| fig.width: 8
#| fig.height: 8
sim_layers <- paste0("sim", c(1L, 34L, 67L, 100L))
terra::plot(my_sim[[sim_layers]], nc = 2, main = sim_layers)
```
