---
title: "SUN(3,3) maximum likelihood with analytic gradients"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output: pdf_document
vignette: >
  %\VignetteIndexEntry{SUN(3,3) maximum likelihood with analytic gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(admvn)
```

This vignette shows how to obtain the MLE of a SUN$(3,3)$ distribution with
analytic gradients from `admvn::dsun_fun()`. Orthant probabilities in
dimension $\le 3$ use specialized bivariate/trivariate CDF routines (as in
`sn`/`mnormt`), and their gradients are analytic. We compare `stats::optim()`
and the C++ trust-region driver `sun_mle()`, and check agreement with
`sn::dsun` at the fitted value.

## Parameterization

A SUN$(3,3)$ observation is $Y = U \mid V > 0$ with $U,V\in\mathbb{R}^3$.
The latent structure is defined on the **correlation scale** of $U$, then
scaled by $D_u=\operatorname{diag}(\nu)$ with $\nu_j>0$ the marginal
standard deviations. Writing $U_\star$ for the standardized latent (unit
variances),

\[
\begin{bmatrix} U_\star \\ V \end{bmatrix}
\sim
\mathcal{N}\!\left(
\begin{bmatrix} 0 \\ 0 \end{bmatrix},
\begin{bmatrix} I & 0 \\ L & I \end{bmatrix}
\begin{bmatrix} C_u & 0 \\ 0 & R \end{bmatrix}
\begin{bmatrix} I & L^\top \\ 0 & I \end{bmatrix}
\right),
\qquad
U = \xi + D_u U_\star.
\]

Equivalently,

\[
\operatorname{Var}
\begin{pmatrix} U \\ V \end{pmatrix}
=
\begin{bmatrix}
D_u C_u D_u & D_u C_u L^\top \\
L C_u D_u & L C_u L^\top + R
\end{bmatrix}.
\]

Excluding $D_u$, the SUN law is therefore determined by a correlation matrix
$C_u$ together with loadings $L$ and residual $R$; entries of $L$ in
$[-1,1]$ are the natural regime.

The free parameter vector has **21** components and is mapped to the `sn`
list `dp = list(xi, Omega, Delta, tau, Gamma)` by `make_sun_params()` as
follows.

**Location.** $\xi = (\xi_1,\xi_2,\xi_3)^\top$.

**Observed-scale covariance** $\Omega = D_u C_u D_u$ with
$D_u=\operatorname{diag}(\nu)$ and $\nu_j>0$ the marginal **standard
deviations**. A correlation $C_u$ comes from unconstrained hyperspherical
coordinates $(\ell_{21},\ell_{31},\ell_{32})$ (unit-row Cholesky +
$\tanh$):

\[
\Omega = D_u C_u D_u,
\qquad
D_u = \operatorname{diag}(\nu_1,\nu_2,\nu_3).
\]

So $\Omega_{jj}=\nu_j^2$.

**Skew loadings.** $L$ multiplies $C_u$ (not $\Omega$). It is parameterized
by its three diagonals, three off-diagonal pair levels
$(L_{12},L_{13},L_{23})$, and three signed asymmetries
$(e_{12},e_{13},e_{23})$: lower triangle gets $L_{ij}+e_{ij}$, upper
gets $L_{ij}-e_{ij}$.

\[
L =
\begin{bmatrix}
L_{11} & L_{12}-e_{12} & L_{13}-e_{13} \\
L_{12}+e_{12} & L_{22} & L_{23}-e_{23} \\
L_{13}+e_{13} & L_{23}+e_{23} & L_{33}
\end{bmatrix}.
\]

($L = 0$ recovers a Gaussian; a small diagonal $L$ is a mild skew start.
Default box constraints keep diagonals in $[-1,1]$ and pair levels /
asymmetries $L_{ij}$, $e_{ij}$ in $[-0.5,0.5]$.)

**Residual correlation** \(C\) and latent \(\Sigma_{vv}\).
Unconstrained free parameters \((a,b,c)\) enter a unit-row Cholesky
(hyperspherical) factor for a residual correlation \(C\):

\[
\begin{aligned}
B_r &=
\begin{bmatrix}
1 & 0 & 0 \\
t_a & \sqrt{1-t_a^2} & 0 \\
t_b & t_c\sqrt{1-t_b^2} & \sqrt{(1-t_b^2)(1-t_c^2)}
\end{bmatrix},
\\
t_a &= \tanh(a),\; t_b = \tanh(b),\; t_c = \tanh(c),
\qquad
C = B_r B_r^\top.
\end{aligned}
\]

With \(M = L C_u L^\top\) and \(D = \operatorname{diag}\!\bigl(1 - \operatorname{diag}(M)\bigr)\)
(requiring \(\operatorname{diag}(M) < 1\)),

\[
R = D^{1/2} C D^{1/2},
\qquad
\Sigma_{vv} = M + R.
\]

Then \(\Sigma_{vv}\) is a correlation matrix
(\(\operatorname{diag}(\Sigma_{vv}) = 1\)) by construction—no post-hoc
\(\omega_v\) / `cov2cor` rescaling of \(V\).

**`sn` cross-covariance quantities.** Writing
\(\Sigma_{uv} = D_u C_u L^\top\),

\[
\Delta = D_u^{-1}\Sigma_{uv} = C_u L^\top,
\qquad
\Gamma = \Sigma_{vv},
\qquad
\tau = 0.
\]

**Parameter order** (names used in code):

\[
\begin{aligned}
&\xi_1,\xi_2,\xi_3,\;
\nu_1,\nu_2,\nu_3,\;
\ell_{21},\ell_{31},\ell_{32},\\
&L_{11},L_{22},L_{33},\;
L_{12},L_{13},L_{23},\;
e_{12},e_{13},e_{23},\;
a,b,c.
\end{aligned}
\]

```{r params}
true_params <- c(
  xi1 = 0, xi2 = 0, xi3 = 0,
  nu1 = 1, nu2 = 1, nu3 = 1,
  ell21 = 0.2, ell31 = 0.1, ell32 = 0.2,
  L11 = 0.4, L22 = 0.4, L33 = 0.35,
  L12 = 0.1, L13 = 0.05, L23 = 0.1,
  e12 = 0, e13 = -0.05, e23 = 0,
  a = 0.3, b = 0.2, c = 0.3
)
true_sun_params <- make_sun_params(true_params)
str(true_sun_params, max.level = 1)
# Sigma_vv = Gamma has unit diagonal by construction
diag(true_sun_params$Gamma)
```

## Simulation

A modest sample keeps the vignette fast while still illustrating the API.

```{r sim, eval = requireNamespace("sn", quietly = TRUE)}
library(sn)
set.seed(0)
n <- 250
ysim <- sn::rsun(n, dp = true_sun_params)
colnames(ysim) <- paste0("Y", 1:3)
head(ysim)
```

## Build a reusable log-likelihood tape

`dsun_fun()` fixes the data and builds per-observation shard tapes. For
SUN$(3,3)$ the orthant CDFs use the specialized $d\le 3$ path (QMC settings
are unused for those values).

```{r tape, eval = requireNamespace("sn", quietly = TRUE)}
# Unused for d<=3 specialized CDF values; kept for API compatibility.
qmc_n_points <- 64L
qmc_n_shifts <- 2L

build_time <- system.time({
  tape <- dsun_fun(
    ysim,
    true_params,
    n_points = qmc_n_points,
    n_shifts = qmc_n_shifts,
    n_threads = 2L
  )
})["elapsed"]

check <- tape$eval(true_params, log = TRUE, deriv = 1L)
sn_ll <- sum(sn::dsun(ysim, dp = true_sun_params, log = TRUE))
c(
  sn = sn_ll,
  admvn = check$value,
  build_sec = unname(build_time),
  n_threads = tape$n_threads
)
```

## MLE with `optim()` and an analytic gradient

`tape$optim_fns()` returns cached `fn`/`gr` that share one `deriv = 1`
evaluation per new parameter value.

```{r mle-setup, eval = requireNamespace("sn", quietly = TRUE)}
start_params <- true_params
start_params[] <- c(
  xi1 = 0.2, xi2 = -0.1, xi3 = 0.1,
  nu1 = 1.4, nu2 = 0.7, nu3 = 1.2,
  ell21 = 0.0, ell31 = 0.3, ell32 = 0.0,
  L11 = 0.7, L22 = 0.8, L33 = 0.9,
  L12 = 0.2, L13 = 0.1, L23 = 0.2,
  e12 = 0, e13 = 0, e23 = 0.1,
  a = 0.1, b = 0.0, c = 0.2
)

bnd <- sun33_bounds()
lower <- bnd$lower
upper <- bnd$upper
# Keep nu floors a bit above the package default for this demo.
lower[4:6] <- pmax(lower[4:6], 1e-3)
# Rescale free parameters so L-BFGS-B does not take wild Omega steps.
parscale <- pmax(abs(as.numeric(start_params)), 0.1)
optim_ctrl <- list(
  fnscale = -1,
  maxit = 40,
  parscale = parscale
)

obj <- tape$optim_fns()
```

```{r mle-admvn, eval = requireNamespace("sn", quietly = TRUE)}
fit_ad_time <- system.time({
  fit_ad <- optim(
    start_params,
    fn = obj$fn,
    gr = obj$gr,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = optim_ctrl
  )
})

fit_ad$convergence
fit_ad$value
unname(fit_ad_time["elapsed"])
```

## MLE with C++ `sun_mle()` (trustOptim)

`sun_mle()` runs a quasi-Newton trust-region loop entirely in C++ (no R
fn/gr round-trips). The `nu` parameters are optimized on the log scale.

```{r mle-trust, eval = requireNamespace("sn", quietly = TRUE) && requireNamespace("trustOptim", quietly = TRUE)}
fit_trust_time <- system.time({
  fit_trust <- sun_mle(
    tape,
    start_params,
    control = list(maxit = 40L, report.freq = 0L, report.level = 0L)
  )
})
fit_trust$value
fit_trust$iterations
unname(fit_trust_time["elapsed"])
```

## Comparison with `sn::dsun`

Score each fit with both `sn::dsun` and `tape$eval` (no separate value-only
`sn` optim — that is much slower and not needed to check agreement). Package
tests (`tests/testthat/test-dsun.R`, `test-dsun44.R`, `test-sun-kl.R`) also
assert that `dsun` / `dsun_fun` / `evaluate_sun33_slice` match `sn::dsun` on
mild parameter values.

```{r compare, eval = requireNamespace("sn", quietly = TRUE)}
sn_ll_at <- function(par) {
  sum(sn::dsun(ysim, dp = make_sun_params(par), log = TRUE))
}
ad_ll_at <- function(par) {
  tape$eval(par, log = TRUE, deriv = 0L)$value
}

compare <- data.frame(
  method = "ad_optim",
  sn_loglik = sn_ll_at(fit_ad$par),
  admvn_loglik = ad_ll_at(fit_ad$par),
  iterations = unname(fit_ad$counts[["gradient"]]),
  elapsed_sec = unname(fit_ad_time["elapsed"]),
  stringsAsFactors = FALSE
)
if (exists("fit_trust")) {
  compare <- rbind(
    data.frame(
      method = "ad_trustOptim",
      sn_loglik = sn_ll_at(fit_trust$par),
      admvn_loglik = ad_ll_at(fit_trust$par),
      iterations = as.numeric(fit_trust$iterations),
      elapsed_sec = unname(fit_trust_time["elapsed"]),
      stringsAsFactors = FALSE
    ),
    compare
  )
}
rownames(compare) <- NULL
compare
```

`sn_loglik` and `admvn_loglik` are \(\sum_i \log f(y_i;\hat\theta)\) at each
method's reported parameter estimate. With the specialized $d\le 3$ CDF, the
two log-densities agree closely.

## Gradient check at the fitted MLE

Compare analytic/`CppAD` partials to central finite differences along a few
one-dimensional slices through $\hat\theta$.

```{r grad-check, eval = requireNamespace("sn", quietly = TRUE), fig.width = 7, fig.height = 5}
par_hat <- as.numeric(if (exists("fit_trust")) fit_trust$par else fit_ad$par)
par_names <- names(true_params)
if (is.null(par_names)) {
  par_names <- paste0("p", seq_along(par_hat))
}
names(par_hat) <- par_names

idx <- match(c("xi1", "nu1", "L11", "a"), par_names)

ll <- function(p) tape$eval(p, log = TRUE, deriv = 0L)$value
ad_grad_j <- function(p, j) tape$eval(p, log = TRUE, deriv = 1L)$gradient[j]
fd_grad_j <- function(p, j, h) {
  p_up <- p
  p_dn <- p
  p_up[j] <- p[j] + h
  p_dn[j] <- p[j] - h
  (ll(p_up) - ll(p_dn)) / (2 * h)
}

op <- par(mfrow = c(2, 2), mar = c(3.2, 3.2, 2.2, 0.8), oma = c(0, 0, 1.5, 0))
on.exit(par(op), add = TRUE)

for (j in idx) {
  center <- par_hat[j]
  span <- pmax(0.15 * abs(center), 0.05)
  grid <- center + seq(-span, span, length.out = 9L)
  if (par_names[j] %in% c("nu1", "nu2", "nu3")) {
    grid <- grid[grid > 1e-3]
  }

  ad_vals <- numeric(length(grid))
  fd_vals <- numeric(length(grid))
  for (k in seq_along(grid)) {
    p <- par_hat
    p[j] <- grid[k]
    h <- pmax(1e-5, 1e-4 * abs(grid[k]))
    ad_vals[k] <- ad_grad_j(p, j)
    fd_vals[k] <- fd_grad_j(p, j, h)
  }

  ylim <- range(c(ad_vals, fd_vals), finite = TRUE)
  plot(
    grid, ad_vals,
    type = "l", lwd = 2, col = "#1b4f72",
    xlab = par_names[j], ylab = expression(partialdiff * ell / partialdiff * theta),
    main = par_names[j],
    ylim = ylim
  )
  lines(grid, fd_vals, lwd = 2, col = "#b03a2e", lty = 2)
  abline(v = center, col = "gray50", lty = 3)
}
legend(
  "topleft",
  legend = c("AD", "finite diff"),
  col = c("#1b4f72", "#b03a2e"),
  lty = c(1, 2), lwd = 2, bty = "n", cex = 0.8
)
```
