---
title: "SUN(3,3) maximum likelihood with analytic gradients"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SUN(3,3) maximum likelihood with analytic gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(admvn)
```

This vignette shows how to obtain the MLE of a SUN$(3,3)$ distribution with
analytic gradients from `admvn::dsun_fun()`. Gradients of the MVN orthant
probabilities use closed-form formulas, so the QMC budget only needs to
resolve the log-likelihood *value*. We compare `stats::optim()`, the C++
trust-region driver `sun_mle()`, and a value-only `sn::dsun` fit.

## Parameterization

The free parameter vector has 21 components and maps to the `sn` list
`dp = list(xi, Omega, Delta, tau, Gamma)` via `make_sun_params()`:

```{r params}
true_params <- c(
  xi1 = 0, xi2 = 0, xi3 = 0,
  nu1 = 1, nu2 = 1, nu3 = 1,
  ell21 = 0.2, ell31 = 0.1, ell32 = 0.2,
  L11 = 1, L12 = 0.5, L13 = 1,
  L21 = 0.5, L22 = 1, L23 = 1.5,
  L31 = 0, L32 = 0.5, L33 = 1,
  a = 0.3, b = 0.2, c = 0.3
)
true_sun_params <- make_sun_params(true_params)
str(true_sun_params, max.level = 1)
```

## Simulation

```{r sim, eval = requireNamespace("sn", quietly = TRUE)}
library(sn)
set.seed(0)
n <- 500
ysim <- sn::rsun(n, dp = true_sun_params)
colnames(ysim) <- paste0("Y", 1:3)
head(ysim)
```

## Build a reusable log-likelihood tape

`dsun_fun()` fixes the data and builds per-observation shard tapes. Because
gradients are analytic, a modest QMC budget (e.g. `400 × 6`) is enough for
optimization.

```{r tape, eval = requireNamespace("sn", quietly = TRUE)}
# Modest QMC: gradients do not differentiate through QMC.
qmc_n_points <- 400L
qmc_n_shifts <- 6L

build_time <- system.time({
  tape <- dsun_fun(
    ysim,
    true_params,
    n_points = qmc_n_points,
    n_shifts = qmc_n_shifts,
    n_threads = 8L
  )
})["elapsed"]

check <- tape$eval(true_params, log = TRUE, deriv = 1L)
sn_ll <- sum(sn::dsun(ysim, dp = true_sun_params, log = TRUE))
c(
  sn = sn_ll,
  admvn = check$value,
  build_sec = unname(build_time),
  n_threads = tape$n_threads
)
```

## MLE with `optim()` and an analytic gradient

`tape$optim_fns()` returns cached `fn`/`gr` that share one `deriv = 1`
evaluation per new parameter value.

```{r mle-admvn, eval = requireNamespace("sn", quietly = TRUE)}
start_params <- true_params
start_params[] <- c(
  xi1 = 0.2, xi2 = -0.1, xi3 = 0.1,
  nu1 = 1.4, nu2 = 0.7, nu3 = 1.2,
  ell21 = 0.0, ell31 = 0.3, ell32 = 0.0,
  L11 = 0.7, L12 = 0.2, L13 = 0.1,
  L21 = 0.2, L22 = 0.8, L23 = 0.3,
  L31 = 0.1, L32 = 0.3, L33 = 0.9,
  a = 0.1, b = 0.0, c = 0.2
)

lower <- c(
  -Inf, -Inf, -Inf,
  1e-3, 1e-3, 1e-3,
  -Inf, -Inf, -Inf,
  rep(-Inf, 9),
  -Inf, -Inf, -Inf
)
upper <- rep(Inf, length(lower))

obj <- tape$optim_fns()

fit_ad_time <- system.time({
  fit_ad <- optim(
    start_params,
    fn = obj$fn,
    gr = obj$gr,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = list(fnscale = -1, maxit = 50, REPORT = 1, trace = 1)
  )
})

fit_ad$convergence
fit_ad$value
unname(fit_ad_time["elapsed"])
```

## MLE with C++ `sun_mle()` (trustOptim)

`sun_mle()` runs a quasi-Newton trust-region loop entirely in C++ (no R
fn/gr round-trips). The `nu` parameters are optimized on the log scale.

```{r mle-trust, eval = requireNamespace("sn", quietly = TRUE) && requireNamespace("trustOptim", quietly = TRUE)}
fit_trust_time <- system.time({
  fit_trust <- sun_mle(
    tape,
    start_params,
    control = list(maxit = 50L, report.freq = 1L, report.level = 1L)
  )
})
fit_trust$value
fit_trust$iterations
unname(fit_trust_time["elapsed"])
```

## Comparison with value-only `sn` MLE

```{r mle-sn, eval = requireNamespace("sn", quietly = TRUE)}
sun_loglik_sn <- function(params, x) {
  dp <- make_sun_params(params)
  sum(sn::dsun(x, dp = dp, log = TRUE))
}

fit_sn_time <- system.time({
  fit_sn <- optim(
    start_params,
    sun_loglik_sn,
    x = ysim,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = list(fnscale = -1, maxit = 50, REPORT = 1, trace = 1)
  )
})

fit_sn$convergence
fit_sn$value
c(
  optim_sec = unname(fit_ad_time["elapsed"]),
  sun_mle_sec = if (exists("fit_trust_time")) unname(fit_trust_time["elapsed"]) else NA,
  sn_sec = unname(fit_sn_time["elapsed"])
)
```

```{r compare, eval = requireNamespace("sn", quietly = TRUE)}
tab <- cbind(
  true = true_params,
  start = start_params,
  mle_optim = fit_ad$par,
  mle_sn = fit_sn$par
)
if (exists("fit_trust")) {
  tab <- cbind(tab, mle_sun_mle = fit_trust$par)
}
tab
```

Small differences between fits are expected: `admvn` approximates the SUN
orthant probabilities with quasi-Monte Carlo, while `sn` uses its own CDF
routine.
