---
title: "SUN(3,3) maximum likelihood with derivatives"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SUN(3,3) maximum likelihood with derivatives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(admvn)
qmc_n_points <- 2000L
qmc_n_shifts <- 12L
```

This vignette reproduces the SUN$(3,3)$ simulation and MLE example from the
supervision notes. We first fit the model with `sn::dsun` (density only), then
repeat the fit using `admvn::dsun_fun`, which builds per-observation CppAD
shard tapes with the data fixed and returns derivatives with respect to the
21-parameter vector.

## Parameterization

The free parameter vector has 21 components and is converted to the `sn` list
`dp = list(xi, Omega, Delta, tau, Gamma)` by `make_sun_params()`:

```{r params}
true_params <- c(
  xi1 = 0, xi2 = 0, xi3 = 0,
  nu1 = 1, nu2 = 1, nu3 = 1,
  ell21 = 0.2, ell31 = 0.1, ell32 = 0.2,
  L11 = 1, L12 = 0.5, L13 = 1,
  L21 = 0.5, L22 = 1, L23 = 1.5,
  L31 = 0, L32 = 0.5, L33 = 1,
  a = 0.3, b = 0.2, c = 0.3
)
true_sun_params <- make_sun_params(true_params)
str(true_sun_params, max.level = 1)
```

## Simulation

```{r sim, eval = requireNamespace("sn", quietly = TRUE)}
library(sn)
set.seed(0)
n <- 2000
ysim <- sn::rsun(n, dp = true_sun_params)
colnames(ysim) <- paste0("Y", 1:3)
head(ysim)
```

The vignette uses `n = 2000` rather than 8000 to keep build time reasonable.

## Log-likelihood check

At the true parameters, `admvn::dsun` and `sn::dsun` agree up to quasi-Monte
Carlo error in the orthant probabilities. The tape fixes the data and treats
only `par` as independent; `deriv = 0` evaluates the value only.

```{r check-density, eval = requireNamespace("sn", quietly = TRUE)}
sn_ll <- sum(sn::dsun(ysim, dp = true_sun_params, log = TRUE))

build_time <- system.time({
  tape_check <- dsun_fun(
    ysim,
    true_params,
    n_points = qmc_n_points,
    n_shifts = qmc_n_shifts
  )
})["elapsed"]

eval_time <- system.time({
  ad_out <- tape_check$eval(true_params, log = TRUE, deriv = 0L)
})["elapsed"]

c(
  sn = sn_ll,
  admvn = ad_out$value,
  build_sec = build_time,
  eval_sec = eval_time,
  n_threads = tape_check$n_threads
)
```

## Derivative check

We slice the log-likelihood along one parameter and compare the analytic
gradient from `admvn` with a finite-difference derivative of the slice.

```{r deriv-check, eval = requireNamespace("sn", quietly = TRUE), fig.width = 9, fig.height = 3.8, fig.cap = "Left: log-likelihood slice along ell21. Right: finite-difference derivative of the slice (grey) versus the admvn analytic gradient (blue)."}
param_name <- "ell21"
base_par <- true_params
grid <- seq(-0.1, 0.5, length.out = 21L)

loglik <- vapply(grid, function(v) {
  par <- base_par
  par[param_name] <- v
  tape_check$eval(par, log = TRUE, deriv = 0L)$value
}, numeric(1))

grad_ad <- vapply(grid, function(v) {
  par <- base_par
  par[param_name] <- v
  tape_check$eval(par, log = TRUE, deriv = 1L)$gradient[param_name]
}, numeric(1))

grad_num <- diff(loglik) / diff(grid)
grid_mid <- 0.5 * (grid[-length(grid)] + grid[-1L])

old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))

plot(
  grid,
  loglik,
  type = "l",
  lwd = 2,
  col = "steelblue",
  xlab = param_name,
  ylab = "log-likelihood",
  main = sprintf("Log-likelihood slice (%s)", param_name)
)
abline(v = true_params[param_name], lty = 3, col = "gray50")

plot(
  grid_mid,
  grad_num,
  type = "l",
  lwd = 2,
  col = "gray40",
  xlab = param_name,
  ylab = expression(paste("partial log-likelihood / partial", theta)),
  main = "Numeric vs AD gradient",
  ylim = range(c(grad_num, grad_ad), na.rm = TRUE)
)
lines(grid, grad_ad, lwd = 2, col = "steelblue", lty = 2)
abline(v = true_params[param_name], lty = 3, col = "gray50")
legend(
  "topright",
  legend = c("finite difference", "admvn gradient"),
  col = c("gray40", "steelblue"),
  lty = c(1, 2),
  lwd = 2,
  bty = "n",
  cex = 0.85
)

par(old_par)
```

## MLE without derivatives (`sn`)

```{r mle-sn, eval = requireNamespace("sn", quietly = TRUE)}
sun_loglik_sn <- function(params, x) {
  dp <- make_sun_params(params)
  sum(sn::dsun(x, dp = dp, log = TRUE))
}

start_params <- true_params
start_params[] <- c(
  xi1 = 0.2, xi2 = -0.1, xi3 = 0.1,
  nu1 = 1.4, nu2 = 0.7, nu3 = 1.2,
  ell21 = 0.0, ell31 = 0.3, ell32 = 0.0,
  L11 = 0.7, L12 = 0.2, L13 = 0.1,
  L21 = 0.2, L22 = 0.8, L23 = 0.3,
  L31 = 0.1, L32 = 0.3, L33 = 0.9,
  a = 0.1, b = 0.0, c = 0.2
)

lower <- c(
  -Inf, -Inf, -Inf,
  1e-3, 1e-3, 1e-3,
  -Inf, -Inf, -Inf,
  rep(-Inf, 9),
  -Inf, -Inf, -Inf
)
upper <- rep(Inf, length(lower))

fit_sn_time <- system.time({
  fit_sn <- optim(
    start_params,
    sun_loglik_sn,
    x = ysim,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = list(fnscale = -1, maxit = 40)
  )
})
fit_sn$convergence
fit_sn$value
fit_sn_time["elapsed"]
```

## MLE with analytic gradient (`admvn`)

The AD fit uses cached fn/gr from `optim_fns()` so each L-BFGS step evaluates
the log-likelihood and gradient in a single CppAD pass. Shard evaluation runs
in parallel when OpenMP is enabled.

```{r mle-admvn, eval = requireNamespace("sn", quietly = TRUE)}
tape_build_time <- system.time({
  tape <- dsun_fun(
    ysim,
    start_params,
    n_points = qmc_n_points,
    n_shifts = qmc_n_shifts
  )
})["elapsed"]

obj <- tape$optim_fns()

fit_ad_time <- system.time({
  fit_ad <- optim(
    start_params,
    obj$fn,
    gr = obj$gr,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = list(fnscale = -1, maxit = 40)
  )
})
fit_ad$convergence
fit_ad$value
c(tape_build_sec = tape_build_time, optim_sec = fit_ad_time["elapsed"])
```

## Comparison

```{r compare, eval = requireNamespace("sn", quietly = TRUE)}
cbind(
  true = true_params,
  start = start_params,
  mle_sn = fit_sn$par,
  mle_admvn = fit_ad$par
)
rbind(
  true_mean = sn::sunMean(dp = true_sun_params),
  mle_sn_mean = sn::sunMean(dp = make_sun_params(fit_sn$par)),
  mle_admvn_mean = sn::sunMean(dp = make_sun_params(fit_ad$par))
)
```

The analytic-gradient fit should recover similar estimates to the `sn`-based fit,
with modest differences attributable to quasi-Monte Carlo error in the orthant
probabilities.
