---
title: "SUN(3,3) maximum likelihood with analytic gradients"
author: "Patrick Brown"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SUN(3,3) maximum likelihood with analytic gradients}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(admvn)
```

This vignette shows how to obtain the MLE of a SUN$(3,3)$ distribution with
analytic gradients from `admvn::dsun_fun()`. Orthant probabilities in
dimension $\le 3$ use specialized bivariate/trivariate CDF routines (as in
`sn`/`mnormt`), and their gradients are analytic, so a small QMC budget is
only a fallback for higher-dimensional cases. We compare `stats::optim()`,
the C++ trust-region driver `sun_mle()`, and a value-only `sn::dsun` fit.

## Parameterization

The free parameter vector has 21 components and maps to the `sn` list
`dp = list(xi, Omega, Delta, tau, Gamma)` via `make_sun_params()`:

```{r params}
true_params <- c(
  xi1 = 0, xi2 = 0, xi3 = 0,
  nu1 = 1, nu2 = 1, nu3 = 1,
  ell21 = 0.2, ell31 = 0.1, ell32 = 0.2,
  L11 = 1, L12 = 0.5, L13 = 1,
  L21 = 0.5, L22 = 1, L23 = 1.5,
  L31 = 0, L32 = 0.5, L33 = 1,
  a = 0.3, b = 0.2, c = 0.3
)
true_sun_params <- make_sun_params(true_params)
str(true_sun_params, max.level = 1)
```

## Simulation

```{r sim, eval = requireNamespace("sn", quietly = TRUE)}
library(sn)
set.seed(0)
n <- 2000
ysim <- sn::rsun(n, dp = true_sun_params)
colnames(ysim) <- paste0("Y", 1:3)
head(ysim)
```

## Build a reusable log-likelihood tape

`dsun_fun()` fixes the data and builds per-observation shard tapes. For
SUN$(3,3)$ the orthant CDFs use the specialized $d\le 3$ path (QMC settings
are unused for those values).

```{r tape, eval = requireNamespace("sn", quietly = TRUE)}
# Unused for d<=3 specialized CDF values; kept for API compatibility.
qmc_n_points <- 64L
qmc_n_shifts <- 2L

build_time <- system.time({
  tape <- dsun_fun(
    ysim,
    true_params,
    n_points = qmc_n_points,
    n_shifts = qmc_n_shifts,
    n_threads = 8L
  )
})["elapsed"]

check <- tape$eval(true_params, log = TRUE, deriv = 1L)
sn_ll <- sum(sn::dsun(ysim, dp = true_sun_params, log = TRUE))
c(
  sn = sn_ll,
  admvn = check$value,
  build_sec = unname(build_time),
  n_threads = tape$n_threads
)
```

## MLE with `optim()` and an analytic gradient

`tape$optim_fns()` returns cached `fn`/`gr` that share one `deriv = 1`
evaluation per new parameter value.

```{r mle-setup, eval = requireNamespace("sn", quietly = TRUE)}
start_params <- true_params
start_params[] <- c(
  xi1 = 0.2, xi2 = -0.1, xi3 = 0.1,
  nu1 = 1.4, nu2 = 0.7, nu3 = 1.2,
  ell21 = 0.0, ell31 = 0.3, ell32 = 0.0,
  L11 = 0.7, L12 = 0.2, L13 = 0.1,
  L21 = 0.2, L22 = 0.8, L23 = 0.3,
  L31 = 0.1, L32 = 0.3, L33 = 0.9,
  a = 0.1, b = 0.0, c = 0.2
)

lower <- c(
  -Inf, -Inf, -Inf,
  1e-3, 1e-3, 1e-3,
  -Inf, -Inf, -Inf,
  rep(-Inf, 9),
  -Inf, -Inf, -Inf
)
upper <- rep(Inf, length(lower))
# Rescale free parameters so L-BFGS-B does not take wild Omega steps.
parscale <- pmax(abs(as.numeric(start_params)), 0.1)
optim_ctrl <- list(
  fnscale = -1,
  maxit = 100,
  parscale = parscale,
  REPORT = 20,
  trace = 2
)

obj <- tape$optim_fns()
```

```{r mle-admvn, eval = requireNamespace("sn", quietly = TRUE)}
fit_ad_time <- system.time({
  fit_ad <- optim(
    start_params,
    fn = obj$fn,
    gr = obj$gr,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = optim_ctrl
  )
})

fit_ad$convergence
fit_ad$value
unname(fit_ad_time["elapsed"])
```

## MLE with C++ `sun_mle()` (trustOptim)

`sun_mle()` runs a quasi-Newton trust-region loop entirely in C++ (no R
fn/gr round-trips). The `nu` parameters are optimized on the log scale.

```{r mle-trust, eval = requireNamespace("sn", quietly = TRUE) && requireNamespace("trustOptim", quietly = TRUE)}
fit_trust_time <- system.time({
  fit_trust <- sun_mle(
    tape,
    start_params,
    control = list(maxit = 500, report.freq = 1L, report.level = 4L)
  )
})
fit_trust$value
fit_trust$iterations
unname(fit_trust_time["elapsed"])
```

## Comparison with value-only `sn` MLE

```{r mle-sn, eval = requireNamespace("sn", quietly = TRUE)}
sun_loglik_sn <- function(params, x) {
  dp <- make_sun_params(params)
  ll <- sum(sn::dsun(x, dp = dp, log = TRUE))
  if (!is.finite(ll)) -1e300 else ll
}

fit_sn_time <- system.time({
  fit_sn <- optim(
    start_params,
    sun_loglik_sn,
    x = ysim,
    method = "L-BFGS-B",
    lower = lower,
    upper = upper,
    control = optim_ctrl
  )
})

fit_sn$convergence
fit_sn$value
```

```{r compare, eval = requireNamespace("sn", quietly = TRUE)}
# Score every fit's MLE with both sn::dsun and admvn (tape$eval) log-likelihoods.
sn_ll_at <- function(par) {
  sum(sn::dsun(ysim, dp = make_sun_params(par), log = TRUE))
}
ad_ll_at <- function(par) {
  tape$eval(par, log = TRUE, deriv = 0L)$value
}

compare <- data.frame(
  method = c("ad_optim", "sn_dsun"),
  sn_loglik = c(sn_ll_at(fit_ad$par), sn_ll_at(fit_sn$par)),
  admvn_loglik = c(ad_ll_at(fit_ad$par), ad_ll_at(fit_sn$par)),
  iterations = c(
    unname(fit_ad$counts[["gradient"]]),
    unname(fit_sn$counts[["function"]])
  ),
  elapsed_sec = c(
    unname(fit_ad_time["elapsed"]),
    unname(fit_sn_time["elapsed"])
  ),
  stringsAsFactors = FALSE
)
if (exists("fit_trust")) {
  compare <- rbind(
    data.frame(
      method = "ad_trustOptim",
      sn_loglik = sn_ll_at(fit_trust$par),
      admvn_loglik = ad_ll_at(fit_trust$par),
      iterations = as.numeric(fit_trust$iterations),
      elapsed_sec = unname(fit_trust_time["elapsed"]),
      stringsAsFactors = FALSE
    ),
    compare
  )
}
rownames(compare) <- NULL
compare
```

`sn_loglik` and `admvn_loglik` are \(\sum_i \log f(y_i;\hat\theta)\) at each
method's reported parameter estimate, evaluated with `sn::dsun` and
`tape$eval(..., log = TRUE)` respectively. With the specialized $d\le 3$ CDF,
the two log-densities agree closely; remaining differences come from optimizer
paths. For `ad_optim` / `sn_dsun`, `iterations` is the reported gradient or
function count from `optim`; for `ad_trustOptim` it is the trust-region
iteration count.

## Gradient check at the `ad_trustOptim` MLE

For each of ten free parameters, a one-dimensional slice through
$\hat\theta$ from `sun_mle()` compares the analytic/`CppAD` partial
derivative with a central finite-difference estimate of the same
component.

```{r grad-check, eval = requireNamespace("sn", quietly = TRUE) && exists("fit_trust"), fig.width = 8, fig.height = 10}
par_hat <- as.numeric(fit_trust$par)
par_names <- names(true_params)
if (is.null(par_names)) {
  par_names <- paste0("p", seq_along(par_hat))
}
names(par_hat) <- par_names

# Ten representative coordinates (location, scale, dependence, skewness).
idx <- match(
  c("xi1", "xi2", "nu1", "nu2", "ell21", "ell32", "L11", "L22", "L33", "a")[1:9],
  par_names
)

ll <- function(p) tape$eval(p, log = TRUE, deriv = 0L)$value
ad_grad_j <- function(p, j) tape$eval(p, log = TRUE, deriv = 1L)$gradient[j]
fd_grad_j <- function(p, j, h) {
  p_up <- p
  p_dn <- p
  p_up[j] <- p[j] + h
  p_dn[j] <- p[j] - h
  (ll(p_up) - ll(p_dn)) / (2 * h)
}

op <- par(mfrow = c(3, 3), mar = c(3.2, 3.2, 2.2, 0.8), oma = c(0, 0, 1.5, 0))
on.exit(par(op), add = TRUE)

for (j in idx) {
  center <- par_hat[j]
  # Relative step that stays positive for nu_* when possible.
  span <- pmax(0.15 * abs(center), 0.05)
  grid <- center + seq(-span, span, length.out = 21L)
  if (par_names[j] %in% c("nu1", "nu2", "nu3")) {
    grid <- grid[grid > 1e-3]
  }

  ad_vals <- numeric(length(grid))
  fd_vals <- numeric(length(grid))
  for (k in seq_along(grid)) {
    p <- par_hat
    p[j] <- grid[k]
    h <- pmax(1e-5, 1e-4 * abs(grid[k]))
    ad_vals[k] <- ad_grad_j(p, j)
    fd_vals[k] <- fd_grad_j(p, j, h)
  }

  ylim <- range(c(ad_vals, fd_vals), finite = TRUE)
  plot(
    grid, ad_vals,
    type = "l", lwd = 2, col = "#1b4f72",
    xlab = par_names[j], ylab = expression(partialdiff * ell / partialdiff * theta),
    main = par_names[j],
    ylim = ylim
  )
  lines(grid, fd_vals, lwd = 2, col = "#b03a2e", lty = 2)
  abline(v = center, col = "gray50", lty = 3)
}
legend(
  "topleft",
  legend = c("AD", "finite diff"),
  col = c("#1b4f72", "#b03a2e"),
  lty = c(1, 2), lwd = 2, bty = "n", cex = 0.8
)
```
